### 2015-04-07 abstract.tex
Initial outline

%!TEX root = paper.tex

The largest HPC centers spend considerable time supporting software configurations for thousands of users, but the complexity of software is quickly outpacing the capabilities of existing management tools. Modern HPC applications may require specific versions of many dependency libraries, and they must build with many compilers and MPI versions. The configuration space is exponential in size, and it is infeasible to maintain a single software stack for all users at a site. Existing tools are not sufficient to manage this complexity. We present Spack, a novel package management tool that allows users to build software configurations on demand. Spack provides a novel, recursive specification syntax, and it allows arbitrarily many software and dependency combinations to coexist on the same system. Spack's build mechanism ensures that each package can find its dependencies, regardless of a user's runtime environment settings. Spack brings order to the complexity of HPC software installation management.

### 2015-04-07 conclusion.tex
Initial outline

%!TEX root = paper.tex

\section{Conclusion}
\label{sec:conclusion}

\todo{.5 pg}

### 2015-04-07 implementation.tex
Initial outline

%!TEX root = paper.tex

\section{Implementation}
\label{sec:implementation}

\todo{.5 page overview}

\subsection{Spack Specs}
	\todo{1 page}

\subsection{Versioned Virtual Dependencies}
	\todo{.5 page}

\subsection{Abstract \& Concrete Specs}
	\todo{1 page}

\subsection{Install Environment}

\subsubsection{Compiler Wrappers \& RPATHs}
	\todo{.5 page}

\subsubsection{Potential for auto-generation of modules (with deps)}
	\todo{.5 page}

### 2015-04-07 introduction.tex
Initial outline

%!TEX root = paper.tex

\section{Introduction}
\label{sec:intro}

\todo{1 page}

\begin{verbatim}
- Why do supercomputing centers need this?

	- specs
		- describe concisely
		- query installed stuff
		- modules require standardized set of names (no order)

	- flexibility (managing combinatoric configuration space)

	- RPATHs instead of (or in addition to) modules
		- pkgs work as built when run
		- spack means there is no module incantation required

	- code teams can build own libs & stacks

	- Contributions:
		1. Flexible mechanism for specifying and querying config space (spec)
		2. Implementation of this in spack
		3. 3 detailed use cases outlining our experiences at LLNL
			a. combinatoric versions
			b. multi-version python installation
			c. site-specific and user-specific policies
\end{verbatim}

### 2015-04-07 motivation.tex
Initial outline

%!TEX root = paper.tex

\section{Motivation}
\label{sec:motivation}

\subsection{Common software installation practices}
\todo{.5 pages}

\subsection{Related work}
\todo{.5 pages}

### 2015-04-07 usecase-combinatoric.tex
Initial outline

%!TEX root = paper.tex

\subsection{Combinatoric Names}
\label{sec:usecase-combinatoric}

\subsubsection{Gperftools packaging}
\todo{.5 page}

\begin{verbatim}
	- combinatoric nightmare
	- compilers x versions
\end{verbatim}

\subsubsection{{\tt mpileaks} packaging}
\todo{.5 page}
\begin{verbatim}
	- compiler x version x MPI x mpi version
\end{verbatim}

### 2015-04-07 usecase-policy.tex
Initial outline

%!TEX root = paper.tex

\subsection{User and Site Policies}
\label{sec:usecase-policy}

\subsubsection{Handling site installs and policies}

\todo{1 pgs}

\begin{verbatim}
Putting things together for the end user
		- Site installs and policies
		- views
		- customizable defaults
		- site-specific policies
			- prefer mvapich, compiler, version
\end{verbatim}

\subsubsection{Allowing the user to override}

\todo{.5 pgs}

\begin{verbatim}
	- user can build his own entire stack without site support
	- user can point to site installs or build entire stack
\end{verbatim}

### 2015-04-07 usecase-python.tex
Initial outline

%!TEX root = paper.tex

\subsection{Support for extensible languages}
\label{sec:usecase-python}

\todo{1 pg}

\begin{verbatim}
- need for special python handling (prefix issues)
- solution of offering "meta package management"
- activate/deactivate python packages
- installs are in own prefix but can be merged into extended prefix
- design is extensible, allows for other anguages to use custom logic (e.g. perl, R)

- Integrating python meta-packages with full spack package
  management allows building optimized numpy stack AND easy python management.
\end{verbatim}

### 2015-04-07 usecases.tex
Initial outline

%!TEX root = paper.tex

\section{Use Cases}
\label{sec:usecases}

\input{usecase-combinatoric}
\input{usecase-python}
\input{usecase-policy}

### 2015-04-09 abstract.tex
Updated build.

@@
 
-The largest HPC centers spend considerable time supporting software configurations for thousands of users, but the complexity of software is quickly outpacing the capabilities of existing management tools.
-Modern HPC applications may require specific versions of many dependency libraries, and they must build with many compilers and MPI versions.
-The configuration space is exponential in size, and it is infeasible to maintain a single software stack for all users at a site.
-Existing tools are not sufficient to manage this complexity.
-We present Spack, a novel package management tool that allows users to build software configurations on demand.
+The largest HPC centers spend considerable time supporting software configurations for thousands of users, but the complexity of HPC software is quickly outpacing the capabilities of existing software management tools.
+Typical scientific applications require specific versions of many dependency libraries, and they must build with many compilers and MPI versions.
+This configuration space is exponential in size, and it is infeasible to maintain a single software stack for all users at a site.
+We present Spack, a novel package management tool that allows many software configurations to be built on demand.
 Spack provides a novel, recursive specification syntax, and it allows arbitrarily many software and dependency combinations to coexist on the same system.
 Spack's build mechanism ensures that each package can find its dependencies, regardless of a user's runtime environment settings.
-Spack brings order to the complexity of HPC software installation management.
+We show through several real-world use cases that Spack brings order to the chaos of HPC software installation management.

### 2015-04-11 abstract.tex
Update outline.

--- 2015-04-07 abstract.tex
@@
 
-The largest HPC centers spend considerable time supporting software configurations for thousands of users, but the complexity of software is quickly outpacing the capabilities of existing management tools.
-Modern HPC applications may require specific versions of many dependency libraries, and they must build with many compilers and MPI versions.
-The configuration space is exponential in size, and it is infeasible to maintain a single software stack for all users at a site.
-Existing tools are not sufficient to manage this complexity.
-We present Spack, a novel package management tool that allows users to build software configurations on demand.
-Spack provides a novel, recursive specification syntax, and it allows arbitrarily many software and dependency combinations to coexist on the same system.
-Spack's build mechanism ensures that each package can find its dependencies, regardless of a user's runtime environment settings.
-Spack brings order to the complexity of HPC software installation management.
+Large HPC centers spend considerable time supporting software for thousands of users, but the complexity of HPC software is quickly outpacing the capabilities of existing management tools.
+Scientific applications require specific versions of many compilers, MPI versions, and dependency libraries.
+This makes a single software stack infeasible, but managing many configurations is difficult because the configuration space is exponential in size.
+%
+Spack is a package management tool used at Lawrence Livermore National Laboratory (LLNL) to manage software complexity.
+It allows many different software configurations to build on demand and to coexist on the same system.
+Spack provides a novel, recursive specification syntax to manage combinatoric builds of packages and dependencies.
+It ensures that installed packages find their runtime dependencies, {\it regardless of the environment}.
+We show through real-world use cases that Spack supports a diverse and demanding user base, bringing order to the chaos of HPC software management.

### 2015-04-11 implementation.tex
Update outline.

@@
 	\todo{.5 page}
+
+\begin{verbatim}
+module disadvantages:
+	- lmod solves a lot of problems
+	- not on all systems
+		- users often run from just a path.
+		- not guaranteed that LD_LIBRARY_PATH, etc will
+		  be set correctly.
+
+- Lmod assumes build & submit environment must match
+	- users often have NO IDEA what environment a package
+	  was built with
+	- multiple apps may be built with different stacks
+	- combining them in a workflow is difficult with modules
+		(need to know right incantation)
+
+	- TACC implements a rather complicated mapping back to
+	    the build env (Lmod CheckExec)
+
+- Spack solves the matrix problem mentioned in the Lmod
+    presentation (slide 98)
+	- hierarchy is nasty -- every branch represents a
+	  greedy choice
+	- might NEED a specific version of ONE dependency
+	  that is not supported in the tree
+	  => SPACK!
+
+- libmesh
+	- boost, petsc, trilinos, grvy, etc.
+\end{verbatim}

### 2015-04-11 introduction.tex
Updated build.
Update outline.

@@
 - Why do supercomputing centers need this?
+
+- Build space on HPC machines is exponential
+	- each new variable multiplies number of build combinations
+	- complicated by the ABI.
+	- often need a new version on demand
+		- cannot just build it.
+
+	- need tools that can sample space.
+
+	- users cannot navigate space, and cumbersome to specify all details
+	- users often only care about a subset of constraints and just want things
+		built.
+	- users need to run code again later, but don't remember complex module
+		setup -- dynamic settings lead to user support calls and difficulties.
+
+	- install procedure is different between users and admins
+		- duplicated effort
+		- need reusable package file.
 

### 2015-04-12 abstract.tex
Intro done.

--- 2015-04-07 abstract.tex
@@
 
-The largest HPC centers spend considerable time supporting software configurations for thousands of users, but the complexity of software is quickly outpacing the capabilities of existing management tools.
-Modern HPC applications may require specific versions of many dependency libraries, and they must build with many compilers and MPI versions.
-The configuration space is exponential in size, and it is infeasible to maintain a single software stack for all users at a site.
-Existing tools are not sufficient to manage this complexity.
-We present Spack, a novel package management tool that allows users to build software configurations on demand.
-Spack provides a novel, recursive specification syntax, and it allows arbitrarily many software and dependency combinations to coexist on the same system.
-Spack's build mechanism ensures that each package can find its dependencies, regardless of a user's runtime environment settings.
-Spack brings order to the complexity of HPC software installation management.
+Large HPC centers spend considerable time supporting software for thousands of users, but the complexity of HPC software is quickly outpacing the capabilities of existing software management tools.
+Scientific applications require specific versions of many compilers, MPI implementations, and dependency libraries, so using a single, standard software stack is infeasible.
+However, managing many configurations is difficult because the configuration space is exponential in size.
+%
+Spack is a package manager used at Lawrence Livermore National Laboratory (LLNL) to manage software complexity.
+Spack provides a novel, recursive specification syntax to invoke parametric builds of packages and dependencies.
+It allows any number of builds to coexist on the same system, and it ensures that installed packages can find their dependencies, {\it regardless of the environment}.
+We show through real-world use cases that Spack supports a diverse and demanding user base, bringing order to HPC software chaos.

### 2015-04-12 introduction.tex
Intro done.

--- 2015-04-07 introduction.tex
@@
 
-\todo{1 page}
+The Livermore Computing (LC) facility at Lawrence Livermore National Laboratory
+(LLNL) supports around 2,500 users on 25 different clusters, ranging 
+in size from a 1.6 teraflop, 256-core cluster to to the
+20 petaflop, 1.6 million-core Sequoia machine, currently ranked first and
+third on the Graph500~\cite{graph500} and Top500~\cite{top500}
+lists, respectively.
+%
+%The massively parallel simulations that run on LC machines support
+%U.S. Department of Energy missions across a range of scientific domains, 
+%including nuclear physics, material science, climate science, mechanical 
+%engineering, geophysics, seismology, and stockpile stewardship.
+%
+The simulations that run on these machines are very complex; some
+depend on specific versions of over 50 dependency libraries.
+They require specific compilers, build options and MPI implementations to
+achieve the best performance, and users may run several
+different such codes in the same environment as part of larger 
+scientific workflows.
 
-\begin{verbatim}
-- Why do supercomputing centers need this?
+To support the diverse needs of applications, administrators and application
+developers 
+frequently build, install, and support many different configurations
+of math and physics libraries, as well as other software.
+Frequently, applications must be rebuilt to fix bugs and to support
+new versions of the OS, MPI, compilers, and other dependencies.
+Unfortunately, building scientific software is notoriously complex, with
+immature build systems that are difficult to adapt to new
+machines~\cite{dubois+:comp-sci-eng,hoste+:pyhpc12,wilson+:corr}.
+Worse, the space of required builds grows combinatorically
+with each new configuration parameter. As a result, LLNL staff
+spend countless hours dealing with build and deployment issues.
 
-	- specs
-		- describe concisely
-		- query installed stuff
-		- modules require standardized set of names (no order)
+Existing package management tools can automate some aspects of the build 
+process~\cite{bsdports,digirolamo:smithy,dolstra+:icfp08,dolstra+:lisa04,hashdist,homebrew,hoste+:pyhpc12,macports,thiruvathukal:gentoo04}.
+For the most part, these systems focus on keeping a single, stable set of 
+packages up to date, and they do not handle installation of multiple
+versions or configurations.  Those that {\it do} handle multiple configurations
+typically require that package files be created for each combination of 
+options~ \cite{digirolamo:smithy,dolstra+:icfp08,dolstra+:lisa04,hashdist,hoste+:pyhpc12}, 
+leading to a profusion of files and maintenance issues.
+To our knowledge, no existing tool allows new configurations to be composed
+and assembled on demand.  Some allow limited forms of 
+composition~\cite{hoste+:pyhpc12,dolstra+:icfp08,dolstra+:lisa04}, but they
+have limited dependency management capabilities, and they still burden
+users with combinatoric naming and version management.
 
-	- flexibility (managing combinatoric configuration space)
+In this paper, we describe our experiences with the {\it Spack} package manager,
+which we have developed at LLNL to manage increasing software complexity.
+Specifically, this paper offers the following contributions:
+\begin{enumerate}
+\item A novel, recursive syntax for specifying, building, and querying
+      HPC packages in a large parameter space;
+\item A build methodology that ensures packages find their dependencies
+      regardless of users' environment; 
+\item Spack: An implementation of these concepts; and
+\item Three use cases detailing LLNL's use of Spack to rapidly compose
+      software configurations, manage Python installations, and implement
+      site-specific build policies.
+\end{enumerate}
 
-	- RPATHs instead of (or in addition to) modules
-		- pkgs work as built when run
-		- spack means there is no module incantation required
-
-	- code teams can build own libs & stacks
-
-	- Contributions:
-		1. Flexible mechanism for specifying and querying config space (spec)
-		2. Implementation of this in spack
-		3. 3 detailed use cases outlining our experiences at LLNL
-			a. combinatoric versions
-			b. multi-version python installation
-			c. site-specific and user-specific policies
-\end{verbatim}
+Spack is in active development at LLNL, and our use cases outline some of
+its first uses in our environment.  Spack solves a problem that is pervasive
+at large, multi-user HPC centers, and we believe that spack and our
+experiences with it can be applied at other HPC facilities to improve
+operational efficiency by simplifying the build and deployment of 
+bleeding-edge scientific software.

### 2015-04-13 abstract.tex
Finish motivation.

@@
 Large HPC centers spend considerable time supporting software for thousands of users, but the complexity of HPC software is quickly outpacing the capabilities of existing software management tools.
-Scientific applications require specific versions of many compilers, MPI implementations, and dependency libraries, so using a single, standard software stack is infeasible.
+Scientific applications require specific versions of compilers, MPI implementations, and dependency libraries, so using a single, standard software stack is infeasible.
 However, managing many configurations is difficult because the configuration space is exponential in size.

### 2015-04-13 introduction.tex
Finish motivation.

@@
 
-Existing package management tools can automate some aspects of the build 
+Existing package management tools automate parts of the build 
 process~\cite{bsdports,digirolamo:smithy,dolstra+:icfp08,dolstra+:lisa04,hashdist,homebrew,hoste+:pyhpc12,macports,thiruvathukal:gentoo04}.
@@
 and assembled on demand.  Some allow limited forms of 
-composition~\cite{hoste+:pyhpc12,dolstra+:icfp08,dolstra+:lisa04}, but they
-have limited dependency management capabilities, and they still burden
-users with combinatoric naming and version management.
+composition~\cite{hoste+:pyhpc12,dolstra+:icfp08,dolstra+:lisa04}, but their
+dependency management is overly rigid, and they burden users with
+combinatoric installation naming and versioning.
 
@@
 \item A build methodology that ensures packages find their dependencies
-      regardless of users' environment; 
+      regardless of users' environments; 
 \item Spack: An implementation of these concepts; and
@@
 Spack is in active development at LLNL, and our use cases outline some of
-its first uses in our environment.  Spack solves a problem that is pervasive
-at large, multi-user HPC centers, and we believe that spack and our
+its first uses in our environment.  Spack solves software problems that are
+pervasive at large, multi-user HPC centers, and we believe that our
 experiences with it can be applied at other HPC facilities to improve

### 2015-04-13 motivation.tex
Updated build.
Finish motivation.

@@
 
-\section{Motivation}
+\section{Common Practice}
 \label{sec:motivation}
 
-\subsection{Common software installation practices}
-\todo{.5 pages}
+This section describes existing tools for managing HPC 
+software and motivates the need for a system like Spack.
 
-\subsection{Related work}
-\todo{.5 pages}
+\paragraph{Meta-build Systems}
+
+{\it Meta-build systems} such as Contractor, WAF, and
+MixDown~\cite{amundson:contractor,epperly+:mixdown,nagy:waf} are closely
+related to package managers, but their focus is on ensuring that a single
+project builds correctly along with all of its dependencies.  They do not address
+larger, multi-user system administration issues such as versioning, naming
+conventions, rapid multi-configuration installation, or maintaining many
+divergent software stacks.
+
+\paragraph{Traditional package managers}
+Package managers automate the installation of complex sets of software packages.
+{\it Binary package managers} such as RPM, yum, and  
+APT~\cite{foster+:rpm03,silva:apt01,yum} are integrated with most
+operating system (OS) distributions, and they are used to ensure that dependencies
+are installed before packages that require them.
+These systems largely solve the problem of managing a single software
+stack, which works well for the baseline OS and drivers, which are 
+common to all applications on a system.
+However, they make the assumption that there is only a single version
+of each package, and most install packages in a single, inflexible location.
+To install multiple configurations, the user must create a custom, combinatoric
+naming scheme to avoid conflicts. They typically require root 
+privileges and do not optimize for specific hardware.
+
+{\it Port systems} such as Gentoo, BSD Ports, MacPorts, and 
+Homebrew~\cite{bsdports,groffen:gentoo-prefix,homebrew,macports,thiruvathukal:gentoo04}
+allow packages to be built on demand, but most suffer from 
+the same versioning and naming issues as traditional package managers.
+Some allow multiple versions to be installed in the same
+prefix~\cite{groffen:gentoo-prefix}, but again the burden is on package
+creators to manage conflicts. This restricts installations to only a few
+configurations.
+
+\paragraph{Virtual Machines and Containers}
+
+Packaging problems arise in HPC because a supercomputer's hardware, OS, and
+filesystem are shared by many users with different requirements.  The classic solution
+to this problem is to use virtual
+machines (VMs)~\cite{barham2003xen,rosenblum1999vmware,smith2005architecture}
+or lightweight virtualization techniques like Linux 
+containers~\cite{felter2014updated,merkel2014docker}. This model allows each user to 
+have her own environment with its own package manager, and it has been extremely 
+successful for servers at cloud data centers.  VMs typically have 
+near-native compute performance, but there are still major performance issues with
+low-level HPC network drivers.  VMs are not well supported on many
+non-Linux operating systems, an issue for the lightweight 
+kernels of bleeding-edge Blue Gene/Q and Cray machines. Security
+is another major concern: a profusion of VMs makes mandatory patching
+difficult when user's environments differ significantly from those of the facility.
+
+\begin{table*}\centering
+\begin{tabular}{|l|l|}
+\hline
+Site           & Naming Convention \\
+\hline
+\hline
+LLNL       & {\tt / usr / global / tools / \$arch / \$package / \$version} \\
+           & {\tt / usr / local~ / tools / \$package-\$compiler-\$build-\$version } \\
+\hline
+Oak Ridge~\cite{jones+:cug08}  & {\tt / \$arch / \$package / \$version / \$build} \\
+\hline
+TACC/Lmod~\cite{mclay:lmod-tutorial}& {\tt / \$compiler-\$comp\_version / \$mpi / \$mpi\_version / \$package / \$version} \\
+\hline
+\hline
+Spack default                  & {\tt / \$arch / \$compiler-\$comp\_version / \$package-\$version-\$options-\$hash} \\
+\hline
+\end{tabular}
+\caption{
+	Software organization of various HPC sites.
+	\label{tab:naming-conventions}
+}
+\end{table*}
+
+\paragraph{Manual and Semi-automated Installation}
+
+To cope with software diversity, many HPC sites use a combination of existing
+packages managers and either manual or semi-automated software installation.  
+For the baseline OS, it is common for sites to maintain traditional binary
+packages using the vendor's package manager.  LLNL maintains a Linux 
+distribution, CHAOS~\cite{chaos} for this purpose.
+
+For more specific builds, many sites adhere to detailed naming conventions
+that encode information in filesystem paths. A sampling of conventions
+at several sites is shown in Table~\ref{tab:naming-conventions}. At LLNL, the
+APT package manager is used for installs in the {\tt /usr/local/tools} 
+filesystem, and {\tt /usr/global/tools} is used for manual installs.
+ORNL uses hand installs but adheres to a strict scripting structure
+that can be used to reproduce each build~\cite{jones+:cug08}.
+TACC relies heavily on locally maintained RPMs for its software installs.
+
+Looking again at the conventions in Table~\ref{tab:naming-conventions},
+we see that most sites use some combination of architecture, compiler version,
+package name, package version, and a custom (up to the author, sometimes
+encoded) build identifier.  TACC and many other sites also include the MPI
+version in the path. None of these naming conventions is defined for
+the entire configuration space, and none defines a way to create new versions
+from arbitrary combinations of dependencies.  MPI is explicitly called out,
+but it is only one of many dependencies that go into a build.
+Sites have made a conscious choice to divide the directory hierarchy
+according to the most common software packages, but they ignore the rest.
+Frequently, desired configurations do not fit the local convention, and
+staff must install them in nonstandard locations.
+
+\paragraph{Environment Modules and RPATHs}
+
+Diverse software versions not only present problems for build and installation; 
+they also complicate the runtime environment. When run, an executable
+must determine the location of its dependency libraries, or it will fail to run.
+This is typically not a problem for statically linked executables, but modern
+operating systems make extensive use of dynamic linking, which has many 
+benefits for software tools and maintenance.
+By default, the dynamic loader on most systems is configured to search only
+system library paths such as {\tt /lib}, {\tt /usr/lib}, and
+{\tt /usr/local/lib}.  If binaries are installed to other locations, the
+{\it user} who runs the program must typically add dependency library paths to
+{\tt LD\_LIBRARY\_PATH} (or a similar environment variable on other OSes).
+Most HPC sites address this problem using {\it environment modules}, which allow
+users to dynamically ``load'' and ``unload'' settings for search paths like
+{\tt LD\_LIBRARY\_PATH}, {\tt PATH}, {\tt MANPATH}, etc.
+
+Environment modules date back to 1991, and there are many implementations 
+available~\cite{dotkit,furlani+:lisa91,furlani+:lisa96,mclay:lmod,mclay:lmod-tutorial}.
+The most advanced implementations, such as Lmod~\cite{mclay:lmod,mclay:lmod-tutorial},
+provide software hierarchies similar to the naming conventions in 
+Table\ref{tab:naming-conventions}, and they allow users to quickly load a software stack
+if they know which one is required.  TACC expects the user to run with the same
+environment that she built with~\cite{mclay:lmod-tutorial}.  But, at LC, the user is
+frequently not the builder of the application, and may not know the particular
+incantation required to load the its environment. Moreover, users frequently run more 
+than one application in the same job, each built by a separate team, and each with
+its own environment. Most LC users are used to running programs with
+specific paths; only 10\% use environment modules extensively.
+
+The alternative to per-user environment settings is to embed library search paths
+in installed binaries at compile time. When set this way, the search path is called
+an {\tt RPATH}. {\tt RPATHs} and environment modules are not mutually exclusive.
+Modules can still be used to set {\tt MANPATH}, {\tt PATH} and variables, but
+adding {\tt RPATHs} ensures that binaries will run correctly regardless of whether
+the right module is loaded. LC installs software with both {\tt RPATHs} and
+{\tt dotkit}~\cite{dotkit} modules.
+
+\paragraph{Modern Package Managers}
+
+Recently, a number of package managers have been developed to manage 
+multi-configuration builds.
+%
+Nix~\cite{dolstra+:icfp08,dolstra+:lisa04}
+is a package manager and an OS distribution that supports the installation of
+arbitrarily many software configurations.  As at most HPC sites, each package
+in Nix installed in a unique prefix, but in Nix, there is no human-readable
+naming convention.  Instead, Nix determines the prefix by hashing the package
+file and all of its dependencies.
+%
+Hashdist~\cite{hashdist} is a meta-build system and package manager for HPC.
+Like Nix, it uses cryptographic versioning, and it archives installations
+in a git repository so that they can be recalled on demand. 
+%
+Both Nix and hashdist use {\tt RPATHs} in their packages to ensure that
+libraries are found correctly.
+
+The EasyBuild~\cite{hoste+:pyhpc12} tool is in production use at
+the University of Ghent.  It allows multiple versions to be installed at once,
+but it requires package authors to manage combinatoric package
+versioning. Rather than setting {\tt RPATHs}, it auto-generates module files
+to manage each package's environment, and it is closely coupled with 
+Lmod~\cite{geimer+:hust14}.  EasyBuild groups
+the compiler, MPI, FFT, and BLAS libraries together in a
+{\it toolchain}, providing limited package composability. 
+Smithy~\cite{digirolamo:smithy} is a similar tool in use at ORNL. It
+can generate module files, but it does not provide any 
+automated dependency management; it only checks whether a package's
+prerequisites have already been installed by the user.
+
+We observed a number of shortcomings in these existing systems. While the
+cryptographic versioning of Nix and Hashdist is appealing, it does not allow
+users to easily navigate the installed software or to find packages consistent
+with their own setup.  In fact, the only way to access software after it is built
+is to recall it by its hash. EasyBuild and Smithy address some of these problems
+by providing modules for the installed packages, but neither has a good system
+for managing the full space of combinatoric names.  EasyBuild attempts to do this by
+lumping versions of common packages together in a tooclhain, but the toolchains
+themselves have cryptic names like {\tt goolf} 
+(\textbf{g}cc, \textbf{o}penmpi, \textbf{o}penblas, \textbf{S}caLAPACK,
+\textbf{F}FTW), and their versions are meaningless. {\tt RPATH} handling, where
+provided, is left up to the package authors, and is not consistently enforced.
+
+The main limitation of existing tools is the lack of build {\it composability}.
+HPC configuration space is combinatoric, and realizing an arbitrary
+combination of compilers, MPI version, build options, and dependency versions
+requires tediously modifying many packages implementation.  
+No existing system parameterizes 
+package files to deal with this issue.  Indeed, the number of package files
+required for existing systems scales with the number of version {\it combinations},
+not the number of packages.  This quickly becomes unmanageable.  Indeed, EasyBuild
+maintains over 3,300 files for only 600 packages.  Other systems have similar
+issues but limit the number of supported configurations.

### 2015-04-13 motivation.tex
rework gaps.

@@
 naming convention.  Instead, Nix determines the prefix by hashing the package
-file and all of its dependencies.
-%
+file and along with its dependencies.
+
 Hashdist~\cite{hashdist} is a meta-build system and package manager for HPC.
@@
 The EasyBuild~\cite{hoste+:pyhpc12} tool is in production use at
-the University of Ghent.  It allows multiple versions to be installed at once,
-but it requires package authors to manage combinatoric package
-versioning. Rather than setting {\tt RPATHs}, it auto-generates module files
+the University of Ghent.  It allows multiple versions to be installed
+at once.  Rather than setting {\tt RPATHs}, it auto-generates module files
 to manage each package's environment, and it is closely coupled with 
-Lmod~\cite{geimer+:hust14}.  EasyBuild groups
-the compiler, MPI, FFT, and BLAS libraries together in a
-{\it toolchain}, providing limited package composability. 
+Lmod~\cite{geimer+:hust14}.  EasyBuild groups the compiler, MPI, FFT, and
+BLAS libraries together in a {\it toolchain}, which can be combined with
+any package file. This provides some degree of composability and
+separates compiler flags and MPI concerns from client packages.
+
 Smithy~\cite{digirolamo:smithy} is a similar tool in use at ORNL. It
@@
 
-We observed a number of shortcomings in these existing systems. While the
-cryptographic versioning of Nix and Hashdist is appealing, it does not allow
-users to easily navigate the installed software or to find packages consistent
-with their own setup.  In fact, the only way to access software after it is built
-is to recall it by its hash. EasyBuild and Smithy address some of these problems
-by providing modules for the installed packages, but neither has a good system
-for managing the full space of combinatoric names.  EasyBuild attempts to do this by
-lumping versions of common packages together in a tooclhain, but the toolchains
-themselves have cryptic names like {\tt goolf} 
-(\textbf{g}cc, \textbf{o}penmpi, \textbf{o}penblas, \textbf{S}caLAPACK,
-\textbf{F}FTW), and their versions are meaningless. {\tt RPATH} handling, where
-provided, is left up to the package authors, and is not consistently enforced.
+\paragraph{Gaps in the state of the art}
+The cryptographic versioning of Nix and Hashdist is very flexible.  It versions
+the package {\it and} its dependency configuration, and can represent any
+configuration. However, there is no way for users to navigate or query
+software. The systems are in a sense ``write-only'':
+Nix does not even offer a way to ask for a package's dependencies.
+%
+EasyBuild and Smithy generate environment modules, which allow querying
+through their own interface. Naming in existing module systems, however, cannot
+handle combinatoric versions. The Lmod authors call this the
+``matrix problem``~\cite{mclay:lmod-tutorial}.
+
+%EasyBuild's attempts to version groups dependencies by adding versions to
+%toolchains, but the naming is difficult to understand.
+%. Compiler, MPI, and 
+%some library versions are lumped together, but the results is cryptic, e.g.
+%{\tt goolf} stands for ``gcc, openmpi, openblas, ScaLAPACK, FFTW'', and
+%its version is meaningless.
+
+Existing tools do not enforce {\tt RPATHs}, leaving them to package authors.
+This may lead to erroneous runs.
 
 The main limitation of existing tools is the lack of build {\it composability}.
-HPC configuration space is combinatoric, and realizing an arbitrary
-combination of compilers, MPI version, build options, and dependency versions
-requires tediously modifying many packages implementation.  
-No existing system parameterizes 
-package files to deal with this issue.  Indeed, the number of package files
-required for existing systems scales with the number of version {\it combinations},
-not the number of packages.  This quickly becomes unmanageable.  Indeed, EasyBuild
-maintains over 3,300 files for only 600 packages.  Other systems have similar
-issues but limit the number of supported configurations.
+The full set of package versions is combinatoric, and realizing an arbitrary
+combination of compiler, MPI version, build options, and dependency versions
+requires tediously modifying many package files.
+%
+Indeed, the number of package files required for existing systems scales with
+the number of version {\it combinations}, not the number of packages.
+This quickly becomes unmanageable.  As an example, the very flexible EasyBuild
+system has over 3,300 files for several permutations of only 600 packages. 
+Other systems have similar issues. A mechanism is needed to {\it parameterize}
+packages so that new builds can be {\it composed} on the fly.

### 2015-04-13 implementation.tex
rework gaps.
Minor updates, start of implementation section.

@@
 
-\section{Implementation}
+%\begin{figure}
+%\includegraphics[width=\columnwidth]{figs/spack-find-output.pdf}
+%\end{figure}
+
+\section{The Spack Package Manager}
 \label{sec:implementation}
 
-\todo{.5 page overview}
+We have developed {\it Spack}, the Supercomputing PACKage manager, based on our
+software needs Lawrence Livermore National Laboratory.
+Spack solves the software matrix problem by allowing {\it parametric} builds.
+Each package in Spack is parameterized by its name, version, architecture, compiler, 
+compiler version, build options, and by its dependencies.
+Like other systems, Spack allows multiple build variants to be installed at once,
+but it provides novel, recursive {\it spec} syntax.  Specs are concise
+expressions for partially constrained software dependence graphs. The same
+spec syntax can be used on the command line to request new build 
+configurations, or to query installed packages over many dimensions.
+
+Spack borrows ideas from Nix to support multiple installations with 
+cryptographic hashes, but it goes further and provides a build environment
+that takes many burdens off of the package writer.  In particular, Spack uses
+compiler wrappers to automatically add {\tt RPATHs} for {\it all} dependencies
+to installed binaries, so that package authors do not have to.  Compiler
+wrappers also give the packager more control over the build: package authors
+can more easily port to new platforms by filtering compiler flags.
 
@@
 	\todo{1 page}
+
+%\begin{figure}
+%\begin{grammar}
+%  <spec-list> ::= \{ <spec> [ <dep-list> ] \}
+%  
+%  <dep-list>  ::= \{ `\textsf \textasciicircum' <spec> \}
+%
+%  <spec>      ::= <id> [ options ]
+%  
+%  <options>   ::= \{ `@' version-list | `+' <variant> \newline
+%                  | `-' <variant> ~~~| `~' <variant> \newline
+%                  | `\%' <compiler> ~| `=' <architecture> \}
+%                  
+%  <variant>      ::= <id>
+%
+%  <architecture> ::= <id>
+%
+%  <compiler>     ::= <id> [ <version-list> ]
+%
+%  <version-list> ::= <version> [ \{ `,' version \} ]
+%
+%  <version>      ::= <id> | <id> `:' | `:' <id> | <id> `:' <id>
+%
+%  <id>           ::= [A-Za-z0-9_][A-Za-z0-9_.-]*  
+%
+%\end{grammar}
+%\end{figure}
 
@@
 \subsection{Abstract \& Concrete Specs}
-	\todo{1 page}
+	\todo{.75 page}
 

### 2015-04-13 usecase-policy.tex
First draft of policy usecase

@@
 
-\subsubsection{Handling site installs and policies}
+Spack makes it easy for package maintainers to create and organize package installations.
+But it also needs to be easy for end-users to find and use those packages.
+Different end-users may have different expectations about how packages should be built and installed, and those expectations are typically shaped by years of site policies, personal preferences, and lingering legacy decisions originallly made on a DEC VAX.
+Rather than try to dictate these expectations, Spack allows both end-users and package maintainers to create custom policies dictating how packages are built and installed at their site.
 
-\todo{1 pgs}
+\subsection{Package Views}
+\label{sec:package-views}
 
-\begin{verbatim}
-Putting things together for the end user
-		- Site installs and policies
-		- views
-		- customizable defaults
-		- site-specific policies
-			- prefer mvapich, compiler, version
-\end{verbatim}
+\todo{fill in citations below}
+While Spack can easily create a dozen installs of a package like MPILeaks across different compilers and MPI implementations, it can be difficult for an end-user to determine which of those dozen packages should be used with their application.
+The module (cite) or dotkit (cite) package managers are commonly used on supercomputer systems to solve this problem, and Spack allows the package administrator to automatically create module and dotkit files for packages Spack installs.
 
-\subsubsection{Allowing the user to override}
+However, even when modules and dotkits are available many users will go directly through the filesystem to access package installs.
+Spack installs packages into paths based on concretized specs, which is ideal for maintaining multiple package installations, but may be difficult for an end-user to navigate.
+A version of MPILeaks, for example, may be installed in a location like {\tt spack/opt/chaos\_5\_x86\_64\_ib/ gcc@4.9.2/mpileaks@1.0-db465029}.
 
-\todo{.5 pgs}
+Spack thus allows the creation of views, which are a symlink-based directory layout of packages.
+Views provide a human-readable directory layout, which can be taylored to fit alongside existing directory layouts or used to create new layouts.
+For example, the above MPILeaks package may have a view that creates a symlink in {\tt /opt/mpileaks-1.0-openmpi} to the spack installation of MPILeaks.
+The same package install may be referenced by multiple symlinks and views, so the above package could also be symlinked from a more generic {\tt /opt/mpileaks-openmpi} symlink (which is useful for users who don't want to hardcode towards a specific version).
+Views can also be used to create symlinks to specific executables or libraries in an install, so a spack-built gcc 4.9.2 install may have a view that creates symlinks from {\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.
 
-\begin{verbatim}
-	- user can build his own entire stack without site support
-	- user can point to site installs or build entire stack
-\end{verbatim}
+Views are configured through a spackconfig file, which can be setup at a site-wide level or for individual users.
+For each package or set of packages, the spackconfig file contains rules describing the symlinks that should point into that package.
+The symlink names can be parameterized.
+For example, the above MPILeaks symlink might have been created by {\tt /opt/\$\{PACKAGE\}-\$\{VERSION\}-\$\{MPINAME\} } rule.
+When spack installs or removes packages, symlinks are automaticaly created, deleted, or updated according to the rules.
+
+Spack's views are a projection from a point in a high-dimension space (the concretized spec, which will have all parameters fully specified) to point in a lower-dimension space (the symlink name, which may only specify a few parameters).
+Thus several package installations may map to the same symlink.
+For example, the above MPILeaks symlink could point to an MPIleaks compiled with gcc or icc--the compiler parameter is not part of the symlink.
+To keep  package installations  consistent and reproducible, Spack has a well-defined mechanism for resolving conflicting symlinks.
+Spack defines a sort order for packages, which arranges packages from ``most desired'' to ``least desired''.
+When a symlink could resolve to multiple package installations, Spack points the symlink at the most desired package.
+
+By default, Spack treats newer versions of packages compiled with newer compilers as more desirable than older packages built with older compilers.
+It has well-defined, but not necessarily meaningful, preferences for deciding between MPI implementations and different compilers.
+A user can override Spack's default sort order through their spackconfig file.
+For example, at one site users may typically use the intel compiler, but some users also use system's default gcc 4.4.7.
+These preferences could be stated by adding the line {\tt compiler\_order = icc,gcc@4.4.7} to the site's spackconfig file, which would cause the ambigious MPILeaks symlink to point to an ICC installation.
+Any compiler left out of the compiler\_order setting is treated as less prefered than explicitly stated compilers.
+Spack can also be configured to treat specific package versions as more prefered than other versions, which can be useful if a new version is unstable and untested, and should not be advertised with symlinks.
+
+\subsection{Package Configurations}
+
+It is also useful to specify how site-specific options for how packages should be built.
+The correct configure line for building packages may be different across sites.
+For example, the Spindle tool can have logging locations and security policies specified on its configure line.
+Spack can make reasonable guesses about Spindle's proper configure line, but it can not created a customized installation for the site.
+One way to solve this would be to edit the Spindle package description in Spack and manually change how configure is invoked.
+This is not ideal, as the Spindle package description is part of the Spack source code, and modifying it makes it difficult to share spack versions and updates across sites.
+
+Instead Spack allows the creation of site-specific variants of packages.
+The build instructions for a spack package are encoded in a Python class, which must inherit from Spack's {\tt Package} class.
+Via the spackconfig file, users can specify additional search directories for finding additional {\tt Package} classes.
+These additional packages can inherit from and replace the Spack's default packages, allowing sites to either tweak or completely replace Spack's build recipes.
+To continue the previous example, a site can write a {\tt LocalSpindle} Python class, which inherits from Spack's {\tt Spindle} class.
+{\tt LocalSpindle} may simply add additional configure flags to the {\tt Spindle} class, while leaving the dependencies and most of the build instructions from its parent class.
+To keep builds deterministic and reproducable, Spack also tracks which {\tt Package} class drove a specific build.

### 2015-04-15 implementation.tex
Checking in what I've got so far -- lots to do.

--- 2015-04-11 implementation.tex
@@
 
-\section{Implementation}
+\section{The Spack Package Manager}
 \label{sec:implementation}
+Based on our experiences at LLNL, we have developed
+{\it Spack}, the Supercomputing PACKage manager.
+Spack is written in Python.  We chose Python because it is
+very widely used int he HPC community, and because it provides
+flexible scripting capabilities. 
+%
+Like some prior systems, Spack supports arbitrarily many software installations.
+It borrows ideas from Nix; packages are installed in their own prefixes,
+and configurations are managed using hashes.
+% 
+Spack adds several unique capabilities that are essential for HPC:
+\begin{enumerate}
+\item {\bf Parametric builds} address the matrix problem:
+      Packages are parameterized by name, version, architecture, compiler, 
+      options, and dependencies.
+\item To manage the configuration space, spack provides a novel, 
+      recursive {\bf spec syntax} for dependency graphs.
+\item Specs support versioned, ABI-incompatible interfaces like MPI through
+      {\bf versioned virtual dependencies}.
+\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
+      consistency on the package author.
+\end{enumerate}
 
-\todo{.5 page overview}
+\subsection{Packages}
 
-\subsection{Spack Specs}
-	\todo{1 page}
+%\subsubsection{Package Files}
+\begin{figure}
+	\includegraphics[width=\columnwidth]{code/mpileaks.pdf}
+	\caption{
+		Spack package for the {\tt mpileaks} tool.
+		\label{fig:mpileaks}
+	}
+\end{figure}
 
-\subsection{Versioned Virtual Dependencies}
-	\todo{.5 page}
+Spack comprises many packages, and each describes how to build a software
+artifact.  Each package is a class written in pure Python, but Spack uses
+a simple, embedded domain-specific language (DSL) to make packaging easier.
+Spack's DSL adds directives such as {\tt depends\_on}, {\tt version},
+{\tt provides}, and {\tt extends}, in order to add metadata to package classes.
 
-\subsection{Abstract \& Concrete Specs}
-	\todo{1 page}
+Figure~\ref{fig:mpileaks} shows the package for the \mpileaks tool.
+Inside the {\tt MpiLeaks} class, the package provides a text description
+and a homepage, as well as 
+a download URL.  Notably, Spack packages can have {\it multiple} version
+directives; each identifies a version and its checksum, allowing it to 
+be downloaded and installed safely. Below this, three {\tt depends\_on}
+directives indicate prerequisites that must be installed before \mpileaks.
 
-\subsection{Install Environment}
+\subsubsection{Installation Environment}
 
-\subsubsection{Compiler Wrappers \& RPATHs}
-	\todo{.5 page}
+Existing package managers leave much up to package authors, and they 
+frequently do
+not always have
+
+Spack has a number of capabilities that make package writing more 
+convenient. It can extrapolate a URL for each version using the {\tt url}
+attribute as a model\footnote{This works for consistently named URLs,
+and it can be specified per-version for packages with inconsistent naming.}.
+It can use the same model to scrape webpages for new versions. 
+Boilerplate for new packages can be generated from a URL.
+
+Each package defines an {\tt install()} method, which is responsible
+for its build. The {\tt install} environment is designed to look like
+custom build scripts that LLNL staff have written for years.  Based on
+past experience, we have tailored Spack's build environment to prevent common
+build errors.
+
+{\bf Familiar command syntax.} While Spack's Python DSL adds power and 
+structure to the build process, LLNL staff are used to writing 
+builds as shell scripts.  In Spack, shell commands can be invoked as
+Python functions, which makes builds readable for package authors.
+The {\tt mpileaks} build in Figure~\ref{fig:mpileaks} uses a 
+shell-like installation procedure with it {\tt configure},
+{\tt make}, and {\tt make install} calls.
+
+{\bf Environment isolation.}
+A common build error is to link in the wrong version of a library because
+of an incorrect build environment.
+%
+For example, there are two versions of the {\tt libelf} library used by
+LLNL performance tools. One is distributed with RedHat Linux, but the
+publicly available version uses the same API but an incompatible ABI.
+Failing to specify the right version can cause tools to crash.
+%
+Most build systems require users to manually specify paths to dependencies
+installed in nonstandard locations.  This includes {\it most} software used
+by HPC users.  However, depending on the package's particular build system,
+libraries may be discovered
+in different ways, either through special environment variables, compiler tests,
+or manual specification in a Makefile or on the {\tt configure} line.
+
+Spack manages the build environment by running each call to {\tt install}
+in a new process.  This gives package authors full control of the build
+process, and they can set build-specific environment variables without 
+interfering with other packages.  It also helps packages find dependencies 
+correctly, by setting
+{\tt PATH}, {\tt PKG\_CONFIG\_PATH}, {\tt CMAKE\_PREFIX\_PATH}, and
+{\tt LD\_LIBRARY\_PATH} to include the dependencies of the current build.
+These variables are commonly used by build systems to locate dependencies,
+and setting them ensures that incorrect libraries are not detected.
+
+{\bf Compiler wrappers and RPATHs.}
+
+The build scripts of existing packages managers are largely compiler-specific.
+Most OS-specific package managers standardize on the same compiler used to build
+the operating system, and offer some support for other compilers as an afterthought.
+EasyBuild allows packages to be built with multiple toolchains, but it again
+requires separate package files for each compiler.  Moreover, existing
+package managers and LLNL's build scripts rely largely on package authors
+to configure link options like {\tt RPATH}.  The setting is not enforced across
+different packages, and enforcing a site-wide policy for link options is 
+difficult.  If a package author forgets to add an {\tt RPATH}, the binary
+can fail to run correctly.  Worse, a package maintainer may not detect this
+after installing, because the package may work perfectly well in {\it her} 
+environment but not that of other users.
+
+Spack goes a step further and uses {\it compiler wrappers} to enforce
+consistent link policies across builds.
+
+ a {\tt RPATHs} are ommitted
+
+To make library finding more consistent,
+
+Also in the build environment, Spack sets the standard environment variables
+{\tt CC}, {\tt CXX}, {\tt F77}, and {\tt FC} to point to custom compiler
+wrapper scripts.  When run, the wrappers insert include ({\tt -I}), 
+library ({\tt -L}), and {\tt RPATH} flags for all of the dependencies of
+the package being built. This causes build tests like those in {\tt configure}
+scripts to succeed.
+
+These variables are used by most most modern build systems
+to select the compilers to use for a build, so they are typically picked up
+by default.
+
+we must still add custom arguments to configuration scripts, as done for
+{\tt callpath} and {\tt adept-utils} in Figure~\ref{fig:mpileaks}, but
+
+These environment
+variables are respected by most modern build systems
+
+ and {\tt -Wl,-rpath} flags 
+aware of the location of the dependencies
+
+ will be invoked
+after all of its prerequisites are built and installed.  takes the burden of
+build consistency {\it off} of package authors by providing compiler 
+wrappers that automatically set {\tt RPATHs} and guarantee that packages
+find their dependencies.  Compiler wrappers also give package authors
+more control over build options like compiler flags.
+
+%The flags used to specify dependency locations are different from package
+%to package, and build scripts frequently get these wrong.
+%For example, in some LLNL packages that depend on the {\tt Silo} library,
+%the {\tt --with-silo} parameter takes a path to {\tt Silo}'s prefix, but
+%in others it takes both the {\tt include} and {\tt lib} subdirectories,
+%separated by a comma.
 
@@
 \end{verbatim}
+
+\subsection{Spack Specs}\label{sec:specs}
+
+  Specs are partial descriptions of full
+dependency graphs, but they are concise.  This allows a user to use
+only a few parameters to quickly identify a configuration to install,
+or to query existing installations and their dependencies.
+
+\todo{1 page}
+
+\subsection{Versioned Virtual Dependencies}
+	\todo{.5 page}
+
+\subsection{Abstract \& Concrete Specs}
+	\todo{.75 page}

### 2015-04-16 implementation.tex
Updated environemnt

--- 2015-04-11 implementation.tex
@@
 
-\section{Implementation}
+\section{The Spack Package Manager}
 \label{sec:implementation}
+Based on our experiences at LLNL, we have developed
+{\it Spack}, the Supercomputing PACKage manager.
+Spack is written in Python.  We chose Python because it is
+very widely used int he HPC community, and because it provides
+flexible scripting capabilities. 
+%
+Like some prior systems, Spack supports arbitrarily many software installations.
+It borrows ideas from Nix; packages are installed in their own prefixes,
+and configurations are managed using hashes.
+% 
+Spack adds several unique capabilities that are essential for HPC:
+\begin{enumerate}
+\item {\bf Parametric builds} address the matrix problem:
+      Packages are parameterized by name, version, architecture, compiler, 
+      options, and dependencies.
+\item To manage the configuration space, spack provides a novel, 
+      recursive {\bf spec syntax} for dependency graphs.
+\item Specs support versioned, ABI-incompatible interfaces like MPI through
+      {\bf versioned virtual dependencies}.
+\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
+      consistency on the package author.
+\end{enumerate}
 
-\todo{.5 page overview}
+\subsection{Packages}
 
-\subsection{Spack Specs}
-	\todo{1 page}
+%\subsubsection{Package Files}
+\begin{figure}
+	\includegraphics[width=\columnwidth]{code/mpileaks.pdf}
+	\caption{
+		Spack package for the {\tt mpileaks} tool.
+		\label{fig:mpileaks}
+	}
+\end{figure}
 
-\subsection{Versioned Virtual Dependencies}
-	\todo{.5 page}
+Spack comprises many packages, and each describes how to build a software
+artifact.  Each package is a class written in pure Python, but Spack uses
+a simple, embedded domain-specific language (DSL) to make packaging easier.
+Spack's DSL adds directives such as {\tt depends\_on}, {\tt version},
+{\tt provides}, and {\tt extends}, in order to add metadata to package classes.
 
-\subsection{Abstract \& Concrete Specs}
-	\todo{1 page}
+Figure~\ref{fig:mpileaks} shows the package for the \mpileaks tool.
+Inside the {\tt MpiLeaks} class, the package provides a text description
+and a homepage, as well as 
+a download URL.  Next, two {\tt version} directives identify known versions
+of the package, and checksums so that they can be downloaded safely.
+Below this, three {\tt depends\_on}
+directives indicate prerequisites that must be installed before \mpileaks.
+Last, each package defines an {\tt install()} method, which contains the
+commands used to build the package.
 
-\subsection{Install Environment}
+Most package managers fix a single version in a package file, but Spack packages
+are {\it parameterized} by version.  In the figure, we see that the {\tt mpileaks}
+package specifies two versions.  A user can safely install either by 
+specifying it on the command line.  In our experience, however, users
+frequently want bleeding-edge versions, and package managers frequently lag 
+behind the latest releases.  Spack therefore has a capability to extrapolate
+the latest URL, using the {\tt url} attribute as a model\footnote{This works
+for consistently named URLs, and it can be specified per-version for packages
+with inconsistent naming.}.
+The user can ask for an unknown version on the command line and still
+successfully download and install it.  Spack can also use the same model to
+scrape webpages and find new versions.
 
-\subsubsection{Compiler Wrappers \& RPATHs}
-	\todo{.5 page}
+\subsubsection{Installation Environment}
 
-\subsubsection{Potential for auto-generation of modules (with deps)}
+%{\bf Reproducibility.}
+%\paragraph{Reproducibility}
+Spack is intended to build a consistent HPC stack for our multi-user
+environment, and reproducible builds are one of our design goals.
+Experience at LLNL has shown that it is vexingly difficult to reproduce
+a build manually.
+%
+Many packages used at LLNL have a profusion of build options, and specifying them 
+correctly often requires tedious experimentation.  This is due to lack of
+build standards and to the diversity of HPC environments.  
+For example, in some packages that depend on the {\tt Silo} library,
+the {\tt ---with-silo} parameter takes a path to {\tt Silo}'s prefix.
+In others, it takes the {\tt include} and {\tt lib} subdirectories,
+separated by a comma.
+The {\tt install()} method in Spack's package files allows us to record
+precise build incantations for later reuse.
+
+%{\bf Environment isolation.}
+\paragraph{Environment isolation}
+In addition to command line issues, we
+frequently encounter errors due to inconsistencies between the environment of
+the package installer and the package user.
+%
+For example, there are two versions of the {\tt libelf} library used by
+LLNL performance tools. One is distributed with RedHat Linux, but another
+publicly available version has the same API but an incompatible ABI.
+Failure to specify the right version at build time has caused many
+inexplicable crashes.
+%
+Spack manages the build environment by running {\tt install} invocation
+in a new process.  It helps packages find dependencies 
+correctly, by setting
+{\tt PATH}, {\tt PKG\_CONFIG\_PATH}, {\tt CMAKE\_PREFIX\_PATH}, and
+{\tt LD\_LIBRARY\_PATH} to include the dependencies of the current build.
+These variables are commonly used by build systems to locate dependencies,
+and setting them helps to ensure that incorrect libraries are not detected.
+The isolated build environment also gives package authors 
+free reign to set build-specific environment variables without interfering
+with other packages.
+
+%{\bf Compiler wrappers and RPATHs.}
+\paragraph{Compiler wrappers and RPATHs}
+Finding compilers at build time is not the only obstacle to reproducible
+behavior.  As mentioned in Section~\ref{sec:motivation}, it is also important
+for binaries to be able to find dependency libraries at {\it runtime}.
+One of the most frequent user errors at LC is improper library configuration.
+Users frequently do not know what libraries a package was built with, and 
+it is difficult for them to construct a suitable {\tt LD\_LIBRARY\_PATH} for
+a package that was built by someone else.  Because of frequent support calls,
+we typically add {\tt RPATHs} to public software installations, so that paths
+to dependencies are embedded in binaries and so that users do not have to know
+this information to run installed software correctly.
+
+Spack manages {\tt RPATH} settings and other build policies with
+{\it compiler wrappers}. 
+In each isolated {\tt install} environment, Spack sets the standard 
+environment variables
+{\tt CC}, {\tt CXX}, {\tt F77}, and {\tt FC} to point to its own compiler
+wrapper scripts.  These variables are used by most build systems to select
+C, C++, and Fortran compilers, so they are generally picked up 
+automatically\footnote{If builds do not respect {\tt CC}, {\tt CXX}, etc.,
+wrappers can typically be added as arguments or inserted into Makefiles
+by {\tt install}.}.
+When run, the wrappers insert include ({\tt -I}), library ({\tt -L}), and 
+{\tt RPATH} ({\tt -Wl,-rpath} or similar) flags into the argument list.
+These point to the {\tt include} and {\tt lib} directories of dependency
+library installations, where needed headers and libraries are typically located.
+After the modified argument list is constructed, the wrappers delegate it
+to the real compiler to execute.
+
+Spack's compiler wrappers have a number of useful effects.  First, they allow
+Spack to transparently parameterize the compiler for most builds. 
+Spack users to easily request a build with a new compiler swapped in.
+Second, they enforce the use of {\tt RPATHs} in
+installed binaries.  This causes applications built by Spack to run correctly
+{\it regardless of the environment}.  Third, because compiler wrappers add 
+header and library search paths for dependencies, header and library detection
+tests run by most build systems succeed automatically, {\it without}
+the need to use special arguments for nonstandard locations.  {\tt configure}
+commands in Spack's {\tt install} function can have fewer arguments, and can
+be written as they would be for system installs.  This reduces complexity
+for package maintainers and enforces consistent, reproducible
+build policies across packages.  Finally, because Spack has control over the 
+wrappers, package authors can programmatically filter the compiler flags
+used by software build systems, a useful last resort when porting to
+bleeding-edge platforms or new, esoteric compilers.
+
+\subsubsection{Environment Module Integration}
 	\todo{.5 page}

### 2015-04-16 implementation-packages.tex
Move packages subsection to separate file.

--- 2015-04-16 implementation.tex
@@
 %!TEX root = paper.tex
-
-\section{The Spack Package Manager}
-\label{sec:implementation}
-Based on our experiences at LLNL, we have developed
-{\it Spack}, the Supercomputing PACKage manager.
-Spack is written in Python.  We chose Python because it is
-very widely used int he HPC community, and because it provides
-flexible scripting capabilities. 
-%
-Like some prior systems, Spack supports arbitrarily many software installations.
-It borrows ideas from Nix; packages are installed in their own prefixes,
-and configurations are managed using hashes.
-% 
-Spack adds several unique capabilities that are essential for HPC:
-\begin{enumerate}
-\item {\bf Parametric builds} address the matrix problem:
-      Packages are parameterized by name, version, architecture, compiler, 
-      options, and dependencies.
-\item To manage the configuration space, spack provides a novel, 
-      recursive {\bf spec syntax} for dependency graphs.
-\item Specs support versioned, ABI-incompatible interfaces like MPI through
-      {\bf versioned virtual dependencies}.
-\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
-      consistency on the package author.
-\end{enumerate}
 
@@
 \end{verbatim}
-
-\subsection{Spack Specs}\label{sec:specs}
-
-  Specs are partial descriptions of full
-dependency graphs, but they are concise.  This allows a user to use
-only a few parameters to quickly identify a configuration to install,
-or to query existing installations and their dependencies.
-
-\todo{1 page}
-
-\subsection{Versioned Virtual Dependencies}
-	\todo{.5 page}
-
-\subsection{Abstract \& Concrete Specs}
-	\todo{.75 page}

### 2015-04-16 implementation.tex
Move packages subsection to separate file.

--- 2015-04-13 implementation.tex
@@
 %!TEX root = paper.tex
-
-%\begin{figure}
-%\includegraphics[width=\columnwidth]{figs/spack-find-output.pdf}
-%\end{figure}
 
@@
 \label{sec:implementation}
+Based on our experiences at LLNL, we have developed
+{\it Spack}, the Supercomputing PACKage manager.
+Spack is written in Python.  We chose Python because it is
+very widely used int he HPC community, and because it provides
+flexible scripting capabilities. 
+%
+Like some prior systems, Spack supports arbitrarily many software installations.
+It borrows ideas from Nix; packages are installed in their own prefixes,
+and configurations are managed using hashes.
+% 
+Spack adds several unique capabilities that are essential for HPC:
+\begin{enumerate}
+\item {\bf Parametric builds} address the matrix problem:
+      Packages are parameterized by name, version, architecture, compiler, 
+      options, and dependencies.
+\item To manage the configuration space, spack provides a novel, 
+      recursive {\bf spec syntax} for dependency graphs.
+\item Specs support versioned, ABI-incompatible interfaces like MPI through
+      {\bf versioned virtual dependencies}.
+\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
+      consistency on the package author.
+\end{enumerate}
 
-We have developed {\it Spack}, the Supercomputing PACKage manager, based on our
-software needs Lawrence Livermore National Laboratory.
-Spack solves the software matrix problem by allowing {\it parametric} builds.
-Each package in Spack is parameterized by its name, version, architecture, compiler, 
-compiler version, build options, and by its dependencies.
-Like other systems, Spack allows multiple build variants to be installed at once,
-but it provides novel, recursive {\it spec} syntax.  Specs are concise
-expressions for partially constrained software dependence graphs. The same
-spec syntax can be used on the command line to request new build 
-configurations, or to query installed packages over many dimensions.
+\input{implementation-packages}
 
-Spack borrows ideas from Nix to support multiple installations with 
-cryptographic hashes, but it goes further and provides a build environment
-that takes many burdens off of the package writer.  In particular, Spack uses
-compiler wrappers to automatically add {\tt RPATHs} for {\it all} dependencies
-to installed binaries, so that package authors do not have to.  Compiler
-wrappers also give the packager more control over the build: package authors
-can more easily port to new platforms by filtering compiler flags.
+\subsection{Spack Specs}\label{sec:specs}
 
-\subsection{Spack Specs}
+  Specs are partial descriptions of full
+dependency graphs, but they are concise.  This allows a user to use
+only a few parameters to quickly identify a configuration to install,
+or to query existing installations and their dependencies.
+
 \todo{1 page}
-
-%\begin{figure}
-%\begin{grammar}
-%  <spec-list> ::= \{ <spec> [ <dep-list> ] \}
-%  
-%  <dep-list>  ::= \{ `\textsf \textasciicircum' <spec> \}
-%
-%  <spec>      ::= <id> [ options ]
-%  
-%  <options>   ::= \{ `@' version-list | `+' <variant> \newline
-%                  | `-' <variant> ~~~| `~' <variant> \newline
-%                  | `\%' <compiler> ~| `=' <architecture> \}
-%                  
-%  <variant>      ::= <id>
-%
-%  <architecture> ::= <id>
-%
-%  <compiler>     ::= <id> [ <version-list> ]
-%
-%  <version-list> ::= <version> [ \{ `,' version \} ]
-%
-%  <version>      ::= <id> | <id> `:' | `:' <id> | <id> `:' <id>
-%
-%  <id>           ::= [A-Za-z0-9_][A-Za-z0-9_.-]*  
-%
-%\end{grammar}
-%\end{figure}
 
@@
 	\todo{.75 page}
-
-\subsection{Install Environment}
-
-\subsubsection{Compiler Wrappers \& RPATHs}
-	\todo{.5 page}
-
-\subsubsection{Potential for auto-generation of modules (with deps)}
-	\todo{.5 page}
-
-\begin{verbatim}
-module disadvantages:
-	- lmod solves a lot of problems
-	- not on all systems
-		- users often run from just a path.
-		- not guaranteed that LD_LIBRARY_PATH, etc will
-		  be set correctly.
-
-- Lmod assumes build & submit environment must match
-	- users often have NO IDEA what environment a package
-	  was built with
-	- multiple apps may be built with different stacks
-	- combining them in a workflow is difficult with modules
-		(need to know right incantation)
-
-	- TACC implements a rather complicated mapping back to
-	    the build env (Lmod CheckExec)
-
-- Spack solves the matrix problem mentioned in the Lmod
-    presentation (slide 98)
-	- hierarchy is nasty -- every branch represents a
-	  greedy choice
-	- might NEED a specific version of ONE dependency
-	  that is not supported in the tree
-	  => SPACK!
-
-- libmesh
-	- boost, petsc, trilinos, grvy, etc.
-\end{verbatim}

### 2015-04-16 abstract.tex
intro updates.

@@
 Large HPC centers spend considerable time supporting software for thousands of users, but the complexity of HPC software is quickly outpacing the capabilities of existing software management tools.
-Scientific applications require specific versions of compilers, MPI implementations, and dependency libraries, so using a single, standard software stack is infeasible.
+Scientific applications require specific versions of compilers, MPI, and other dependency libraries, so using a single, standard software stack is infeasible.
 However, managing many configurations is difficult because the configuration space is exponential in size.
-%
-Spack is a package manager used at Lawrence Livermore National Laboratory (LLNL) to manage software complexity.
+
+We introduce Spack, a package manager used at Lawrence Livermore National Laboratory (LLNL) to manage this complexity.
 Spack provides a novel, recursive specification syntax to invoke parametric builds of packages and dependencies.

### 2015-04-16 introduction.tex
intro updates.

@@
 %
-The simulations that run on these machines are very complex; some
-depend on specific versions of over 50 dependency libraries.
+The simulation software that runs on these machines is very complex; some
+codes depend on specific versions of over 50 dependency libraries.
 They require specific compilers, build options and MPI implementations to
 achieve the best performance, and users may run several
-different such codes in the same environment as part of larger 
+different codes in the same environment as part of larger 
 scientific workflows.
@@
 dependency management is overly rigid, and they burden users with
-combinatoric installation naming and versioning.
+the combinatoric problems of naming and versioning.
 
@@
 Spack is in active development at LLNL, and our use cases outline some of
-its first uses in our environment.  Spack solves software problems that are
-pervasive at large, multi-user HPC centers, and we believe that our
-experiences with it can be applied at other HPC facilities to improve
+its first applications in our environment.  Spack solves software problems 
+that are pervasive at large, multi-user HPC centers, and we believe that our
+experiences with it can be applied at other HPC facilities.  Spack improves
 operational efficiency by simplifying the build and deployment of 

### 2015-04-16 implementation-concretization.tex
updates through specs.

--- 2015-04-16 implementation.tex
@@
 
-\section{The Spack Package Manager}
-\label{sec:implementation}
-Based on our experiences at LLNL, we have developed
-{\it Spack}, the Supercomputing PACKage manager.
-Spack is written in Python.  We chose Python because it is
-very widely used int he HPC community, and because it provides
-flexible scripting capabilities. 
-%
-Like some prior systems, Spack supports arbitrarily many software installations.
-It borrows ideas from Nix; packages are installed in their own prefixes,
-and configurations are managed using hashes.
-% 
-Spack adds several unique capabilities that are essential for HPC:
-\begin{enumerate}
-\item {\bf Parametric builds} address the matrix problem:
-      Packages are parameterized by name, version, architecture, compiler, 
-      options, and dependencies.
-\item To manage the configuration space, spack provides a novel, 
-      recursive {\bf spec syntax} for dependency graphs.
-\item Specs support versioned, ABI-incompatible interfaces like MPI through
-      {\bf versioned virtual dependencies}.
-\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
-      consistency on the package author.
-\end{enumerate}
-
-\input{implementation-packages}
-
-\subsection{Spack Specs}\label{sec:specs}
-
-  Specs are partial descriptions of full
-dependency graphs, but they are concise.  This allows a user to use
-only a few parameters to quickly identify a configuration to install,
-or to query existing installations and their dependencies.
-
-\todo{1 page}
-
-\subsection{Versioned Virtual Dependencies}
-	\todo{.5 page}
-
-\subsection{Abstract \& Concrete Specs}
+\subsection{Abstract Specs \& Concretization}
 	\todo{.75 page}

### 2015-04-16 implementation-environment.tex
updates through specs.

--- 2015-04-16 implementation-packages.tex
@@
 
-\subsection{Packages}
-
-%\subsubsection{Package Files}
-\begin{figure}
-	\includegraphics[width=\columnwidth]{code/mpileaks.pdf}
-	\caption{
-		Spack package for the {\tt mpileaks} tool.
-		\label{fig:mpileaks}
-	}
-\end{figure}
-
-Spack comprises many packages, and each describes how to build a software
-artifact.  Each package is a class written in pure Python, but Spack uses
-a simple, embedded domain-specific language (DSL) to make packaging easier.
-Spack's DSL adds directives such as {\tt depends\_on}, {\tt version},
-{\tt provides}, and {\tt extends}, in order to add metadata to package classes.
-
-Figure~\ref{fig:mpileaks} shows the package for the \mpileaks tool.
-Inside the {\tt MpiLeaks} class, the package provides a text description
-and a homepage, as well as 
-a download URL.  Next, two {\tt version} directives identify known versions
-of the package, and checksums so that they can be downloaded safely.
-Below this, three {\tt depends\_on}
-directives indicate prerequisites that must be installed before \mpileaks.
-Last, each package defines an {\tt install()} method, which contains the
-commands used to build the package.
-
-Most package managers fix a single version in a package file, but Spack packages
-are {\it parameterized} by version.  In the figure, we see that the {\tt mpileaks}
-package specifies two versions.  A user can safely install either by 
-specifying it on the command line.  In our experience, however, users
-frequently want bleeding-edge versions, and package managers frequently lag 
-behind the latest releases.  Spack therefore has a capability to extrapolate
-the latest URL, using the {\tt url} attribute as a model\footnote{This works
-for consistently named URLs, and it can be specified per-version for packages
-with inconsistent naming.}.
-The user can ask for an unknown version on the command line and still
-successfully download and install it.  Spack can also use the same model to
-scrape webpages and find new versions.
-
-\subsubsection{Installation Environment}
+\subsection{Installation Environment}
 
@@
 For example, in some packages that depend on the {\tt Silo} library,
-the {\tt ---with-silo} parameter takes a path to {\tt Silo}'s prefix.
+the {\tt ---with-silo} parameter takes a path to f{\tt Silo}'s prefix.
 In others, it takes the {\tt include} and {\tt lib} subdirectories,
@@
 
+TACC expects the user to run with the same
+environment that she built with~\cite{mclay:lmod-tutorial}.
+
+But, at LC, the user is
+frequently not the builder of the application, and may not know the particular
+incantation required to load the its environment. Moreover, users frequently run more 
+than one application in the same job, each built by a separate team, and each with
+its own environment. Most LC users are used to running programs with
+specific paths; only 10\% use environment modules extensively.
+
+\begin{figure}\centering
+	\includegraphics[width=.6\linewidth]{specs/rpaths.pdf}
+	\newline
+	\verb|spack install mpileaks ^mpich|\newline
+	\verb|spack install mpileaks ^openmpi|
+	\caption{
+		Installing {\tt mpileaks} with MPICH (solid), then with OpenMPI (dashed).  The {\tt dyninst} subtree is
+		reused, but {\tt mpileaks} and {\tt callpath} must be rebuilt.
+	}
+\end{figure}
+
 \begin{verbatim}

### 2015-04-16 implementation-packages.tex
intro updates.
updates through specs.

@@
 
-Spack comprises many packages, and each describes how to build a software
-artifact.  Each package is a class written in pure Python, but Spack uses
-a simple, embedded domain-specific language (DSL) to make packaging easier.
-Spack's DSL adds directives such as {\tt depends\_on}, {\tt version},
-{\tt provides}, and {\tt extends}, in order to add metadata to package classes.
+In Spack, packages are build scripts that describe how to build software
+artifacts.  Each package is a class written in pure Python to describe
+{\it specific} build instructions, and it must extend the {\tt Package}
+base class, which implements the more general parts of the build process.
+Spack implements a simple, embedded domain-specific language (DSL) to make
+packaging easier; it adds special directives such as {\tt depends\_on} and
+{\tt version} that add metadata to the class.
 
-Figure~\ref{fig:mpileaks} shows the package for the \mpileaks tool.
+Figure~\ref{fig:mpileaks} shows the package for \mpileaks, an LLNL-developed
+tool developed for finding leaks in MPI programs.
 Inside the {\tt MpiLeaks} class, the package provides a text description
@@
 a download URL.  Next, two {\tt version} directives identify known versions
-of the package, and checksums so that they can be downloaded safely.
+of the package, and MD5 checksums ensure they can be downloaded safely.
 Below this, three {\tt depends\_on}
@@
 Last, each package defines an {\tt install()} method, which contains the
-commands used to build the package.
-
-Most package managers fix a single version in a package file, but Spack packages
-are {\it parameterized} by version.  In the figure, we see that the {\tt mpileaks}
-package specifies two versions.  A user can safely install either by 
-specifying it on the command line.  In our experience, however, users
-frequently want bleeding-edge versions, and package managers frequently lag 
-behind the latest releases.  Spack therefore has a capability to extrapolate
-the latest URL, using the {\tt url} attribute as a model\footnote{This works
-for consistently named URLs, and it can be specified per-version for packages
-with inconsistent naming.}.
-The user can ask for an unknown version on the command line and still
-successfully download and install it.  Spack can also use the same model to
-scrape webpages and find new versions.
-
-\subsubsection{Installation Environment}
-
-%{\bf Reproducibility.}
-%\paragraph{Reproducibility}
-Spack is intended to build a consistent HPC stack for our multi-user
-environment, and reproducible builds are one of our design goals.
-Experience at LLNL has shown that it is vexingly difficult to reproduce
-a build manually.
-%
-Many packages used at LLNL have a profusion of build options, and specifying them 
-correctly often requires tedious experimentation.  This is due to lack of
-build standards and to the diversity of HPC environments.  
-For example, in some packages that depend on the {\tt Silo} library,
-the {\tt ---with-silo} parameter takes a path to {\tt Silo}'s prefix.
-In others, it takes the {\tt include} and {\tt lib} subdirectories,
-separated by a comma.
-The {\tt install()} method in Spack's package files allows us to record
-precise build incantations for later reuse.
-
-%{\bf Environment isolation.}
-\paragraph{Environment isolation}
-In addition to command line issues, we
-frequently encounter errors due to inconsistencies between the environment of
-the package installer and the package user.
-%
-For example, there are two versions of the {\tt libelf} library used by
-LLNL performance tools. One is distributed with RedHat Linux, but another
-publicly available version has the same API but an incompatible ABI.
-Failure to specify the right version at build time has caused many
-inexplicable crashes.
-%
-Spack manages the build environment by running {\tt install} invocation
-in a new process.  It helps packages find dependencies 
-correctly, by setting
-{\tt PATH}, {\tt PKG\_CONFIG\_PATH}, {\tt CMAKE\_PREFIX\_PATH}, and
-{\tt LD\_LIBRARY\_PATH} to include the dependencies of the current build.
-These variables are commonly used by build systems to locate dependencies,
-and setting them helps to ensure that incorrect libraries are not detected.
-The isolated build environment also gives package authors 
-free reign to set build-specific environment variables without interfering
-with other packages.
-
-%{\bf Compiler wrappers and RPATHs.}
-\paragraph{Compiler wrappers and RPATHs}
-Finding compilers at build time is not the only obstacle to reproducible
-behavior.  As mentioned in Section~\ref{sec:motivation}, it is also important
-for binaries to be able to find dependency libraries at {\it runtime}.
-One of the most frequent user errors at LC is improper library configuration.
-Users frequently do not know what libraries a package was built with, and 
-it is difficult for them to construct a suitable {\tt LD\_LIBRARY\_PATH} for
-a package that was built by someone else.  Because of frequent support calls,
-we typically add {\tt RPATHs} to public software installations, so that paths
-to dependencies are embedded in binaries and so that users do not have to know
-this information to run installed software correctly.
-
-Spack manages {\tt RPATH} settings and other build policies with
-{\it compiler wrappers}. 
-In each isolated {\tt install} environment, Spack sets the standard 
-environment variables
-{\tt CC}, {\tt CXX}, {\tt F77}, and {\tt FC} to point to its own compiler
-wrapper scripts.  These variables are used by most build systems to select
-C, C++, and Fortran compilers, so they are generally picked up 
-automatically\footnote{If builds do not respect {\tt CC}, {\tt CXX}, etc.,
-wrappers can typically be added as arguments or inserted into Makefiles
-by {\tt install}.}.
-When run, the wrappers insert include ({\tt -I}), library ({\tt -L}), and 
-{\tt RPATH} ({\tt -Wl,-rpath} or similar) flags into the argument list.
-These point to the {\tt include} and {\tt lib} directories of dependency
-library installations, where needed headers and libraries are typically located.
-After the modified argument list is constructed, the wrappers delegate it
-to the real compiler to execute.
-
-Spack's compiler wrappers have a number of useful effects.  First, they allow
-Spack to transparently parameterize the compiler for most builds. 
-Spack users to easily request a build with a new compiler swapped in.
-Second, they enforce the use of {\tt RPATHs} in
-installed binaries.  This causes applications built by Spack to run correctly
-{\it regardless of the environment}.  Third, because compiler wrappers add 
-header and library search paths for dependencies, header and library detection
-tests run by most build systems succeed automatically, {\it without}
-the need to use special arguments for nonstandard locations.  {\tt configure}
-commands in Spack's {\tt install} function can have fewer arguments, and can
-be written as they would be for system installs.  This reduces complexity
-for package maintainers and enforces consistent, reproducible
-build policies across packages.  Finally, because Spack has control over the 
-wrappers, package authors can programmatically filter the compiler flags
-used by software build systems, a useful last resort when porting to
-bleeding-edge platforms or new, esoteric compilers.
-
-\subsubsection{Environment Module Integration}
-	\todo{.5 page}
-
-\begin{verbatim}
-module disadvantages:
-	- lmod solves a lot of problems
-	- not on all systems
-		- users often run from just a path.
-		- not guaranteed that LD_LIBRARY_PATH, etc will
-		  be set correctly.
-
-- Lmod assumes build & submit environment must match
-	- users often have NO IDEA what environment a package
-	  was built with
-	- multiple apps may be built with different stacks
-	- combining them in a workflow is difficult with modules
-		(need to know right incantation)
-
-	- TACC implements a rather complicated mapping back to
-	    the build env (Lmod CheckExec)
-
-- Spack solves the matrix problem mentioned in the Lmod
-    presentation (slide 98)
-	- hierarchy is nasty -- every branch represents a
-	  greedy choice
-	- might NEED a specific version of ONE dependency
-	  that is not supported in the tree
-	  => SPACK!
-
-- libmesh
-	- boost, petsc, trilinos, grvy, etc.
-\end{verbatim}
+commands used to build the package.  Spack's DSL allows shell
+commands to be invoked as Python functions. Here, the {\tt install()} 
+method invokes the familiar {\tt configure}, {\tt make}, and
+{\tt make install} build commands, as a shell script would.

### 2015-04-16 implementation-specs.tex
updates through specs.

%!TEX root = paper.tex

\begin{table*}\centering
\begin{tabular}{|r|p{2.4in}|p{4in}|}
\hline
& {\bf Spec} & {\bf Meaning} \\
\hline
\hline
1&\small\verb|mpileaks|                         & \small {\tt mpileaks} package, no constraints. \\\hline
2&\small\verb|mpileaks@1.1.2|                   & \small {\tt mpileaks} package, version 1.1.2. \\\hline
3&\small\verb|mpileaks@1.1.2 %gcc|              & \small {\tt mpileaks} package, version 1.1.2, built with {\tt gcc} at the default version. \\\hline
4&\small\verb|mpileaks@1.1.2 %intel@14.1 +debug| & \small {\tt mpileaks} package, version 1.1.2, built with Intel compiler version 14.1, \newline with the ``debug'' build option. \\\hline
5&\small\verb|mpileaks@1.1.2 =bgq|              & \small {\tt mpileaks} package, version 1.1.2, built for the Blue Gene/Q platform (BG/Q). \\\hline
6&\small\verb|mpileaks@1.1.2 ^mvapich2@1.9|     & \small {\tt mpileaks} package version 1.1.2, using {\tt mvapich2}  version 1.9 for MPI. \\\hline
7&\small\verb|mpileaks @1.2:1.4 %gcc@4.7.5 -debug =bgq| \newline
      \verb|  ^callpath @1.1 %gcc@4.7.2| \newline
      \verb|  ^openmpi @1.4.7|                & \small%
      {\tt mpileaks} at any version between 1.2 and 1.4 (inclusive), built with gcc 4.7.5, 
      without the debug option, for BG/Q, linked with {\tt callpath} version 1.1
      and building {\tt callpath} with {\tt gcc} version 4.7.2, linked with {\tt openmpi} version 1.4.7.    \\
\hline
\end{tabular}
\caption{
	Example Spack build specs and their meanings.
	\label{tab:specs}
}
\end{table*}

\begin{figure}
	\subfigure[{\tt mpileaks}]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks.pdf}
		\label{fig:specs-mpileaks}
	}
	\subfigure[{\tt mpileaks \^{}callpath\@1.0+debug \^{}libelf\@0.8.11}]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-abstract.pdf}
		\label{fig:specs-mpileaks-abstract}
	}
	\subfigure[Concrete version of (b)]{
		\centering
		\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
		\label{fig:specs-mpileaks-concrete}
	}
	\caption{
		Specs for the {\tt mpileaks} package.
		\label{fig:specs}
	}
\end{figure}

\subsection{Spack Specs}\label{sec:specs}

The package shown in Figure~\ref{fig:mpileaks} seems simple, but even this short script
allows Spack to build many versions of the {\tt mpileaks} package.  This is because builds
in Spack are parameterized.  In traditional port systems, code is structured to build a single
version of a package, but in Spack, each package file is a {\it template} that can build 
in many different ways.  Spack's representation of a configuration is a build specification, 
or {\tt spec}.

\subsubsection{Spec structure}
To understand how specs works, consider the package structure.  The  {\tt Mpileaks} package
is a class, and its metadata ({\tt version}, {\tt depends\_on}, etc.) statically describe
its relationships with other packages.  In this case, there are two direct dependencies: 
the {\tt callpath} library and {\tt mpi}.  Spack recursively inspects the class definitions
for each of these packages to construct a graph of their relationships.  
To avoid issues with incompatible ABIs, and to guarantee a consistent build, the DAG is constructed
so that there is only ever one version of any particular package.

For {\tt mpileaks}, 
the initial graph is shown in Figure~\ref{fig:specs-mpileaks}.  This graph is the in-memory
representation of a spec.

Each package, or node, in the spec has a number of configuration parameters that control
how the package will be built.  These are the version, compiler to build with, compiler version, 
named build options (or {\it variants} like +debug or -debug), and the architecture to
build for.  When Spack runs to install {\tt mpileaks}, it constructs a package object for each node
in the spec DAG, and it traverses the DAG in a bottom-up fashion.  At each node, it
invokes the package's {\tt install()} method, passing in a sub-DAG rooted at the
package to be installed.  This is the {\tt spec} parameter in Figure~\ref{fig:mpileaks}.
Package maintainers are responsible for examining the spec's configuration and adjusting the 
build if necessary.

\subsubsection{Spec syntax}
A spec DAG has many degrees of freedom, and it is not reasonable to expect users to
understand all of them.  In our experience at LLNL, users typically have a small number of 
build constraints that they care about, and the others only confuse them.  This is part of why
the HPC software ecosystem is so difficult to manage: there are simply too many parameters.
However, the small set of important build constraints can be very specific, so we need the
ability to specify all aspects of the configuration space if necessary.  For this reason, 
we introduce a telescoping syntax for package specs that allows users of Spack to
specify {\it only those constraints they care about}, but to be {\it specific when necessary}.
This is the true power of our syntax---specs are concise enough to use on the command line,
and users can request installs or query existing installs by their structure.

Table~\ref{tab:specs} shows examples of specs, ranging from very simple to very complex.
Next to each is a description of the spec in plain language.  In the most basic case,
shown on line 1 in the table, a user can simply name the package they want to use, and have it
built and installed a default configuration.  On the command line this would take the form:
\newline\newline
{\tt spack install mpileaks}
\newline

\noindent
If the user cares to specify the desired verson of {\it mpileaks}, she can do so
by supplying {\tt @<version>} (line 2). Versions can be precise (e.g. {\tt @4.5.1}) or 
they can be ambiguous (e.g., @

To specify the compiler (line 3), the user
simply adds {\tt \%} followed by its name, and if a version {\it of the compiler} 
is desired, the same {\tt @} syntax can be used.  Here, {\tt gcc} refers to the full
gcc compiler suite: C, C++, Fortran 77, and Fortran 90 compilers.

To handle build-time options like debug flags and optional components, spack
allows named flags, or {\it variants} to be associated with packages.  
In the example (line 4), {\tt mpileaks} can be built with a {\tt debug} option by
adding {\tt +debug}.  The option can be similarly disabled with {\tt -debug}.  
This allows package authors to provide package-specific build options without 
having to extend Spack.  We use names for these to simplify the spec
syntax, and to prevent Spack's configuration space from becoming too fine-grained.
It would be violate our goal of conciseness, for example, to include detailed 
compiler flags in the spec syntax, but known sets of flags can be exported using
named variants.

To support cross-compilation, Spack includes the platform in the package spec (line 5).
Platforms begin with {\tt =} and take names like {\tt linux-ppc64} or {\tt bbq}.  They are
specified per-package; this allows front-end tools to depend on their back-end measurement
libraries with a {\it different} architecture on cross-compiled machines.

Lines 6 and 7 in the table show the key feature that enables Spack's flexibility.
The user can specify all of the above information not only for the package being
installed, but also for its {\it dependencies}.  To do this, the user needs only supply 
\verb|^| and the dependency's name.  If we need to build a new version with a specific
version of {\tt mvapich2}, we can simply add, e.g., \verb|^mvapich2@1.9|
to the spec, and it will build with that MPI version instead of the default.
This can also be done for multiple libraries in the same spec (line 7).  
As mentioned, Spack guarantees that there is only one version of any package in 
the spec DAG.  So, within the same DAG, each dependency can be uniquely identified by 
only its package name.  This frees the user from having to worry about DAG connectivity;
the user only needs to know that a package depends, somehow, indirectly on {\tt mvapich}
to add this constrain
uniquely identifies a dependency, and dependency constraints can be specified in any
 order.

\subsubsection{Constraining dependencies}

\subsubsection{Build specialization}

\subsubsection{Version extrapolation}

In our experience, however, users
frequently want bleeding-edge versions, and package managers frequently lag 
behind the latest releases.  Spack therefore has a capability to extrapolate
the latest URL, using the {\tt url} attribute as a model\footnote{This works
for consistently named URLs, and it can be specified per-version for packages
with inconsistent naming.}.
The user can ask for an unknown version on the command line and still
successfully download and install it.  Spack can also uses the same model to
scrape webpages and find new versions.

\subsubsection{Querying installed packages}

### 2015-04-16 implementation-virtual-deps.tex
updates through specs.

--- 2015-04-16 implementation.tex
@@
 %!TEX root = paper.tex
-
-\section{The Spack Package Manager}
-\label{sec:implementation}
-Based on our experiences at LLNL, we have developed
-{\it Spack}, the Supercomputing PACKage manager.
-Spack is written in Python.  We chose Python because it is
-very widely used int he HPC community, and because it provides
-flexible scripting capabilities. 
-%
-Like some prior systems, Spack supports arbitrarily many software installations.
-It borrows ideas from Nix; packages are installed in their own prefixes,
-and configurations are managed using hashes.
-% 
-Spack adds several unique capabilities that are essential for HPC:
-\begin{enumerate}
-\item {\bf Parametric builds} address the matrix problem:
-      Packages are parameterized by name, version, architecture, compiler, 
-      options, and dependencies.
-\item To manage the configuration space, spack provides a novel, 
-      recursive {\bf spec syntax} for dependency graphs.
-\item Specs support versioned, ABI-incompatible interfaces like MPI through
-      {\bf versioned virtual dependencies}.
-\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
-      consistency on the package author.
-\end{enumerate}
-
-\input{implementation-packages}
-
-\subsection{Spack Specs}\label{sec:specs}
-
-  Specs are partial descriptions of full
-dependency graphs, but they are concise.  This allows a user to use
-only a few parameters to quickly identify a configuration to install,
-or to query existing installations and their dependencies.
-
-\todo{1 page}
 
@@
 	\todo{.5 page}
-
-\subsection{Abstract \& Concrete Specs}
-	\todo{.75 page}

### 2015-04-16 implementation.tex
updates through specs.

@@
 and configurations are managed using hashes.
-% 
-Spack adds several unique capabilities that are essential for HPC:
+
+\noindent
+Spack adds several essential capabilities for HPC:
 \begin{enumerate}
-\item {\bf Parametric builds} address the matrix problem:
+\item {\bf Parametric builds} address the combinatorial problem:
       Packages are parameterized by name, version, architecture, compiler, 
@@
 \input{implementation-packages}
+\input{implementation-specs}
+\input{implementation-virtual-deps}
+\input{implementation-concretization}
 
-\subsection{Spack Specs}\label{sec:specs}
-
-  Specs are partial descriptions of full
-dependency graphs, but they are concise.  This allows a user to use
-only a few parameters to quickly identify a configuration to install,
-or to query existing installations and their dependencies.
-
-\todo{1 page}
-
-\subsection{Versioned Virtual Dependencies}
-	\todo{.5 page}
-
-\subsection{Abstract \& Concrete Specs}
-	\todo{.75 page}
+\input{implementation-environment}

### 2015-04-16 motivation.tex
updates through specs.

@@
 
-This section describes existing tools for managing HPC 
-software and motivates the need for a system like Spack.
-
 \paragraph{Meta-build Systems}
-
 {\it Meta-build systems} such as Contractor, WAF, and
@@
 related to package managers, but their focus is on ensuring that a single
-project builds correctly along with all of its dependencies.  They do not address
+{\it project} builds correctly along with all of its dependencies.  They do not address
 larger, multi-user system administration issues such as versioning, naming
@@
 are installed before packages that require them.
-These systems largely solve the problem of managing a single software
-stack, which works well for the baseline OS and drivers, which are 
+These tools largely solve the problem of managing a {\it single} software
+stack.  This works well for the baseline OS and drivers, which are 
 common to all applications on a system.
-However, they make the assumption that there is only a single version
+These tools make the assumption that there is only a single version
 of each package, and most install packages in a single, inflexible location.
@@
 Homebrew~\cite{bsdports,groffen:gentoo-prefix,homebrew,macports,thiruvathukal:gentoo04}
-allow packages to be built on demand, but most suffer from 
+build packages from source instead of installing from a pre-built binary.
+Most existing port systems suffer from 
 the same versioning and naming issues as traditional package managers.
@@
 
-Looking again at the conventions in Table~\ref{tab:naming-conventions},
+From the conventions in Table~\ref{tab:naming-conventions},
 we see that most sites use some combination of architecture, compiler version,
 package name, package version, and a custom (up to the author, sometimes
-encoded) build identifier.  TACC and many other sites also include the MPI
-version in the path. None of these naming conventions is defined for
-the entire configuration space, and none defines a way to create new versions
-from arbitrary combinations of dependencies.  MPI is explicitly called out,
-but it is only one of many dependencies that go into a build.
-Sites have made a conscious choice to divide the directory hierarchy
-according to the most common software packages, but they ignore the rest.
-Frequently, desired configurations do not fit the local convention, and
-staff must install them in nonstandard locations.
+encoded) build identifier.  TACC and many other sites also explicitly
+include the MPI version in the path. MPI is explicitly called out
+because it is one of the most common software packages for HPC.
+However, it is only one of many dependencies that go into a build.
+None of these naming conventions covers the entire configuration
+space, and none has a way to represent, e.g., two builds that are identical
+save for the version of a particular dependency library.  In our experience
+at LLNL, naming conventions like these have not succeeded because there
+are more desired configurations than we can represent with a practical
+directory hierarchy. Staff frequently install nonconforming packages 
+in nonstandard locations with ambiguous names.
 
@@
 Diverse software versions not only present problems for build and installation; 
-they also complicate the runtime environment. When run, an executable
-must determine the location of its dependency libraries, or it will fail to run.
-This is typically not a problem for statically linked executables, but modern
-operating systems make extensive use of dynamic linking, which has many 
-benefits for software tools and maintenance.
+they also complicate the runtime environment. When launched, an executable
+must determine the location of its dependency libraries, or it will not run.
+Statically linked binaries do not have this issue, but modern
+operating systems make extensive use of dynamic linking.
 By default, the dynamic loader on most systems is configured to search only
@@
 {\it user} who runs the program must typically add dependency library paths to
-{\tt LD\_LIBRARY\_PATH} (or a similar environment variable on other OSes).
-Most HPC sites address this problem using {\it environment modules}, which allow
-users to dynamically ``load'' and ``unload'' settings for search paths like
-{\tt LD\_LIBRARY\_PATH}, {\tt PATH}, {\tt MANPATH}, etc.
+{\tt LD\_LIBRARY\_PATH} (or a similar environment variable) so that the loader
+can find them.
 
-Environment modules date back to 1991, and there are many implementations 
+Many HPC sites address this problem using {\it environment modules}, which allow
+users to dynamically ``load'' and ``unload'' such settings using simple commands.
+Environment emerged in 1991, and there are many implementations 
 available~\cite{dotkit,furlani+:lisa91,furlani+:lisa96,mclay:lmod,mclay:lmod-tutorial}.
@@
 provide software hierarchies similar to the naming conventions in 
-Table\ref{tab:naming-conventions}, and they allow users to quickly load a software stack
-if they know which one is required.  TACC expects the user to run with the same
-environment that she built with~\cite{mclay:lmod-tutorial}.  But, at LC, the user is
-frequently not the builder of the application, and may not know the particular
-incantation required to load the its environment. Moreover, users frequently run more 
-than one application in the same job, each built by a separate team, and each with
-its own environment. Most LC users are used to running programs with
-specific paths; only 10\% use environment modules extensively.
+Table~\ref{tab:naming-conventions}. They allow users to quickly load a software stack
+if they know which one is required.
 
@@
 
-Existing tools do not enforce {\tt RPATHs}, leaving them to package authors.
-This may lead to erroneous runs.
+%Existing tools do not enforce {\tt RPATHs}, leaving them to package authors.
+%This may lead to erroneous runs.
 

### 2015-04-16 implementation-specs.tex
update specs.

@@
 
-\noindent
-If the user cares to specify the desired verson of {\it mpileaks}, she can do so
-by supplying {\tt @<version>} (line 2). Versions can be precise (e.g. {\tt @4.5.1}) or 
-they can be ambiguous (e.g., @
+If the user wants more specificity, {\tt mpileaks} can be augmented with additional
+qualifiers.  Spack install will handle selecting a build configuration that satisfies
+the user's request.  We describe additional qualifiers below.
 
+{\bf Versions.}
+Versions are specified by supplying {\tt @<version>} after the package name (line 2). 
+Versions can be precise (e.g. {\tt @4.5.1}) or 
+they can be ambiguous (e.g., @4) or a colon-separated, inclusive range (e.g., @4.4:4.5.1),
+and Spack will pick an available compiler from the specified range.
+The package in Figure~\ref{fig:mpileaks} lists two ``safe'' versions with checksums, but
+in our experience users frequently want bleeding-edge versions.  Package managers
+frequently lag behind the latest releases. 
+Spack has a capability to extrapolate URLs from versions,
+using the package's {\tt url} attribute as a model\footnote{This works
+for packages with consistently named URLs}.  The user can request a specific
+version on the command line, even if it is unknown to Spack,
+and Spack will attempt to install it.  Spack also uses the same
+model to scrape webpages and find new versions as they become available.
+
+{\bf Compilers.}
 To specify the compiler (line 3), the user
@@
 is desired, the same {\tt @} syntax can be used.  Here, {\tt gcc} refers to the full
-gcc compiler suite: C, C++, Fortran 77, and Fortran 90 compilers.
+gcc compiler suite: C, C++, Fortran 77, and Fortran 90 compilers.  Spack can auto-detect
+compiler toolchains if they are in the user's PATH or they can be registered manually
+through a configuration file.
 
+{\bf Variants.}
 To handle build-time options like debug flags and optional components, spack
@@
 
+{\bf Cross-compileation.}
 To support cross-compilation, Spack includes the platform in the package spec (line 5).
@@
 
+{\bf Dependencies.}
 Lines 6 and 7 in the table show the key feature that enables Spack's flexibility.
@@
 
-In our experience, however, users
-frequently want bleeding-edge versions, and package managers frequently lag 
-behind the latest releases.  Spack therefore has a capability to extrapolate
-the latest URL, using the {\tt url} attribute as a model\footnote{This works
-for consistently named URLs, and it can be specified per-version for packages
-with inconsistent naming.}.
-The user can ask for an unknown version on the command line and still
-successfully download and install it.  Spack can also uses the same model to
-scrape webpages and find new versions.
-
 \subsubsection{Querying installed packages}

### 2015-04-16 implementation-environment.tex
GLL edits

@@
 For example, in some packages that depend on the {\tt Silo} library,
-the {\tt ---with-silo} parameter takes a path to f{\tt Silo}'s prefix.
+the {\tt ---with-silo} parameter takes a path to {\tt Silo}'s installation prefix.
 In others, it takes the {\tt include} and {\tt lib} subdirectories,
@@
 \paragraph{Environment isolation}
-In addition to command line issues, we
+In addition to command-line issues, we
 frequently encounter errors due to inconsistencies between the environment of
@@
 For example, there are two versions of the {\tt libelf} library used by
-LLNL performance tools. One is distributed with RedHat Linux, but another
+LLNL performance tools. One is distributed with RedHat Linux, while another
 publicly available version has the same API but an incompatible ABI.
@@
 %
-Spack manages the build environment by running {\tt install} invocation
+Spack manages the build environment by running the {\tt install} invocation
 in a new process.  It helps packages find dependencies 
@@
 These variables are commonly used by build systems to locate dependencies,
-and setting them helps to ensure that incorrect libraries are not detected.
+and setting them helps to ensure that incorrect libraries are detected.
 The isolated build environment also gives package authors 
@@
 Spack's compiler wrappers have a number of useful effects.  First, they allow
-Spack to transparently parameterize the compiler for most builds. 
+Spack to transparently parameterize the compiler for most builds and for
 Spack users to easily request a build with a new compiler swapped in.

### 2015-04-16 implementation-packages.tex
GLL edits

@@
 Figure~\ref{fig:mpileaks} shows the package for \mpileaks, an LLNL-developed
-tool developed for finding leaks in MPI programs.
+tool for finding leaks in MPI programs.
 Inside the {\tt MpiLeaks} class, the package provides a text description
@@
 of the package, and MD5 checksums ensure they can be downloaded safely.
-Below this, three {\tt depends\_on}
+Below this, two {\tt depends\_on}
 directives indicate prerequisites that must be installed before \mpileaks.

### 2015-04-16 implementation-specs.tex
GLL edits

@@
 they can be ambiguous (e.g., @4) or a colon-separated, inclusive range (e.g., @4.4:4.5.1),
-and Spack will pick an available compiler from the specified range.
+and Spack will pick an available version from the specified range.
 The package in Figure~\ref{fig:mpileaks} lists two ``safe'' versions with checksums, but
@@
 syntax, and to prevent Spack's configuration space from becoming too fine-grained.
-It would be violate our goal of conciseness, for example, to include detailed 
+It would violate our goal of conciseness, for example, to include detailed 
 compiler flags in the spec syntax, but known sets of flags can be exported using
@@
 the user only needs to know that a package depends, somehow, indirectly on {\tt mvapich}
-to add this constrain
+to add this constraint
 uniquely identifies a dependency, and dependency constraints can be specified in any

### 2015-04-16 usecase-combinatoric.tex
Fix typos.  Add gperftools section.

@@
 \subsubsection{Gperftools packaging}
-\todo{.5 page}
 
-\begin{verbatim}
-	- combinatoric nightmare
-	- compilers x versions
-\end{verbatim}
+The gperftools tool from Google has gained popularity among some developers for its high-performance thread-safe heap and its lite-weight profilers.
+Unfortunately, two issues made it difficult to maintain gperftools installations at LLNL.
+First, gperftools is a C++ library.
+Since C++ does not have standard ABI it needs to be re-built with each compiler and compiler version.
+Second, building gperftools on atypical architectures (such as Blue Gene/Q) requires patches and a complicated configure lines that change with each compiler.
+The first issue motivated application developers to try and maintain their own builds of gperftools that worked with their prefered compilers--a task that was made difficult by the second issue.
+
+Spack presented a solution to both problems.
+Site administrators can use Spack to easily maintain a central install of gperftools across the combinations of versions and compilers.
+Spack's gperftools package also serves as a central repository for the knowledge of how to build gperftools on each platform and compiler combination.
+
+\begin{figure}
+  \includegraphics[width=\columnwidth]{code/gperftools.pdf}
+  \caption{
+    Simplified install routine for the gperftools tool.
+    \label{fig:gperftools}
+  }
+\end{figure}
+
+Figure~\ref{fig:gperftools} illustrates per-compiler and platform build rules with a simplified version of the install routine for the gperftools tool (Spack's real install routine for gperftools includes other compilers and more options per compiler).
+It applies a patch if gperftools 2.4 is built with the XLC compiler, and it selects the correct configure line based on the current platform and compiler.
 

### 2015-04-16 motivation.tex
reword mixdown section.

@@
 {\it Meta-build systems} such as Contractor, WAF, and
-MixDown~\cite{amundson:contractor,epperly+:mixdown,nagy:waf} are closely
-related to package managers, but their focus is on ensuring that a single
-{\it project} builds correctly along with all of its dependencies.  They do not address
-larger, multi-user system administration issues such as versioning, naming
-conventions, rapid multi-configuration installation, or maintaining many
-divergent software stacks.
+MixDown~\cite{amundson:contractor,epperly+:mixdown,epperly+:mixdown-report,nagy:waf} are
+related to package managers, but they focus on ensuring that a single
+package builds correctly with its dependencies.  MixDown notably provides excellent
+features for ensuring consistent compiler flags in a build.
+However, these systems do not provide facilities to manage
+large package repositories or combinatoric versioning.
 

### 2015-04-17 implementation-concretization.tex
Lots of edits.

@@
 	\todo{.75 page}
+
+Before it builds a package, Spack ensures the following conditions:
+\newline
+
+\begin{tabular}{l}
+(1) All dependencies are present in the DAG. \\
+(2) Parameters are set on all packages in the DAG. \\
+\end{tabular}\newline
+
+\noindent
+The install process then constructs a package object for each node in the spec DAG
+and traverses the DAG in a bottom-up fashion.  At each node, it invokes the package's
+{\tt install()} method, using a sub-DAG rooted at the package to be installed as the {\tt spec}
+parameter to {\tt install()}. Package maintainers are responsible for examining the spec's
+configuration and adjusting the build if necessary.
+
+In our experience, much complexity in build scripts arises from the need for the script to
+query the environment and determine how to build.  In spack, build scripts are free from this 
+concern, because the package author is guaranteed that a package's build spec is concrete
+by the time {\tt install()} is called. There is thus no need to perform complicated state checks,
+just to query the spec for its configuration.

### 2015-04-17 implementation-packages.tex
Fix typos.  Add gperftools section.
Lots of edits.

@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{code/mpileaks.pdf}
+\begin{minted}[linenos,
+               numbersep=5pt,
+			   fontsize=\scriptsize,
+               frame=lines,
+               framesep=2mm]{python}
+from spack import *
+
+class Mpileaks(Package):
+    """Tool to detect and report leaked MPI objects like
+       MPI_Requests and MPI_Datatypes."""
+
+    homepage = "https://github.com/hpc/mpileaks"
+    url = homepage + "/releases/download/v1.0/mpileaks-1.0.tar.gz"
+
+    version('1.0', '8838c574b39202a57d7c2d68692718aa')
+    version('1.1', '4282eddb08ad8d36df15b06d4be38bcb')
+
+    depends_on("mpi")
+    depends_on("callpath")
+
+    def install(self, spec, prefix):
+        configure("--prefix=" + prefix,
+                  "--with-callpath=" + spec['callpath'].prefix)
+         make()
+         make("install")
+\end{minted}
 	\caption{
@@
 
-In Spack, packages are build scripts that describe how to build software
-artifacts.  Each package is a class written in pure Python to describe
-{\it specific} build instructions, and it must extend the {\tt Package}
-base class, which implements the more general parts of the build process.
-Spack implements a simple, embedded domain-specific language (DSL) to make
-packaging easier; it adds special directives such as {\tt depends\_on} and
-{\tt version} that add metadata to the class.
+In Spack, packages are Python scripts that build software artifacts.  
+Each package is a class that extends a generic {\tt Package}
+base class.  {\tt Package} implements the bulk of the build process, but
+subclasses provide their own {\tt install()} method to handle the
+specifics of particular packages. Additionally,
+Spack implements an embedded domain-specific language (DSL).  It provides
+special directives such as {\tt depends\_on} and {\tt version} that add
+metadata to the package class.
 
-Figure~\ref{fig:mpileaks} shows the package for \mpileaks, an LLNL-developed
-tool for finding leaks in MPI programs.
-Inside the {\tt MpiLeaks} class, the package provides a text description
-and a homepage, as well as 
-a download URL.  Next, two {\tt version} directives identify known versions
-of the package, and MD5 checksums ensure they can be downloaded safely.
+Figure~\ref{fig:mpileaks} shows the package for \mpileaks, a tool developed
+at LLNL to find memory leaks in MPI programs.
+The {\tt MpiLeaks} class first provides simple metadata: a text description,
+a homepage, and a download URL.  
+Next, two {\tt version} directives identify known versions and provide
+MD5 checksums used to verify downloads.
 Below this, two {\tt depends\_on}
-directives indicate prerequisites that must be installed before \mpileaks.
-Last, each package defines an {\tt install()} method, which contains the
-commands used to build the package.  Spack's DSL allows shell
-commands to be invoked as Python functions. Here, the {\tt install()} 
-method invokes the familiar {\tt configure}, {\tt make}, and
-{\tt make install} build commands, as a shell script would.
+directives indicate prerequisite packages that must be installed before \mpileaks.
+Last, the {\tt install()} method contains the commands used to build
+the tool.  Spack's DSL allows shell commands to be invoked as Python functions,
+and the {\tt install()} method invokes {\tt configure},
+{\tt make}, and {\tt make install} commands as a shell script would.

### 2015-04-17 implementation-specs.tex
Fix typos.  Add gperftools section.
Lots of edits.

@@
 %!TEX root = paper.tex
+
+\begin{figure}
+	\subfigure[Spec for {\tt mpileaks}]{
+		\centering
+		\includegraphics[width=\columnwidth]{specs/mpileaks.pdf}
+		\label{fig:specs-mpileaks}
+	}
+	\subfigure[{\tt mpileaks@2.3}]{
+		\centering
+		\includegraphics[width=\columnwidth]{specs/mpileaks-version}
+		\label{fig:specs-mpileaks-version}
+	}
+	\subfigure[{\tt mpileaks@2.3 \^{}callpath@1.0+debug \^{}libelf@0.8.11}]{
+		\centering
+		\includegraphics[width=\columnwidth]{specs/mpileaks-abstract.pdf}
+		\label{fig:specs-mpileaks-abstract}
+	}
+	\subfigure[Spec (b), concretized.]{
+		\centering
+		\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
+		\label{fig:specs-mpileaks-concrete}
+	}
+	\caption{
+		Specs for the {\tt mpileaks} package.
+		\label{fig:specs}
+	}
+\end{figure}
+
+\subsection{Spack Specs}\label{sec:specs}
+
+Using the simple script in Figure~\ref{fig:mpileaks}, Spack can build many different
+versions and configurations of the {\tt mpileaks} package.  In traditional port systems, 
+package code is structured to build a single version of a package, but in Spack, each
+package file is a {\it template} that can be built in many different ways.  The build is
+{\it parameterized} so that it can be configured many different ways.  
+Spack calls a single build configuration a {\it spec},
+ and the {\tt spec} object passed to {\tt install()} 
+is how the Spack system encapsulates build parameters for package authors.
+
+\subsubsection{Structure}
+To understand how specs work, consider the {\tt mpileaks} package structure.  
+Metadata in the {\tt Mpileaks} class ({\tt version}, {\tt depends\_on}, etc.) describe
+its relationships with other packages.  There are two direct dependencies: 
+the {\tt callpath} library and {\tt mpi}.  Spack recursively inspects the class definitions
+for each dependency and constructs a graph of their relationships.  The result
+is a directed, acyclic graph (DAG) of dependencies\footnote{Spack currently disallows
+circular dependencies.}.  
+%
+To guarantee a consistent build, and to avoid ABI incompatibility, the DAG
+is constructed so that there is only {\it one} version of any particular package.  Note
+the distinction: while Spack can install arbitrarily many configurations of any package,
+no two configurations of the same package will ever appear in the same build DAG.
+
+DAGs for {\tt mpileaks} are shown in Figure~\ref{fig:specs}.
+Each node is a package and has five configuration parameters that control
+how it will be built: 1) the package version, 2) the compiler to
+build with, 3) the compiler version, 4) named compile-time build options, or {\it variants},
+and 5) the target architecture.
+
+\subsubsection{Configuration Complexity}
+A spec DAG has many degrees of freedom, and it is not reasonable to expect users to
+understand or specify all of them.  In our experience at LLNL, the typical user
+only cares about a small number of build constraints (if any), and does not know enough to
+specify the rest. For example, a user may know that a certain version of a library like
+{\tt boost} is required, but only cares that other build parameters are set so that
+the build will succeed.
+%
+Configuration complexity makes the HPC software ecosystem difficult to manage: there are
+simply too many parameters. However, the small set of important build constraints can be
+very specific, so we have two competing concerns.  We need the ability to specify details
+of the configuration space, without the complexity of remembering all of them.
+
+\subsubsection{Spec Syntax}\label{sec:syntax}
+
+\begin{figure}
+\begin{grammar}
+  <spec>         ::= <id> [ constraints ]
+
+  <constraints>   ::= \{ `@' <version-list> | `+' <variant> \newline
+                   | `-' <variant> ~~~| `~' <variant> \newline
+                   | `\%' <compiler> ~| `=' <architecture> \} \newline
+                  [ <dep-list> ]
+
+  <dep-list>  ::= \{ `\textsf \textasciicircum' <spec> \}
+
+  <version-list> ::= <version> [ \{ `,' <version> \} ]
+
+  <version>      ::= <id> | <id> `:' | `:' <id> | <id> `:' <id>
+
+  <compiler>     ::= <id> [ <version-list> ]
+
+  <variant>      ::= <id>
+
+  <architecture> ::= <id>
+
+  <id>           ::= [A-Za-z0-9_][A-Za-z0-9_.-]*
+\end{grammar}
+\caption{
+	EBNF grammar for spec expressions.
+	\label{fig:grammar}
+}
+\end{figure}
 
@@
 \caption{
-	Example Spack build specs and their meanings.
+	Spack build spec syntax examples and their meaning.
 	\label{tab:specs}
@@
 
-\begin{figure}
-	\subfigure[{\tt mpileaks}]{
-		\centering
-		\includegraphics[width=\columnwidth]{specs/mpileaks.pdf}
-		\label{fig:specs-mpileaks}
-	}
-	\subfigure[{\tt mpileaks \^{}callpath\@1.0+debug \^{}libelf\@0.8.11}]{
-		\centering
-		\includegraphics[width=\columnwidth]{specs/mpileaks-abstract.pdf}
-		\label{fig:specs-mpileaks-abstract}
-	}
-	\subfigure[Concrete version of (b)]{
-		\centering
-		\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
-		\label{fig:specs-mpileaks-concrete}
-	}
-	\caption{
-		Specs for the {\tt mpileaks} package.
-		\label{fig:specs}
-	}
-\end{figure}
+We have developed a syntax for specs that allows users to specify
+only those constraints they care about, but to be specific when necessary.
+Our syntax is expressive enough to represent software DAGs  but concise enough to use
+on the command line. The spec syntax is recursively defined to allow users to specify
+parameters on dependencies as well as on the root of the DAG.  The EBNF grammar we use
+to implement this in Spack is shown in Figure~\ref{fig:grammar}.
 
-\subsection{Spack Specs}\label{sec:specs}
+We begin with an example.%Let us explain our syntax with an example.
+Consider the case where a user wants to install the {\tt mpileaks} package, but knows
+nothing about its structure.  To install the package, the user writes:
+%
+\begin{minted}[fontsize=\scriptsize]{bash}
+$ spack install mpileaks
+\end{minted}
+%
+Here, {\tt mpileaks} is the simplest possible spec---a single identifier.
+Spack parses it and converts it into the DAG shown in Figure~\ref{fig:specs-mpileaks}.
+Note that even though the spec contains no dependency information, it is still 
+converted to a full DAG, based on the directives supplied in package files. Since there
+are no constraints on the nodes, Spack has considerable leeway for how to build the package,
+and we say that the package is {\it unconstrained}.
+%
+Now, suppose the user wants a specific version of {\tt mpileaks}.  This can be requested
+with a version constraint after the package name:
+%
+\begin{minted}[fontsize=\scriptsize]{bash}
+$ spack install mpileaks@2.3
+\end{minted}
+%
+We see from Figure~\ref{fig:specs-mpileaks-version} that the specific version constraint is
+placed on the {\tt mpileaks} node in the DAG, but the rest of the DAG remains unconstrained.
+If the user does not need a specific version of {\tt mpileaks}, but does require
+particular minimu version, then the user could use {\it version range} syntax.
+and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would use
+{\tt @2.3:2.5.6} to designate the range. In these cases, the user can save build time if
+Spack already has a version installed that satisfies the constraint -- Spack will just take
+the already built instead of building a new one.
 
-The package shown in Figure~\ref{fig:mpileaks} seems simple, but even this short script
-allows Spack to build many versions of the {\tt mpileaks} package.  This is because builds
-in Spack are parameterized.  In traditional port systems, code is structured to build a single
-version of a package, but in Spack, each package file is a {\it template} that can build 
-in many different ways.  Spack's representation of a configuration is a build specification, 
-or {\tt spec}.
+Figure~\ref{fig:specs-mpileaks-abstract} shows the recursive nature of spec syntax:
+%
+\begin{minted}[fontsize=\scriptsize]{bash}
+$ spack install mpileaks@2.3 ^callpath@1.0+debug ^libelf@0.8.11
+\end{minted}
+%
+The caret (\verb|^|) denotes constraints for a particular dependency.  In the DAG,
+we now see that there are now version constraints on {\tt callpath} and {\tt libelf},
+and the user has asked for the debug variant of the {\tt callpath} library.
 
-\subsubsection{Spec structure}
-To understand how specs works, consider the package structure.  The  {\tt Mpileaks} package
-is a class, and its metadata ({\tt version}, {\tt depends\_on}, etc.) statically describe
-its relationships with other packages.  In this case, there are two direct dependencies: 
-the {\tt callpath} library and {\tt mpi}.  Spack recursively inspects the class definitions
-for each of these packages to construct a graph of their relationships.  
-To avoid issues with incompatible ABIs, and to guarantee a consistent build, the DAG is constructed
-so that there is only ever one version of any particular package.
+Recall that Spack guarantees there will be only a single version of any package in 
+the spec DAG.  Therefore, within the same DAG, each dependency can be uniquely identified by 
+only its package name.  The user does not have to think about DAG connectivity to add
+constraints.  She need only know that a package depends, somehow on {\tt callpath}. 
+For the same reason, the constraint order does not matter; dependency constraints
+can appear in arbitrary order.
 
-For {\tt mpileaks}, 
-the initial graph is shown in Figure~\ref{fig:specs-mpileaks}.  This graph is the in-memory
-representation of a spec.
-
-Each package, or node, in the spec has a number of configuration parameters that control
-how the package will be built.  These are the version, compiler to build with, compiler version, 
-named build options (or {\it variants} like +debug or -debug), and the architecture to
-build for.  When Spack runs to install {\tt mpileaks}, it constructs a package object for each node
-in the spec DAG, and it traverses the DAG in a bottom-up fashion.  At each node, it
-invokes the package's {\tt install()} method, passing in a sub-DAG rooted at the
-package to be installed.  This is the {\tt spec} parameter in Figure~\ref{fig:mpileaks}.
-Package maintainers are responsible for examining the spec's configuration and adjusting the 
-build if necessary.
-
-\subsubsection{Spec syntax}
-A spec DAG has many degrees of freedom, and it is not reasonable to expect users to
-understand all of them.  In our experience at LLNL, users typically have a small number of 
-build constraints that they care about, and the others only confuse them.  This is part of why
-the HPC software ecosystem is so difficult to manage: there are simply too many parameters.
-However, the small set of important build constraints can be very specific, so we need the
-ability to specify all aspects of the configuration space if necessary.  For this reason, 
-we introduce a telescoping syntax for package specs that allows users of Spack to
-specify {\it only those constraints they care about}, but to be {\it specific when necessary}.
-This is the true power of our syntax---specs are concise enough to use on the command line,
-and users can request installs or query existing installs by their structure.
-
-Table~\ref{tab:specs} shows examples of specs, ranging from very simple to very complex.
-Next to each is a description of the spec in plain language.  In the most basic case,
-shown on line 1 in the table, a user can simply name the package they want to use, and have it
-built and installed a default configuration.  On the command line this would take the form:
-\newline\newline
-{\tt spack install mpileaks}
-\newline
-
-If the user wants more specificity, {\tt mpileaks} can be augmented with additional
-qualifiers.  Spack install will handle selecting a build configuration that satisfies
-the user's request.  We describe additional qualifiers below.
+Table~\ref{tab:specs} shows further examples of specs, ranging from very
+simple to more complex. From these examples, we can see that Spack offers constraint
+notation to cover the rest of the HPC package parameter space.
 
 {\bf Versions.}
-Versions are specified by supplying {\tt @<version>} after the package name (line 2). 
-Versions can be precise (e.g. {\tt @4.5.1}) or 
-they can be ambiguous (e.g., @4) or a colon-separated, inclusive range (e.g., @4.4:4.5.1),
-and Spack will pick an available version from the specified range.
+The version constraint, denoted with {\tt @}, was already covered above. Versions
+can be precise ({\tt @2.5.1}) or denote a potentially open-ended 
+range ({\tt @2.5:}, {\tt @2.5:4.4}).
+
 The package in Figure~\ref{fig:mpileaks} lists two ``safe'' versions with checksums, but
@@
 {\bf Compilers.}
-To specify the compiler (line 3), the user
-simply adds {\tt \%} followed by its name, and if a version {\it of the compiler} 
-is desired, the same {\tt @} syntax can be used.  Here, {\tt gcc} refers to the full
-gcc compiler suite: C, C++, Fortran 77, and Fortran 90 compilers.  Spack can auto-detect
-compiler toolchains if they are in the user's PATH or they can be registered manually
+With a compiler constraint (shown on line 3) the user
+simply adds {\tt \%} followed by its name, along with an optional compiler version 
+specifier.  Spack compiler names, e.g. {\tt gcc}, refer to the full compiler {\it toolchain},
+i.e. the C, C++, Fortran 77, and Fortran 90 compilers.  Spack can auto-detect
+compiler toolchains if they are in the user's {\tt PATH} or they can be registered manually
 through a configuration file.
@@
 {\bf Variants.}
-To handle build-time options like debug flags and optional components, spack
-allows named flags, or {\it variants} to be associated with packages.  
-In the example (line 4), {\tt mpileaks} can be built with a {\tt debug} option by
-adding {\tt +debug}.  The option can be similarly disabled with {\tt -debug}.  
-This allows package authors to provide package-specific build options without 
-having to extend Spack.  We use names for these to simplify the spec
-syntax, and to prevent Spack's configuration space from becoming too fine-grained.
+To handle build options like compiler flags or optional components, specs can
+have named flags, or {\it variants}.  Variants are associated with the package,
+so the {\tt mpileaks} package implementor must check the spec and handle the cases
+where debug is enabled ({\tt +debug}) and disabled (with {\tt -debug}
+or {\tt ~debug}).  The names simplify the versioning and prevent
+Spack's configuration space from becoming too fine-grained.
 It would violate our goal of conciseness, for example, to include detailed 
-compiler flags in the spec syntax, but known sets of flags can be exported using
-named variants.
+compiler flags in the spec syntax, but known sets of flags can simply be named.
 
-{\bf Cross-compileation.}
+{\bf Cross-compilation.}
 To support cross-compilation, Spack includes the platform in the package spec (line 5).
-Platforms begin with {\tt =} and take names like {\tt linux-ppc64} or {\tt bbq}.  They are
+Platforms begin with {\tt =} and take names like {\tt linux-ppc64} or {\tt bgq}.  They are
 specified per-package; this allows front-end tools to depend on their back-end measurement
@@
 
-{\bf Dependencies.}
-Lines 6 and 7 in the table show the key feature that enables Spack's flexibility.
-The user can specify all of the above information not only for the package being
-installed, but also for its {\it dependencies}.  To do this, the user needs only supply 
-\verb|^| and the dependency's name.  If we need to build a new version with a specific
-version of {\tt mvapich2}, we can simply add, e.g., \verb|^mvapich2@1.9|
-to the spec, and it will build with that MPI version instead of the default.
-This can also be done for multiple libraries in the same spec (line 7).  
-As mentioned, Spack guarantees that there is only one version of any package in 
-the spec DAG.  So, within the same DAG, each dependency can be uniquely identified by 
-only its package name.  This frees the user from having to worry about DAG connectivity;
-the user only needs to know that a package depends, somehow, indirectly on {\tt mvapich}
-to add this constraint
-uniquely identifies a dependency, and dependency constraints can be specified in any
-order.
+\subsubsection{Constraints in packages}
 
-\subsubsection{Constraining dependencies}
+So far, we have shown examples of specs being used to request constraints from the
+command line, when {\tt spack install} is invoked.  However, the user is not the only
+source of constraints.  Applications may require specific versions of dependencies,
+and it is often desirable to write constraints like this into a package file.  For
+example, the ROSE compiler only builds with a certain version of the {\tt boost} library.
+At first glance, the {\tt depends\_on()} directive in Figure~\ref{fig:mpileaks} look
+like they take simple package names.  However, a package name is also a spec, and
+the same constraint syntax usable from the command line can be applied inside directives.
+So, the ROSE compiler can simply write:
+%
+\begin{lstlisting}
+depends_on(`boost@1.54.0')
+\end{lstlisting}
+%
+This constraint will be incorporated into the initial DAG node generated from
+the ROSE package.
+
+Spack also accepts spec syntax for conditionally patching packages.
+Spack offers a {\tt patch} directive that uses specs as predicates in an
+optional {\tt when} parameter.
+These concise directives in the Python package ensure that very specific patches
+are only applied to the Python source code when it is being built on Blue Gene/Q,
+with the appropriate compiler:
+%
+\begin{lstlisting}
+patch(`python-bgq-xlc.patch',   when=`=bgq%xlc')
+patch(`python-bgq-clang.patch', when=`=bgq%clang')
+\end{lstlisting}
+%
+The {\tt when} clause is matched against the Python package's spec
+to determine whether the patch will be applied.
+
+Constraint tests can also be used directly within package source code.
+For example, in the {\tt install} method, packages can test the versions of dependencies
+using constraint syntax:
+%
+\begin{lstlisting}
+  if '' in spec:
+  elif '
+  '
+\end{lstlisting}
+%
 
@@
 
-\subsubsection{Version extrapolation}
+In some cases, simple conditional logic
 
-\subsubsection{Querying installed packages}
+In our experience maintaining packages over time, we have found tha

### 2015-04-17 implementation-virtual-deps.tex
Lots of edits.

@@
 	\todo{.5 page}
+
+{\bf Dependencies.}
+Lines 6 and 7 in the table show the key feature that enables Spack's flexibility.
+The user can specify all of the above information not only for the package being
+installed, but also for its {\it dependencies}.  To do this, the user needs only supply 
+\verb|^| and the dependency's name.  If we need to build a new version with a specific
+version of {\tt mvapich2}, we can simply add, e.g., \verb|^mvapich2@1.9|
+to the spec, and it will build with that MPI version instead of the default.
+This can also be done for multiple libraries in the same spec (line 7).
+
+\subsubsection{Querying installed packages}

### 2015-04-17 implementation.tex
GLL edits
Fix typos.  Add gperftools section.
Lots of edits.

@@
 {\it Spack}, the Supercomputing PACKage manager.
-Spack is written in Python.  We chose Python because it is
-very widely used int he HPC community, and because it provides
-flexible scripting capabilities. 
+Spack is written in Python.  We chose Python for its flexibility
+its wide use in the HPC community.
 %
-Like some prior systems, Spack supports arbitrarily many software installations.
-It borrows ideas from Nix; packages are installed in their own prefixes,
-and configurations are managed using hashes.
+Like prior systems, Spack supports an arbitrary number of software
+installations, and like Nix it can identify them with hashes.  Unlike any
+prior system, Spack provides a language to specify and manage the
+combinatorial space of HPC software configurations.
 
 \noindent
-Spack adds several essential capabilities for HPC:
+Spack has the following unique capabilities:
 \begin{enumerate}
-\item {\bf Parametric builds} address the combinatorial problem:
-      Packages are parameterized by name, version, architecture, compiler, 
-      options, and dependencies.
-\item To manage the configuration space, spack provides a novel, 
+\item Packages are {\bf parametrized}, allowing new builds to be composed
+      on the fly for different versions, architecture, compiler, options, 
+      and dependencies.
+\item To manage this large parameter space, spack provides a novel, 
       recursive {\bf spec syntax} for dependency graphs.
@@
       {\bf versioned virtual dependencies}.
-\item Spack builds with {\bf compiler wrappers} that reduce the burden of build
-      consistency on the package author.
+\item Spack builds with {\bf compiler wrappers} that enforce build
+      consistency and simplify package writing.
 \end{enumerate}

### 2015-04-17 usecase-combinatoric.tex
Fix line wrapping in gperftools
Lots of edits.

@@
 
-The gperftools tool from Google has gained popularity among some developers for its high-performance thread-safe heap and its lite-weight profilers.
-Unfortunately, two issues made it difficult to maintain gperftools installations at LLNL.
-First, gperftools is a C++ library.
-Since C++ does not have standard ABI it needs to be re-built with each compiler and compiler version.
-Second, building gperftools on atypical architectures (such as Blue Gene/Q) requires patches and a complicated configure lines that change with each compiler.
-The first issue motivated application developers to try and maintain their own builds of gperftools that worked with their prefered compilers--a task that was made difficult by the second issue.
+The gperftools tool from Google has gained popularity among some developers for
+its high-performance thread-safe heap and its lite-weight profilers.  
+Unfortunately, two issues made it difficult to maintain gperftools installations 
+at LLNL.  First, gperftools is a C++ library.  Since C++ does not have standard 
+ABI it needs to be re-built with each compiler and compiler version.  Second, 
+building gperftools on atypical architectures (such as Blue Gene/Q) requires 
+patches and a complicated configure lines that change with each compiler.  The 
+first issue motivated application developers to try and maintain their own builds 
+of gperftools that worked with their prefered compilers--a task that was made 
+difficult by the second issue.
 
-Spack presented a solution to both problems.
-Site administrators can use Spack to easily maintain a central install of gperftools across the combinations of versions and compilers.
-Spack's gperftools package also serves as a central repository for the knowledge of how to build gperftools on each platform and compiler combination.
+Spack presented a solution to both problems.  Package administrators can use Spack to 
+easily maintain a central install of gperftools across the combinations of versions 
+and compilers.  Spack's gperftools package also serves as a central repository for 
+the knowledge of how to build gperftools on each platform and compiler combination.
 
 \begin{figure}
-  \includegraphics[width=\columnwidth]{code/gperftools.pdf}
+\begin{minted}[linenos,
+               numbersep=5pt,
+			   fontsize=\scriptsize,
+               frame=lines,
+               framesep=2mm]{python}
+patch("patch.gpeftools2.4_xlc", when="@2.4 %xlc")
+
+def install(self, spec, prefix):
+    if spec.architecture == "bgq" and self.compiler.satisfies("xlc"):
+        configure("--prefix=" + prefix,  "LDFLAGS=-qnostaticlink")
+    elif spec.architecture == "bgq":
+        configure("--prefix=" + prefix,  "LDFLAGS=-dynamic")
+    else:
+        configure("--prefix=" + prefix)
+
+    make()
+    make("install")
+\end{minted}
   \caption{
-    Simplified install routine for the gperftools tool.
+    Simplified install routine for gperftools.
     \label{fig:gperftools}
@@
 
-Figure~\ref{fig:gperftools} illustrates per-compiler and platform build rules with a simplified version of the install routine for the gperftools tool (Spack's real install routine for gperftools includes other compilers and more options per compiler).
-It applies a patch if gperftools 2.4 is built with the XLC compiler, and it selects the correct configure line based on the current platform and compiler.
+Figure~\ref{fig:gperftools} illustrates per-compiler and platform build rules with 
+a simplified version of the install routine for the gperftools tool (Spack's real 
+install routine for gperftools includes other compilers and more options per
+compiler).  It applies a patch if gperftools 2.4 is built with the XLC compiler, 
+and it selects the correct configure line based on the current platform and compiler.
 

### 2015-04-17 implementation-specs.tex
add specialization.

@@
 %
-\begin{lstlisting}
-depends_on(`boost@1.54.0')
-\end{lstlisting}
+\begin{minted}[fontsize=\scriptsize]{python}
+    depends_on('boost@1.54.0')
+\end{minted}
 %
@@
 
-Spack also accepts spec syntax for conditionally patching packages.
-Spack offers a {\tt patch} directive that uses specs as predicates in an
-optional {\tt when} parameter.
-These concise directives in the Python package ensure that very specific patches
-are only applied to the Python source code when it is being built on Blue Gene/Q,
-with the appropriate compiler:
+Spack's {\tt patch} directive also accepts spec syntax as a predicates in an
+optional {\tt when} parameter.  For example, these concise directives in the 
+Python package ensure that specific patches are applied to the Python source
+code when it is being built on Blue Gene/Q, with the appropriate compiler:
 %
-\begin{lstlisting}
-patch(`python-bgq-xlc.patch',   when=`=bgq%xlc')
-patch(`python-bgq-clang.patch', when=`=bgq%clang')
-\end{lstlisting}
+\begin{minted}[fontsize=\scriptsize]{python}
+    patch('python-bgq-xlc.patch',   when='=bgq%xlc')
+    patch('python-bgq-clang.patch', when='=bgq%clang')
+\end{minted}
 %
-The {\tt when} clause is matched against the Python package's spec
-to determine whether the patch will be applied.
-
-Constraint tests can also be used directly within package source code.
-For example, in the {\tt install} method, packages can test the versions of dependencies
-using constraint syntax:
+Constraints in the {\tt when} clause are matched against the Python package spec.
+Outside of directives, constraints can be used directly
+with the {\tt spec} object in the {\tt install} method:
 %
-\begin{lstlisting}
-  if '' in spec:
-  elif '
-  '
-\end{lstlisting}
+\begin{minted}[fontsize=\scriptsize]{python}
+  def install(self, spec, prefix):
+      if spec.satisfies('%gcc'):
+          # Handle gcc
+      elif spec.satisfies('%xlc'):
+          # Handle XL compilers
+      ...
+\end{minted}
 %
@@
 
-In some cases, simple conditional logic
+In our experience maintaining packages at LLNL, we have sometimes had to change
+entire package build scripts due to large changes in the way certain packages build.
+This can be very cumbersome, and it is difficult to maintain both the old and new
+version of a build script, but we must if we want to keep installing older versions
+for users who rely on them.
 
-In our experience maintaining packages over time, we have found tha
+\begin{figure}
+\begin{minted}[linenos,
+               numbersep=5pt,
+			   fontsize=\scriptsize,
+               frame=lines,
+               framesep=2mm]{python}
+class Dyninst(Package):
+    """"Dyninst binary instrumentation and analysis tool."""
+    url="http://www.paradyn.org/release8.1/DyninstAPI-8.1.1.tgz"
+    version('8.1.1', 'd1a04e995b7aa70960cd1d1fac8bd6ac')
+    depends_on("libelf")                                                                            
+    depends_on("libdwarf")                                                                          
+    depends_on("boost@1.42:")
+
+    # default build version uses cmake                                                                        
+    def install(self, spec, prefix):
+        with working_dir('spack-build', create=True):                                               
+            cmake('..', *std_cmake_args)
+            make()                                                                                  
+            make("install")
+
+    # when version is 8.1 or lower, use configure.                                                                                                    
+    @when('@:8.1')                                                                                  
+    def install(self, spec, prefix):                                                                
+        configure("--prefix=" + prefix)                                                             
+        make()                                                                                      
+        make("install")     
+\end{minted}
+\caption{
+    Specialized {\tt install} method in Dyninst. 
+	\label{fig:specialization}
+}
+\end{figure}
+
+For cases like this, Spack provides functionality that allows Python functions to have
+multiple definitions, with some specialized for particular configurations of the package.
+This allows us to have two separate implementations of {\tt install} or {\it any} method 
+in a package class. Figure~\ref{fig:specialization} shows how this is used in the Dyninst
+package.  The {\tt @when} directive is a Python decorator: a higher order function that 
+takes a function definition as a parameter and returns a new function to put in its place.
+Here, we replace the function with a callable multi-function dispatch object, and we 
+integrate the predicate check into the function dispatch mechanism.  Here, the
+{\tt when} condition is true when Dyninst is at version 8.1 or lower, and in those cases
+the package will use the {\tt configure}-based build.  By default if no predicate matches,
+{\tt install} will simply use the default CMake-based implementation. The simple {\tt @when}
+annotation allows us to maintain our old build code alongside the new version without 
+accumulating complex logic in a single {\tt install} function.

### 2015-04-17 introduction.tex
Lots of edits.
tweaks to introduction

@@
 (LLNL) supports around 2,500 users on 25 different clusters, ranging 
-in size from a 1.6 teraflop, 256-core cluster to to the
+in size from a 1.6 teraflop, 256-core cluster to the
 20 petaflop, 1.6 million-core Sequoia machine, currently ranked first and
@@
 
-To support the diverse needs of applications, administrators and application
-developers 
+To support the diverse needs of applications, system administrators 
+and application developers 
 frequently build, install, and support many different configurations
@@
 Frequently, applications must be rebuilt to fix bugs and to support
-new versions of the OS, MPI, compilers, and other dependencies.
+new versions of the operating system (OS), Message Passing Interface (MPI)
+implementation, compilers, and other dependencies. 
 Unfortunately, building scientific software is notoriously complex, with
@@
 \begin{enumerate}
-\item A novel, recursive syntax for specifying, building, and querying
-      HPC packages in a large parameter space;
+\item A novel, recursive syntax for concisely specifying and querying
+      the large parameter space of HPC packages;
 \item A build methodology that ensures packages find their dependencies
@@
 \item Spack: An implementation of these concepts; and
-\item Three use cases detailing LLNL's use of Spack to rapidly compose
-      software configurations, manage Python installations, and implement
-      site-specific build policies.
+\item Three use cases that detail LLNL's use of Spack to compose
+      software configurations rapidly, to manage Python installations, and 
+      to implement site-specific build policies.
 \end{enumerate}
 
-Spack is in active development at LLNL, and our use cases outline some of
+\noindent Spack is in active development at LLNL, and our use cases outline some of
 its first applications in our environment.  Spack solves software problems 
-that are pervasive at large, multi-user HPC centers, and we believe that our
-experiences with it can be applied at other HPC facilities.  Spack improves
+that are pervasive at large, multi-user HPC centers, and our experiences 
+are relevant to the full range of HPC facilities.  Spack improves
 operational efficiency by simplifying the build and deployment of 

### 2015-04-17 motivation.tex
various changes to common practice to improve readability and to stress state of the practice issues more

@@
 APT~\cite{foster+:rpm03,silva:apt01,yum} are integrated with most
-operating system (OS) distributions, and they are used to ensure that dependencies
+OS distributions, and they are used to ensure that dependencies
 are installed before packages that require them.
 These tools largely solve the problem of managing a {\it single} software
-stack.  This works well for the baseline OS and drivers, which are 
+stack, which works well for the baseline OS and drivers, which are 
 common to all applications on a system.
-These tools make the assumption that there is only a single version
-of each package, and most install packages in a single, inflexible location.
+These tools assume that each package only has a single version
+and most of the tools install packages in a single, inflexible location.
 To install multiple configurations, the user must create a custom, combinatoric
@@
 prefix~\cite{groffen:gentoo-prefix}, but again the burden is on package
-creators to manage conflicts. This restricts installations to only a few
-configurations.
+creators to manage conflicts. This burden effectively restricts installations 
+to few configurations.
 
@@
 Packaging problems arise in HPC because a supercomputer's hardware, OS, and
-filesystem are shared by many users with different requirements.  The classic solution
-to this problem is to use virtual
+filesystem are shared by many users with different requirements.  The classic 
+solution to this problem is to use virtual
 machines (VMs)~\cite{barham2003xen,rosenblum1999vmware,smith2005architecture}
 or lightweight virtualization techniques like Linux 
-containers~\cite{felter2014updated,merkel2014docker}. This model allows each user to 
-have her own environment with its own package manager, and it has been extremely 
-successful for servers at cloud data centers.  VMs typically have 
-near-native compute performance, but there are still major performance issues with
-low-level HPC network drivers.  VMs are not well supported on many
+containers~\cite{felter2014updated,merkel2014docker}. This model allows each
+ user to have a personalized environment with its own package manager, and it 
+has been extremely successful for servers at cloud data centers. VMs typically
+have near-native compute performance but low-level HPC network drivers still 
+exhibit major performance issues. VMs are not well supported on many
 non-Linux operating systems, an issue for the lightweight 
@@
 is another major concern: a profusion of VMs makes mandatory patching
-difficult when user's environments differ significantly from those of the facility.
+difficult when user's environments differ significantly from those of 
+the facility. Perhaps most importantly, significant differences in
+versioning leaves the majority of the installation burden on end users.
 
@@
 packages managers and either manual or semi-automated software installation.  
-For the baseline OS, it is common for sites to maintain traditional binary
+For the baseline OS, many sites maintain traditional binary
 packages using the vendor's package manager. LLNL maintains a Linux 
@@
 For more specific builds, many sites adhere to detailed naming conventions
-that encode information in filesystem paths. A sampling of conventions
-at several sites is shown in Table~\ref{tab:naming-conventions}. At LLNL, the
-APT package manager is used for installs in the {\tt /usr/local/tools} 
-filesystem, and {\tt /usr/global/tools} is used for manual installs.
+that encode information in file system paths. 
+Table~\ref{tab:naming-conventions} shows a sampling of conventions
+at several sites. At LLNL, we use the APT package manager for installs 
+in the {\tt /usr/local/tools} file system; we use {\tt /usr/global/tools} 
+for manual installs.
 ORNL uses hand installs but adheres to a strict scripting structure
@@
 save for the version of a particular dependency library.  In our experience
-at LLNL, naming conventions like these have not succeeded because there
-are more desired configurations than we can represent with a practical
+at LLNL, naming conventions like these have not succeeded because the full
+set of users want more configurations than we can represent with a practical
 directory hierarchy. Staff frequently install nonconforming packages 
@@
 system library paths such as {\tt /lib}, {\tt /usr/lib}, and
-{\tt /usr/local/lib}.  If binaries are installed to other locations, the
-{\it user} who runs the program must typically add dependency library paths to
+{\tt /usr/local/lib}.  If binaries are installed in other locations, the
+{\it user} who runs the program (often not the user or staff member who
+installed the binary) must typically add dependency library paths to
 {\tt LD\_LIBRARY\_PATH} (or a similar environment variable) so that the loader
@@
 
-Many HPC sites address this problem using {\it environment modules}, which allow
-users to dynamically ``load'' and ``unload'' such settings using simple commands.
-Environment emerged in 1991, and there are many implementations 
+Many HPC sites address this problem using {\it environment modules}, which 
+allow users to ``load'' and ``unload'' such settings dynamically using simple 
+commands. Environment emerged in 1991, and there are many implementations 
 available~\cite{dotkit,furlani+:lisa91,furlani+:lisa96,mclay:lmod,mclay:lmod-tutorial}.
@@
 provide software hierarchies similar to the naming conventions in 
-Table~\ref{tab:naming-conventions}. They allow users to quickly load a software stack
-if they know which one is required.
+Table~\ref{tab:naming-conventions}. They allow users to load a software stack
+quickly if they know which one is required.
 
-The alternative to per-user environment settings is to embed library search paths
-in installed binaries at compile time. When set this way, the search path is called
-an {\tt RPATH}. {\tt RPATHs} and environment modules are not mutually exclusive.
-Modules can still be used to set {\tt MANPATH}, {\tt PATH} and variables, but
-adding {\tt RPATHs} ensures that binaries will run correctly regardless of whether
-the right module is loaded. LC installs software with both {\tt RPATHs} and
-{\tt dotkit}~\cite{dotkit} modules.
+The alternative to per-user environment settings is to embed library search 
+paths in installed binaries at compile time. When set this way, the search 
+path is called an {\tt RPATH}. {\tt RPATHs} and environment modules are not 
+mutually exclusive. Modules can still be used to set {\tt MANPATH}, {\tt PATH}
+and variables, but adding {\tt RPATHs} ensures that binaries run correctly 
+regardless of whether the right module is loaded. LC installs software with 
+both {\tt RPATHs} and {\tt dotkit}~\cite{dotkit} modules.
 
@@
 arbitrarily many software configurations.  As at most HPC sites, each package
-in Nix installed in a unique prefix, but in Nix, there is no human-readable
+in Nix is installed in a unique prefix, but Nix does not have a human-readable
 naming convention.  Instead, Nix determines the prefix by hashing the package
-file and along with its dependencies.
+file along with its dependencies.
 
@@
 Lmod~\cite{geimer+:hust14}.  EasyBuild groups the compiler, MPI, FFT, and
-BLAS libraries together in a {\it toolchain}, which can be combined with
+BLAS libraries together in a {\it toolchain} that can be combined with
 any package file. This provides some degree of composability and
@@
 
-\paragraph{Gaps in the state of the art}
+\paragraph{Gaps in current common practice}
 The cryptographic versioning of Nix and Hashdist is very flexible. It versions
 the package {\it and} its dependency configuration, and can represent any
-configuration. However, there is no way for users to navigate or query
-software. The systems are in a sense ``write-only'':
+configuration. However, users cannot navigate or query the installed software.
+The systems are in a sense ``write-only'':
 Nix does not even offer a way to ask for a package's dependencies.
@@
 Indeed, the number of package files required for existing systems scales with
-the number of version {\it combinations}, not the number of packages.
-This quickly becomes unmanageable.  As an example, the very flexible EasyBuild
+the number of version {\it combinations}, not the number of packages, which.
+quickly becomes unmanageable.  As an example, the very flexible EasyBuild
 system has over 3,300 files for several permutations of only 600 packages. 
-Other systems have similar issues. A mechanism is needed to {\it parameterize}
-packages so that new builds can be {\it composed} on the fly.
+Other systems have similar issues. HPC sites need a mechanism to 
+{\it parameterize} packages so that new builds can be {\it composed} on the fly.

### 2015-04-17 implementation-concretization.tex

@@
 	\todo{.75 page}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
+	\caption{
+		Concretized version of specs from Figure~\ref{fig:specs-mpileaks}.
+		\label{fig:specs-mpileaks-concrete}
+	}
+\end{figure}
+
+{\bf Dependencies.}
+Lines 6 and 7 in the table show the key feature that enables Spack's flexibility.
+The user can specify all of the above information not only for the package being
+installed, but also for its {\it dependencies}.  To do this, the user needs only supply 
+\verb|^| and the dependency's name.  If we need to build a new version with a specific
+version of {\tt mvapich2}, we can simply add, e.g., \verb|^mvapich2@1.9|
+to the spec, and it will build with that MPI version instead of the default.
+This can also be done for multiple libraries in the same spec (line 7).
 

### 2015-04-17 implementation-virtual-deps.tex

--- 2015-04-16 implementation-virtual-deps.tex
@@
 \subsection{Versioned Virtual Dependencies}
-	\todo{.5 page}
+
+\begin{figure}
+%
+\begin{subfigure}{\linewidth}
+	\begin{minted}[numbersep=5pt,
+	              fontsize=\scriptsize,
+	              frame=lines,
+                  framesep=2mm]{python}
+class Mvapich2(Package):    
+    provides('mpi@:2.2', when='@1.9')
+    provides('mpi@:3.0', when='@2.0')
+\end{minted}
+	\caption{%
+		{\tt mvapich2} provides {\tt mpi} interface with specific versions.
+		\label{fig:virtual-deps-provides}
+	}
+\end{subfigure}
+%
+\begin{subfigure}{\linewidth}
+	\begin{minted}[numbersep=5pt,
+	              fontsize=\scriptsize,
+	              frame=lines,
+                  framesep=2mm]{python}
+class Mpileaks(Package):    
+    depends_on('mpi')
+\end{minted}
+	\caption{%
+		{\tt mpileaks} depends on any MPI version.
+		\label{fig:virtual-deps-mpileaks}
+	}
+\end{subfigure}
+%
+\begin{subfigure}{\linewidth}
+	\begin{minted}[numbersep=5pt,
+	              fontsize=\scriptsize,
+	              frame=lines,
+                  framesep=2mm]{python}
+class Gerris(Package):
+    depends_on('mpi@2:')
+\end{minted}
+	\caption{%
+		Gerris CFD solver requires MPI at version 2 or higher.
+		\label{fig:virtual-deps-depends}
+	}
+\end{subfigure}
+%
+\caption{
+	Virtual dependencies.
+	\label{fig:virtual-deps}
+}
+\end{figure}
+
+Many libraries share a common source interface and can be interchanged for
+one another in a build.
+The archetypal example in HPC is the MPI interface, which has a number of
+open source implementations (MPICH, OpenMPI, MVAPICH, etc.) and vendor-specific
+implementations. An application that can be built with one MPI can generally be
+built with another.  BLAS implementations such as ATLAS, LAPACK-BLAS, MKL, etc.
+are another common example.
+
+At LLNL, we frequently need to build tools with many versions of MPI to support
+the many different applications that are run at our center.
+However, MPI has no ABI, so applications cannot simply be re-linked;
+they must be recompiled and reinstalled. 
+Complicating matters, the MPI interface is versioned, and some
+packages need later versions of MPI to run correctly.  Often, MPI
+{\it implementation} versions do not correspond obviously to MPI
+interface versions, and determining the right version of 
+MPI to pair with an application can be tedious and tricky.
+
+To allow rapid composition of libraries by interface, Spack supports
+{\it virtual dependencies}.  A virtual dependency is an abstract name
+representing a library interface or capability instead of a library 
+implementation.  Packages that need this capability need not depend on 
+a specific implementation; they can depend on the virtual name, for which
+the user or Spack can select an an implementation at build time.
+Other package managers support the notion of virtual dependencies, but Spack
+adds {\it versioning} to its interfaces.  This allows concepts like MPI
+interface versions and BLAS levels to be represented directly, and Spack can
+handle the details of managing complex constraints while the user focuses on 
+composing software.
+
+Figure~\ref{fig:virtual-deps-provides} shows how packages can provide
+virtual interfaces in Spack.  The spec syntax is used to concisely
+associate a range of {\tt mpi} versions (up to 2.2) with Mvapich version 1.9.
+In Figure~\ref{fig:virtual-deps-mpileaks}, we see that {\tt mpileaks} requires
+MPI but is unconstrained on version.  Since this spec satisfies the constraints in 
+the {\tt mvapich2} package, {\tt Mvapich2} can be used to satisfy the {\tt mpi}
+dependency in {\tt mpileaks}.  Similarly, the Gerris CFD library needs MPI
+at version 2 or higher -- either {\tt mvapich2} 1.9 or 2.0 can be used to satisfy
+this constrained dependency.

### 2015-04-17 implementation-concretization.tex
Add concretization fig
Concretization section.

--- 2015-04-16 implementation-concretization.tex
@@
 
-\subsection{Abstract Specs \& Concretization}
-	\todo{.75 page}
+\subsection{Concretization}
+
+We have discussed Spack's internal software DAG model, and we have shown
+how the spec syntax can be used to quickly specify a partially constrained
+software DAG.  We say such a DAG is {\it abstract}, i.e. it could potentially describe
+more than one unique software configuration. Before Spack builds a spec, it
+must ensure the folioing conditions:
+\begin{enumerate}
+\item No package in the spec DAG is missing dependencies.
+\item No package in the spec DAG is virtual.
+\item All parameters are set for all packages in the DAG.
+\end{enumerate}
+If a spec meets all of these criteria, we say it is {\it concrete}.
+The central component of the Spack build that allows it to reduce a 
+highly unconstrained abstract description to a concrete build, is
+called {\it concretization}.
+
+\begin{figure*}
+	\centering
+	\includegraphics[width=\textwidth]{figs/concretization.pdf}
+	\caption{
+		Concretization resolves an abstract spec DAG into a concrete spec that can be installed.
+		\label{fig:concretization}
+	}
+\end{figure*}
+
+Spack's concretization algorithm is shown in Figure~\ref{fig:concretization}.
+The process starts when a user invokes {\tt spack install} and requests that
+a spec be built.  Spack converts the user's spec to an abstract spec DAG.
+It then builds a {\it separate} spec DAG for any constraints that are encoded
+in directives in package files.
+
+Spack intersects the constraints of the two DAGs package by package.  It checks
+each parameter for inconsistencies.  Inconsistencies can arise if, for example,
+the user inadvertently requests two versions of the same package, or if a
+package file explicitly requests a different version from what the user requested.
+Likewise, if the package and the user specified different compilers, variants,
+or platforms for particular packages, this will cause the user to be notified
+of the contradiction. If the user or package specified version ranges, they are
+intersected, and if the ranges do not overlap, an error will be raised.
+When the intersection succeeds, Spack has a single spec DAG with the merged
+constraints of the user and the package files.
+
+The next part of the process is iterative.
+If any node in the DAG is a virtual dependency, spack replaces it with with a
+suitable concrete interface provider.  It does this by building a reverse
+index from virtual packages to providers using the {\tt provides when} 
+directives discussed in Section~\ref{sec:virtual}. If multiple providers
+could satisfy the constraints on the virtual spec, 
+Spack will consult site and user policies to select the ``best'' possible
+provider.  The selected provider may {\it itself} have virtual dependencies,
+so Spack repeats this process until there are no more virtual packages
+in the DAG.
+
+With the now non-virtual DAG, Spack again consults site and user preferences for 
+variants, compilers, and versions to resolve any remaining abstract nodes.
+Adding a variant {\it may} cause a package to depend on more libraries. For example,
+if the site default behavior is to have a particular package build with {\tt +python},
+and this causes it to depend on Python libraries, then the cycle begins again.
+Spack currently avoids cycles by using a greedy algorithm.  It will not backtrack and
+try other options if the first policy choice leads to an inconsistency.  In future
+work we will investigate adding a SAT solver to Spack's concretization process to
+handle these cases, but in practice, we have found complex policy interactions to be
+very rare.
+
+When the process completes, it outputs a fully concrete Spec DAG.  An example 
+concrete DAG with architectures, compiler, versions, and variants, and all 
+dependencies resolved is shown in Figure~\ref{fig:specs-mpileaks-concrete}.
+The concrete DAG fulfills Spack's guarantees.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
+	\caption{
+		Concretized version of specs from Figure~\ref{fig:specs-mpileaks}.
+		\label{fig:specs-mpileaks-concrete}
+	}
+\end{figure}
+
+At install time, Spack constructs a package object for each node in the spec DAG
+and traverses the DAG in a bottom-up fashion.  At each node, it invokes the package's
+{\tt install()} method, using a sub-DAG rooted at the package to be installed as the {\tt spec}
+parameter to {\tt install()}. Package maintainers are responsible for examining the 
+concrete spec and adjusting the build if necessary.
+
+In our experience, much complexity in build scripts arises from the need for the script to
+query the environment and disambiguate complex environment settings.  If the build process is
+ill-defined, and the desired build configuration is often nonspecific, and it is these
+ambiguities that cause many build systems to fall over when faced with too many configuration
+options.
+
+In spack, build scripts are free from this concern, because the package author
+is guaranteed that a package's build spec is concrete at {\tt install()} time.
+There is thus no need to perform complicated state checks in spack build scripts;
+the {\tt spec} itself has the full configuration.  Other multi-configuration systems like Nix, 
+EasyBuild, and hashdist require the package author to write a mostly concrete build
+spec up front, and there is no way to set concretization policies in the way that Spack
+allows them.  We believe that this puts undue burden on the package author.  It also makes
+the task of separating site policies from user policies difficult.

### 2015-04-17 implementation-environment.tex
Fix typos.  Add gperftools section.
Module section

@@
 Spack's compiler wrappers have a number of useful effects.  First, they allow
-Spack to transparently parameterize the compiler for most builds and for
+Spack to transparently parametrize the compiler for most builds and for
 Spack users to easily request a build with a new compiler swapped in.
@@
 \subsubsection{Environment Module Integration}
-	\todo{.5 page}
+\label{sec:envmodule}
+In addition to build-time environments, Spack can also assist in create run-time
+environments.  A package may need environment variables like {\tt PATH}, 
+{\tt LD_LIBRARY_PATH}, or {\tt MANPATH} set before they can be used.  Some users
+may prefer to manually point environment variables and paths at packages, and 
+Section~\ref{sec:package-views} discusses how Spack can assist those users.  
+As discussed in Section~\ref{sec:motivation}, many sites rely on environment 
+modules to set their runtime environment.  Spack can automatically create 
+dotkit~\cite{dotkit} and Module configuration files for its packages, allowing 
+users to setup their runtime environment using familiar systems.
 
-TACC expects the user to run with the same
-environment that she built with~\cite{mclay:lmod-tutorial}.
-
-But, at LC, the user is
-frequently not the builder of the application, and may not know the particular
-incantation required to load the its environment. Moreover, users frequently run more 
-than one application in the same job, each built by a separate team, and each with
-its own environment. Most LC users are used to running programs with
-specific paths; only 10\% use environment modules extensively.
+Future versions of Spack may also allow the creation of Lmod~\cite{mclay:lmod} 
+module files.  Lmod allows users to load packages using a heirarchy of module 
+load operations.  For example, a user might load a {\tt gcc} module and a 
+{\tt mvapich2} module, and a subsequent load of the {\tt MPILeaks} package would 
+load the appropriate version for that compiler and MPI implementation.  Spack 
+understands where packages would fit in the Lmod hierarchy, allowing it to create 
+Lmod module files.
 
@@
 
-\begin{verbatim}
-module disadvantages:
-	- lmod solves a lot of problems
-	- not on all systems
-		- users often run from just a path.
-		- not guaranteed that LD_LIBRARY_PATH, etc will
-		  be set correctly.
-
-- Lmod assumes build & submit environment must match
-	- users often have NO IDEA what environment a package
-	  was built with
-	- multiple apps may be built with different stacks
-	- combining them in a workflow is difficult with modules
-		(need to know right incantation)
-
-	- TACC implements a rather complicated mapping back to
-	    the build env (Lmod CheckExec)
-
-- Spack solves the matrix problem mentioned in the Lmod
-    presentation (slide 98)
-	- hierarchy is nasty -- every branch represents a
-	  greedy choice
-	- might NEED a specific version of ONE dependency
-	  that is not supported in the tree
-	  => SPACK!
-
-- libmesh
-	- boost, petsc, trilinos, grvy, etc.
 \end{verbatim}

### 2015-04-17 usecase-policy.tex
Module section

@@
 
-\todo{fill in citations below}
 While Spack can easily create a dozen installs of a package like MPILeaks across different compilers and MPI implementations, it can be difficult for an end-user to determine which of those dozen packages should be used with their application.
-The module (cite) or dotkit (cite) package managers are commonly used on supercomputer systems to solve this problem, and Spack allows the package administrator to automatically create module and dotkit files for packages Spack installs.
+As discussed in Section~\ref{sec:envmodule}, environment modules are commonly used on supercomputer systems to solve this problem, and Spack allows the package administrator to automatically create module and dotkit files for packages Spack installs.
 

### 2015-04-18 implementation-specs.tex
GLL edits

@@
 \begin{figure}
-	\subfigure[Spec for {\tt mpileaks}]{
+	\begin{subfigure}{\linewidth}
 		\centering
 		\includegraphics[width=\columnwidth]{specs/mpileaks.pdf}
+		\caption{
+			Spec for {\tt mpileaks}
 			\label{fig:specs-mpileaks}
 		}
-	\subfigure[{\tt mpileaks@2.3}]{
+	\end{subfigure}
+%
+	\begin{subfigure}{\linewidth}
 		\centering
 		\includegraphics[width=\columnwidth]{specs/mpileaks-version}
+		\caption{
+			{\tt mpileaks@2.3}
 			\label{fig:specs-mpileaks-version}
 		}
-	\subfigure[{\tt mpileaks@2.3 \^{}callpath@1.0+debug \^{}libelf@0.8.11}]{
+	\end{subfigure}
+%
+	\begin{subfigure}{\linewidth}
 		\centering
 		\includegraphics[width=\columnwidth]{specs/mpileaks-abstract.pdf}
+		\caption{
+			{\tt mpileaks@2.3 \^{}callpath@1.0+debug \^{}libelf@0.8.11}
 			\label{fig:specs-mpileaks-abstract}
 		}
-	\subfigure[Spec (b), concretized.]{
-		\centering
-		\includegraphics[width=\columnwidth]{specs/mpileaks-concrete.pdf}
-		\label{fig:specs-mpileaks-concrete}
-	}
+	\end{subfigure}
+%
 	\caption{
-		Specs for the {\tt mpileaks} package.
-		\label{fig:specs}
+		Constraints applied to {\tt mpileaks} specs.
 	}
@@
 
-We begin with an example.%Let us explain our syntax with an example.
+We begin with a simple example.
 Consider the case where a user wants to install the {\tt mpileaks} package, but knows
-nothing about its structure.  To install the package, the user writes:
+nothing about its structure.  To install the package, the user invokes the {\tt spack install}
+command:
 %
@@
 If the user does not need a specific version of {\tt mpileaks}, but does require
-particular minimu version, then the user could use {\it version range} syntax.
+particular minimum version, then the user could use {\it version range} syntax.
 and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would use
 {\tt @2.3:2.5.6} to designate the range. In these cases, the user can save build time if
-Spack already has a version installed that satisfies the constraint -- Spack will just take
-the already built instead of building a new one.
+Spack already has a version installed that satisfies the constraint -- Spack will just use
+the previously-built installation instead of building a new one.
 
-Figure~\ref{fig:specs-mpileaks-abstract} shows the recursive nature of spec syntax:
+Figure~\ref{fig:specs-mpileaks-abstract} shows the recursive nature of the spec syntax:
 %
@@
 The caret (\verb|^|) denotes constraints for a particular dependency.  In the DAG,
-we now see that there are now version constraints on {\tt callpath} and {\tt libelf},
-and the user has asked for the debug variant of the {\tt callpath} library.
+we now see that there are version constraints on {\tt callpath} and {\tt libelf},
+and that the user has asked for the debug variant of the {\tt callpath} library.
 
@@
 only its package name.  The user does not have to think about DAG connectivity to add
-constraints.  She need only know that a package depends, somehow on {\tt callpath}. 
+constraints.  She need only know that a package depends, somehow, on {\tt callpath}. 
 For the same reason, the constraint order does not matter; dependency constraints
@@
 It would violate our goal of conciseness, for example, to include detailed 
-compiler flags in the spec syntax, but known sets of flags can simply be named.
+compiler flags in spec syntax, but known sets of flags can simply be named.
 
@@
 example, the ROSE compiler only builds with a certain version of the {\tt boost} library.
-At first glance, the {\tt depends\_on()} directive in Figure~\ref{fig:mpileaks} look
+At first glance, the {\tt depends\_on()} directives in Figure~\ref{fig:mpileaks} look
 like they take simple package names.  However, a package name is also a spec, and
@@
 
-Spack's {\tt patch} directive also accepts spec syntax as a predicates in an
+Spack's {\tt patch} directive also accepts spec syntax as a predicate in an
 optional {\tt when} parameter.  For example, these concise directives in the 
@@
 takes a function definition as a parameter and returns a new function to put in its place.
-Here, we replace the function with a callable multi-function dispatch object, and we 
+We replace the function with a callable multi-function dispatch object, and we 
 integrate the predicate check into the function dispatch mechanism.  Here, the
@@
 the package will use the {\tt configure}-based build.  By default if no predicate matches,
-{\tt install} will simply use the default CMake-based implementation. The simple {\tt @when}
+{\tt install} will use the default CMake-based implementation. The simple {\tt @when}
 annotation allows us to maintain our old build code alongside the new version without 

### 2015-04-18 usecase-combinatoric.tex
GLL edits

@@
 building gperftools on atypical architectures (such as Blue Gene/Q) requires
-patches and a complicated configure lines that change with each compiler.  The 
+patches and a complicated configure line that change with each compiler.  The
 first issue motivated application developers to try and maintain their own builds

### 2015-04-18 usecase-policy.tex
GLL edits

@@
 
-While Spack can easily create a dozen installs of a package like MPILeaks across different compilers and MPI implementations, it can be difficult for an end-user to determine which of those dozen packages should be used with their application.
+While Spack can easily create a dozen installations of a package like MPILeaks across different compilers and MPI implementations, it can be difficult for an end-user to determine which of those dozen packages should be used with their application.
 As discussed in Section~\ref{sec:envmodule}, environment modules are commonly used on supercomputer systems to solve this problem, and Spack allows the package administrator to automatically create module and dotkit files for packages Spack installs.
@@
 
-It is also useful to specify how site-specific options for how packages should be built.
+It is also useful to specify site-specific options for how packages should be built.
 The correct configure line for building packages may be different across sites.
 For example, the Spindle tool can have logging locations and security policies specified on its configure line.
-Spack can make reasonable guesses about Spindle's proper configure line, but it can not created a customized installation for the site.
+Spack can make reasonable guesses about Spindle's proper configure line, but it cannot create a customized installation for the site.
 One way to solve this would be to edit the Spindle package description in Spack and manually change how configure is invoked.

### 2015-04-18 implementation-concretization.tex
GLL edits
Updated concretization section.

@@
 how the spec syntax can be used to quickly specify a partially constrained
-software DAG.  We say such a DAG is {\it abstract}, i.e. it could potentially describe
-more than one unique software configuration. Before Spack builds a spec, it
-must ensure the folioing conditions:
+software DAG.  We say such a DAG is {\it abstract}, as it could potentially
+describe more than one software configuration. Before Spack builds a spec, it
+must ensure the following conditions:
 \begin{enumerate}
@@
 If a spec meets all of these criteria, we say it is {\it concrete}.
-The central component of the Spack build that allows it to reduce a 
-highly unconstrained abstract description to a concrete build, is
-called {\it concretization}.
+{\it Concretization} is the central component of the Spack build
+process that allows it to reduce a highly unconstrained abstract
+description to a concrete build.
 
@@
 	\caption{
-		Concretization resolves an abstract spec DAG into a concrete spec that can be installed.
+		Spack's concretization process.
 		\label{fig:concretization}
@@
 Spack's concretization algorithm is shown in Figure~\ref{fig:concretization}.
-The process starts when a user invokes {\tt spack install} and requests that
-a spec be built.  Spack converts the user's spec to an abstract spec DAG.
+The process starts when a user invokes {\tt spack install} and requests that a
+spec be built.  Spack converts the spec to an abstract DAG.
 It then builds a {\it separate} spec DAG for any constraints that are encoded
@@
 If any node in the DAG is a virtual dependency, spack replaces it with with a
-suitable concrete interface provider.  It does this by building a reverse
+suitable interface provider.  It does this by building a reverse
 index from virtual packages to providers using the {\tt provides when} 
@@
 and this causes it to depend on Python libraries, then the cycle begins again.
-Spack currently avoids cycles by using a greedy algorithm.  It will not backtrack and
-try other options if the first policy choice leads to an inconsistency.  In future
-work we will investigate adding a SAT solver to Spack's concretization process to
-handle these cases, but in practice, we have found complex policy interactions to be
-very rare.
+Spack currently avoids an exhaustive search by using a greedy algorithm.  
+It will not backtrack to try other options if its first policy choice leads
+to an inconsistency.  Rather, it will raise an error and the user must resolve
+the issue by being more explicit.  The user might toggle a variant, or she might
+force the build to use a {\it particular} MPI implementation
+by supplying, e.g. \verb|^openmpi| or \verb|^mpich|.  In future
+work we will investigate adding a SAT solver to Spack's concretization process
+to present users with better options.  In our experience, however, complex policy
+interactions like these are very rare.
 
-When the process completes, it outputs a fully concrete Spec DAG.  An example 
-concrete DAG with architectures, compiler, versions, and variants, and all 
+When the concretization process completes, it outputs a fully concrete spec DAG. 
+A concrete DAG with architectures, compiler, versions, and variants, and all 
 dependencies resolved is shown in Figure~\ref{fig:specs-mpileaks-concrete}.
-The concrete DAG fulfills Spack's guarantees.
+This fulfills Spack's guarantees.
 
@@
 	\caption{
-		Concretized version of specs from Figure~\ref{fig:specs-mpileaks}.
+		Concretized spec from Figure~\ref{fig:specs-mpileaks}.
 		\label{fig:specs-mpileaks-concrete}
@@
 and traverses the DAG in a bottom-up fashion.  At each node, it invokes the package's
-{\tt install()} method, using a sub-DAG rooted at the package to be installed as the {\tt spec}
-parameter to {\tt install()}. Package maintainers are responsible for examining the 
-concrete spec and adjusting the build if necessary.
+{\tt install} method.  For the {\tt spec} parameter to install, it uses a sub-DAG
+rooted at current node, which is also a concrete spec.  Package authors are
+responsible for querying the spec in {\tt install} and handling different configurations.
 
-In our experience, much complexity in build scripts arises from the need for the script to
-query the environment and disambiguate complex environment settings.  If the build process is
-ill-defined, and the desired build configuration is often nonspecific, and it is these
-ambiguities that cause many build systems to fall over when faced with too many configuration
-options.
+\subsubsection{Directory Layout}\label{sec:directory-layout}
+\begin{figure}\centering
+   \includegraphics[width=\linewidth]{specs/rpaths.pdf}
+   \caption{
+       {\tt mpileaks} built with {\tt mpich}, then {\tt openmpi}.
+       \label{fig:reuse}
+   }
+\end{figure}
 
-In spack, build scripts are free from this concern, because the package author
-is guaranteed that a package's build spec is concrete at {\tt install()} time.
-There is thus no need to perform complicated state checks in spack build scripts;
-the {\tt spec} itself has the full configuration.  Other multi-configuration systems like Nix, 
-EasyBuild, and hashdist require the package author to write a mostly concrete build
-spec up front, and there is no way to set concretization policies in the way that Spack
-allows them.  We believe that this puts undue burden on the package author.  It also makes
-the task of separating site policies from user policies difficult.
+We mentioned in Section~\ref{sec:packages} that each unique configuration is
+guaranteed a unique install prefix. Spack uses the concrete {\it spec} 
+to generate a unique path, shown in Table~\ref{tab:naming-conventions}.
+To prevent the directory name from growing too long, Spack uses a SHA hash of
+dependencies' specs as the last directory component.  This
+does {\it not} mean that Spack rebuilds every library for each new configuration.
+If two configurations share a sub-DAG, then the sub-DAG's configuration will
+be reused.  Figure~\ref{fig:reuse} shows how the {\tt dyninst} sub-DAG is used for
+both the {\tt mpich} and {\tt openmpi} builds of {\tt mpileaks}.
+
+\subsubsection{Reproducibility}
+For reproducibility and to preserve provenance, Spack stores a file containing
+the concrete spec in each
+install prefix. The spec can be examined later to understand how the installed
+package was built.  It an also be used to reproduce the build on another machine
+or at another site.
+
+\subsubsection{Site policies and build complexity}
+
+Our experience with manual installs helped us understand that much of the complexity
+of building HPC software comes from the size of the build parameter space.  
+As mentioned, there are only a few parameters most users care about, and the rest
+only serve to add complexity to the build.  When doing manual builds, LLNL staff
+tend to make arbitrary choices about many of the secondary build parameters,
+or they would include logic in build scripts to make many of these choices.
+In either scenario, publicly installed software was generally not installed
+in a consistent way.
+
+Concretization provides two benefits.  First, it allows users and staff to 
+request builds with a minimal spec expression, while still providing a
+mechanism for the site and the user to make consistent, repeatable choices about
+other build parameters.  For example, the site or the user can set default versions
+to use for {\it any} library that is not specified explicitly.
+%
+Second, concretization reduces the burden of packaging software because package authors
+do not have to make these decisions.  There is no
+need for packages to contain complicated checks, or for packages to be overly
+specific about versions.  Other multi-configuration systems like Nix, EasyBuild,
+and hashdist require the package author to write a mostly concrete build spec in advance.
+We believe that this puts undue burden on the package author, and it makes the task
+of changing site policies within a software stack difficult.  Spack
+separates these concerns.

### 2015-04-18 motivation.tex
edits to motivation

@@
 To cope with software diversity, many HPC sites use a combination of existing
-packages managers and either manual or semi-automated software installation.  
+package managers and either manual or semi-automated software installation.  
 For the baseline OS, many sites maintain traditional binary
 packages using the vendor's package manager. LLNL maintains a Linux 
-distribution, CHAOS~\cite{chaos} for this purpose.
+distribution, CHAOS~\cite{chaos} for this purpose, which is managed by RPM.
 
@@
 they also complicate the runtime environment. When launched, an executable
-must determine the location of its dependency libraries, or it will not run.
+must determine the location of its dependency libraries, or it will not run,
+or even worse, it may find the wrong dependencies and subtly produce incorrect results.
 Statically linked binaries do not have this issue, but modern
@@
 {\tt /usr/local/lib}.  If binaries are installed in other locations, the
-{\it user} who runs the program (often not the user or staff member who
+{\it user} who runs the program (often not the same person who
 installed the binary) must typically add dependency library paths to
@@
 mutually exclusive. Modules can still be used to set {\tt MANPATH}, {\tt PATH}
-and variables, but adding {\tt RPATHs} ensures that binaries run correctly 
+and other variables, but adding {\tt RPATHs} ensures that binaries run correctly 
 regardless of whether the right module is loaded. LC installs software with 
@@
 
-Hashdist~\cite{hashdist} is a meta-build system and package manager for HPC.
+HashDist~\cite{hashdist} is a meta-build system and package manager for HPC.
 Like Nix, it uses cryptographic versioning, and it archives installations
@@
 %
-Both Nix and hashdist use {\tt RPATHs} in their packages to ensure that
+Both Nix and HashDist use {\tt RPATHs} in their packages to ensure that
 libraries are found correctly.
@@
 \paragraph{Gaps in current common practice}
-The cryptographic versioning of Nix and Hashdist is very flexible. It versions
+The cryptographic versioning of Nix and HashDist is very flexible. It versions
 the package {\it and} its dependency configuration, and can represent any
@@
 EasyBuild and Smithy generate environment modules, which allow querying
-through their own interface. Naming in existing module systems, however, cannot
+through their own interface. Naming schemes used in existing module systems, however, cannot
 handle combinatoric versions. The Lmod authors call this the
@@
 Indeed, the number of package files required for existing systems scales with
-the number of version {\it combinations}, not the number of packages, which.
+the number of version {\it combinations}, not the number of packages, which
 quickly becomes unmanageable.  As an example, the very flexible EasyBuild

### 2015-04-18 implementation-packages.tex

@@
 
-\subsection{Packages}
+\subsection{Packages}\label{sec:packages}
 
@@
 base class.  {\tt Package} implements the bulk of the build process, but
-subclasses provide their own {\tt install()} method to handle the
-specifics of particular packages. Additionally,
-Spack implements an embedded domain-specific language (DSL).  It provides
-special directives such as {\tt depends\_on} and {\tt version} that add
-metadata to the package class.
+subclasses provide their own {\tt install} method to handle the
+specifics of particular packages. The subclass does {\it not} have to
+handle managing the install location.  Rather, Spack passes the {\tt install}
+method a {\tt prefix} parameter.  The package implementor must ensure that
+her {\tt install} function installs the package {\it into} the {\tt prefix},
+but Spack ensures that {\tt prefix} is computed in such a way that it is
+unique for every configuration of a package.  To further simplify 
+package implementation, Spack implements an embedded domain-specific
+language (DSL).  The DSL provides special directives such as {\tt depends\_on},
+{\tt version}, {\tt provides}, and {\tt patch} that add metadata
+to the package class.
 

### 2015-04-18 implementation-specs.tex

@@
 
+%class Dyninst(Package):
+%    """"Dyninst binary instrumentation and analysis tool."""
+%    url="http://www.paradyn.org/release8.1/DyninstAPI-8.1.1.tgz"
+%    version('8.1.1', 'd1a04e995b7aa70960cd1d1fac8bd6ac')
+%    depends_on("libelf")
+%    depends_on("libdwarf")
+%    depends_on("boost@1.42:")
+%
 \begin{figure}
@@
                framesep=2mm]{python}
-class Dyninst(Package):
-    """"Dyninst binary instrumentation and analysis tool."""
-    url="http://www.paradyn.org/release8.1/DyninstAPI-8.1.1.tgz"
-    version('8.1.1', 'd1a04e995b7aa70960cd1d1fac8bd6ac')
-    depends_on("libelf")                                                                            
-    depends_on("libdwarf")                                                                          
-    depends_on("boost@1.42:")
-
-    # default build version uses cmake                                                                        
-    def install(self, spec, prefix):
+    def install(self, spec, prefix):  # default build uses cmake
         with working_dir('spack-build', create=True):
@@
 
-    # when version is 8.1 or lower, use configure.                                                                                                    
-    @when('@:8.1')                                                                                  
+    @when('@:8.1')                    # <= 8.1 uses autotools
     def install(self, spec, prefix):

### 2015-04-18 implementation-virtual-deps.tex
Concretization section.

@@
 
-\subsection{Versioned Virtual Dependencies}
+\subsection{Versioned Virtual Dependencies}\label{sec:virtual}
 
 \begin{figure}
-%
-\begin{subfigure}{\linewidth}
 	\begin{minted}[numbersep=5pt,
@@
                   framesep=2mm]{python}
-class Mvapich2(Package):    
-    provides('mpi@:2.2', when='@1.9')
-    provides('mpi@:3.0', when='@2.0')
+ # providers of mpi                       # mpi dependents
+ class Mvapich2(Package):                 class Mpileaks(Package):
+     provides('mpi@:2.2', when='@1.9')        depends_on('mpi')
+     provides('mpi@:3.0', when='@2.0')           ...
+     ...                                  class Gerris(Package):
+ class Mpich(Package):                        depends_on('mpi@2:')
+     provides('mpi@:3', when='@3:')           ...
+     provides('mpi@:1', when='@1:')
+     ...
 \end{minted}
-	\caption{%
-		{\tt mvapich2} provides {\tt mpi} interface with specific versions.
-		\label{fig:virtual-deps-provides}
-	}
-\end{subfigure}
-%
-\begin{subfigure}{\linewidth}
-	\begin{minted}[numbersep=5pt,
-	              fontsize=\scriptsize,
-	              frame=lines,
-                  framesep=2mm]{python}
-class Mpileaks(Package):    
-    depends_on('mpi')
-\end{minted}
-	\caption{%
-		{\tt mpileaks} depends on any MPI version.
-		\label{fig:virtual-deps-mpileaks}
-	}
-\end{subfigure}
-%
-\begin{subfigure}{\linewidth}
-	\begin{minted}[numbersep=5pt,
-	              fontsize=\scriptsize,
-	              frame=lines,
-                  framesep=2mm]{python}
-class Gerris(Package):
-    depends_on('mpi@2:')
-\end{minted}
-	\caption{%
-		Gerris CFD solver requires MPI at version 2 or higher.
-		\label{fig:virtual-deps-depends}
-	}
-\end{subfigure}
-%
 \caption{
@@
 
-Figure~\ref{fig:virtual-deps-provides} shows how packages can provide
+Figure~\ref{fig:virtual-deps} shows how packages provide
 virtual interfaces in Spack.  The spec syntax is used to concisely
-associate a range of {\tt mpi} versions (up to 2.2) with Mvapich version 1.9.
-In Figure~\ref{fig:virtual-deps-mpileaks}, we see that {\tt mpileaks} requires
-MPI but is unconstrained on version.  Since this spec satisfies the constraints in 
-the {\tt mvapich2} package, {\tt Mvapich2} can be used to satisfy the {\tt mpi}
-dependency in {\tt mpileaks}.  Similarly, the Gerris CFD library needs MPI
-at version 2 or higher -- either {\tt mvapich2} 1.9 or 2.0 can be used to satisfy
-this constrained dependency.
+associate ranges of {\tt mpi} versions {\tt mvapich2} and {\tt mpich}
+The {\tt mpileaks} package requires {\tt mpi} but it does not constrain the version.
+Any version of {\tt mvapich2} or {\tt mpich} could be used to to satisfy the {\tt mpi}
+constraint. The Gerris CFD library, however, needs MPI version 2 or higher.  So any
+version except {\tt mpich} 1.x could be used to satisfy the constrained dependency.
+
+%{\bf Dependencies.}
+%Lines 6 and 7 in the table show the key feature that enables Spack's flexibility.
+%The user can specify all of the above information not only for the package being
+%installed, but also for its {\it dependencies}.  To do this, the user needs only supply 
+%\verb|^| and the dependency's name.  If we need to build a new version with a specific
+%version of {\tt mvapich2}, we can simply add, e.g., \verb|^mvapich2@1.9|
+%to the spec, and it will build with that MPI version instead of the default.
+%This can also be done for multiple libraries in the same spec (line 7).

### 2015-04-18 implementation-packages.tex
edit 3.1
props to homebrew

@@
 package implementation, Spack implements an embedded domain-specific
-language (DSL).  The DSL provides special directives such as {\tt depends\_on},
-{\tt version}, {\tt provides}, and {\tt patch} that add metadata
+language (DSL), similar to that of Homebrew~\cite{homebrew}.
+The DSL provides special directives such as {\tt depends\_on},
+{\tt version}, and {\tt provides} that add metadata
 to the package class.
@@
 Figure~\ref{fig:mpileaks} shows the package for \mpileaks, a tool developed
-at LLNL to find memory leaks in MPI programs.
+at LLNL to find leaks of MPI objects in programs.
 The {\tt MpiLeaks} class first provides simple metadata: a text description,

### 2015-04-18 implementation-environment.tex
Fixbuild.
Updated concretization section.
Trim a bit.

@@
 For example, in some packages that depend on the {\tt Silo} library,
-the {\tt ---with-silo} parameter takes a path to {\tt Silo}'s installation prefix.
+the \verb|--with-silo| parameter takes a path to {\tt Silo}'s installation prefix.
 In others, it takes the {\tt include} and {\tt lib} subdirectories,
 separated by a comma.
-The {\tt install()} method in Spack's package files allows us to record
+The {\tt install} method in Spack's package files allows us to record
 precise build incantations for later reuse.
@@
 Failure to specify the right version at build time has caused many
-inexplicable crashes.
+unexpected crashes.
 %
-Spack manages the build environment by running the {\tt install} invocation
+Spack manages the build environment by running each {\tt install} invocation
 in a new process.  It helps packages find dependencies 
@@
 These variables are commonly used by build systems to locate dependencies,
-and setting them helps to ensure that incorrect libraries are detected.
+and setting them helps to ensure that correct libraries are detected.
 The isolated build environment also gives package authors 
@@
 automatically\footnote{If builds do not respect {\tt CC}, {\tt CXX}, etc.,
-wrappers can typically be added as arguments or inserted into Makefiles
+wrappers can be added as arguments or inserted into Makefiles
 by {\tt install}.}.
@@
 These point to the {\tt include} and {\tt lib} directories of dependency
-library installations, where needed headers and libraries are typically located.
-After the modified argument list is constructed, the wrappers delegate it
-to the real compiler to execute.
+library installations, where needed headers and libraries are located.
+The wrappers then invoke the real compiler with the modified arguments.
 
 Spack's compiler wrappers have a number of useful effects.  First, they allow
-Spack to transparently parametrize the compiler for most builds and for
-Spack users to easily request a build with a new compiler swapped in.
-Second, they enforce the use of {\tt RPATHs} in
+Spack to transparently switch compilers in most builds.  This is how
+compiler options like {\tt \%gcc} are implemented.  Second, they enforce the
+use of {\tt RPATHs} in
 installed binaries.  This causes applications built by Spack to run correctly
@@
 \label{sec:envmodule}
-In addition to build-time environments, Spack can also assist in create run-time
-environments.  A package may need environment variables like {\tt PATH}, 
-{\tt LD_LIBRARY_PATH}, or {\tt MANPATH} set before they can be used.  Some users
-may prefer to manually point environment variables and paths at packages, and 
-Section~\ref{sec:package-views} discusses how Spack can assist those users.  
+In addition to managing the build-time environment, Spack can assist in managing
+the run-time environment.  A package may need environment variables like {\tt PATH}, 
+{\tt LD_LIBRARY_PATH}, or {\tt MANPATH} set before they can be used.  
 As discussed in Section~\ref{sec:motivation}, many sites rely on environment 
-modules to set their runtime environment.  Spack can automatically create 
+modules to set up the runtime environment.  Spack can automatically create simple
 dotkit~\cite{dotkit} and Module configuration files for its packages, allowing 
 users to setup their runtime environment using familiar systems.  
-
+%
 Future versions of Spack may also allow the creation of Lmod~\cite{mclay:lmod} 
-module files.  Lmod allows users to load packages using a heirarchy of module 
-load operations.  For example, a user might load a {\tt gcc} module and a 
-{\tt mvapich2} module, and a subsequent load of the {\tt MPILeaks} package would 
-load the appropriate version for that compiler and MPI implementation.  Spack 
-understands where packages would fit in the Lmod hierarchy, allowing it to create 
-Lmod module files.
-
-\begin{figure}\centering
-	\includegraphics[width=.6\linewidth]{specs/rpaths.pdf}
-	\newline
-	\verb|spack install mpileaks ^mpich|\newline
-	\verb|spack install mpileaks ^openmpi|
-	\caption{
-		Installing {\tt mpileaks} with MPICH (solid), then with OpenMPI (dashed).  The {\tt dyninst} subtree is
-		reused, but {\tt mpileaks} and {\tt callpath} must be rebuilt.
-	}
-\end{figure}
-
-\end{verbatim}
+hierarchies, as discussed in Section~\ref{sec:env-rpath}. Spack's rich 
+dependency information would allow such hierarchies to be generated automatically.

### 2015-04-18 implementation-packages.tex
homebrew moved up.
flow.
liked the old way after all.

@@
 package implementation, Spack implements an embedded domain-specific
-language (DSL), similar to that of Homebrew~\cite{homebrew}.
+language (DSL).
 The DSL provides special directives such as {\tt depends\_on},
-{\tt version}, and {\tt provides} that add metadata
+{\tt version}, {\tt patch}, and {\tt provides} that add metadata
 to the package class.

### 2015-04-18 implementation-specs.tex
edits to specs

@@
 	}
+	\label{fig:specs}
 \end{figure}
@@
 DAGs for {\tt mpileaks} are shown in Figure~\ref{fig:specs}.
-Each node is a package and has five configuration parameters that control
+Each node represents a package, and each package has five configuration parameters that control
 how it will be built: 1) the package version, 2) the compiler to
@@
 If the user does not need a specific version of {\tt mpileaks}, but does require
-particular minimum version, then the user could use {\it version range} syntax.
+particular minimum version, then the user could use {\it version range} syntax
 and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would use
@@
 The version constraint, denoted with {\tt @}, was already covered above. Versions
-can be precise ({\tt @2.5.1}) or denote a potentially open-ended
-range ({\tt @2.5:}, {\tt @2.5:4.4}).
+can be precise ({\tt @2.5.1}) or denote a range ({\tt @2.5:4.4}),
+which may be open-ended ({\tt @2.5:}).
 
@@
 i.e. the C, C++, Fortran 77, and Fortran 90 compilers.  Spack can auto-detect
-compiler toolchains if they are in the user's {\tt PATH} or they can be registered manually
+compiler toolchains if they are in the user's {\tt PATH}, or they can be registered manually
 through a configuration file.
@@
 where debug is enabled ({\tt +debug}) and disabled (with {\tt -debug}
-or {\tt ~debug}).  The names simplify the versioning and prevent
+or {\tt \textasciitilde \ignorespaces debug}).  The names simplify the versioning and prevent
 Spack's configuration space from becoming too fine-grained.
@@
 integrate the predicate check into the function dispatch mechanism.  Here, the
-{\tt when} condition is true when Dyninst is at version 8.1 or lower, and in those cases
+{\tt @when} condition is true when Dyninst is at version 8.1 or lower, and in those cases
 the package will use the {\tt configure}-based build.  By default if no predicate matches,

### 2015-04-18 implementation-virtual-deps.tex
edits to virtual deps

@@
 At LLNL, we frequently need to build tools with many versions of MPI to support
-the many different applications that are run at our center.
+the many different applications that run at our center.
 However, MPI has no ABI, so applications cannot simply be re-linked;
@@
 virtual interfaces in Spack.  The spec syntax is used to concisely
-associate ranges of {\tt mpi} versions {\tt mvapich2} and {\tt mpich}
-The {\tt mpileaks} package requires {\tt mpi} but it does not constrain the version.
+associate ranges of {\tt mpi} versions for the {\tt mvapich2} and {\tt mpich} packages.
+The {\tt mpileaks} package requires {\tt mpi}, but it does not constrain the version.
 Any version of {\tt mvapich2} or {\tt mpich} could be used to to satisfy the {\tt mpi}

### 2015-04-18 implementation-environment.tex
edits to env

@@
 header and library search paths for dependencies, header and library detection
-tests run by most build systems succeed automatically, {\it without}
+tests that are run by most build systems succeed automatically, {\it without}
 the need to use special arguments for nonstandard locations.  {\tt configure}
@@
 the run-time environment.  A package may need environment variables like {\tt PATH}, 
-{\tt LD_LIBRARY_PATH}, or {\tt MANPATH} set before they can be used.  
+{\tt LD_LIBRARY_PATH}, or {\tt MANPATH} set before it can be used.  
 As discussed in Section~\ref{sec:motivation}, many sites rely on environment 
-modules to set up the runtime environment.  Spack can automatically create simple
+modules to setup the runtime environment.  Spack can automatically create simple
 dotkit~\cite{dotkit} and Module configuration files for its packages, allowing 

### 2015-04-18 usecase-combinatoric.tex
Started Python section.

@@
 
-\subsection{Combinatoric Names}
+\subsection{Combinatoric Naming}
 \label{sec:usecase-combinatoric}
+%\subsubsection{Gperftools and MPILeaks}
 
-\subsubsection{Gperftools packaging}
-
-The gperftools tool from Google has gained popularity among some developers for
-its high-performance thread-safe heap and its lite-weight profilers.
-Unfortunately, two issues made it difficult to maintain gperftools installations
-at LLNL.  First, gperftools is a C++ library.  Since C++ does not have standard
-ABI it needs to be re-built with each compiler and compiler version.  Second,
-building gperftools on atypical architectures (such as Blue Gene/Q) requires
-patches and a complicated configure line that change with each compiler.  The
-first issue motivated application developers to try and maintain their own builds
-of gperftools that worked with their prefered compilers--a task that was made
-difficult by the second issue.
+{\tt Gperftools} is a suite of tools from Google that has gained popularity among
+developers for its high-performance thread-safe heap and its lightweight profilers.
+Unfortunately, two issues made it difficult to maintain {\tt gperftools} installations
+at LLNL.  First, {\tt gperftools} is a C++ library.  Since C++ does not have standard
+ABI, it needs to be re-built with each compiler and compiler version used by client
+code.  Second, building {\tt gperftools} on bleeding-edge architectures
+(such as Blue Gene/Q) requires patches and a complicated configure lines that
+change with each compiler.  An LLNL application team tried to maintain their own
+builds of {\tt gperftools} that worked with their preferred compilers, but this
+task was complicated by the second issue.
 
 Spack presented a solution to both problems.  Package administrators can use Spack to
-easily maintain a central install of gperftools across the combinations of versions
-and compilers.  Spack's gperftools package also serves as a central repository for
-the knowledge of how to build gperftools on each platform and compiler combination.
-
+easily maintain a central install of {\tt gperftools} across the combinations of
+versions and compilers.  Spack's {\tt gperftools} package also serves as a central
+institutional knowledge repository for builds, as its package files encode
+the {\tt gperftools} patches and configure line for each platform and compiler combination.
+We are investigating whether we can move some of these settings into compiler and
+architecture descriptors to further simplify Spack's build template for cross-platform
+installations.
 \begin{figure}
@@
 Figure~\ref{fig:gperftools} illustrates per-compiler and platform build rules with
-a simplified version of the install routine for the gperftools tool (Spack's real
-install routine for gperftools includes other compilers and more options per
-compiler).  It applies a patch if gperftools 2.4 is built with the XLC compiler,
-and it selects the correct configure line based on the current platform and compiler.
+a simplified version of the install routine for {\tt gperftools} (The full
+{\tt gperftools} install routine includes other compilers and more options).  It applies a patch if {\tt gperftools} 2.4 is built with the XL compiler,
+and it selects the correct configure line based on the spec's platform and compiler.
 
-\subsubsection{{\tt mpileaks} packaging}
-\todo{.5 page}
-\begin{verbatim}
-	- compiler x version x MPI x mpi version
-\end{verbatim}
+LLNL's {\tt mpileaks} tool, which we have used as a running example in this paper,
+has an even larger configuration space than {\tt gperftools}.  It must be built
+for different compilers, compiler versions, MPI implementations, and MPI implementation
+versions. As with {\tt gperftools}, we have been able to install many different
+configurations of MPILeaks using Spack.  Moreover, Spack's virtual dependency system
+allows us to
+quickly compose a new MPILeaks build when a new MPI is deployed at LLNL, {\it without}
+modifying the MPILeaks package itself. Though we simplified its dependency structure
+here, the {\tt callpath} library can be built with three different stack trace libraries:
+{\tt dyninst}, {\tt libunwind}, or {\tt glibc}.  For some codes we may soon need to build
+different versions of MPILeaks with different {\tt callpath} configurations.

### 2015-04-18 usecase-python.tex
Started Python section.

@@
 \label{sec:usecase-python}
+
+Python is becoming increasingly popular for HPC applications, 
+both due to its flexibility as a language and due to its excellent support
+for calling into native numerical code and achieving high performance. 
+Python applications frequently make native calls into functions in
+compiled numerical libraries like NumPy, FFTW, ATLAS, and other linear
+algebra libraries.
+
+LC maintains a number of Python installations for production application teams,
+and maintenance of these repositories has been increasingly complex for
+some time.  LC must maintain different Python installations containing different
+versions of Python modules for different code teams, and the modules must be able
+to make calls into optimized native libraries, for maximum performance.
+Python itself was easy to package with Spack; we built a package for it like any
+other and it compiled and ran
+
+they must also call into high performance math libraries. Python installations
+suffer from many of the environment problems we mentioned in Sections~\ref{env-rpath}.
+
+Python
+
+ \cite{eby:setuptools}
+
+Python's package management
+
+To support code teams using Python, LC maintains a number of
+
+both for rapid prototyping
+
+- LC maintains python installations for existing codes.
+
+- maintain separate instances for different code teams
+	- each code team needs to have particular python packages installed
+	- need fast builds of math libraries to go with python installs
+
+- model for python package installation doesn't really fit spack
+
+- implemented best of both worlds solution:
+	- combine spack's flexibility with python meta-package-manager convenience
+
+	- many other languages fit this model; could use it w/them.
 

### 2015-04-18 usecases.tex
flow.
Started Python section.

@@
 
+We have recently using Spack in production at 
+Livermore Computing.  We have already outlined a number of the
+experiences that drove us to develop Spack.
+In this section, we describe real-world use cases that Spack
+has been used to address.  In some cases, Spack had to be 
+adapted to meet production needs, and we describe how its
+flexibility has allowed us to put together solutions quickly.
+
 \input{usecase-combinatoric}

### 2015-04-18 usecase-combinatoric.tex
edits to usecase combo

@@
 Unfortunately, two issues made it difficult to maintain {\tt gperftools} installations
-at LLNL.  First, {\tt gperftools} is a C++ library.  Since C++ does not have standard
-ABI, it needs to be re-built with each compiler and compiler version used by client
-code.  Second, building {\tt gperftools} on bleeding-edge architectures
-(such as Blue Gene/Q) requires patches and a complicated configure lines that
-change with each compiler.  An LLNL application team tried to maintain their own
-builds of {\tt gperftools} that worked with their preferred compilers, but this
+at LLNL.  First, {\tt gperftools} is a C++ library.  Since C++ does not define a standard
+ABI, {\tt gperftools} must be re-built with each compiler and compiler version used by client
+applications.  Second, building {\tt gperftools} on bleeding-edge architectures
+(such as Blue Gene/Q) requires patches and complicated configure lines that
+change with each compiler.  One LLNL application team tried to maintain their own
+builds of {\tt gperftools} for their preferred compilers, but this
 task was complicated by the second issue.
@@
 Spack presented a solution to both problems.  Package administrators can use Spack to
-easily maintain a central install of {\tt gperftools} across the combinations of
+easily maintain a central install of {\tt gperftools} across combinations of
 versions and compilers.  Spack's {\tt gperftools} package also serves as a central
 institutional knowledge repository for builds, as its package files encode
-the {\tt gperftools} patches and configure line for each platform and compiler combination.
+the patches and configure lines required for each platform and compiler combination.
 We are investigating whether we can move some of these settings into compiler and
@@
 Figure~\ref{fig:gperftools} illustrates per-compiler and platform build rules with
-a simplified version of the install routine for {\tt gperftools} (The full
-{\tt gperftools} install routine includes other compilers and more options).  It applies a patch if {\tt gperftools} 2.4 is built with the XL compiler,
+a simplified version of the install routine for {\tt gperftools}.
+%(the full {\tt gperftools} install routine includes other compilers and more options).
+The build applies a patch if {\tt gperftools} 2.4 is built with the XL compiler,
 and it selects the correct configure line based on the spec's platform and compiler.
@@
 versions. As with {\tt gperftools}, we have been able to install many different
-configurations of MPILeaks using Spack.  Moreover, Spack's virtual dependency system
+configurations of {\tt mpileaks} using Spack.  Moreover, Spack's virtual dependency system
 allows us to
-quickly compose a new MPILeaks build when a new MPI is deployed at LLNL, {\it without}
-modifying the MPILeaks package itself. Though we simplified its dependency structure
-here, the {\tt callpath} library can be built with three different stack trace libraries:
-{\tt dyninst}, {\tt libunwind}, or {\tt glibc}.  For some codes we may soon need to build
-different versions of MPILeaks with different {\tt callpath} configurations.
+quickly compose a new {\tt mpileaks} build when a new MPI library is deployed at LLNL, {\it without}
+modifying the {\tt mpileaks} package itself.
+%Though we simplified its dependency structure
+%here, the {\tt callpath} library can be built with three different stack trace libraries:
+%{\tt dyninst}, {\tt libunwind}, or {\tt glibc}.  For some codes we may soon need to build
+%different versions of {\tt mpileaks} with different {\tt callpath} configurations.

### 2015-04-18 usecase-combinatoric.tex
edits to usecase combo

@@
 %(the full {\tt gperftools} install routine includes other compilers and more options).
-The build applies a patch if {\tt gperftools} 2.4 is built with the XL compiler,
+The package applies a patch if {\tt gperftools} 2.4 is built with the XL compiler,
 and it selects the correct configure line based on the spec's platform and compiler.

### 2015-04-18 usecase-combinatoric.tex
edits to usecase combo
replace LLNL with LC in usecase combo

@@
 Unfortunately, two issues made it difficult to maintain {\tt gperftools} installations
-at LLNL.  First, {\tt gperftools} is a C++ library.  Since C++ does not define a standard
+at LC.  First, {\tt gperftools} is a C++ library.  Since C++ does not define a standard
 ABI, {\tt gperftools} must be re-built with each compiler and compiler version used by client
@@
 (such as Blue Gene/Q) requires patches and complicated configure lines that
-change with each compiler.  One LLNL application team tried to maintain their own
+change with each compiler.  One application team tried to maintain their own
 builds of {\tt gperftools} for their preferred compilers, but this
-task was complicated by the second issue.
+task was complicated due to the second issue.
 
@@
 easily maintain a central install of {\tt gperftools} across combinations of
-versions and compilers.  Spack's {\tt gperftools} package also serves as a central
+compilers and compiler versions.  Spack's {\tt gperftools} package also serves as a central
 institutional knowledge repository for builds, as its package files encode
@@
 
-LLNL's {\tt mpileaks} tool, which we have used as a running example in this paper,
+LC's {\tt mpileaks} tool, which we have used as a running example in this paper,
 has an even larger configuration space than {\tt gperftools}.  It must be built
@@
 allows us to
-quickly compose a new {\tt mpileaks} build when a new MPI library is deployed at LLNL, {\it without}
+quickly compose a new {\tt mpileaks} build when a new MPI library is deployed at LC, {\it without}
 modifying the {\tt mpileaks} package itself.

### 2015-04-18 usecase-policy.tex
edits to usecase policy

@@
 
-While Spack can easily create a dozen installations of a package like MPILeaks across different compilers and MPI implementations, it can be difficult for an end-user to determine which of those dozen packages should be used with their application.
-As discussed in Section~\ref{sec:envmodule}, environment modules are commonly used on supercomputer systems to solve this problem, and Spack allows the package administrator to automatically create module and dotkit files for packages Spack installs.
+While Spack can easily create a dozen installations of a package like {\tt mpileaks} for different compilers and MPI implementations, it can be difficult for an end-user to determine which of those dozen packages should be used with their application.
+As discussed in Section~\ref{sec:envmodule}, environment modules are commonly used on supercomputer systems to solve this problem, and Spack allows the package administrator to automatically create module and dotkit files for packages that Spack installs.
 
-However, even when modules and dotkits are available many users will go directly through the filesystem to access package installs.
+However, even when modules and dotkits are available, many users will go directly through the filesystem to access package installs.
 Spack installs packages into paths based on concretized specs, which is ideal for maintaining multiple package installations, but may be difficult for an end-user to navigate.
-A version of MPILeaks, for example, may be installed in a location like {\tt spack/opt/chaos\_5\_x86\_64\_ib/ gcc@4.9.2/mpileaks@1.0-db465029}.
+A version of {\tt mpileaks}, for example, may be installed in a location like {\tt spack/opt/chaos\_5\_x86\_64\_ib/ gcc@4.9.2/mpileaks@1.0-db465029}.
 
@@
 Views provide a human-readable directory layout, which can be taylored to fit alongside existing directory layouts or used to create new layouts.
-For example, the above MPILeaks package may have a view that creates a symlink in {\tt /opt/mpileaks-1.0-openmpi} to the spack installation of MPILeaks.
-The same package install may be referenced by multiple symlinks and views, so the above package could also be symlinked from a more generic {\tt /opt/mpileaks-openmpi} symlink (which is useful for users who don't want to hardcode towards a specific version).
-Views can also be used to create symlinks to specific executables or libraries in an install, so a spack-built gcc 4.9.2 install may have a view that creates symlinks from {\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.
+For example, the above {\tt mpileaks} package may have a view that creates a symlink in {\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpilleaks}.
+The same package install may be referenced by multiple symlinks and views, so the above package could also be symlinked from a more generic {\tt /opt/mpileaks-openmpi} symlink (which is useful for users who don't want to hardcode a specific version).
+Views can also be used to create symlinks to specific executables or libraries in an install, so a Spack-built {\tt gcc 4.9.2} install may have a view that creates symlinks from {\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.
 
-Views are configured through a spackconfig file, which can be setup at a site-wide level or for individual users.
-For each package or set of packages, the spackconfig file contains rules describing the symlinks that should point into that package.
+Views are configured through a {\tt spackconfig} file, which can be setup at a site-wide level or for individual users.
+For each package or set of packages, the {\tt spackconfig} file contains rules describing the symlinks that should point into that package.
 The symlink names can be parameterized.
-For example, the above MPILeaks symlink might have been created by {\tt /opt/\$\{PACKAGE\}-\$\{VERSION\}-\$\{MPINAME\} } rule.
-When spack installs or removes packages, symlinks are automaticaly created, deleted, or updated according to the rules.
+For example, the above {\tt mpileaks} symlink might have been created by rule like {\tt /opt/\$\{PACKAGE\}-\$\{VERSION\}-\$\{MPINAME\} }.
+When Spack installs or removes packages, symlinks are automaticaly created, deleted, or updated according to these rules.
 
-Spack's views are a projection from a point in a high-dimension space (the concretized spec, which will have all parameters fully specified) to point in a lower-dimension space (the symlink name, which may only specify a few parameters).
+Spack's views are a projection from a point in a high-dimension space (the concretized spec, which fully specifies all parameters) to point in a lower-dimension space (the symlink name, which may only specify a few parameters).
 Thus several package installations may map to the same symlink.
-For example, the above MPILeaks symlink could point to an MPIleaks compiled with gcc or icc--the compiler parameter is not part of the symlink.
+For example, the above {\tt mpileaks} symlink could point to an {\tt mpileaks} compiled with {\tt gcc} or {\tt icc} -- the compiler parameter is not part of the symlink.
 To keep  package installations  consistent and reproducible, Spack has a well-defined mechanism for resolving conflicting symlinks.
@@
 It has well-defined, but not necessarily meaningful, preferences for deciding between MPI implementations and different compilers.
-A user can override Spack's default sort order through their spackconfig file.
-For example, at one site users may typically use the intel compiler, but some users also use system's default gcc 4.4.7.
-These preferences could be stated by adding the line {\tt compiler\_order = icc,gcc@4.4.7} to the site's spackconfig file, which would cause the ambigious MPILeaks symlink to point to an ICC installation.
+A user can override Spack's default sort order through their {\tt spackconfig} file.
+For example, at one site users may typically use the Intel compiler, but some users also use the system's default {\tt gcc 4.4.7}.
+These preferences could be stated by adding the line {\tt compiler\_order = icc,gcc@4.4.7} to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} symlink to point to an {\tt icc} installation.
 Any compiler left out of the compiler\_order setting is treated as less prefered than explicitly stated compilers.
@@
 One way to solve this would be to edit the Spindle package description in Spack and manually change how configure is invoked.
-This is not ideal, as the Spindle package description is part of the Spack source code, and modifying it makes it difficult to share spack versions and updates across sites.
+This is not ideal, as the Spindle package description is part of the Spack source code, and modifying it makes it difficult to share Spack versions and updates across sites.
 
 Instead Spack allows the creation of site-specific variants of packages.
-The build instructions for a spack package are encoded in a Python class, which must inherit from Spack's {\tt Package} class.
-Via the spackconfig file, users can specify additional search directories for finding additional {\tt Package} classes.
+The build instructions for a Spack package are encoded in a Python class, which must inherit from Spack's {\tt Package} class.
+Via the {\tt spackconfig} file, users can specify additional search directories for finding additional {\tt Package} classes.
 These additional packages can inherit from and replace the Spack's default packages, allowing sites to either tweak or completely replace Spack's build recipes.

### 2015-04-18 usecase-python.tex
Add Python use case.

--- 2015-04-07 usecase-python.tex
@@
 
-\subsection{Support for extensible languages}
+\subsection{Support for interpreted languages}
 \label{sec:usecase-python}
 
-\todo{1 pg}
+Python is becoming increasingly popular for HPC applications, 
+due to its flexibility as a language and due to its excellent support
+calling into compiled numerical libraries for enhanced performance.
+Even though Python is an interpreted language, it is frequently used
+as a friendlier interface to compiled libraries like FFTW, ATLAS, and
+other linear libraries.  This is how many LLNL code teams use Python.
+%
+LC supports Python installations for several application teams,
+and maintenance of these repositories has grown increasingly complex over
+time.  The problems are similar to those that drove us to create
+Spack: different teams want different Python libraries with many different
+configurations.
 
-\begin{verbatim}
-- need for special python handling (prefix issues)
-- solution of offering "meta package management"
-- activate/deactivate python packages
-- installs are in own prefix but can be merged into extended prefix
-- design is extensible, allows for other anguages to use custom logic (e.g. perl, R)
+Existing Python package managers are either language-specific like 
+{\tt easy\_install}~\cite{eby:setuptools}, 
+or they do not effectively support building from source~\cite{anaconda,conda}, let
+alone multi-configuration builds like Spack's.  More glaringly, the default
+model for Python extensions is to install them {\it directly} into the Python
+interpreter's prefix, but this makes having multiple versions installed impossible.
+The alternate way to install Python extensions is to install them in their own prefix,
+but in this mode users must manually add each package to their {\tt PYTHONPATH}. 
+We wanted a way to  easily manage many different versions and {\it also} allow teams
+to have a baseline set of Python packages available by default in every instance.
 
-- Integrating python meta-packages with full spack package
-  management allows building optimized numpy stack AND easy python management.
-\end{verbatim}
+To support this mode of operation, we added the concept of {\tt extension} packages
+to Spack. Python modules now use the {\tt extends('python')} directive instead of
+{\tt depends\_on('python')}, and they install into their own prefix like any other
+package.  However, we added the capability to {\tt activate} or {\tt deactivate} an
+extension within the Python extension it depends on.  When this happens, each file
+in the extension prefix is symbolically linked into Python's prefix, as though it were
+installed wihin the Python instance. The activate command notifies the user and
+fails to activate if any file conflict would arise from this operation.
+Deactivating removes the symbolic links and restores the Python installation to its 
+pristine state.
+
+There were additional complications because many Python packages {\it install their own
+package manager} if they do not find {\tt easy\_install} in the Python installation.
+There are also no fewer than three ways that Python package add themselves to the
+interpreter's default path, and some of them conflict. We modified Spack so that 
+extendable packages like Python can supply custom code in their package file that is
+run on activation and deactivation to handle any necessary language-specific tasks.
+Python uses this to merge conflicting files generated by the {\tt easy\_install} tool.
+
+The end result is that Python extensions can now be installed automatically
+in their own prefixes, and they but they can be composed with a wide range
+of bleeding-edge libraries that other package managers do not handle.  To experiment
+with these extensions, users can load environment modules generated by Spack.
+If they want a particular version to be available without any special environment 
+settings, they can activate it within the Python instance.
+
+Spack now essentially implements a ``meta package-manager'' for each Python
+instance, and it can live together with Spack's normal installation model.
+This has allowed us to efficiently support our application teams, for whom we can
+now rapidly construct custom Python installations.  We have also reduced
+the amount of time LLNL staff spend installing Python modules.
+Because Spack packages can extend the activation and deactivation mechanisms,
+we believe the same mechanism could be used with other 
+interpreted languages that have similar extension models, such as R, Ruby, or Lua.

### 2015-04-18 usecase-policy.tex
add newlines so git diff is easier to see
subsubsection 4.4 and 4.5 into 4.3
updates to policy

@@
 Spack installs packages into paths based on concretized specs, which is ideal for maintaining multiple package installations, but may be difficult for an end-user to navigate.
-A version of {\tt mpileaks}, for example, may be installed in a location like {\tt spack/opt/chaos\_5\_x86\_64\_ib/ gcc@4.9.2/mpileaks@1.0-db465029}.
-
-Spack thus allows the creation of views, which are a symlink-based directory layout of packages.
-Views provide a human-readable directory layout, which can be taylored to fit alongside existing directory layouts or used to create new layouts.
-For example, the above {\tt mpileaks} package may have a view that creates a symlink in {\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpilleaks}.
-The same package install may be referenced by multiple symlinks and views, so the above package could also be symlinked from a more generic {\tt /opt/mpileaks-openmpi} symlink (which is useful for users who don't want to hardcode a specific version).
-Views can also be used to create symlinks to specific executables or libraries in an install, so a Spack-built {\tt gcc 4.9.2} install may have a view that creates symlinks from {\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.
+A version of {\tt mpileaks}, for example, may be installed in a location like:
+%
+\begin{minted}[fontsize=\scriptsize]{bash}
+    spack/opt/chaos_5_x86_64_ib/gcc@4.9.2/mpileaks@1.0-db465029
+\end{minted}
+%
+Spack thus allows the creation of views, which are a symbolic-link based directory layout of packages.
+Views provide a human-readable directory layout, which can be tailored to fit alongside existing directory layouts or used to create new layouts.
+For example, the above {\tt mpileaks} package may have a view that creates a link in {\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpilleaks}.
+The same package install may be referenced by multiple links and views, so the above package could also be linked from a more generic {\tt /opt/mpileaks-openmpi} link (which is useful for users who don't want to hardcode a specific version).
+Views can also be used to create links to specific executables or libraries in an install, so a Spack-built {\tt gcc 4.9.2} install may have a view that creates links from {\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.
 
 Views are configured through a {\tt spackconfig} file, which can be setup at a site-wide level or for individual users.
-For each package or set of packages, the {\tt spackconfig} file contains rules describing the symlinks that should point into that package.
-The symlink names can be parameterized.
-For example, the above {\tt mpileaks} symlink might have been created by rule like {\tt /opt/\$\{PACKAGE\}-\$\{VERSION\}-\$\{MPINAME\} }.
-When Spack installs or removes packages, symlinks are automaticaly created, deleted, or updated according to these rules.
+For each package or set of packages, the {\tt spackconfig} file contains rules describing the links that should point into that package.
+The link names can be parameterized.
+For example, the above {\tt mpileaks} link might have been created by rule like:
+%
+\begin{minted}[fontsize=\scriptsize]{bash}
+    /opt/${PACKAGE}-${VERSION}-${MPINAME}
+\end{minted}
+%
+When Spack installs or removes packages, symbolic links are automaticaly created, deleted, or updated according to these rules.
 
-Spack's views are a projection from a point in a high-dimension space (the concretized spec, which fully specifies all parameters) to point in a lower-dimension space (the symlink name, which may only specify a few parameters).
-Thus several package installations may map to the same symlink.
-For example, the above {\tt mpileaks} symlink could point to an {\tt mpileaks} compiled with {\tt gcc} or {\tt icc} -- the compiler parameter is not part of the symlink.
-To keep  package installations  consistent and reproducible, Spack has a well-defined mechanism for resolving conflicting symlinks.
+Spack's views are a projection from a point in a high-dimensional space (the directory layout shown in Table~\ref{tab:naming-conventions}, which fully specifies all parameters) to point in a lower-dimension space (the link name, which may only specify a few parameters).
+Thus several package installations may map to the same symbolic link.
+For example, the above {\tt mpileaks} link could point to an {\tt mpileaks} compiled with {\tt gcc} or {\tt icc} -- the compiler parameter is not part of the link.
+To keep  package installations  consistent and reproducible, Spack has a well-defined mechanism for resolving conflicting links.
 Spack defines a sort order for packages, which arranges packages from ``most desired'' to ``least desired''.
@@
 For example, at one site users may typically use the Intel compiler, but some users also use the system's default {\tt gcc 4.4.7}.
-These preferences could be stated by adding the line {\tt compiler\_order = icc,gcc@4.4.7} to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} symlink to point to an {\tt icc} installation.
-Any compiler left out of the compiler\_order setting is treated as less prefered than explicitly stated compilers.
-Spack can also be configured to treat specific package versions as more prefered than other versions, which can be useful if a new version is unstable and untested, and should not be advertised with symlinks.
+These preferences could be stated by adding the line:
+%
+\begin{minted}[fontsize=\scriptsize]{bash}
+    compiler_order = icc,gcc@4.4.7
+\end{minted}
+%
+to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} symlink to point to an {\tt icc} installation.
+Any compiler left out of the {\tt compiler\_order} setting is treated as less preferred than those explicitly mentioned.
+Spack also supports allowing specific package versions and other spec options as more preferred than others.
+This is useful for deploying unstable or untested software within Spack before moving it into wider public view.
 
-\subsection{Package Configurations}
+\subsection{Site-specific Package Repositories}
 
-It is also useful to specify site-specific options for how packages should be built.
-The correct configure line for building packages may be different across sites.
-For example, the Spindle tool can have logging locations and security policies specified on its configure line.
-Spack can make reasonable guesses about Spindle's proper configure line, but it cannot create a customized installation for the site.
-One way to solve this would be to edit the Spindle package description in Spack and manually change how configure is invoked.
-This is not ideal, as the Spindle package description is part of the Spack source code, and modifying it makes it difficult to share Spack versions and updates across sites.
+By default, Spack stores its package files in a mainline repository that is present when users
+first run Spack.  At many sites, packages may build sensitive, proprietary software, or they 
+may have patches that are not useful outside of a certain company or organization.  Putting
+this type of code back into a public repository does not often make sense, and if it makes the
+mainline less stable, it can actually make sharing code between sites more difficult.
 
-Instead Spack allows the creation of site-specific variants of packages.
+To support our own private packages, and to support those of LLNL code teams, Spack allows he creation of site-specific variants of packages.
+
 The build instructions for a Spack package are encoded in a Python class, which must inherit from Spack's {\tt Package} class.

### 2015-04-18 usecase-policy.tex

--- 2015-04-18 usecase-policy.tex
@@
 But it also needs to be easy for end-users to find and use those packages.
-Different end-users may have different expectations about how packages should be built and installed, and those expectations are typically shaped by years of site policies, personal preferences, and lingering legacy decisions originallly made on a DEC VAX.
-Rather than try to dictate these expectations, Spack allows both end-users and package maintainers to create custom policies dictating how packages are built and installed at their site.
+Different end-users may have different expectations about how packages should be built and installed,
+and those expectations are typically shaped by years of site policies, personal preferences,
+and lingering legacy decisions originallly made on a DEC VAX.
+Rather than try to dictate these expectations,
+Spack allows both end-users and package maintainers to create custom policies dictating
+how packages are built and installed at their site.
 
-\subsection{Package Views}
+\subsubsection{Package Views}
 \label{sec:package-views}
 
-While Spack can easily create a dozen installations of a package like {\tt mpileaks} for different compilers and MPI implementations, it can be difficult for an end-user to determine which of those dozen packages should be used with their application.
-As discussed in Section~\ref{sec:envmodule}, environment modules are commonly used on supercomputer systems to solve this problem, and Spack allows the package administrator to automatically create module and dotkit files for packages that Spack installs.
+While Spack can easily create a dozen installations of a package like {\tt mpileaks}
+for different compilers and MPI implementations,
+it can be difficult for an end-user to determine which of those dozen packages should be used with their application.
+As discussed in Section~\ref{sec:envmodule},
+environment modules are commonly used on supercomputer systems to solve this problem,
+and Spack allows the package administrator to automatically create module and dotkit files for packages that Spack installs.
 
-However, even when modules and dotkits are available, many users will go directly through the filesystem to access package installs.
-Spack installs packages into paths based on concretized specs, which is ideal for maintaining multiple package installations, but may be difficult for an end-user to navigate.
-A version of {\tt mpileaks}, for example, may be installed in a location like {\tt spack/opt/chaos\_5\_x86\_64\_ib/ gcc@4.9.2/mpileaks@1.0-db465029}.
+However, even when modules and dotkits are available,
+many users will go directly through the filesystem to access package installs.
+Spack installs packages into paths based on concretized specs,
+which is ideal for maintaining multiple package installations, but may be difficult for an end-user to navigate.
+A version of {\tt mpileaks}, for example, may be installed in a location like
+{\tt spack/opt/chaos\_5\_x86\_64\_ib/ gcc@4.9.2/mpileaks@1.0-db465029}.
 
 Spack thus allows the creation of views, which are a symlink-based directory layout of packages.
-Views provide a human-readable directory layout, which can be taylored to fit alongside existing directory layouts or used to create new layouts.
-For example, the above {\tt mpileaks} package may have a view that creates a symlink in {\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpilleaks}.
-The same package install may be referenced by multiple symlinks and views, so the above package could also be symlinked from a more generic {\tt /opt/mpileaks-openmpi} symlink (which is useful for users who don't want to hardcode a specific version).
-Views can also be used to create symlinks to specific executables or libraries in an install, so a Spack-built {\tt gcc 4.9.2} install may have a view that creates symlinks from {\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.
+Views provide a human-readable directory layout,
+which can be taylored to fit alongside existing directory layouts or used to create new layouts.
+For example, the above {\tt mpileaks} package may have a view that creates a symlink in
+{\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpilleaks}.
+The same package install may be referenced by multiple symlinks and views,
+so the above package could also be symlinked from a more generic {\tt /opt/mpileaks-openmpi}
+symlink (which is useful for users who don't want to hardcode a specific version).
+Views can also be used to create symlinks to specific executables or libraries in an install,
+so a Spack-built {\tt gcc 4.9.2} install may have a view that creates symlinks from
+{\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.
 
 Views are configured through a {\tt spackconfig} file, which can be setup at a site-wide level or for individual users.
-For each package or set of packages, the {\tt spackconfig} file contains rules describing the symlinks that should point into that package.
+For each package or set of packages, the {\tt spackconfig} file contains rules
+describing the symlinks that should point into that package.
 The symlink names can be parameterized.
-For example, the above {\tt mpileaks} symlink might have been created by rule like {\tt /opt/\$\{PACKAGE\}-\$\{VERSION\}-\$\{MPINAME\} }.
-When Spack installs or removes packages, symlinks are automaticaly created, deleted, or updated according to these rules.
+For example, the above {\tt mpileaks} symlink might have been created by rule like
+{\tt /opt/\$\{PACKAGE\}-\$\{VERSION\}-\$\{MPINAME\} }.
+When Spack installs or removes packages,
+symlinks are automaticaly created, deleted, or updated according to these rules.
 
-Spack's views are a projection from a point in a high-dimension space (the concretized spec, which fully specifies all parameters) to point in a lower-dimension space (the symlink name, which may only specify a few parameters).
+Spack's views are a projection from a point in a high-dimension space
+(the concretized spec, which fully specifies all parameters) to point in a lower-dimension space
+(the symlink name, which may only specify a few parameters).
 Thus several package installations may map to the same symlink.
-For example, the above {\tt mpileaks} symlink could point to an {\tt mpileaks} compiled with {\tt gcc} or {\tt icc} -- the compiler parameter is not part of the symlink.
-To keep  package installations  consistent and reproducible, Spack has a well-defined mechanism for resolving conflicting symlinks.
+For example, the above {\tt mpileaks} symlink could point to an {\tt mpileaks} compiled with
+{\tt gcc} or {\tt icc} -- the compiler parameter is not part of the symlink.
+To keep  package installations  consistent and reproducible,
+Spack has a well-defined mechanism for resolving conflicting symlinks.
 Spack defines a sort order for packages, which arranges packages from ``most desired'' to ``least desired''.
-When a symlink could resolve to multiple package installations, Spack points the symlink at the most desired package.
+When a symlink could resolve to multiple package installations,
+Spack points the symlink at the most desired package.
 
-By default, Spack treats newer versions of packages compiled with newer compilers as more desirable than older packages built with older compilers.
-It has well-defined, but not necessarily meaningful, preferences for deciding between MPI implementations and different compilers.
+By default, Spack treats newer versions of packages compiled with newer compilers
+as more desirable than older packages built with older compilers.
+It has well-defined, but not necessarily meaningful,
+preferences for deciding between MPI implementations and different compilers.
 A user can override Spack's default sort order through their {\tt spackconfig} file.
-For example, at one site users may typically use the Intel compiler, but some users also use the system's default {\tt gcc 4.4.7}.
-These preferences could be stated by adding the line {\tt compiler\_order = icc,gcc@4.4.7} to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} symlink to point to an {\tt icc} installation.
+For example, at one site users may typically use the Intel compiler,
+but some users also use the system's default {\tt gcc 4.4.7}.
+These preferences could be stated by adding the line
+{\tt compiler\_order = icc,gcc@4.4.7} to the site's {\tt spackconfig} file,
+which would cause the ambigious {\tt mpileaks} symlink to point to an {\tt icc} installation.
 Any compiler left out of the compiler\_order setting is treated as less prefered than explicitly stated compilers.
-Spack can also be configured to treat specific package versions as more prefered than other versions, which can be useful if a new version is unstable and untested, and should not be advertised with symlinks.
+Spack can also be configured to treat specific package versions as more prefered than other versions,
+which can be useful if a new version is unstable and untested, and should not be advertised with symlinks.
 
-\subsection{Package Configurations}
+By default, Spack stores its package files in a mainline repository that is present when users
+first run Spack.  At many sites, packages may build sensitive, proprietary software, or they 
+may have patches that are not useful outside of a certain company or organization.  Putting
+this type of code back into a public repository does not often make sense, and if it makes the
+mainline less stable, it can actually make sharing code between sites more difficult.
 
-It is also useful to specify site-specific options for how packages should be built.
-The correct configure line for building packages may be different across sites.
-For example, the Spindle tool can have logging locations and security policies specified on its configure line.
-Spack can make reasonable guesses about Spindle's proper configure line, but it cannot create a customized installation for the site.
-One way to solve this would be to edit the Spindle package description in Spack and manually change how configure is invoked.
-This is not ideal, as the Spindle package description is part of the Spack source code, and modifying it makes it difficult to share Spack versions and updates across sites.
+To support our own private packages, and to support those of LLNL code teams, Spack allows he creation of site-specific variants of packages.
 
-Instead Spack allows the creation of site-specific variants of packages.
 The build instructions for a Spack package are encoded in a Python class, which must inherit from Spack's {\tt Package} class.

### 2015-04-18 usecase-python.tex
edits to usecase python

--- 2015-04-07 usecase-python.tex
@@
 
-\subsection{Support for extensible languages}
+\subsection{Support for interpreted languages}
 \label{sec:usecase-python}
 
-\todo{1 pg}
+Python is becoming increasingly popular for HPC applications, 
+due to its flexibility as a language and its excellent support
+for calling into fast, compiled numerical libraries.
+Python is an interpreted language, but one can use it
+as a friendlier interface to compiled libraries like FFTW, ATLAS, and
+other linear algebra libraries.  Many LLNL code teams use Python in this manner.
+%
+LC supports Python installations for several application teams,
+and maintenance of these repositories has grown increasingly complex over
+time.  The problems are similar to those that drove us to create
+Spack: different teams want different Python libraries with different
+configurations.
 
-\begin{verbatim}
-- need for special python handling (prefix issues)
-- solution of offering "meta package management"
-- activate/deactivate python packages
-- installs are in own prefix but can be merged into extended prefix
-- design is extensible, allows for other anguages to use custom logic (e.g. perl, R)
+Existing Python package managers are either language-specific like 
+{\tt easy\_install}~\cite{eby:setuptools}, 
+or they do not effectively support building from source~\cite{anaconda,conda}, let
+alone multi-configuration builds like Spack's.  More glaringly, the default
+model for Python extensions is to install them {\it directly} into the Python
+interpreter's prefix, but this makes it impossible to install multiple versions.
+An alternative is to install Python extensions in their own prefix,
+but then users must manually add each package to their {\tt PYTHONPATH}. 
+We wanted a way to  easily manage many different versions and {\it also} allow teams
+to have a baseline set of Python packages available by default.
 
-- Integrating python meta-packages with full spack package
-  management allows building optimized numpy stack AND easy python management.
-\end{verbatim}
+To support this mode of operation, we added the concept of {\tt extension} packages
+to Spack. Python modules now use the {\tt extends('python')} directive instead of
+{\tt depends\_on('python')}.
+Each module installs into its own prefix like any other package,
+and each module depends on a particular Python installation.
+Additionally, we added the capability to {\tt activate} and {\tt deactivate} an
+extension within its dependent Python installation.  When this happens, each file
+in the module prefix is symbolically linked into the Python installation prefix, as though it were
+directly installed within the Python instance. If any file conflict would arise from this operation,
+the activate operation fails and notifies the user.
+The deactivate operation removes the symbolic links and restores the Python installation to its 
+pristine state.
+
+There were additional complications because many Python packages {\it install their own
+package manager} if they do not find {\tt easy\_install} in the Python installation.
+There are also no fewer than three ways that Python packages add themselves to the
+interpreter's default path, and some of these conflict. We modified Spack so that 
+extendable packages, like Python, can supply custom code in their package file that is
+executed during activation and deactivation to handle any necessary language-specific tasks.
+Python uses this feature to merge conflicting files that are generated by the {\tt easy\_install} tool.
+
+The end result is that Python extensions can now be installed automatically
+in their own prefixes, but they can be composed with a wide range
+of bleeding-edge libraries that other package managers do not handle.  To experiment
+with these extensions, users can load environment modules generated by Spack.
+If they want a particular version to be available without any special environment 
+settings, they can activate it within the Python instance.
+
+Spack now essentially implements a ``meta package-manager'' for each Python
+instance, and it can live together with Spack's normal installation model.
+This has allowed us to efficiently support our application teams, for whom we can
+now rapidly construct custom Python installations.  We have also reduced
+the amount of time that LC staff spend installing Python modules.
+Because Spack packages can extend the activation and deactivation mechanisms,
+we believe the same mechanism could be used with other 
+interpreted languages with similar extension models, such as R, Ruby, or Lua.

### 2015-04-18 usecase-policy.tex

@@
 A version of {\tt mpileaks}, for example, may be installed in a location like
-{\tt spack/opt/chaos\_5\_x86\_64\_ib/ gcc@4.9.2/mpileaks@1.0-db465029}.
-
-Spack thus allows the creation of views, which are a symlink-based directory layout of packages.
+%
+\begin{minted}[fontsize=\small]{bash}
+    spack/opt/chaos_5_x86_64_ib/ gcc@4.9.2/mpileaks@1.0-db465029
+\end{minted}
+%
+Spack thus allows the creation of views, which are a symbolic-link based directory layout of packages.
 Views provide a human-readable directory layout,
@@
 For example, the above {\tt mpileaks} symlink might have been created by rule like
-{\tt /opt/\$\{PACKAGE\}-\$\{VERSION\}-\$\{MPINAME\} }.
+%
+\begin{minted}[fontsize=\small]{bash}
+    /opt/${PACKAGE}-${VERSION}-${MPINAME}
+\end{minted}
+%
 When Spack installs or removes packages,
@@
 but some users also use the system's default {\tt gcc 4.4.7}.
-These preferences could be stated by adding the line
-{\tt compiler\_order = icc,gcc@4.4.7} to the site's {\tt spackconfig} file,
-which would cause the ambigious {\tt mpileaks} symlink to point to an {\tt icc} installation.
-Any compiler left out of the compiler\_order setting is treated as less prefered than explicitly stated compilers.
-Spack can also be configured to treat specific package versions as more prefered than other versions,
+These preferences could be stated by adding:
+%
+\begin{minted}[fontsize=\small]{bash}
+    compiler_order = icc,gcc@4.4.7
+\end{minted}
+%
+to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} symlink to point to an {\tt icc} installation.
+Any compiler left out of the compiler\_order setting is treated as less preferred than explicitly stated compilers.
+Spack can also be configured to treat specific package versions as more preferred than other versions,
 which can be useful if a new version is unstable and untested, and should not be advertised with symlinks.
@@
 
-To support our own private packages, and to support those of LLNL code teams, Spack allows he creation of site-specific variants of packages.
-
-The build instructions for a Spack package are encoded in a Python class, which must inherit from Spack's {\tt Package} class.
+To support our own private packages, and to support those of LLNL code teams, Spack allows the creation of site-specific variants of packages.
 Via the {\tt spackconfig} file, users can specify additional search directories for finding additional {\tt Package} classes.
-These additional packages can inherit from and replace the Spack's default packages, allowing sites to either tweak or completely replace Spack's build recipes.
+%
+The additional packages are like the {\tt mpileaks} package shown in Figure~\ref{fig:mpileaks}.
+However, the extension packages can extend from not only {\tt Package}, but also any of Spack's built-in packages.
+Custom packages can inherit from and replace the Spack's default packages, allowing sites to either tweak or completely replace Spack's build recipes.
 To continue the previous example, a site can write a {\tt LocalSpindle} Python class, which inherits from Spack's {\tt Spindle} class.

### 2015-04-18 usecase-python.tex
clarify "when this" --> "During the activate operation"

@@
 Additionally, we added the capability to {\tt activate} and {\tt deactivate} an
-extension within its dependent Python installation.  When this happens, each file
+extension within its dependent Python installation.  During the activate operation, each file
 in the module prefix is symbolically linked into the Python installation prefix, as though it were

### 2015-04-18 conclusion.tex
Submitted version

@@
 
-\todo{.5 pg}
+The complexity of managing HPC software is rapidly increasing, and it
+will continue unabated without effective tools.
+In this paper, we reviewed the state of software management tools 
+across a number of HPC sites, with particular focus on Livermore
+Computing (LC). While tools exist that can handle multi-configuration
+installs, none of them addresses the combinatorial nature
+of the software configuration space directly. None of them allows
+a user to rapidly {\it compose} new parametric builds.
+
+We introduced Spack, a package manager in development at LLNL, that
+provides truly {\it parameterized} builds.  Spack implements
+a novel, recursive {\it spec} syntax that simplifies the process of working
+with large software configuration spaces, and it builds software 
+so that it will run correctly, regardless of the environment.
+We outlined a number of additional unique features of Spack, including
+versioned virtual dependencies, and Spack's {\it concretization} process,
+which converts an underspecified build DAG into a buildable spec.
+We showed through several use cases that Spack is already
+increasing operational efficiency and simplifying software management
+at LLNL, and that it can be rapidly adapted to new production use cases.

### 2015-04-18 implementation.tex
Added versioned vdeps
GLL edits
Updated concretization section.
Submitted version

@@
 Spack is written in Python.  We chose Python for its flexibility
-its wide use in the HPC community.
+and its wide use in the HPC community.
 %
@@
 \begin{enumerate}
-\item Packages are {\bf parametrized}, allowing new builds to be composed
+\item Packages are {\bf parameterized}, allowing new builds to be composed
       on the fly for different versions, architecture, compiler, options, 
@@
 \input{implementation-concretization}
-
 \input{implementation-environment}

### 2015-04-18 motivation.tex
Trim a bit.
neg vspace in table 1 and 2
Submitted version

@@
 To cope with software diversity, many HPC sites use a combination of existing
-package managers and either manual or semi-automated software installation.  
+package managers and either manual or semi-automated installation.
 For the baseline OS, many sites maintain traditional binary
@@
 that encode information in file system paths.
-Table~\ref{tab:naming-conventions} shows a sampling of conventions
+Table~\ref{tab:naming-conventions} shows conventions
 at several sites. At LLNL, we use the APT package manager for installs
@@
 ORNL uses hand installs but adheres to a strict scripting structure
-that can be used to reproduce each build~\cite{jones+:cug08}.
-TACC relies heavily on locally maintained RPMs for its software installs.
+to reproduce each build~\cite{jones+:cug08}.
+TACC relies heavily on locally maintained RPMs for its installs.
 
@@
 
-\paragraph{Environment Modules and RPATHs}
+\paragraph{Environment Modules and RPATHs}\label{sec:env-rpath}
 

### 2015-04-18 usecase-policy.tex
Submitted version

@@
 
-While Spack can easily create a dozen installations of a package like {\tt mpileaks}
-for different compilers and MPI implementations,
-it can be difficult for an end-user to determine which of those dozen packages should be used with their application.
+While Spack can easily create many installations of a package like {\tt mpileaks}
+for different compilers and MPI implementations, end users may find Spack's directory layout
+confusing, and they may not be able to find the right libraries.
 As discussed in Section~\ref{sec:envmodule},
 environment modules are commonly used on supercomputer systems to solve this problem,
-and Spack allows the package administrator to automatically create module and dotkit files for packages that Spack installs.
+and Spack allows the package author to automatically create module and dotkit files for packages that Spack installs.
 
@@
 \begin{minted}[fontsize=\small]{bash}
-    spack/opt/chaos_5_x86_64_ib/ gcc@4.9.2/mpileaks@1.0-db465029
+spack/opt/chaos_5_x86_64_ib/gcc@4.9.2/mpileaks@1.0-db465029
 \end{minted}
@@
 which can be taylored to fit alongside existing directory layouts or used to create new layouts.
-For example, the above {\tt mpileaks} package may have a view that creates a symlink in
+For example, the above {\tt mpileaks} package may have a view that creates a link in
 {\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpilleaks}.
-The same package install may be referenced by multiple symlinks and views,
-so the above package could also be symlinked from a more generic {\tt /opt/mpileaks-openmpi}
-symlink (which is useful for users who don't want to hardcode a specific version).
-Views can also be used to create symlinks to specific executables or libraries in an install,
-so a Spack-built {\tt gcc 4.9.2} install may have a view that creates symlinks from
+The same package install may be referenced by multiple links and views,
+so the above package could also be linked from a more generic {\tt /opt/mpileaks-openmpi}
+link (which is useful for users who don't want to hardcode a specific version).
+Views can also be used to create symbolic links to specific executables or libraries in an install,
+so a Spack-built {\tt gcc 4.9.2} install may have a view that creates links from
 {\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.
 
-Views are configured through a {\tt spackconfig} file, which can be setup at a site-wide level or for individual users.
+Views are configured through a {\tt spackconfig} file, which can be set up at the site or user level.
 For each package or set of packages, the {\tt spackconfig} file contains rules
-describing the symlinks that should point into that package.
-The symlink names can be parameterized.
-For example, the above {\tt mpileaks} symlink might have been created by rule like
+describing the links that should point into that package.
+The link names can be parameterized.
+For example, the above {\tt mpileaks} symbolic link might have been created by a rule like:
 %
@@
 %
-When Spack installs or removes packages,
-symlinks are automaticaly created, deleted, or updated according to these rules.
+On installation or removal,
+links are automatically created, deleted, or updated according to these rules.
 
@@
 (the concretized spec, which fully specifies all parameters) to point in a lower-dimension space
-(the symlink name, which may only specify a few parameters).
-Thus several package installations may map to the same symlink.
-For example, the above {\tt mpileaks} symlink could point to an {\tt mpileaks} compiled with
-{\tt gcc} or {\tt icc} -- the compiler parameter is not part of the symlink.
+(the link name, which may only specify a few parameters).
+Thus several package installations may map to the same link.
+For example, the above {\tt mpileaks} link could point to an {\tt mpileaks} compiled with
+{\tt gcc} or {\tt icc} -- the compiler parameter is not part of the link.
 To keep  package installations  consistent and reproducible,
-Spack has a well-defined mechanism for resolving conflicting symlinks.
+Spack has a well-defined mechanism for resolving conflicting links.
 Spack defines a sort order for packages, which arranges packages from ``most desired'' to ``least desired''.
-When a symlink could resolve to multiple package installations,
-Spack points the symlink at the most desired package.
+When a symbolic link could resolve to multiple package installations,
+Spack points the link at the most desired package.
 
@@
 %
-to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} symlink to point to an {\tt icc} installation.
+to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} link to point to an {\tt icc} installation.
 Any compiler left out of the compiler\_order setting is treated as less preferred than explicitly stated compilers.
 Spack can also be configured to treat specific package versions as more preferred than other versions,
-which can be useful if a new version is unstable and untested, and should not be advertised with symlinks.
+which can be useful if a new version is unstable and untested, and should not be advertised with symbolic links.
 
@@
 {\tt LocalSpindle} may simply add additional configure flags to the {\tt Spindle} class, while leaving the dependencies and most of the build instructions from its parent class.
-To keep builds deterministic and reproducable, Spack also tracks which {\tt Package} class drove a specific build.
+For reproducibility, Spack also tracks the {\tt Package} class that drove a specific build.

### 2015-04-18 usecase-python.tex
Submitted version

@@
 The end result is that Python extensions can now be installed automatically
-in their own prefixes, but they can be composed with a wide range
+in their own prefixes, and they can be composed with a wide range
 of bleeding-edge libraries that other package managers do not handle.  To experiment

### 2015-04-19 conclusion.tex
Actual submitted version

@@
 The complexity of managing HPC software is rapidly increasing, and it
-will continue unabated without effective tools.
+will continue unabated without better tools.
 In this paper, we reviewed the state of software management tools 
@@
 so that it will run correctly, regardless of the environment.
-We outlined a number of additional unique features of Spack, including
-versioned virtual dependencies, and Spack's {\it concretization} process,
+We outlined a number of Spack's unique features, including
+versioned virtual dependencies, and the {\it concretization} process,
 which converts an underspecified build DAG into a buildable spec.
-We showed through several use cases that Spack is already
-increasing operational efficiency and simplifying software management
-at LLNL, and that it can be rapidly adapted to new production use cases.
+
+We showed through three use cases that Spack is already
+increasing operational efficiency 
+in production at LLNL.  We believe that the software management
+techniques implemented in
+Spack are applicable to a broad range of HPC facilities.
+Spack is available online at: http://bit.ly/spack-git.

### 2015-06-12 rebuttal.txt
Added initial rebuttal draft.

We thank the reviewers for their thoughtful comments. We will fix all
grammatical errors and phrasing issues. Below we address the major
concerns of the reviewers.

Reviewer 2 asks for new scientific contributions. We first note that
this is a State of the Practice (SOP) submission. From the SC15
website:

    A SOP paper can describe a first-of-its-kind technology or
    methodology or can capture a unique perspective (or experience) on
    issues, challenge, and solutions for dealing with aspects of
    unprecedented scale and complexity, particularly the experiences
    and knowledge that can be generalized to wide ranges of systems
    and usages [...] how the experience generalizes to wider
    applicability should be explored.

    (http://sc15.supercomputing.org/program/technical-papers/papers-faq)

This paper matches this description, as well as these SOP topics from
the SC15 CFP:

  1. Deployment experiences of large-scale infrastructures &
     facilities.

  2. User support experiences with large-scale and novel machines.

Further, Spack offers novel research contributions. In particular:

  a. No other tool offers a concise, grammar to specify constraints on
     elements in a dependency DAG.

  b. A novel concretization process to translate the relaxed spec
     grammar into a full build specification.

  c. First-class support for versionsed dependencies, versioned
     virtual dependencies, versioned compilers, and build constraints.

To our knowledge, no other system gives the user this much control
over parameterized builds, and no other system handles the full
combinatoric complexity or constraint management for software
dependency DAGs on modern HPC systems.

Reviewer 2 asks for quantitative measures of success.  Using other
tools, it was impossible to automatically build an LLNL production
code in all supported configurations. Using Spack's parameterized
builds, we can build the code with 50-60 dependencies in 20
configurations on 2 architectures, nightly. Each such build replaces
what was previously a multi-day effort.

Reviewers 2 and 3 ask for limitations and downsides. Spack can require
more disk space than a purely environment-module-based system, as we
require otherwise identical packages with different dependencies to be
built separately. In our view, the significant reduction in complexity
for the end user makes this worthwhile.

Reviewer 2 asks for barriers to entry for end users. Recursive syntax
has not been a barrier. As designed, Spack requires only a package
name in most cases, and constraints allow specificity *when
needed*. End users with mid- or large-size projects have readily
adopted Spack.  Python has been a barrier for some and a major
attraction for others. For those who would otherwise create local
workarounds, Python's flexibility has allowed them to extend local
Spack packages to suit their needs. Such customizations frequently
guide our priorities for new core features.

Reviewer 3 asks for performance overhead. Spack commands run quickly;
concretization takes no more than seconds even for packages with many
dependencies. Spack builds run faster than typical manual builds
because spack automatically stages them in a fast tmp
filesystem. Overhead of our compiler wrapper is negligible.

Reviewer 3 mentions the ROCKS distribution, which formerly used RPM
and now uses Anaconda. Both are binary package managers and are cited
in Section 2. We will explicitly mention ROCKS in the final paper.

Reviewer 4 asks what remains for future work. Spack uses a greedy
constraint solver, which is sufficient for current packages. We may
need full SAT solving in the future. Architecture support is currently
LLNL-specific, and we plan to support a broader range of OS and system
types. Currently no system supports allows packages to depend on
particular compiler features, e.g. support for C++11 language features
or support for specific OpenMP versions. We plan to add this
capability to our already robust compiler support.

We will clarify all of the points above in the final paper.

Finally, Reviewer 1 mentions the need to grow a broader community for
Spack. We wholeheartedly agree. Outside LLNL, we have received
contributions from Argonne National Laboratory, Los Alamos National
Laboratory, Illinois Institute of Technology, Stowers Institute for
Medical Research, UC Davis, and U. Canterbury.  Via weekly
teleconferences and a public mailing list, we actively collaborating
with these organizations, as well as groups at Sandia and Intel.

We are actively working to grow the base of contributors, as a tool
like Spack will only survive in the long term with community support.
We submitted to SC15 for greater visibility in the HPC community, and
because SC is a well-known top-tier venue for HPC tools.

### 2015-06-12 rebuttal.txt
Spelling.
small tweaks to rebuttal

@@
 
-  c. First-class support for versionsed dependencies, versioned
+  c. First-class support for versioned dependencies, versioned
      virtual dependencies, versioned compilers, and build constraints.
@@
 built separately. In our view, the significant reduction in complexity
-for the end user makes this worthwhile.
+for the end user justifies this cost.
 
@@
 LLNL-specific, and we plan to support a broader range of OS and system
-types. Currently no system supports allows packages to depend on
-particular compiler features, e.g. support for C++11 language features
-or support for specific OpenMP versions. We plan to add this
-capability to our already robust compiler support.
+types. Currently no system allows packages to depend on particular 
+compiler features, e.g., support for C++11 language features or support 
+for specific OpenMP versions. We plan to add this capability to our 
+already robust compiler support.
 

### 2015-06-12 rebuttal.txt
rebuttal fixes.

@@
 constraint solver, which is sufficient for current packages. We may
-need full SAT solving in the future. Architecture support is currently
-LLNL-specific, and we plan to support a broader range of OS and system
-types. Currently no system allows packages to depend on particular 
-compiler features, e.g., support for C++11 language features or support 
-for specific OpenMP versions. We plan to add this capability to our 
-already robust compiler support.
+need full SAT solving in the future. We have used Spack on Linux
+(primarily RHEL), Blue Gene, and Cray machines. Packages build on all
+of these, but we are continuing to add core features to reduce the
+amount of architecture-specific conditional logic in package
+scripts. We also plan to increase build robustness by deploying a
+continuous testing system. Currently no system allows packages to
+depend on particular compiler features, e.g., support for C++11
+language features or support for specific OpenMP versions. We plan to
+add this capability to our already robust compiler support.
 

### 2015-06-12 rebuttal.txt
rebuttal update.

@@
 
-Reviewer 2 asks for quantitative measures of success.  Using other
-tools, it was impossible to automatically build an LLNL production
-code in all supported configurations. Using Spack's parameterized
-builds, we can build the code with 50-60 dependencies in 20
-configurations on 2 architectures, nightly. Each such build replaces
-what was previously a multi-day effort.
+Reviewer 2 asks for quantitative measures of success.  A large LLNL
+production code previously used another tool to manage its 50-60
+dependencies.  Even with tool support, builds were a multi-day manual
+task requiring a specialist.  After moving to Spack the team was able
+to start building across 20 configurations and 2 architectures as
+automatic nightly builds.
 

### 2015-07-13 abstract.tex
Submitted abstract for program.

@@
 Scientific applications require specific versions of compilers, MPI, and other dependency libraries, so using a single, standard software stack is infeasible.
-However, managing many configurations is difficult because the configuration space is exponential in size.
+However, managing many configurations is difficult because the configuration space is combinatorial in size.
 
-We introduce Spack, a package manager used at Lawrence Livermore National Laboratory (LLNL) to manage this complexity.
+We introduce Spack, a tool used at Lawrence Livermore National Laboratory to manage this complexity.
 Spack provides a novel, recursive specification syntax to invoke parametric builds of packages and dependencies.
 It allows any number of builds to coexist on the same system, and it ensures that installed packages can find their dependencies, {\it regardless of the environment}.
-We show through real-world use cases that Spack supports a diverse and demanding user base, bringing order to HPC software chaos.
+We show through real-world use cases that Spack supports diverse and demanding applications, bringing order to HPC software chaos.

### 2015-07-13 implementation-specs.tex
neg vspace in table 1 and 2
Submitted version
Address reviewer comment on awkward sentence.

@@
 \begin{figure}
-\begin{grammar}
+\begin{grammar}\small
   <spec>         ::= <id> [ constraints ]
@@
 only its package name.  The user does not have to think about DAG connectivity to add
-constraints.  She need only know that a package depends, somehow, on {\tt callpath}.
+constraints.  She only needs to know that a package somehow depends on {\tt callpath}.
 For the same reason, the constraint order does not matter; dependency constraints
@@
 {\bf Versions.}
-The version constraint, denoted with {\tt @}, was already covered above. Versions
+Version constraints, denoted with {\tt @}, were already covered above. Versions
 can be precise ({\tt @2.5.1}) or denote a range ({\tt @2.5:4.4}),

### 2015-07-18 introduction.tex
Change name of tex file to 'spack-paper-sc15'
combinatoric -> combinatorial

@@
-%!TEX root = paper.tex
+%!TEX root = spack-paper-sc15.tex
 
@@
 To support the diverse needs of applications, system administrators
-and application developers 
-frequently build, install, and support many different configurations
-of math and physics libraries, as well as other software.
-Frequently, applications must be rebuilt to fix bugs and to support
-new versions of the operating system (OS), Message Passing Interface (MPI)
-implementation, compilers, and other dependencies. 
-Unfortunately, building scientific software is notoriously complex, with
-immature build systems that are difficult to adapt to new
+and application developers frequently build, install, and support many
+different configurations of math and physics libraries, as well as
+other software.  Frequently, applications must be rebuilt to fix bugs
+and to support new versions of the operating system (OS), Message
+Passing Interface (MPI) implementation, compilers, and other
+dependencies.  Unfortunately, building scientific software is
+notoriously complex, with immature build systems that are difficult to
+adapt to new
 machines~\cite{dubois+:comp-sci-eng,hoste+:pyhpc12,wilson+:corr}.
-Worse, the space of required builds grows combinatorically
-with each new configuration parameter. As a result, LLNL staff
-spend countless hours dealing with build and deployment issues.
+Worse, the space of required builds grows combinatorially with each
+new configuration parameter. As a result, LLNL staff spend countless
+hours dealing with build and deployment issues.
 
@@
 dependency management is overly rigid, and they burden users with
-the combinatoric problems of naming and versioning.
+the combinatorial naming and versioning problems.
 

### 2015-07-18 motivation.tex
Change name of tex file to 'spack-paper-sc15'
combinatoric -> combinatorial

@@
 However, these systems do not provide facilities to manage
-large package repositories or combinatoric versioning.
+large package repositories or combinatorial versioning.
 
@@
 and most of the tools install packages in a single, inflexible location.
-To install multiple configurations, the user must create a custom, combinatoric
+To install multiple configurations, the user must create a custom, combinatorial
 naming scheme to avoid conflicts. They typically require root
@@
 through their own interface. Naming schemes used in existing module systems, however, cannot
-handle combinatoric versions. The Lmod authors call this the
+handle combinatorial versions. The Lmod authors call this the
 ``matrix problem``~\cite{mclay:lmod-tutorial}.
@@
 The main limitation of existing tools is the lack of build {\it composability}.
-The full set of package versions is combinatoric, and realizing an arbitrary
+The full set of package versions is combinatorial, and realizing an arbitrary
 combination of compiler, MPI version, build options, and dependency versions

### 2015-07-18 motivation.tex
Pass over motivation.

@@
 creators to manage conflicts. This burden effectively restricts installations
-to few configurations.
+to a few configurations.
 
@@
 non-Linux operating systems, an issue for the lightweight
-kernels of bleeding-edge Blue Gene/Q and Cray machines. Security
-is another major concern: a profusion of VMs makes mandatory patching
-difficult when user's environments differ significantly from those of
-the facility. Perhaps most importantly, significant differences in
-versioning leaves the majority of the installation burden on end users.
+kernels of bleeding-edge Blue Gene/Q and Cray machines. 
+Finally, each VM still uses a traditional package manager,
+so running many configurations still requires a large number of VMs.
+For facilities, this is a security concern, as a profusion of VMs makes 
+mandatory patching difficult.  It is also tedious for users to manage a large
+number of VM environments.
 
@@
 packages using the vendor's package manager. LLNL maintains a Linux
-distribution, CHAOS~\cite{chaos} for this purpose, which is managed by RPM.
+distribution, CHAOS~\cite{chaos} for this purpose, which is managed using RPM.
 
@@
 at several sites. At LLNL, we use the APT package manager for installs
-in the {\tt /usr/local/tools} file system; we use {\tt /usr/global/tools}
+in the {\tt /usr/local/tools} file system and {\tt /usr/global/tools}
 for manual installs.
@@
 save for the version of a particular dependency library.  In our experience
-at LLNL, naming conventions like these have not succeeded because the full
-set of users want more configurations than we can represent with a practical
+at LLNL, naming conventions like these have not succeeded because
+users want more configurations than we can represent with a practical
 directory hierarchy. Staff frequently install nonconforming packages
@@
 they also complicate the runtime environment. When launched, an executable
-must determine the location of its dependency libraries, or it will not run,
-or even worse, it may find the wrong dependencies and subtly produce incorrect results.
+must determine the location of its dependency libraries, or it will not run.
+Even worse, it may find the wrong dependencies and subtly produce incorrect results.
 Statically linked binaries do not have this issue, but modern
@@
 {\tt /usr/local/lib}.  If binaries are installed in other locations, the
-{\it user} who runs the program (often not the same person who
-installed the binary) must typically add dependency library paths to
+{\it user} who runs the program must typically add dependency library paths to
 {\tt LD\_LIBRARY\_PATH} (or a similar environment variable) so that the loader
-can find them.
+can find them.  The user is often not the same person who installed the binary,
+so determining what paths to add can be difficult, especially for inexperienced users.
 
@@
 allow users to ``load'' and ``unload'' such settings dynamically using simple
-commands. Environment emerged in 1991, and there are many implementations
-available~\cite{dotkit,furlani+:lisa91,furlani+:lisa96,mclay:lmod,mclay:lmod-tutorial}.
-The most advanced implementations, such as Lmod~\cite{mclay:lmod,mclay:lmod-tutorial},
+commands. Environment modules emerged in 1991, and there are many implementations~\cite{dotkit,furlani+:lisa91,furlani+:lisa96,mclay:lmod,mclay:lmod-tutorial}.
+The most advanced of these, such as Lmod~\cite{mclay:lmod,mclay:lmod-tutorial},
 provide software hierarchies similar to the naming conventions in
@@
 mutually exclusive. Modules can still be used to set {\tt MANPATH}, {\tt PATH}
-and other variables, but adding {\tt RPATHs} ensures that binaries run correctly
+and other variables, while adding {\tt RPATHs} ensures that binaries run correctly
 regardless of whether the right module is loaded. LC installs software with
@@
 
-Recently, a number of package managers have been developed to manage
+Recently, a number of HPC package managers have emerged that manage
 multi-configuration builds.
 %
+Smithy~\cite{digirolamo:smithy} is an installation tool in use at ORNL. It
+can generate module files, but it does not provide any
+automated dependency management; it only checks whether a package's
+prerequisites have already been installed by the user.
+
 Nix~\cite{dolstra+:icfp08,dolstra+:lisa04}
@@
 naming convention.  Instead, Nix determines the prefix by hashing the package
-file along with its dependencies.
-
-HashDist~\cite{hashdist} is a meta-build system and package manager for HPC.
-Like Nix, it uses cryptographic versioning, and it archives installations
-in a git repository so that they can be recalled on demand.
-%
-Both Nix and HashDist use {\tt RPATHs} in their packages to ensure that
-libraries are found correctly.
+file along with its dependencies. Nix package files are written in a custom
+functional language designed for packaging.
 
 The EasyBuild~\cite{hoste+:pyhpc12} tool is in production use at
-the University of Ghent.  It allows multiple versions to be installed
-at once.  Rather than setting {\tt RPATHs}, it auto-generates module files
+the University of Ghent and several other sites.  It allows multiple versions
+to be installed at once.  Rather than setting {\tt RPATHs}, it
+auto-generates module files
 to manage each package's environment, and it is closely coupled with
@@
 
-Smithy~\cite{digirolamo:smithy} is a similar tool in use at ORNL. It
-can generate module files, but it does not provide any
-automated dependency management; it only checks whether a package's
-prerequisites have already been installed by the user.
+HashDist~\cite{hashdist} is a meta-build system and package manager for HPC.
+Of existing solutions, it is probably the most similar to spack.
+Like Nix, it uses cryptographic versioning and stores installations in unique
+directories.
+%
+Both Nix and HashDist use {\tt RPATHs} in their packages to ensure that
+libraries are found correctly.
 
-\paragraph{Gaps in current common practice}
+\paragraph{Gaps in current practice}
 The cryptographic versioning of Nix and HashDist is very flexible. It versions
@@
 The systems are in a sense ``write-only'':
-Nix does not even offer a way to ask for a package's dependencies.
+Nix does not offer a way to view a package's dependencies.
 %
@@
 requires tediously modifying many package files.
-%
-Indeed, the number of package files required for existing systems scales with
-the number of version {\it combinations}, not the number of packages, which
+Indeed, the number of package files required for most existing systems scales
+with the number of version {\it combinations}, not the number of packages, which
 quickly becomes unmanageable.  As an example, the very flexible EasyBuild
 system has over 3,300 files for several permutations of only 600 packages.
-Other systems have similar issues. HPC sites need a mechanism to
-{\it parameterize} packages so that new builds can be {\it composed} on the fly.
+To create a slightly different dependency graph, an entire new package file
+hierarchy must be constructed.  HashDist a more robust support for composition,
+but it does not have a first-class parameters for package versions, compilers,
+or versioned interfaces.
+%
+HPC sites need better mechanisms to {\it parameterize} packages so that new
+builds can be {\it composed} in response to user needs.
+%

### 2015-07-18 motivation.tex
Tighten motivation.

@@
 related to package managers, but they focus on ensuring that a single
-package builds correctly with its dependencies.  MixDown notably provides excellent
+package builds with its dependencies.  MixDown notably provides excellent
 features for ensuring consistent compiler flags in a build.
@@
 build packages from source instead of installing from a pre-built binary.
-Most existing port systems suffer from
+Most port systems suffer from
 the same versioning and naming issues as traditional package managers.
@@
 distribution, CHAOS~\cite{chaos} for this purpose, which is managed using RPM.
-
-For more specific builds, many sites adhere to detailed naming conventions
+%
+For custom builds, many sites adhere to detailed naming conventions
 that encode information in file system paths.
-Table~\ref{tab:naming-conventions} shows conventions
-at several sites. At LLNL, we use the APT package manager for installs
+Table~\ref{tab:naming-conventions} shows several sites' conventions.
+LLNL uses the APT package manager for installs
 in the {\tt /usr/local/tools} file system and {\tt /usr/global/tools}
 for manual installs.
-ORNL uses hand installs but adheres to a strict scripting structure
+ORNL uses hand installs but adheres to strict scripting conventions
 to reproduce each build~\cite{jones+:cug08}.
-TACC relies heavily on locally maintained RPMs for its installs.
+TACC relies heavily on locally maintained RPMs.
 
@@
 path is called an {\tt RPATH}. {\tt RPATHs} and environment modules are not
-mutually exclusive. Modules can still be used to set {\tt MANPATH}, {\tt PATH}
-and other variables, while adding {\tt RPATHs} ensures that binaries run correctly
+mutually exclusive. Modules can still be used to set {\tt MANPATH}, {\tt PATH},
+etc., while adding {\tt RPATHs} ensures that binaries run correctly
 regardless of whether the right module is loaded. LC installs software with
@@
 
-HashDist~\cite{hashdist} is a meta-build system and package manager for HPC.
-Of existing solutions, it is probably the most similar to spack.
-Like Nix, it uses cryptographic versioning and stores installations in unique
-directories.
+HashDist~\cite{hashdist} is a meta-build system and package manager
+for HPC.  Of the existing solutions, it is the most similar to spack.
+Like Nix, it uses cryptographic versioning and stores installations in
+unique directories.
 %

### 2015-07-18 motivation.tex
Tighten more.

@@
 the package {\it and} its dependency configuration, and can represent any
-configuration. However, users cannot navigate or query the installed software.
-The systems are in a sense ``write-only'':
-Nix does not offer a way to view a package's dependencies.
+configuration. However, users cannot navigate or easily query the installed
+software.
+%The systems are in a sense ``write-only'':
+%Nix does not offer a way to view a package's dependencies.
 %
-EasyBuild and Smithy generate environment modules, which allow querying
-through their own interface. Naming schemes used in existing module systems, however, cannot
-handle combinatorial versions. The Lmod authors call this the
-``matrix problem``~\cite{mclay:lmod-tutorial}.
+EasyBuild and Smithy generate environment modules, and some querying is
+possible through the module system.  Naming schemes used in existing module
+systems, however, cannot handle combinatorial versions. The Lmod authors
+call this the ``matrix problem''~\cite{mclay:lmod-tutorial}.
 
@@
 with the number of version {\it combinations}, not the number of packages, which
-quickly becomes unmanageable.  As an example, the very flexible EasyBuild
+quickly becomes unmanageable.  As an example, the EasyBuild
 system has over 3,300 files for several permutations of only 600 packages.
-To create a slightly different dependency graph, an entire new package file
-hierarchy must be constructed.  HashDist a more robust support for composition,
-but it does not have a first-class parameters for package versions, compilers,
+A slightly different dependency graph requires an entire new package file
+hierarchy.  HashDist a more robust support for composition,
+but it does not have a first-class parameters for versions, compilers,
 or versioned interfaces.

### 2015-07-19 implementation-concretization.tex
edits to concretization
Change name of tex file to 'spack-paper-sc15'
Pass over implementation.

@@
 If a spec meets all of these criteria, we say it is {\it concrete}.
-{\it Concretization} is the central component of the Spack build
-process that allows it to reduce a highly unconstrained abstract
+{\it Concretization} is the central component of Spack's build
+process that allows it to reduce an unconstrained abstract
 description to a concrete build.
@@
 
-Spack intersects the constraints of the two DAGs package by package.  It checks
-each parameter for inconsistencies.  Inconsistencies can arise if, for example,
+Spack intersects the constraints of the two DAGs package by package, and it
+checks each parameter for inconsistencies.  Inconsistencies can arise if, for example,
 the user inadvertently requests two versions of the same package, or if a
-package file explicitly requests a different version from what the user requested.
+package file explicitly requests a different version than the user requested.
 Likewise, if the package and the user specified different compilers, variants,
 or platforms for particular packages, this will cause the user to be notified
-of the contradiction. If the user or package specified version ranges, they are
+of the conflict. If the user or package specified version ranges, they are
 intersected, and if the ranges do not overlap, an error will be raised.
@@
 variants, compilers, and versions to resolve any remaining abstract nodes.
-Adding a variant {\it may} cause a package to depend on more libraries. For example,
-if the site default behavior is to have a particular package build with {\tt +python},
-and this causes it to depend on Python libraries, then the cycle begins again.
+Adding a variant like {\tt +mpi} may cause a package to depend on more
+libraries (as in Section~\ref{sec:constraints}). The concretization algorithm
+evaluates conditions from {\tt when} clauses on the DAG, and if there are
+new libraries or other changes, the cycle begins again.
 Spack currently avoids an exhaustive search by using a greedy algorithm.  
@@
 force the build to use a {\it particular} MPI implementation
-by supplying, e.g. \verb|^openmpi| or \verb|^mpich|.  In future
-work we will investigate adding a SAT solver to Spack's concretization process
-to present users with better options.  In our experience, however, complex policy
-interactions like these are very rare.
+by supplying \verb|^openmpi| or \verb|^mpich|.  We leave automatic constraint
+space exploration for future work.
 
 When the concretization process completes, it outputs a fully concrete spec DAG. 
-A concrete DAG with architectures, compiler, versions, and variants, and all 
-dependencies resolved is shown in Figure~\ref{fig:specs-mpileaks-concrete}.
+Figure~\ref{fig:specs-mpileaks-concrete} shows a concrete DAG with architectures, 
+compilers, versions, variants, and all dependencies resolved.
 This fulfills Spack's guarantees.
@@
 and traverses the DAG in a bottom-up fashion.  At each node, it invokes the package's
-{\tt install} method.  For the {\tt spec} parameter to install, it uses a sub-DAG
-rooted at current node, which is also a concrete spec.  Package authors are
+{\tt install} method.  For the {\tt spec} parameter to {\tt install}, it passes
+a sub-DAG rooted at current node, which is also a concrete spec.  Package authors are
 responsible for querying the spec in {\tt install} and handling different configurations.
 
-\subsubsection{Directory Layout}\label{sec:directory-layout}
+\subsubsection{Shared sub-DAGs}\label{sec:directory-layout}
 \begin{figure}\centering
@@
 \subsubsection{Reproducibility}
-For reproducibility and to preserve provenance, Spack stores a file containing
-the concrete spec in each
-install prefix. The spec can be examined later to understand how the installed
-package was built.  It an also be used to reproduce the build on another machine
-or at another site.
+For reproducibility, and to preserve provenance, Spack stores a number of 
+files in the installation directory that document how the installed package was
+built.  These include the {\tt package.py} file used to build, a build log
+containing output and error messages, and a file containing the complete
+concrete spec for the package {\it and} its dependencies. The spec file can be
+used later to reproduce the build, even if concretization preferences have changed.
 
@@
 of building HPC software comes from the size of the build parameter space.
-As mentioned, there are only a few parameters most users care about, and the rest
-only serve to add complexity to the build.  When doing manual builds, LLNL staff
-tend to make arbitrary choices about many of the secondary build parameters,
-or they would include logic in build scripts to make many of these choices.
-In either scenario, publicly installed software was generally not installed
-in a consistent way.
+As mentioned, there are only a few parameters that typical users care about.
+The rest add unneeded complexity.  When building manually, LLNL staff
+tend to make arbitrary choices about the secondary build parameters,
+or they add logic to build scripts to make these choices.
+Manually built software is generally not installed in a consistent manner.
 
@@
 other build parameters.  For example, the site or the user can set default versions
-to use for {\it any} library that is not specified explicitly.
+to use for any library that is not specified explicitly.
 %
-Second, concretization reduces the burden of packaging software because package authors
-do not have to make these decisions.  There is no
+Second, concretization reduces the burden of packaging software, because
+package authors do not have to make these decisions.  There is no
 need for packages to contain complicated checks, or for packages to be overly
 specific about versions.  Other multi-configuration systems like Nix, EasyBuild,
-and hashdist require the package author to write a mostly concrete build spec in advance.
+and HashDist require the package author to write a mostly concrete build spec in advance.
 We believe that this puts undue burden on the package author, and it makes the task

### 2015-07-19 implementation-specs.tex
Use times, except for EBNF where computer modern looks better.
Change name of tex file to 'spack-paper-sc15'
Pass over implementation.

@@
 package code is structured to build a single version of a package, but in Spack, each
-package file is a {\it template} that can be built in many different ways.  The build is
-{\it parameterized} so that it can be configured many different ways.
-Spack calls a single build configuration a {\it spec},
- and the {\tt spec} object passed to {\tt install()}
-is how the Spack system encapsulates build parameters for package authors.
+package file is a {\it template} that can be configured and built in many
+different ways, according to a set of {\it parameters}.
+Spack calls a single build configuration a {\it spec}. The {\tt spec} object
+passed to {\tt install()} is how Spack communicates dependencies and parameters to package authors.
 
@@
 for each dependency and constructs a graph of their relationships.  The result
-is a directed, acyclic graph (DAG) of dependencies\footnote{Spack currently disallows
-circular dependencies.}.
+is a directed, acyclic graph (DAG) of dependencies.\footnote{Spack currently disallows
+circular dependencies.}
 %
-To guarantee a consistent build, and to avoid ABI incompatibility, the DAG
+To guarantee a consistent build, and to avoid ABI compatibility issues, the DAG
 is constructed so that there is only {\it one} version of any particular package.  Note
@@
 specify the rest. For example, a user may know that a certain version of a library like
-{\tt boost} is required, but only cares that other build parameters are set so that
+{\tt boost}~\cite{boost} is required, but only cares that other build parameters are set so that
 the build will succeed.
@@
 very specific, so we have two competing concerns.  We need the ability to specify details
-of the configuration space, without the complexity of remembering all of them.
+of the configuration, without the complexity of remembering all of it.
 
 \subsubsection{Spec Syntax}\label{sec:syntax}
-
 \begin{figure}
+{\fontfamily{cmr}\selectfont
 \begin{grammar}\small
-  <spec>         ::= <id> [ constraints ]
+  <spec>         ::= <id> [ <constraints> ]
 
@@
 
-  <id>           ::= [A-Za-z0-9_][A-Za-z0-9_.-]*
+  <id>           ::= "[A-Za-z0-9_][A-Za-z0-9_.-]*"
 \end{grammar}
+} % End fontfamily{cmr}
 \caption{
@@
 Consider the case where a user wants to install the {\tt mpileaks} package, but knows
-nothing about its structure.  To install the package, the user invokes the {\tt spack install}
-command:
+nothing about its structure.  To install the package, the user invokes {\tt spack install}:
 %
@@
 Note that even though the spec contains no dependency information, it is still
-converted to a full DAG, based on the directives supplied in package files. Since there
-are no constraints on the nodes, Spack has considerable leeway for how to build the package,
-and we say that the package is {\it unconstrained}.
+converted to a full DAG, based on directives supplied in package files. Since there
+are no constraints on the nodes, Spack has considerable leeway when building the
+package, and we say that the package is {\it unconstrained}.
 %
@@
 and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would use
-{\tt @2.3:2.5.6} to designate the range. In these cases, the user can save build time if
-Spack already has a version installed that satisfies the constraint -- Spack will just use
+{\tt @2.3:2.5.6} to designate the range. In these cases, the user can save time if
+Spack already has a version installed that satisfies the spec -- Spack will use
 the previously-built installation instead of building a new one.
@@
 The caret (\verb|^|) denotes constraints for a particular dependency.  In the DAG,
-we now see that there are version constraints on {\tt callpath} and {\tt libelf},
-and that the user has asked for the debug variant of the {\tt callpath} library.
+we see that there are now version constraints on {\tt callpath} {\it and} {\tt libelf},
+and that the user has requested the {\tt debug} variant of the {\tt callpath} library.
 
-Recall that Spack guarantees there will be only a single version of any package in
-the spec DAG.  Therefore, within the same DAG, each dependency can be uniquely identified by
-only its package name.  The user does not have to think about DAG connectivity to add
-constraints.  She only needs to know that a package somehow depends on {\tt callpath}.
+Recall that Spack guarantees that there is only a single version of any package in
+the spec DAG.  Within the same DAG, each dependency can be uniquely
+identified by its package name alone.  Therefore, the user need not think about DAG 
+connectivity to add constraints.  She need only know that {\tt mpileaks}
+somehow depends on {\tt callpath}.
 For the same reason, the constraint order does not matter; dependency constraints
@@
 The package in Figure~\ref{fig:mpileaks} lists two ``safe'' versions with checksums, but
-in our experience users frequently want bleeding-edge versions.  Package managers
+in our experience users frequently want bleeding-edge versions, and package managers
 frequently lag behind the latest releases.
 Spack has a capability to extrapolate URLs from versions,
-using the package's {\tt url} attribute as a model\footnote{This works
-for packages with consistently named URLs}.  The user can request a specific
+using the package's {\tt url} attribute as a model.\footnote{This works
+for packages with consistently named URLs.}  The user can request a specific
 version on the command line, even if it is unknown to Spack,
@@
 
-\subsubsection{Constraints in packages}
+\subsubsection{Constraints in packages}\label{sec:constraints}
 
@@
 This constraint will be incorporated into the initial DAG node generated from
-the ROSE package.
-
-Spack's {\tt patch} directive also accepts spec syntax as a predicate in an
-optional {\tt when} parameter.  For example, these concise directives in the
-Python package ensure that specific patches are applied to the Python source
-code when it is being built on Blue Gene/Q, with the appropriate compiler:
+the {\tt ROSE} package.  Dependencies can also be {\it conditional}.  For example,
+the {\tt ROSE} package builds with different versions of boost, depending on the
+compiler version. To handle this, directives also accept spec syntax as a
+predicate in an optional {\tt when} parameter:
 %
 \begin{minted}[fontsize=\scriptsize]{python}
-    patch('python-bgq-xlc.patch',   when='=bgq%xlc')
+    depends_on('boost@1.47.0', when='%gcc@:4')
+    depends_on('boost@1.54.0', when='%gcc@5:')
+\end{minted}
+%
+The same notation is used to optionally build with libraries like MPI:
+%
+\begin{minted}[fontsize=\scriptsize]{python}
+    depends_on('mpi', when='+mpi')
+\end{minted}
+%
+and to ensure that specific patches are applied to Python source code
+when it is built on Blue Gene/Q, with different compilers:
+%
+\begin{minted}[fontsize=\scriptsize]{python}
+    patch('python-bgq-xlc.patch',   when='=bgq%xl')
     patch('python-bgq-clang.patch', when='=bgq%clang')
@@
 %
-Constraints in the {\tt when} clause are matched against the Python package spec.
-Outside of directives, constraints can be used directly
-with the {\tt spec} object in the {\tt install} method:
+Constraints in the {\tt when} clause are matched against the package spec.
+Outside of directives, constraints can also be tested directly on the {\tt spec}
+object in the {\tt install} method:
 %
@@
           # Handle gcc
-      elif spec.satisfies('%xlc'):
+      elif spec.satisfies('%xl'):
           # Handle XL compilers
@@
 
-In our experience maintaining packages at LLNL, we have sometimes had to change
+In our experience maintaining packages at LLNL, we often have to change
 entire package build scripts due to large changes in the way certain packages build.
 This can be very cumbersome, and it is difficult to maintain both the old and new
-version of a build script, but we must if we want to keep installing older versions
-for users who rely on them.
+version of a build script, which we must do for users who rely on the old version.
 
@@
 For cases like this, Spack provides functionality that allows Python functions to have
-multiple definitions, with some specialized for particular configurations of the package.
+multiple definitions, each specialized for particular configurations of the package.
 This allows us to have two separate implementations of {\tt install} or {\it any} method
@@
 package.  The {\tt @when} directive is a Python decorator: a higher order function that
-takes a function definition as a parameter and returns a new function to put in its place.
+takes a function definition as a parameter and returns a new function to replace it.
 We replace the function with a callable multi-function dispatch object, and we

### 2015-07-19 implementation-virtual-deps.tex
Change name of tex file to 'spack-paper-sc15'
Pass over implementation.

@@
 
-Many libraries share a common source interface and can be interchanged for
-one another in a build.
+Many libraries share a common source interface and can be interchanged 
+within a build.
 The archetypal example in HPC is the MPI interface, which has a number of
@@
 implementations. An application that can be built with one MPI can generally be
-built with another.  BLAS implementations such as ATLAS, LAPACK-BLAS, MKL, etc.
-are another common example.
+built with another.  Another example is the Basic Linear Algebra
+Subroutines (BLAS), which has many fungible implementations
+(ATLAS, LAPACK-BLAS, MKL, etc.).
 
+Neither MPI nor BLAS has a standard ABI, so applications cannot simply be
+re-linked with a new version. They must be recompiled and reinstalled. 
 At LLNL, we frequently need to build tools with many versions of MPI to support
 the many different applications that run at our center.
-However, MPI has no ABI, so applications cannot simply be re-linked;
-they must be recompiled and reinstalled. 
 Complicating matters, the MPI interface is versioned, and some
@@
 
-To allow rapid composition of libraries by interface, Spack supports
+To allow rapid composition of libraries using an interface, Spack supports
 {\it virtual dependencies}.  A virtual dependency is an abstract name
@@
 adds {\it versioning} to its interfaces.  This allows concepts like MPI
-interface versions and BLAS levels to be represented directly, and Spack can
-handle the details of managing complex constraints while the user focuses on 
-composing software.
+interface versions and BLAS levels to be represented directly.  Spack 
+handles the details of managing and checking complex versioning constraints.
 
@@
 Any version of {\tt mvapich2} or {\tt mpich} could be used to to satisfy the {\tt mpi}
-constraint. The Gerris CFD library, however, needs MPI version 2 or higher.  So any
+constraint. The Gerris computational fluid dynamics library, however, needs MPI version 2 or higher.  So any
 version except {\tt mpich} 1.x could be used to satisfy the constrained dependency.

### 2015-07-19 implementation.tex
Change name of tex file to 'spack-paper-sc15'
Pass over implementation.

@@
 \begin{enumerate}
-\item Packages are {\bf parameterized}, allowing new builds to be composed
-      on the fly for different versions, architecture, compiler, options, 
-      and dependencies.
+\item Packages are explicitly {\bf parameterized} by version,
+      architecture, compiler, other options, and dependencies,
+      allowing easy composition of new builds.
 \item To manage this large parameter space, spack provides a novel, 
-      recursive {\bf spec syntax} for dependency graphs.
+      recursive {\bf spec syntax} for dependency graphs and constraints.
 \item Specs support versioned, ABI-incompatible interfaces like MPI through

### 2015-07-19 limitations.tex
Pass over implementation.

%!TEX root = spack-paper-sc15.tex

\subsection{Limitations and Future Work}

In future work we will investigate adding a SAT solver to Spack's
concretization process to present users with better options.  In our
experience, however, complex policy interactions like these are very
rare.

### 2015-07-19 implementation-environment.tex
Change name of tex file to 'spack-paper-sc15'
Install environment revisions.

@@
 
-%{\bf Reproducibility.}
-%\paragraph{Reproducibility}
-Spack is intended to build a consistent HPC stack for our multi-user
+Spack is designed to build a consistent HPC stack for our multi-user
 environment, and reproducible builds are one of our design goals.
@@
 %
-Many packages used at LLNL have a profusion of build options, and specifying them 
-correctly often requires tedious experimentation.  This is due to lack of
-build standards and to the diversity of HPC environments.  
+Many packages used at LLNL have a profusion of build options, and specifying
+them correctly often requires tedious experimentation.  This is due to lack of
+build standards, as well as the diversity of HPC environments.
 For example, in some packages that depend on the {\tt Silo} library,
@@
 separated by a comma.
-The {\tt install} method in Spack's package files allows us to record
-precise build incantations for later reuse.
+The {\tt install} method in Spack's package files encodes precise build
+incantations for later reuse.
 
-%{\bf Environment isolation.}
-\paragraph{Environment isolation}
-In addition to command-line issues, we
-frequently encounter errors due to inconsistencies between the environment of
-the package installer and the package user.
+\subsubsection{Environment isolation}
+In addition to inconsistencies that arise from building manually, we
+frequently encounter errors due to inconsistencies between the package
+installation environment and the environment of the package user.
 %
-For example, there are two versions of the {\tt libelf} library used by
-LLNL performance tools. One is distributed with RedHat Linux, while another
-publicly available version has the same API but an incompatible ABI.
-Failure to specify the right version at build time has caused many
+For example, LLNL performance tools use two versions of the {\tt libelf}
+library. One is distributed with RedHat Linux, while the
+other is publicly available, and they have the same API but an incompatible ABI.
+Specifying the wrong version at build time has caused many
 unexpected crashes.
-%
-Spack manages the build environment by running each {\tt install} invocation
-in a new process.  It helps packages find dependencies 
-correctly, by setting
+
+Spack manages the build environment by running each call to {\tt install}
+in its own process.  In this environment, Spack sets
 {\tt PATH}, {\tt PKG\_CONFIG\_PATH}, {\tt CMAKE\_PREFIX\_PATH}, and
@@
 These variables are commonly used by build systems to locate dependencies,
-and setting them helps to ensure that correct libraries are detected.
+and setting them helps to ensure that the correct libraries are detected by
+build systems.
 The isolated build environment also gives package authors 
@@
 
-%{\bf Compiler wrappers and RPATHs.}
-\paragraph{Compiler wrappers and RPATHs}
-Finding compilers at build time is not the only obstacle to reproducible
+\subsubsection{Compiler wrappers and RPATHs}
+Finding dependencies at build time is not the only obstacle to reproducible
 behavior.  As mentioned in Section~\ref{sec:motivation}, it is also important
 for binaries to be able to find dependency libraries at {\it runtime}.
-One of the most frequent user errors at LC is improper library configuration.
+One of the most common user errors at LC is improper library configuration.
 Users frequently do not know what libraries a package was built with, and 
 it is difficult for them to construct a suitable {\tt LD\_LIBRARY\_PATH} for
-a package that was built by someone else.  Because of frequent support calls,
+a package that was built by someone else.  To reduce the number of support calls,
 we typically add {\tt RPATHs} to public software installations, so that paths
-to dependencies are embedded in binaries and so that users do not have to know
-this information to run installed software correctly.
+to dependencies are embedded in binaries. This way, users do not need to know
+details of dependencies to run installed software correctly.
 
@@
 C, C++, and Fortran compilers, so they are generally picked up 
-automatically\footnote{If builds do not respect {\tt CC}, {\tt CXX}, etc.,
+automatically.\footnote{If builds do not respect {\tt CC}, {\tt CXX}, etc.,
 wrappers can be added as arguments or inserted into Makefiles
-by {\tt install}.}.
+by {\tt install}.}
 When run, the wrappers insert include ({\tt -I}), library ({\tt -L}), and 
@@
 header and library search paths for dependencies, header and library detection
-tests that are run by most build systems succeed automatically, {\it without}
+tests that are run by most build systems succeed automatically, without
 the need to use special arguments for nonstandard locations.  {\tt configure}
@@
 be written as they would be for system installs.  This reduces complexity
-for package maintainers and enforces consistent, reproducible
+for package maintainers and helps to ensure consistent, reproducible
 build policies across packages.  Finally, because Spack has control over the 
@@
 the run-time environment.  A package may need environment variables like {\tt PATH}, 
-{\tt LD_LIBRARY_PATH}, or {\tt MANPATH} set before it can be used.  
+{\tt MANPATH}, or {\tt PKG\_CONFIG\_PATH} set before it can be used.  
 As discussed in Section~\ref{sec:motivation}, many sites rely on environment 
@@
 dotkit~\cite{dotkit} and Module configuration files for its packages, allowing 
-users to setup their runtime environment using familiar systems.  
+users to setup their runtime environment using familiar systems.  While Spack
+packages do not {\tt LD\_LIBRARY\_PATH} to run, we set it in our generated module
+files as well, because it can be used by build systems to find libraries, as well
+as by dependent packages that do not use {\tt RPATH}.
 %

### 2015-07-19 implementation-packages.tex
Change name of tex file to 'spack-paper-sc15'
Pass over implementation.
Install environment revisions.

@@
 Figure~\ref{fig:mpileaks} shows the package for \mpileaks, a tool developed
-at LLNL to find leaks of MPI objects in programs.
+at LLNL to find un-released MPI handle objects in parallel programs.
 The {\tt MpiLeaks} class first provides simple metadata: a text description,
@@
 directives indicate prerequisite packages that must be installed before \mpileaks.
-Last, the {\tt install()} method contains the commands used to build
+Last, the {\tt install} method contains the commands used to build
 the tool.  Spack's DSL allows shell commands to be invoked as Python functions,
-and the {\tt install()} method invokes {\tt configure},
+and the {\tt install} method invokes {\tt configure},
 {\tt make}, and {\tt make install} commands as a shell script would.

### 2015-07-19 implementation-environment.tex
Tighten.

@@
 
-Spack's compiler wrappers have a number of useful effects.  First, they allow
+Spack's compiler wrappers have a number of benefits.  First, they allow
 Spack to transparently switch compilers in most builds.  This is how
@@
 header and library search paths for dependencies, header and library detection
-tests that are run by most build systems succeed automatically, without
+tests used by most build systems succeed automatically, without
 the need to use special arguments for nonstandard locations.  {\tt configure}

### 2015-07-19 usecase-combinatorial.tex
combinatoric -> combinatorial
Edits through use cases.

--- 2015-04-18 usecase-combinatoric.tex
@@
-%!TEX root = paper.tex
+%!TEX root = spack-paper-sc15.tex
 
-\subsection{Combinatoric Naming}
-\label{sec:usecase-combinatoric}
+\subsection{Combinatorial Naming}
+\label{sec:usecase-combinatorial}
 %\subsubsection{Gperftools and MPILeaks}
@@
 at LC.  First, {\tt gperftools} is a C++ library.  Since C++ does not define a standard
-ABI, {\tt gperftools} must be re-built with each compiler and compiler version used by client
-applications.  Second, building {\tt gperftools} on bleeding-edge architectures
+ABI, {\tt gperftools} must be rebuilt with each compiler and compiler version used by client applications.  Second, building {\tt gperftools} on bleeding-edge architectures
 (such as Blue Gene/Q) requires patches and complicated configure lines that
 change with each compiler.  One application team tried to maintain their own
-builds of {\tt gperftools} for their preferred compilers, but this
-task was complicated due to the second issue.
+builds of {\tt gperftools}, but the maintenance burden soon became too great.
 
-Spack presented a solution to both problems.  Package administrators can use Spack to
-easily maintain a central install of {\tt gperftools} across combinations of
-compilers and compiler versions.  Spack's {\tt gperftools} package also serves as a central
-institutional knowledge repository for builds, as its package files encode
-the patches and configure lines required for each platform and compiler combination.
-We are investigating whether we can move some of these settings into compiler and
-architecture descriptors to further simplify Spack's build template for cross-platform
-installations.
 \begin{figure}
@@
 
+Spack presented a solution to both problems.  Package administrators can use Spack to
+easily maintain a central install of {\tt gperftools} across combinations of
+compilers and compiler versions.  Spack's {\tt gperftools} package also serves as a central
+institutional knowledge repository, as its package files encode
+the patches and configure lines required for each platform and compiler combination.
+%
 Figure~\ref{fig:gperftools} illustrates per-compiler and platform build rules with
@@
 
-LC's {\tt mpileaks} tool, which we have used as a running example in this paper,
+LC's {\tt mpileaks} tool, which has been a running example in this paper,
 has an even larger configuration space than {\tt gperftools}.  It must be built
-for different compilers, compiler versions, MPI implementations, and MPI implementation
+for different compilers, compiler versions, MPI implementations, and MPI
 versions. As with {\tt gperftools}, we have been able to install many different
-configurations of {\tt mpileaks} using Spack.  Moreover, Spack's virtual dependency system
-allows us to
-quickly compose a new {\tt mpileaks} build when a new MPI library is deployed at LC, {\it without}
-modifying the {\tt mpileaks} package itself.
+configurations of {\tt mpileaks} using Spack.  Moreover, Spack's virtual dependency 
+system allows us to quickly compose a new {\tt mpileaks} build when a new MPI library is 
+deployed at LC, {\it without} modifying the {\tt mpileaks} package itself.
 %Though we simplified its dependency structure

### 2015-07-19 usecase-policy.tex
Change name of tex file to 'spack-paper-sc15'
Edits through use cases.

@@
 
-Spack makes it easy for package maintainers to create and organize package installations.
-But it also needs to be easy for end-users to find and use those packages.
-Different end-users may have different expectations about how packages should be built and installed,
-and those expectations are typically shaped by years of site policies, personal preferences,
-and lingering legacy decisions originallly made on a DEC VAX.
-Rather than try to dictate these expectations,
-Spack allows both end-users and package maintainers to create custom policies dictating
-how packages are built and installed at their site.
+Spack makes it easy to create and organize package installations,
+but it must also be easy for end-users to find and use packages on an HPC
+system. Different end-users have different expectations about how packages
+should be built and installed, and at LLNL those expectations are shaped
+by years of site policies, personal preferences, and lingering legacy
+decisions originally made on a DEC VAX.
+Spack provides mechanisms that allow users to customize it for the needs
+of particular sites and users.
+
+%Rather than try to dictate these expectations,
+%Spack allows both end-users and package maintainers to create custom policies dictating
+%how packages are built and installed at their site.
 
@@
 While Spack can easily create many installations of a package like {\tt mpileaks}
-for different compilers and MPI implementations, end users may find Spack's directory layout
-confusing, and they may not be able to find the right libraries.
-As discussed in Section~\ref{sec:envmodule},
-environment modules are commonly used on supercomputer systems to solve this problem,
-and Spack allows the package author to automatically create module and dotkit files for packages that Spack installs.
-
-However, even when modules and dotkits are available,
-many users will go directly through the filesystem to access package installs.
-Spack installs packages into paths based on concretized specs,
-which is ideal for maintaining multiple package installations, but may be difficult for an end-user to navigate.
-A version of {\tt mpileaks}, for example, may be installed in a location like
+for different compilers and MPI implementations, end users may find Spack's directory layout confusing, and they may not be able to find libraries intuitively.
+Spack installs packages into paths based on concretized specs.
+This is ideal for maintaining multiple package installations, 
+but it may be difficult for an end-user to navigate.
+For example, an {\tt mpileaks} installation prefix might be:
 %
 \begin{minted}[fontsize=\small]{bash}
-spack/opt/chaos_5_x86_64_ib/gcc@4.9.2/mpileaks@1.0-db465029
+spack/opt/linux-x86_64/gcc-4.9.2/mpileaks-1.0-db465029
 \end{minted}
-%
+
+As discussed in Section~\ref{sec:envmodule},
+environment modules are commonly used on HPC systems to solve this problem,
+and Spack allows the package author to automatically create environment modules
+for packages that Spack installs.
+Even when modules are available, many users still
+navigate the filesystem to access packages.
+
 Spack thus allows the creation of views, which are a symbolic-link based directory layout of packages.
 Views provide a human-readable directory layout,
-which can be taylored to fit alongside existing directory layouts or used to create new layouts.
+which can be adapted to match legacy directory layouts that users rely on.
 For example, the above {\tt mpileaks} package may have a view that creates a link in
-{\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpilleaks}.
+{\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpileaks}.
 The same package install may be referenced by multiple links and views,
 so the above package could also be linked from a more generic {\tt /opt/mpileaks-openmpi}
-link (which is useful for users who don't want to hardcode a specific version).
+link. This is useful for users who always want to use the latest version.
 Views can also be used to create symbolic links to specific executables or libraries in an install,
-so a Spack-built {\tt gcc 4.9.2} install may have a view that creates links from
+so a Spack-built {\tt gcc@4.9.2} install may have a view that creates links from
 {\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.
 
-Views are configured through a {\tt spackconfig} file, which can be set up at the site or user level.
-For each package or set of packages, the {\tt spackconfig} file contains rules
+Views are configured using configuration files, which can be set up at the site or
+user level.
+For each package or set of packages, configuration file contains rules
 describing the links that should point into that package.
@@
 %
-On installation or removal,
+On installation and removal,
 links are automatically created, deleted, or updated according to these rules.
 
-Spack's views are a projection from a point in a high-dimension space
-(the concretized spec, which fully specifies all parameters) to point in a lower-dimension space
-(the link name, which may only specify a few parameters).
-Thus several package installations may map to the same link.
+Spack's views are a projection from a point in a high-dimensional space
+(the concretized spec, which fully specifies all parameters) to point
+in a lower-dimensional space
+(the link name, which may only contain a few parameters).
+Thus, several installations may map to the same link.
 For example, the above {\tt mpileaks} link could point to an {\tt mpileaks} compiled with
@@
 To keep  package installations consistent and reproducible,
-Spack has a well-defined mechanism for resolving conflicting links.
-Spack defines a sort order for packages, which arranges packages from ``most desired'' to ``least desired''.
-When a symbolic link could resolve to multiple package installations,
-Spack points the link at the most desired package.
-
-By default, Spack treats newer versions of packages compiled with newer compilers
-as more desirable than older packages built with older compilers.
-It has well-defined, but not necessarily meaningful,
-preferences for deciding between MPI implementations and different compilers.
-A user can override Spack's default sort order through their {\tt spackconfig} file.
-For example, at one site users may typically use the Intel compiler,
-but some users also use the system's default {\tt gcc 4.4.7}.
+Spack has a well-defined mechanism for resolving conflicting links;
+it uses a combination of internal default policies and user- or site-defined
+policies to define an {\it order} of preference for different parameters.
+By default, Spack prefers newer versions of packages compiled with newer compilers
+to older packages built with older compilers. It has well-defined, but not 
+necessarily meaningful, order of preference for deciding between MPI
+implementations and different compilers.
+The default policies can be overridden in configuration files, by either users
+or by sites. For example, at one site users may typically use the Intel compiler,
+but some users also use the system's default {\tt gcc@4.4.7}.
 These preferences could be stated by adding:
@@
 %
-to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} link to point to an {\tt icc} installation.
-Any compiler left out of the compiler\_order setting is treated as less preferred than explicitly stated compilers.
-Spack can also be configured to treat specific package versions as more preferred than other versions,
-which can be useful if a new version is unstable and untested, and should not be advertised with symbolic links.
+to the site's {\tt spackconfig} file, which would cause the ambiguous
+{\tt mpileaks} link to point to an installation compiled with {\tt icc}.
+Any compiler not in the compiler\_order setting is treated as less preferred
+than those explicitly provided.
+%
+In a similar manner, Spack can be configured to give specific package
+configurations priority over others.
+This can be useful if a new version is unstable and untested.
 
-By default, Spack stores its package files in a mainline repository that is present when users
-first run Spack.  At many sites, packages may build sensitive, proprietary software, or they 
-may have patches that are not useful outside of a certain company or organization.  Putting
-this type of code back into a public repository does not often make sense, and if it makes the
+\subsubsection{External Package Repositories}
+By default, Spack stores its package files in a mainline repository
+that is present when users
+first run Spack.  At many sites, packages may build sensitive,
+proprietary software, or they 
+may have patches that are not useful outside of a certain company or
+organization.  Putting this type of code back into a public repository
+does not often make sense, and if it makes the
 mainline less stable, it can actually make sharing code between sites more difficult.
 
-To support our own private packages, and to support those of LLNL code teams, Spack allows the creation of site-specific variants of packages.
-Via the {\tt spackconfig} file, users can specify additional search directories for finding additional {\tt Package} classes.
+To support our own private packages, and to support those of LLNL code teams,
+Spack allows the creation of site-specific variants of packages.
+Via configuration files, users can specify additional search
+directories for finding additional {\tt Package} classes.
 %
-The additional packages are like the {\tt mpileaks} package shown in Figure~\ref{fig:mpileaks}.
-However, the extension packages can extend from not only {\tt Package}, but also any of Spack's built-in packages.
-Custom packages can inherit from and replace the Spack's default packages, allowing sites to either tweak or completely replace Spack's build recipes.
-To continue the previous example, a site can write a {\tt LocalSpindle} Python class, which inherits from Spack's {\tt Spindle} class.
-{\tt LocalSpindle} may simply add additional configure flags to the {\tt Spindle} class, while leaving the dependencies and most of the build instructions from its parent class.
-For reproducibility, Spack also tracks the {\tt Package} class that drove a specific build.
+The additional packages are like the {\tt mpileaks} package shown in Figure~\ref{fig:mpileaks}.  However, the extension packages can extend from not
+only {\tt Package}, but also any of Spack's built-in packages.
+Custom packages can inherit from and replace Spack's default
+packages, so other sites to either tweak or completely replace Spack's
+build recipes.  To continue the previous example, a site can write
+a {\tt LocalSpindle} Python class, which inherits from Spack's
+{\tt Spindle} class.  {\tt LocalSpindle} may simply add additional
+configure flags to the {\tt Spindle} class, while leaving the
+dependencies and most of the build instructions from its parent
+class.  For reproducibility, Spack also tracks the {\tt Package}
+class that drove a specific build.

### 2015-07-19 usecase-python.tex
Change name of tex file to 'spack-paper-sc15'
Edits through use cases.

--- 2015-04-18 usecase-python.tex
@@
 LC supports Python installations for several application teams,
-and maintenance of these repositories has grown increasingly complex over
+and maintaining these repositories has grown increasingly complex over
 time. The problems are similar to those that drove us to create
@@
 
-Existing Python package managers are either language-specific like 
-{\tt easy\_install}~\cite{eby:setuptools}, 
-or they do not effectively support building from source~\cite{anaconda,conda}, let
-alone multi-configuration builds like Spack's.  More glaringly, the default
-model for Python extensions is to install them {\it directly} into the Python
-interpreter's prefix, but this makes it impossible to install multiple versions.
-An alternative is to install Python extensions in their own prefix,
-but then users must manually add each package to their {\tt PYTHONPATH}. 
-We wanted a way to  easily manage many different versions and {\it also} allow teams
-to have a baseline set of Python packages available by default.
+Existing Python package managers are either language-specific~\cite{eby:setuptools}, 
+or they do not support building from source~\cite{anaconda,conda}. None
+handles multi-configuration builds.  More glaringly, Python extensions
+are usually installed into the Python interpreter's prefix.
+This makes it impossible to install multiple versions.\footnote{{\tt setuptools}
+has support for multiple versions via {\tt pkg_resources}, 
+but this requires modifications to client code.}
+Installing each extension in its own prefix enables combinatorial versioning,
+but it requires users to add packages to the {\tt PYTHONPATH} variable at runtime.
 
+Per Spack's design philosophy, we wanted a way to easily manage many different
+versions, but {\it also} to provide a baseline set of extensions {\it without}
+requiring environment settings.
+%
 To support this mode of operation, we added the concept of {\tt extension} packages
-to Spack. Python modules now use the {\tt extends('python')} directive instead of
+to Spack. Python modules use the {\tt extends('python')} directive instead of
 {\tt depends\_on('python')}.
@@
 and each module depends on a particular Python installation.
-Additionally, we added the capability to {\tt activate} and {\tt deactivate} an
-extension within its dependent Python installation.  When this happens, each file
-in the module prefix is symbolically linked into the Python installation prefix, as though it were
-directly installed within the Python instance. If any file conflict would arise from this operation,
-the activate operation fails and notifies the user.
-The deactivate operation removes the symbolic links and restores the Python installation to its 
-pristine state.
+But extensions can be {\tt activated} or {\tt deactivated} an
+within the dependent Python installation.  The {\tt activate} operation
+symbolically links each file in the extension prefix into the Python
+installation prefix, as though it were installed directly. If any file
+conflict would arise from this operation, {\tt activate} fails.
+Similarly, the {\tt deactivate} operation removes the symbolic links and restores
+the Python installation to its  pristine state.
 
 There were additional complications because many Python packages {\it install their own
-package manager} if they do not find {\tt easy\_install} in the Python installation.
-There are also no fewer than three ways that Python packages add themselves to the
-interpreter's default path, and some of these conflict. We modified Spack so that 
-extendable packages, like Python, can supply custom code in their package file that is
-executed during activation and deactivation to handle any necessary language-specific tasks.
-Python uses this feature to merge conflicting files that are generated by the {\tt easy\_install} tool.
+package manager} if they do not find one in the Python installation.
+There are also many ways that Python packages add themselves to the
+interpreter's default path, and some of them conflict. We modified
+Spack so that extendable packages, like Python, can supply custom code
+in the package file that that specializes {\tt activate} and {\tt deactivate}
+for the particular package. Python uses this feature to merge conflicting
+files during activation.  The end result is that Python extensions can
+be installed automatically in their own prefixes, and they can be composed
+with a wide range of bleeding-edge libraries that other package managers do
+not handle.
 
-The end result is that Python extensions can now be installed automatically
-in their own prefixes, but they can be composed with a wide range
-of bleeding-edge libraries that other package managers do not handle.  To experiment
-with these extensions, users can load environment modules generated by Spack.
-If they want a particular version to be available without any special environment 
-settings, they can activate it within the Python instance.
+%To experiment with these extensions, users can load environment
+%modules generated by Spack. If they want a particular version to be available
+%without any special environment settings, they can activate it within the Python instance.
 
-Spack now essentially implements a ``meta package-manager'' for each Python
-instance, and it can live together with Spack's normal installation model.
+Spack essentially implements a ``meta package-manager'' for each Python
+instance, which can coexist with Spack's normal installation model.
 This has allowed us to efficiently support our application teams, for whom we can

### 2015-07-19 usecase-ares.tex
Start ARES section.

%!TEX root = spack-paper-sc15.tex

\subsection{The ARES Multi-physics Code}
\label{sec:ares}

\begin{figure*}
	\includegraphics[width=\textwidth]{figs/ares-dot/ares-fig.pdf}
	\caption{
		Dependencies of ARES, colored by type of package.
		\label{fig:ares}
	}
\end{figure*}

\begin{figure}
\begin{tabular}
\end{figure}

For our final use case, we describe our experiences with ARES, a multi-physics code used
in production at LLNL. ARES~\cite{ares1,ares2} is a 1, 2 and 3-dimensional radiation
hydrodynamics code capable of running small serial to large massively parallel simulations.
It is used primarily in munitions modeling and inertial confinement fusion simulations.
%
ARES runs on commodity Linux clusters and on Blue Gene/Q systems at LLNL.  It also
runs on the Cielo Cray XE6 system at Los Alamos National Laboratory (LANL), and it is being
ported to LANL's forthcoming Trinity Cray XC30 machine, using Trinitite, a smaller version
of the full system.  The Trinity machine will consist of two partitions; 
one using Intel Haswell processors and another using Intel Knights Landing processors.
Currently, only the Haswell partition is deployed on Trinitite.

ARES comprises 47 packages, with complex dependency relationships.  The DAG for
the current production configuration of ARES is shown in Figure~\ref{fig:ares}.
At the top is ARES itself.  ARES depends on 11 LLNL physics packages (red), 
4 LLNL math/meshing libraries (gold), and 8 LLNL utility libraries (green).
The utility libraries handle tasks including logging, I/O, and performance
measurement. In addition, ARES uses 23 external software packages, including MPI, BLAS, 
Python, and many other libraries.  Together, these packages ares written in a diverse
set of languages (C, C++, Fortran, Python, tcl) using MPI and OpenMP for parallelism. 
In the configuration we describe here, we have configured Spack to build ARES with
external MPI implementations, depending on the host system. This is a common
configuration, because the vendor- or site-supplied MPI installation is often
optimized and integrated tightly with network drivers on the host system. MPI is shown
as a virtual dependency in the figure, as the implementation differs according to the
host machine.

### 2015-07-20 usecase-ares.tex
Update ARES use case.

@@
 
-\begin{figure}
-\begin{tabular}
-\end{figure}
+\begin{table}\centering
+\begin{tabular}{|r|c|c|c|c|}
+\hline
+\multirow{2}{*}{} & \multicolumn{2}{|c|}{\bf Linux} & {\bf BG/Q}     & {\bf Cray XE6} \\\cline{2-5}
+                  & {\it MVAPICH} & {\it OpenMPI}   & {\it BG/Q MPI} & {\it Cray MPI} \\\hline
+{\bf GCC}         & {\tt C P L D} &                 &                &                \\\hline
+{\bf Intel}       & {\tt C P L D} & {\tt C P L D}   &                &                \\\hline
+{\bf PGI}         & {\tt C P L D} &                 &                & {\tt C P L D}  \\\hline
+{\bf Clang}       & {\tt C P L D} &                 & {\tt C P L D}  &                \\\hline
+{\bf XL}          &               &                 & {\tt C P L D}  &                \\\hline
+\end{tabular}
+\caption{
+	Configurations of ARES built with Spack: \newline
+	(C)urrent and 
+	(P)revious production, (L)ite, and (D)evelopment).
+	\label{tab:ares-configs}
+}
+\end{table}
 
-For our final use case, we describe our experiences with ARES, a multi-physics code used
-in production at LLNL. ARES~\cite{ares1,ares2} is a 1, 2 and 3-dimensional radiation
-hydrodynamics code capable of running small serial to large massively parallel simulations.
-It is used primarily in munitions modeling and inertial confinement fusion simulations.
+For our final use case, we describe our experiences using Spack to build ARES.
+ARES~\cite{ares1,ares2} is a 1, 2 and 3-dimensional radiation hydrodynamics code,
+developed for production use at LLNL.  It is capable of running both small, serial
+and large, massively parallel jobs. ARES is used primarily in munitions modeling
+and inertial confinement fusion simulations.
 %
-ARES runs on commodity Linux clusters and on Blue Gene/Q systems at LLNL.  It also
-runs on the Cielo Cray XE6 system at Los Alamos National Laboratory (LANL), and it is being
-ported to LANL's forthcoming Trinity Cray XC30 machine, using Trinitite, a smaller version
-of the full system.  The Trinity machine will consist of two partitions; 
+At LLNL, it runs on commodity Linux clusters and on Blue Gene/Q systems.
+It also runs on the Cielo Cray XE6 system at Los Alamos National Laboratory (LANL), and
+it is being ported to LANL's forthcoming Trinity Cray XC30 machine on Trinitite,
+a smaller version of the full system.  The Trinity machine will consist of two partitions; 
 one using Intel Haswell processors and another using Intel Knights Landing processors.
@@
 Python, and many other libraries.  Together, these packages ares written in a diverse
-set of languages (C, C++, Fortran, Python, tcl) using MPI and OpenMP for parallelism. 
+set of languages (C, C++, Fortran, Python, tcl, etc.) using MPI and OpenMP for parallelism.
+
 In the configuration we describe here, we have configured Spack to build ARES with
@@
 as a virtual dependency in the figure, as the implementation differs according to the
-host machine.
+host machine.  ARES builds its own Python version in order to run on machines
+where Python is not well supported, like Blue Gene/Q.  In particular, ARES builds 
+a version of Python 2.7 for Blue Gene/Q, which is not supported by the native
+software stack.
+
+When we first began developing Spack packages for ARES and its dependencies,
+the code managed its software stack using the MixDown meta-build system.  
+Thus, the ARES team already had some experience supporting automated builds of
+dependencies. We developed Spack packages for all of the LLNL packages shown in
+Figure~\ref{fig:ares}.
+Many of the external packages were already available in Spack when ARES was
+packaged, but some, such as Python, required modifications to support the
+new platforms and compilers.
+
+Table~\ref{tab:ares-configs} shows the configurations of ARES that we have packaged.
+The rows and columns show architectures, compilers, and MPI versions.
+Currently, we use the ARES Spack package to build four different configurations
+of ARES: the current and previous production versions, a ``lite'' version that includes
+a smaller set of features and dependencies, and a development version.  We denote these
+with C, P, L, and D, respectively.  Each cell in the table indicates the 
+ARES configurations built for an architecture, compiler, and MPI combination.
+Each different configuration of ARES requires a slightly different set of
+dependencies and dependency versions, but all are supported through one common
+ARES package using conditional logic on versions and variants.
+
+Altogether, the initial packaging effort for ARES took roughly two months,
+half time, for an experienced build engineer.  With 4 configurations and 8 
+architecture-compiler-MPI combinations, this is a total of 64 different builds.
+Prior to using Spack, only the Linux/Intel configurations were automated.
+In our discussions with the ARES team, they listed a number of key features
+that enabled the new new platform builds to be automated:
+\begin{enumerate}
+\item Spack's version tracking and optional dependencies were required to
+      build the four configurations with correct libraries;
+\item The spec syntax allowed build scripts to concisely test compiler, 
+      compiler version, and dependency versions: a necessity
+      for handling the particulars of different architectures;
+\item Patching packages for particular platforms was
+      necessary to build many packages; and 
+\item Using a DSL embedded in Python was a significant benefit;
+      certain packages required custom scripting to patch.
+\end{enumerate}
+
+In addition to these immediate benefits, the ARES team mentioned
+two potential long-term payoffs. First, using Spack allowed
+the team to test with Clang.  This compiler is not currently used in
+production, but we expect it to be used in production on future LLNL machines.
+Testing with Clang revealed many incompatibilities, which were patched with
+Spack. The team is communicating these changes back to library developers,
+who will integrate them in future versions.  In this case, build automation
+has allowed more testing. This helps both ARES and LLNL library developers
+build more robust software.
+%
+Second, many of the libraries used by ARES are also used by other LLNL code
+teams.  The LLNL code teams have begun creating an internal repository of
+Spack build recipes to be shared.  Leveraging such a repository will
+make packaging the next code significantly easier.

### 2015-07-20 implementation.tex
Name change spack-paper-sc15 -> spack-sc15
clarify unique features.

--- 2015-04-18 implementation.tex
@@
-%!TEX root = paper.tex
+%!TEX root = spack-sc15.tex
 
@@
 \noindent
-Spack has the following unique capabilities:
+Spack provides the following unique features:
 \begin{enumerate}
-\item Packages are {\bf parameterized}, allowing new builds to be composed
-      on the fly for different versions, architecture, compiler, options, 
-      and dependencies.
-\item To manage this large parameter space, spack provides a novel, 
-      recursive {\bf spec syntax} for dependency graphs.
-\item Specs support versioned, ABI-incompatible interfaces like MPI through
-      {\bf versioned virtual dependencies}.
-\item Spack builds with {\bf compiler wrappers} that enforce build
+\item Packages explicitly {\bf parameterized} by version, platform,
+      compiler, options, and dependencies, to allow easy composition.
+\item A novel, recursive {\bf spec syntax} for dependency graphs and constraints,
+      which aids in managing the build parameter space.
+\item {\bf Versioned virtual dependencies} to handle versioned, 
+      ABI-incompatible interfaces like MPI.
+\item A novel {\bf concretization} process that translates an abstract build
+      specification into a full, concrete build specification.
+\item A build environment with {\bf compiler wrappers} that enforce build
       consistency and simplify package writing.

### 2015-07-20 limitations.tex
Edits through use cases.
Name change spack-paper-sc15 -> spack-sc15
Limitations section.

@@
-%!TEX root = spack-paper-sc15.tex
+%!TEX root = spack-sc15.tex
 
@@
 
-In future work we will investigate adding a SAT solver to Spack's
-concretization process to present users with better options.  In our
-experience, however, complex policy interactions like these are very
-rare.
+There are currently several limitations to Spack as implemented.
+As mentioned in Section~\ref{sec:concretization}, the concretization
+algorithm makes choices greedily. For example, if a package {\tt P} depends on 
+both {\tt hwloc@1.9} and {\tt mpi}, and if the concretization algorithm
+chooses an MPI implementation for {\tt P} that depends strictly on
+{\tt hwloc@1.8}, this will lead to a conflict on {\tt hwloc}, and Spack
+will raise an error and the user must resolve the issue.
+Our implementation will not backtrack to find an MPI version
+that does not conflict. So far, cases like this have been rare.
+However, we plan to add better constraint satisfaction solving to
+Spack in future work.
+
+While Spack supports different architectures as part of the 
+configuration space, we do not currently have ways to factor common
+preferences (like {\tt configure} arguments and architecture-specific
+compiler flags) out of packages and into separate architecture
+descriptions.  This leads to some clutter in the package files when
+too many per-platform conditions accumulate.  We are in the process
+of adding features that would further simplify Spack's build templates
+for cross-platform installations.
+
+Spack can require more disk space than a module-based
+system, as we require otherwise identical packages with different
+dependencies to be built separately. In our view, the significant
+reduction in complexity for the end user justifies this cost.
+
+The use of Python has been a barrier for some users, and
+a major attraction for others. For those who would otherwise create local
+workarounds, Python's flexibility has allowed them to extend local
+Spack packages to suit their needs. Such customizations frequently
+guide our priorities for new core features. We believe that the
+benefits of using a popular, rapidly emerging HPC language for our
+embedded DSL outweigh any costs of the learning curve.
+
+In addition to the future directions already mentioned,
+our main near-term goal is to grow a community around Spack and to
+build a larger base of contributors.  On the technical side, we will
+increase build robustness by deploying a continuous testing system
+at LLNL.  Finally, to support the growing number of
+HPC languages, runtimes, and programming models, we plan to add
+capabilities to Spack that allow packages to depend on particular
+compiler features.  More and more, our codes are beginning to rely
+on advanced compiler capabilities, like {\tt C++11} language features,
+OpenMP versions, and GPU compute capabilities. Ideally, Spack
+will find suitable compilers and ensure ABI consistency when many
+such features are in use.

### 2015-07-20 figs/concretization-overhead/concretization-overhead.py
Add concretization time figure.

@@ -1,0 +1,32 @@
+#!/usr/bin/env spack-python
+import time
+import numpy as np
+import spack
+from external import yaml
+from spack.spec import Spec
+
+# prime the timer
+s = Spec('qt')
+s.concretize()
+
+reps = 10
+timings = []
+
+for name in spack.db.all_package_names():
+    print name
+    specs = [Spec(s) for s in [name] * reps]
+    times = []
+
+    for s in specs:
+        start = time.clock()
+        s.concretize()
+        end = time.clock()
+        times.append(end - start)
+
+    record = (name, s.size, times)
+    print record
+    timings.append(record)
+
+
+with open('timings.yaml', 'w') as stream:
+    yaml.dump(stream)

### 2015-07-20 figs/concretization-overhead/timings.py

@@ -1,0 +1,29 @@
+#!/usr/bin/env spack-python
+
+from external import yaml
+import matplotlib.pyplot as plt
+
+files = { 'macbook.yaml' : 'MacBook Pro, Haswell, 2.6Ghz',
+          'merl.yaml'    : 'Linux, SandyBridge, 2.6Ghz'}
+
+
+styles = ['bd', 'r^']
+
+for i, (f, label) in enumerate(files.items()):
+    with open(f) as stream:
+        timings = yaml.load(stream)
+
+    timings.sort(key=lambda x:int(x[1]))
+
+    timings = [(name, size, sum(t)/len(t)) for name, size, t in timings]
+    sizes = [t[1] for t in timings]
+    times = [t[2] for t in timings]
+
+    plt.plot(sizes, times, styles[i], label=label)
+
+plt.ylabel('Time (seconds)', fontsize=18)
+plt.xlabel('Package DAG size (nodes)', fontsize=18)
+
+plt.legend(loc=2)
+
+plt.savefig('concretization-times.pdf')

### 2015-07-20 figs/concretization-overhead/package-sizes.py
Package size concretization timings, histogram

@@ -1,0 +1,20 @@
+#!/usr/bin/env spack-python
+
+from external import yaml
+import matplotlib.pyplot as plt
+import numpy as np
+
+with open('merl.yaml') as stream:
+    timings = yaml.load(stream)
+
+sizes = [t[1] for t in timings]
+
+#plt.yscale('log')
+
+plt.axes().set_aspect(.25)
+plt.hist(sizes, 25)
+
+plt.ylabel('Count', fontsize=18)
+plt.xlabel('Package DAG size (nodes)', fontsize=18)
+
+plt.savefig('package-sizes.pdf')

### 2015-07-20 figs/concretization-overhead/timings.py

@@ -9,6 +9,8 @@ files = { 'macbook.yaml' : 'MacBook Pro, Haswell, 2.6Ghz',
 
 styles = ['bd', 'r^']
 
+plt.axes().set_aspect(3)
+
 for i, (f, label) in enumerate(files.items()):
     with open(f) as stream:
         timings = yaml.load(stream)
@@ -23,7 +25,6 @@ for i, (f, label) in enumerate(files.items()):
 
 plt.ylabel('Time (seconds)', fontsize=18)
 plt.xlabel('Package DAG size (nodes)', fontsize=18)
-
 plt.legend(loc=2)
 
 plt.savefig('concretization-times.pdf')

### 2015-07-20 implementation-concretization.tex
Name change spack-paper-sc15 -> spack-sc15
Limitations section.
Concretization runtime section.

@@
-%!TEX root = spack-paper-sc15.tex
+%!TEX root = spack-sc15.tex
 
 \subsection{Concretization}
+\label{sec:concretization}
 
@@
 
-When the concretization process completes, it outputs a fully concrete spec DAG. 
-Figure~\ref{fig:specs-mpileaks-concrete} shows a concrete DAG with architectures, 
-compilers, versions, variants, and all dependencies resolved.
-This fulfills Spack's guarantees.
-
 \begin{figure}
@@
 \end{figure}
+When the concretization process completes, it outputs a fully concrete spec DAG.
+Figure~\ref{fig:specs-mpileaks-concrete} shows a concrete DAG with architectures,
+compilers, versions, variants, and all dependencies resolved.
+This fulfills Spack's guarantees.
 
@@
 
-\subsubsection{Shared sub-DAGs}\label{sec:directory-layout}
+\subsubsection{Concretization Runtime}
+\label{sec:concretization-overhead}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figs/concretization-overhead/concretization-times.pdf}
+	\caption{
+		Concretization running time for 245 packages.
+		\label{fig:concretization-time}
+	}
+\end{figure}
+
+Figure~\ref{fig:concretization-time} shows the running time of concretization vs.
+the size of a package DAG (in nodes).  We generated this plot by running {\tt concretize}
+on all 245 packages in our internal version of Spack.  We see that for all but the 9 largest
+packages, the algorithm takes less than 2 seconds on a Sandy Bridge cluster node, and less
+than 1 second on a Haswell MacBook Pro.  For larger DAGs, we begin to see superlinear
+scaling, but even for 50 nodes or more, concretization takes less than 6 seconds.
+%
+This performance is expected from a greedy, fixed-point method.
+The algorithm repeats, evaluating conditions and adding dependencies and constraints, until
+there are no more changes to be made.
+%
+%  Additionally, our current compiler concretization
+% policy requires a search in the DAG; it prefers the compiler of the nearest ancestor
+% if the current node's compiler is not set. 
+%
+More generally, concretization is an instance of the constraint
+satisfaction problem, which is NP-complete. Our greedy solver is sufficient for the
+packages we have built so far, and its running time is currently negligible. We do
+not expect to see software packages with thousands of dependencies in the near
+future, and we do not expect concretization to become a bottleneck, even if we
+use a full constraint solver.
+
+\subsubsection{Shared sub-DAGs}
+\label{sec:directory-layout}
+
 \begin{figure}\centering
@@
 \subsubsection{Reproducibility}
+
 For reproducibility, and to preserve provenance, Spack stores a number of

### 2015-07-20 figs/concretization-overhead/package-sizes.py

@@ -10,11 +10,11 @@ with open('merl.yaml') as stream:
 sizes = [t[1] for t in timings]
 
 #plt.yscale('log')
+plt.axes().set_aspect(.15)
+plt.hist(sizes, 20)
 
-plt.axes().set_aspect(.25)
-plt.hist(sizes, 25)
 
-plt.ylabel('Count', fontsize=18)
-plt.xlabel('Package DAG size (nodes)', fontsize=18)
+plt.ylabel('Count')
+plt.xlabel('Package DAG size (nodes)')
 
 plt.savefig('package-sizes.pdf')

### 2015-07-20 figs/concretization-overhead/timings.py

@@ -4,7 +4,7 @@ from external import yaml
 import matplotlib.pyplot as plt
 
 files = { 'macbook.yaml' : 'MacBook Pro, Haswell, 2.6Ghz',
-          'merl.yaml'    : 'Linux, SandyBridge, 2.6Ghz'}
+          'merl.yaml'    : 'Linux, Sandy Bridge, 2.6Ghz'}
 
 
 styles = ['bd', 'r^']
@@ -23,8 +23,8 @@ for i, (f, label) in enumerate(files.items()):
 
     plt.plot(sizes, times, styles[i], label=label)
 
-plt.ylabel('Time (seconds)', fontsize=18)
-plt.xlabel('Package DAG size (nodes)', fontsize=18)
+plt.ylabel('Time (seconds)')
+plt.xlabel('Package DAG size (nodes)')
 plt.legend(loc=2)
 
 plt.savefig('concretization-times.pdf')

### 2015-07-20 figs/concretization-overhead/package-sizes.py
Add wrapper overhead figure

@@ -1,4 +1,4 @@
-#!/usr/bin/env spack-python
+#!/usr/bin/env python
 
 from external import yaml
 import matplotlib.pyplot as plt

### 2015-07-20 figs/concretization-overhead/timings.py

@@ -1,6 +1,6 @@
-#!/usr/bin/env spack-python
+#!/usr/bin/env python
 
-from external import yaml
+import yaml
 import matplotlib.pyplot as plt
 
 files = { 'macbook.yaml' : 'MacBook Pro, Haswell, 2.6Ghz',

### 2015-07-20 figs/wrappers/wrapper-overhead.py

@@ -1,0 +1,53 @@
+#!/usr/bin/env python
+import sys
+import yaml
+import numpy as np
+import matplotlib.pyplot as plt
+
+with open('overhead.yaml') as f:
+    data = yaml.load(f)
+
+N = len(data)
+
+keys = sorted(data.keys())
+
+wrapper_means = []
+no_wrap_means = []
+for pkg in keys:
+    wrapper = [float(f) for f in data[pkg]['wrapper']]
+    no_wrap = [float(f) for f in data[pkg]['no_wrapper']]
+
+    wrapper_means.append(sum(wrapper) / N)
+    no_wrap_means.append(sum(no_wrap) / N)
+
+ind = np.arange(N)  # the x locations for the groups
+width = 0.35       # the width of the bars
+
+fig, ax = plt.subplots()
+
+
+wrapper_rects = ax.bar(ind,       wrapper_means, width, color='r')
+no_wrap_rects = ax.bar(ind+width, no_wrap_means, width, color='b')
+
+# add some text for labels, title and axes ticks
+ax.set_xlabel('Libraries')
+ax.set_ylabel('Build time (s)')
+ax.set_xticks(ind+width)
+ax.set_xticklabels(keys)
+
+ax.legend( (wrapper_rects[0], no_wrap_rects[0]), ('Wrapper', 'No Wrapper') )
+
+def autolabel(rects):
+    # attach some text labels
+    for rect in rects:
+        height = rect.get_height()
+        ax.text(rect.get_x()+rect.get_width()/2., 1.05*height,
+                '%d'%int(height),
+                ha='center', va='bottom')
+
+#autolabel(wrapper_rects)
+#autolabel(no_wrap_rects)
+
+ax.set_aspect(.25 * 1/ax.get_data_ratio())
+
+plt.savefig('wrapper-overhead.pdf')

### 2015-07-20 implementation-environment.tex
Name change spack-paper-sc15 -> spack-sc15
Updated wrapper overhead figure

@@
 
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figs/wrappers/wrapper-overhead.pdf}
+	\caption{
+		Build times with and without Spack's compiler wrappers.
+		\label{fig:wrapper-overhead}
+	}
+\end{figure}
+
+Spack's compiler wrappers incur a small amount of overhead because they intercept
+calls to compilers within a build.  Figure~\ref{fig:wrapper-overhead} shows build time
+with and without the wrappers for seven different packages.
+
 \subsubsection{Environment Module Integration}

### 2015-07-20 figs/wrappers/wrapper-overhead.py

@@ -17,10 +17,17 @@ for pkg in keys:
     wrapper = [float(f) for f in data[pkg]['wrapper']]
     no_wrap = [float(f) for f in data[pkg]['no_wrapper']]
 
-    wrapper_means.append(sum(wrapper) / N)
-    no_wrap_means.append(sum(no_wrap) / N)
+    wrapper_means.append(sum(wrapper) / len(wrapper))
+    no_wrap_means.append(sum(no_wrap) / len(no_wrap))
 
-ind = np.arange(N)  # the x locations for the groups
+wrap = np.array(wrapper_means)
+orig = np.array(no_wrap_means)
+overhead = (wrap - orig) / orig
+for k, z in zip(keys, overhead):
+    print k, overhead
+
+ind = np.arange(N,dtype=float)  # the x locations for the groups
+ind += .1
 width = 0.35       # the width of the bars
 
 fig, ax = plt.subplots()
@@ -39,14 +46,17 @@ ax.legend( (wrapper_rects[0], no_wrap_rects[0]), ('Wrapper', 'No Wrapper') )
 
 def autolabel(rects):
     # attach some text labels
-    for rect in rects:
+    for i, rect in enumerate(rects):
         height = rect.get_height()
-        ax.text(rect.get_x()+rect.get_width()/2., 1.05*height,
-                '%d'%int(height),
+        ax.text(rect.get_x()+rect.get_width()/2. + .1, 1.05*height,
+                #'%d'%int(height),
+                '%.1f%%' % (overhead[i] * 100),
                 ha='center', va='bottom')
 
-#autolabel(wrapper_rects)
+autolabel(wrapper_rects)
 #autolabel(no_wrap_rects)
+
+plt.gca().set_ylim(top=400)
 
 ax.set_aspect(.25 * 1/ax.get_data_ratio())
 

### 2015-07-20 conclusion.tex
Change name of tex file to 'spack-paper-sc15'
Name change spack-paper-sc15 -> spack-sc15
Build environment figure text.

@@
 of the software configuration space directly. None of them allows
-a user to rapidly {\it compose} new parametric builds.
+a user to rapidly {\it compose} new builds with version, compiler,
+and dependency parameters.
 
@@
 We outlined a number of Spack's unique features, including
-versioned virtual dependencies, and the {\it concretization} process,
-which converts an underspecified build DAG into a buildable spec.
+versioned virtual dependencies, and its novel {\it concretization}
+process, which converts an abstract build DAG into a concrete,
+buildable spec.
 
-We showed through three use cases that Spack is already
-increasing operational efficiency 
-in production at LLNL.  We believe that the software management
-techniques implemented in
-Spack are applicable to a broad range of HPC facilities.
-Spack is available online at: http://bit.ly/spack-git.
+We showed through four use cases that Spack is already increasing
+operational efficiency in production at LLNL.  We believe that the
+software management techniques implemented in Spack are applicable
+to a broad range of HPC facilities.
+Spack is available online at http://github.com/scalability-llnl/spack.

### 2015-07-20 implementation-environment.tex
Build environment figure text.

--- 2015-07-19 implementation-environment.tex
@@
 
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figs/wrappers/wrapper-overhead.pdf}
+	\caption{
+		Build times with and without compiler wrappers\newline
+		(\% overhead above bars).
+		\label{fig:wrapper-overhead}
+	}
+\end{figure}
+
+Argument parsing and indirection cause Spack's compiler wrappers to incur a small
+but noticeable performance overhead.  Figure~\ref{fig:wrapper-overhead} shows build times
+with and without the wrappers for seven different packages. Above each set of bars,
+the percent overhead is written. The {\tt dyninst} and {\tt lapack} packages
+use CMake to build, while the rest of the packages use autotools.
+The relative overheads range from negligible (-0.4\%) in the case of {\tt dyninst},
+and they are more exaggerated for shorter builds like {\tt mpileaks}, which
+incurs 12.3\%.
+
 \subsubsection{Environment Module Integration}

### 2015-07-20 introduction.tex
Name change spack-paper-sc15 -> spack-sc15
Build environment figure text.

@@
 \item Spack: An implementation of these concepts; and
-\item Three use cases that detail LLNL's use of Spack to compose
+\item Four use cases that detail LLNL's use of Spack to comcpose
       software configurations rapidly, to manage Python installations, and
-      to implement site-specific build policies.
+      to implement site build policies, and to build an LLNL production code.
 \end{enumerate}

### 2015-07-20 introduction.tex
various small tweaks to intro

@@
 and to support new versions of the operating system (OS), Message
-Passing Interface (MPI) implementation, compilers, and other
+Passing Interface (MPI) implementation, compiler, and other
 dependencies.  Unfortunately, building scientific software is
@@
 dependency management is overly rigid, and they burden users with
-the combinatorial naming and versioning problems.
+combinatorial naming and versioning problems.
 
-In this paper, we describe our experiences with the {\it Spack} package manager,
+This paper describes our experiences with the {\it Spack} package manager,
 which we have developed at LLNL to manage increasing software complexity.
-Specifically, this paper offers the following contributions:
+It makes the following contributions:
 \begin{enumerate}

### 2015-07-20 motivation.tex
Name change spack-paper-sc15 -> spack-sc15
Add ROCKS reference.
tweaks to motivation

@@
 Package managers automate the installation of complex sets of software packages.
-{\it Binary package managers} such as RPM, yum, and
-APT~\cite{foster+:rpm03,silva:apt01,yum} are integrated with most
+{\it Binary package managers} such as RPM, yum, and APT~\cite{foster+:rpm03,silva:apt01,yum} are integrated with most
 OS distributions, and they are used to ensure that dependencies
 are installed before packages that require them.
+Others, such as Anaconda~\cite{anaconda,conda}, use the same approach but
+are designed to run on top of a host OS.
 These tools largely solve the problem of managing a {\it single} software
@@
 so running many configurations still requires a large number of VMs.
-For facilities, this is a security concern, as a profusion of VMs makes
-mandatory patching difficult.  It is also tedious for users to manage a large
-number of VM environments.
+For facilities, this profusion of VMs is a security concern since it 
+complicates mandatory patching. Also, managing  a large number of VM 
+environments is tedious for users.
 
@@
 distribution, CHAOS~\cite{chaos} for this purpose, which is managed using RPM.
+The popular ROCKS~\cite{rocks} cluster distribution uses RPM and Anaconda
+in a similar fashion.
 %
@@
 %
-Smithy~\cite{digirolamo:smithy} is an installation tool in use at ORNL. It
+ORNL uses the Smithy~\cite{digirolamo:smithy} installation tool. It
 can generate module files, but it does not provide any
@@
 
-Nix~\cite{dolstra+:icfp08,dolstra+:lisa04}
-is a package manager and an OS distribution that supports the installation of
-arbitrarily many software configurations.  As at most HPC sites, each package
-in Nix is installed in a unique prefix, but Nix does not have a human-readable
-naming convention.  Instead, Nix determines the prefix by hashing the package
-file along with its dependencies. Nix package files are written in a custom
-functional language designed for packaging.
+The Nix~\cite{dolstra+:icfp08,dolstra+:lisa04} package manager and OS 
+distribution supports installation of arbitrarily many software 
+configurations.  As at most HPC sites, it installs each package
+in a unique prefix but it does not have a human-readable naming 
+convention.  Instead, Nix determines the prefix by hashing the package
+file and its dependencies.
+%% BRONIS: I'm not sure what the following sentence really adds
+%%  Nix package files are written in a custom
+%% functional language designed for packaging.
 
@@
 \paragraph{Gaps in current practice}
-The cryptographic versioning of Nix and HashDist is very flexible. It versions
-the package {\it and} its dependency configuration, and can represent any
-configuration. However, users cannot navigate or easily query the installed
+The flexible cryptographic versioning of Nix and HashDist manages 
+the package {\it and} its dependency configuration and can represent any
+configuration. However, users cannot easily navigate or query the installed
 software.
@@
 %
-EasyBuild and Smithy generate environment modules, and some querying is
-possible through the module system.  Naming schemes used in existing module
-systems, however, cannot handle combinatorial versions. The Lmod authors
-call this the ``matrix problem''~\cite{mclay:lmod-tutorial}.
+EasyBuild and Smithy generate environment modules, which supports some 
+querying. Naming schemes used in existing module systems, however, cannot 
+handle combinatorial versions, which the Lmod authors call the 
+``matrix problem''~\cite{mclay:lmod-tutorial}.
 
@@
 The main limitation of existing tools is the lack of build {\it composability}.
-The full set of package versions is combinatorial, and realizing an arbitrary
-combination of compiler, MPI version, build options, and dependency versions
-requires tediously modifying many package files.
-Indeed, the number of package files required for most existing systems scales
-with the number of version {\it combinations}, not the number of packages, which
-quickly becomes unmanageable.  As an example, the EasyBuild
-system has over 3,300 files for several permutations of only 600 packages.
-A slightly different dependency graph requires an entire new package file
-hierarchy.  HashDist a more robust support for composition,
-but it does not have a first-class parameters for versions, compilers,
-or versioned interfaces.
+The full set of package versions is combinatorial, and arbitrary combinations 
+of compiler, MPI version, build options, and dependency versions requires 
+tedious modifications of many package files. Indeed, the number of package 
+files required for most existing systems scales with the number of version 
+{\it combinations}, not the number of packages, which quickly becomes 
+unmanageable.  As an example, the EasyBuild system has over 3,300 files for 
+several permutations of only 600 packages. A slightly different dependency 
+graph requires an entire new package file hierarchy.  HashDist supports 
+composition more robustly but does not have first-class parameters for 
+versions, compilers, or versioned interfaces.
 %

### 2015-07-20 limitations.tex
various small tweaks

--- 2015-07-19 limitations.tex
@@
 
-In future work we will investigate adding a SAT solver to Spack's
-concretization process to present users with better options.  In our
-experience, however, complex policy interactions like these are very
-rare.
+Spack's current implementation has several limitations.
+As mentioned in Section~\ref{sec:concretization}, we use a greedy 
+concretization algorithm. For example, if package {\tt P} depends on 
+both {\tt hwloc@1.9} and {\tt mpi}, and if the algorithm
+chooses an MPI implementation for {\tt P} that depends strictly on
+{\tt hwloc@1.8}, a conflict on {\tt hwloc} arises. In this case, Spack 
+raises an error and the user must resolve the issue. Our implementation 
+does not backtrack to find an MPI version that does not conflict. These 
+cases have been rare so far. However, we plan to add better constraint
+satisfaction solving to Spack in future work.
+
+While Spack supports different architectures as part of the 
+configuration space, we cannot currently factor common preferences (like 
+{\tt configure} arguments and architecture-specific compiler flags) out of 
+packages and into separate architecture descriptions, which leads to some 
+clutter in the package files when too many per-platform conditions accumulate.
+We are adding features that will further simplify Spack's build templates
+for cross-platform installations.
+
+Spack can require more disk space than a module-based system, as otherwise 
+identical packages with different dependencies must be built separately. In 
+our view, the significant reduction in complexity for the end user justifies 
+this cost.
+
+The use of Python has been a barrier for some users, and
+a major attraction for others. For those who would otherwise create local
+workarounds, Python's flexibility has allowed them to extend local
+Spack packages to suit their needs. Such customizations frequently
+guide our priorities for new core features. The
+benefits of using a popular, rapidly emerging HPC language for our
+embedded DSL outweigh any costs of the learning curve.
+
+In the near term, we are actively working to grow a community around Spack 
+and to build a larger base of contributors. On the technical side, we will
+increase build robustness by deploying a continuous testing system
+at LLNL.  Finally, to support the growing number of
+HPC languages, runtimes, and programming models, we will add
+capabilities to Spack that allow packages to depend on particular
+compiler features.  More and more, our codes are beginning to rely
+on advanced compiler capabilities, like {\tt C++11} language features,
+OpenMP versions, and GPU compute capabilities. Ideally, Spack
+will find suitable compilers and ensure ABI consistency when many
+such features are in use.

### 2015-07-20 usecase-ares.tex
Name change spack-paper-sc15 -> spack-sc15
Limitations section.
several wording tweaks

@@
 
-\begin{figure*}
+For our final use case, we describe our experiences using Spack to build ARES.
+ARES~\cite{ares1,ares2} is a 1, 2 and 3-dimensional radiation hydrodynamics code,
+developed for production use at LLNL.  It can run both small, serial
+and large, massively parallel jobs. ARES is used primarily in munitions modeling
+and inertial confinement fusion simulations.
+%
+At LLNL, it runs on commodity Linux clusters and on Blue Gene/Q systems.
+It also runs on the Cielo Cray XE6 system at Los Alamos National Laboratory (LANL), and
+it is being ported to LANL's forthcoming Trinity Cray XC30 machine on Trinitite,
+a smaller version of the full system.  The Trinity machine will consist of two partitions;
+one using Intel Haswell processors and another using Intel Knights Landing processors.
+Currently, only the Haswell partition is deployed on Trinitite.
+
+\begin{figure*}[t]
 	\includegraphics[width=\textwidth]{figs/ares-dot/ares-fig.pdf}
@@
 \end{figure*}
+
+ARES comprises 47 packages, with complex dependency relationships. 
+Figure~\ref{fig:ares} shows the DAG for the current production configuration 
+of ARES. At the top is ARES itself.  ARES depends on 11 LLNL physics packages 
+(red), 4 LLNL math/meshing libraries (gold), and 8 LLNL utility libraries 
+(green). The utility libraries handle tasks including logging, I/O, and 
+performance measurement. ARES also uses 23 external software packages, 
+including MPI, BLAS, Python, and many other libraries.  Together, these 
+packages ares written in a diverse set of languages including C, C++, 
+Fortran, Python and tcl and uses MPI and OpenMP for parallelism.
+
+We have configured Spack to build ARES with external MPI implementations, 
+depending on the host system. This configuration exploits the vendor- 
+or site-supplied MPI installation that often uses host-specific optimized 
+network drivers. MPI is shown as a virtual dependency in the figure, as 
+the implementation differs according to the host machine.  ARES builds its 
+own Python version in order to run on machines where Python is not well 
+supported, like Blue Gene/Q.  In particular, ARES builds a version of 
+Python 2.7 for Blue Gene/Q, which the native software stack does not support.
+
+Prior to using Spack, ARES managed its software stack with MixDown.
+Thus, the ARES team already had some experience supporting 
+automated builds of dependencies. We developed Spack packages for the LLNL 
+packages in Figure~\ref{fig:ares}. Many of the external packages were already 
+available in Spack, but some, such as Python, required modifications to support
+the new platforms and compilers.
 
@@
 
-For our final use case, we describe our experiences using Spack to build ARES.
-ARES~\cite{ares1,ares2} is a 1, 2 and 3-dimensional radiation hydrodynamics code,
-developed for production use at LLNL.  It is capable of running both small, serial
-and large, massively parallel jobs. ARES is used primarily in munitions modeling
-and inertial confinement fusion simulations.
-%
-At LLNL, it runs on commodity Linux clusters and on Blue Gene/Q systems.
-It also runs on the Cielo Cray XE6 system at Los Alamos National Laboratory (LANL), and
-it is being ported to LANL's forthcoming Trinity Cray XC30 machine on Trinitite,
-a smaller version of the full system.  The Trinity machine will consist of two partitions; 
-one using Intel Haswell processors and another using Intel Knights Landing processors.
-Currently, only the Haswell partition is deployed on Trinitite.
-
-ARES comprises 47 packages, with complex dependency relationships.  The DAG for
-the current production configuration of ARES is shown in Figure~\ref{fig:ares}.
-At the top is ARES itself.  ARES depends on 11 LLNL physics packages (red), 
-4 LLNL math/meshing libraries (gold), and 8 LLNL utility libraries (green).
-The utility libraries handle tasks including logging, I/O, and performance
-measurement. In addition, ARES uses 23 external software packages, including MPI, BLAS, 
-Python, and many other libraries.  Together, these packages ares written in a diverse
-set of languages (C, C++, Fortran, Python, tcl, etc.) using MPI and OpenMP for parallelism.
-
-In the configuration we describe here, we have configured Spack to build ARES with
-external MPI implementations, depending on the host system. This is a common
-configuration, because the vendor- or site-supplied MPI installation is often
-optimized and integrated tightly with network drivers on the host system. MPI is shown
-as a virtual dependency in the figure, as the implementation differs according to the
-host machine.  ARES builds its own Python version in order to run on machines
-where Python is not well supported, like Blue Gene/Q.  In particular, ARES builds 
-a version of Python 2.7 for Blue Gene/Q, which is not supported by the native
-software stack.
-
-When we first began developing Spack packages for ARES and its dependencies,
-the code managed its software stack using the MixDown meta-build system.  
-Thus, the ARES team already had some experience supporting automated builds of
-dependencies. We developed Spack packages for all of the LLNL packages shown in
-Figure~\ref{fig:ares}.
-Many of the external packages were already available in Spack when ARES was
-packaged, but some, such as Python, required modifications to support the
-new platforms and compilers.
-
 Table~\ref{tab:ares-configs} shows the configurations of ARES that we have packaged.
 The rows and columns show architectures, compilers, and MPI versions.
-Currently, we use the ARES Spack package to build four different configurations
-of ARES: the current and previous production versions, a ``lite'' version that includes
-a smaller set of features and dependencies, and a development version.  We denote these
-with C, P, L, and D, respectively.  Each cell in the table indicates the 
+The ARES Spack package supports four different code configurations:
+the current and previous production versions, a ``lite'' version that includes
+a smaller set of features and dependencies, and a development version.  We 
+denote these as C, P, L, and D. Each cell in the table indicates the
 ARES configurations built for an architecture, compiler, and MPI combination.
-Each different configuration of ARES requires a slightly different set of
-dependencies and dependency versions, but all are supported through one common
-ARES package using conditional logic on versions and variants.
+Each configuration requires a slightly different set of dependencies and 
+dependency versions, but one common ARES package supports all of them with 
+conditional logic on versions and variants.
 
 Altogether, the initial packaging effort for ARES took roughly two months,
-half time, for an experienced build engineer.  With 4 configurations and 8 
-architecture-compiler-MPI combinations, this is a total of 64 different builds.
-Prior to using Spack, only the Linux/Intel configurations were automated.
-In our discussions with the ARES team, they listed a number of key features
-that enabled the new new platform builds to be automated:
+half time, for an experienced build engineer. It supports 64 different builds
+(4 configurations and 8 architecture-compiler-MPI combinations). Prior to 
+using Spack, only the Linux/Intel configurations were automated. The ARES 
+team listed a number of key features that enabled the increased automation:
 \begin{enumerate}
@@
       compiler version, and dependency versions: a necessity
-      for handling the particulars of different architectures;
+      for handling the different architectures;
 \item Patching packages for particular platforms was
@@
 
-In addition to these immediate benefits, the ARES team mentioned
-two potential long-term payoffs. First, using Spack allowed
-the team to test with Clang.  This compiler is not currently used in
-production, but we expect it to be used in production on future LLNL machines.
-Testing with Clang revealed many incompatibilities, which were patched with
-Spack. The team is communicating these changes back to library developers,
-who will integrate them in future versions.  In this case, build automation
-has allowed more testing. This helps both ARES and LLNL library developers
-build more robust software.
-%
-Second, many of the libraries used by ARES are also used by other LLNL code
-teams.  The LLNL code teams have begun creating an internal repository of
-Spack build recipes to be shared.  Leveraging such a repository will
-make packaging the next code significantly easier.
+The ARES team also mentioned two potential long-term payoffs. First, Spack 
+allows the team to test with Clang.  This compiler is not currently used in
+production but probably will be on future LLNL machines. Testing with Clang 
+revealed many incompatibilities, which were patched with Spack. The team is 
+communicating these changes back to library developers, who will integrate 
+them in future versions. In this case, build automation has allowed more 
+testing, which helps both ARES and LLNL library developers build more robust 
+software. Second, other LLNL code teams use many libraries that ARES uses.
+The LLNL code teams have begun creating an internal repository of Spack 
+build recipes.  Leveraging this repository will make packaging the next 
+code significantly easier.

### 2015-07-21 implementation-concretization.tex
Tighten text.

@@
 spec be built.  Spack converts the spec to an abstract DAG.
-It then builds a {\it separate} spec DAG for any constraints that are encoded
-in directives in package files.
-
+It then builds a {\it separate} spec DAG for any constraints encoded
+by directives in package files.
+%
 Spack intersects the constraints of the two DAGs package by package, and it
@@
 the user inadvertently requests two versions of the same package, or if a
-package file explicitly requests a different version than the user requested.
+package file depends on a different version than the user requested.
 Likewise, if the package and the user specified different compilers, variants,
-or platforms for particular packages, this will cause the user to be notified
-of the conflict. If the user or package specified version ranges, they are
-intersected, and if the ranges do not overlap, an error will be raised.
-When the intersection succeeds, Spack has a single spec DAG with the merged
+or platforms for particular packages, Spack will stop and notify
+the user of the conflict. If the user or package specifies version ranges,
+they are intersected, and if the ranges do not overlap, an error is raised.
+When the intersection succeeds, Spack has a single DAG with the merged
 constraints of the user and the package files.
@@
 The next part of the process is iterative.
-If any node in the DAG is a virtual dependency, spack replaces it with with a
+If any node is a virtual dependency, spack replaces it with with a
 suitable interface provider.  It does this by building a reverse
 index from virtual packages to providers using the {\tt provides when}
-directives discussed in Section~\ref{sec:virtual}. If multiple providers
-could satisfy the constraints on the virtual spec,
-Spack will consult site and user policies to select the ``best'' possible
+directives (Section~\ref{sec:virtual}). If multiple providers
+satisfy the virtual spec's constraints,
+Spack consults site and user policies to select the ``best'' possible
 provider.  The selected provider may {\it itself} have virtual dependencies,
-so Spack repeats this process until there are no more virtual packages
-in the DAG.
+this process is repeated until there are no more virtual packages in the DAG.
 
@@
 \end{figure}
-When the concretization process completes, it outputs a fully concrete spec DAG.
+On completion, the concretization outputs a fully concrete spec DAG.
 Figure~\ref{fig:specs-mpileaks-concrete} shows a concrete DAG with architectures,
@@
 This fulfills Spack's guarantees.
-
+%
 At install time, Spack constructs a package object for each node in the spec DAG
@@
 {\tt install} method.  For the {\tt spec} parameter to {\tt install}, it passes
-a sub-DAG rooted at current node, which is also a concrete spec.  Package authors are
-responsible for querying the spec in {\tt install} and handling different configurations.
+a sub-DAG rooted at current node (also a concrete spec).  Package authors must
+query the spec in {\tt install} to handle different configurations.
 
@@
 Figure~\ref{fig:concretization-time} shows the running time of concretization vs.
-the size of a package DAG (in nodes).  We generated this plot by running {\tt concretize}
-on all 245 packages in our internal version of Spack.  We see that for all but the 9 largest
-packages, the algorithm takes less than 2 seconds on a Sandy Bridge cluster node, and less
-than 1 second on a Haswell MacBook Pro.  For larger DAGs, we begin to see superlinear
-scaling, but even for 50 nodes or more, concretization takes less than 6 seconds.
+package DAG size in nodes.  We generated this plot by running {\tt concretize}
+on all of Spack's 245 packages.  We see that for all but the 9 largest
+packages, concretization takes less than 2 seconds on a Sandy Bridge machine, 
+and less than 1 second on a Haswell MacBook Pro.  For larger DAGs, we begin
+to see superlinear scaling, but even for 50 nodes or more, concretization
+takes less than 6 seconds.
 %

### 2015-07-21 implementation-specs.tex
Install environment revisions.
Name change spack-paper-sc15 -> spack-sc15
Tighten text.

@@
 Spack calls a single build configuration a {\it spec}. The {\tt spec} object
-passed to {\tt install()} is how Spack communicates dependencies and parameters to package authors.
+passed to the {\tt install} method is how Spack communicates dependencies and parameters to package authors.
 
 \subsubsection{Structure}
-To understand how specs work, consider the {\tt mpileaks} package structure.
+To understand specs, consider the {\tt mpileaks} package structure.
 Metadata in the {\tt Mpileaks} class ({\tt version}, {\tt depends\_on}, etc.) describe
@@
 for each dependency and constructs a graph of their relationships.  The result
-is a directed, acyclic graph (DAG) of dependencies.\footnote{Spack currently disallows
+is a directed, acyclic graph (DAG).\footnote{Spack currently disallows
 circular dependencies.}
 %
-To guarantee a consistent build, and to avoid ABI compatibility issues, the DAG
-is constructed so that there is only {\it one} version of any particular package.  Note
+To guarantee a consistent build and to avoid ABI compatibility issues, we construct
+the DAG so that there is only {\it one} version of any package.  Note
 the distinction: while Spack can install arbitrarily many configurations of any package,
@@
 Here, {\tt mpileaks} is the simplest possible spec---a single identifier.
-Spack parses it and converts it into the DAG shown in Figure~\ref{fig:specs-mpileaks}.
+Spack converts it into the DAG shown in Figure~\ref{fig:specs-mpileaks}.
 Note that even though the spec contains no dependency information, it is still
 converted to a full DAG, based on directives supplied in package files. Since there
-are no constraints on the nodes, Spack has considerable leeway when building the
-package, and we say that the package is {\it unconstrained}.
+are no constraints on the nodes, Spack has leeway when building the
+package, and we say that it is {\it unconstrained}.
 %
-Now, suppose the user wants a specific version of {\tt mpileaks}.  This can be requested
-with a version constraint after the package name:
+Suppose instead that the user wants a specific {\tt mpileaks} version.
+She can request it with a version constraint after the package name:
 %
@@
 The caret (\verb|^|) denotes constraints for a particular dependency.  In the DAG,
-we see that there are now version constraints on {\tt callpath} {\it and} {\tt libelf},
-and that the user has requested the {\tt debug} variant of the {\tt callpath} library.
+there are now version constraints on {\tt callpath} {\it and} {\tt libelf},
+and the user has requested the {\tt debug} variant of the {\tt callpath} library.
 
@@
 Table~\ref{tab:specs} shows further examples of specs, ranging from very
-simple to more complex. From these examples, we can see that Spack offers constraint
-notation to cover the rest of the HPC package parameter space.
+simple to more complex. These examples show how Spack's constraint
+notation covers the rest of the HPC package parameter space.
 
 {\bf Versions.}
-Version constraints, denoted with {\tt @}, were already covered above. Versions
+Version constraints are denoted with {\tt @}. Versions
 can be precise ({\tt @2.5.1}) or denote a range ({\tt @2.5:4.4}),
 which may be open-ended ({\tt @2.5:}).
+%
+The package in Figure~\ref{fig:mpileaks} lists two ``safe'' versions with
+checksums, but
+in our experience users frequently want bleeding-edge versions, and package managers
+often lag behind the latest releases.
+Spack can extrapolate URLs from versions,
+using the package's {\tt url} attribute as a model.\footnote{This works
+for packages with consistently named URLs.}  If the user requests a specific
+version on the command line that is unknown to Spack,
+Spack will attempt to fetch and install it.  Spack also uses the same
+model to scrape webpages and find new versions as they become available.
 
-The package in Figure~\ref{fig:mpileaks} lists two ``safe'' versions with checksums, but
-in our experience users frequently want bleeding-edge versions, and package managers
-frequently lag behind the latest releases.
-Spack has a capability to extrapolate URLs from versions,
-using the package's {\tt url} attribute as a model.\footnote{This works
-for packages with consistently named URLs.}  The user can request a specific
-version on the command line, even if it is unknown to Spack,
-and Spack will attempt to install it.  Spack also uses the same
-model to scrape webpages and find new versions as they become available.
+\begin{figure}
+\begin{minted}[linenos,
+               numbersep=5pt,
+			   fontsize=\scriptsize,
+               frame=lines,
+               framesep=2mm]{python}
+    def install(self, spec, prefix):  # default build uses cmake
+        with working_dir('spack-build', create=True):
+            cmake('..', *std_cmake_args)
+            make()
+            make("install")
+
+    @when('@:8.1')                    # <= 8.1 uses autotools
+    def install(self, spec, prefix):
+        configure("--prefix=" + prefix)
+        make()
+        make("install")
+\end{minted}
+\caption{
+    Specialized {\tt install} method in Dyninst.
+	\label{fig:specialization}
+}
+\end{figure}
 
@@
 With a compiler constraint (shown on line 3) the user
-simply adds {\tt \%} followed by its name, along with an optional compiler version
-specifier.  Spack compiler names, e.g. {\tt gcc}, refer to the full compiler {\it toolchain},
-i.e. the C, C++, Fortran 77, and Fortran 90 compilers.  Spack can auto-detect
-compiler toolchains if they are in the user's {\tt PATH}, or they can be registered manually
+simply adds {\tt \%} followed by its name, with an optional compiler version
+specifier.  Spack compiler names like {\tt gcc} refer to the full compiler {\it toolchain},
+i.e. the C, C++, and Fortran 77 and 90 compilers.  Spack can auto-detect
+compiler toolchains in the user's {\tt PATH}, or they can be registered manually
 through a configuration file.
@@
 {\bf Variants.}
-To handle build options like compiler flags or optional components, specs can
+To handle options like compiler flags or optional components, specs can
 have named flags, or {\it variants}.  Variants are associated with the package,
-so the {\tt mpileaks} package implementor must check the spec and handle the cases
-where debug is enabled ({\tt +debug}) and disabled (with {\tt -debug}
-or {\tt \textasciitilde \ignorespaces debug}).  The names simplify the versioning and prevent
-Spack's configuration space from becoming too fine-grained.
-It would violate our goal of conciseness, for example, to include detailed
-compiler flags in spec syntax, but known sets of flags can simply be named.
+so the {\tt mpileaks} package implementor must explicitly handle cases
+where debug is enabled ({\tt +debug}) or disabled ({\tt -debug}
+or {\tt \textasciitilde \ignorespaces debug}).  Names simplify versioning
+and prevent the configuration space from becoming too large.
+For example, including detailed compiler flags in spec syntax
+would violate our goal of conciseness, but known sets of flags can simply be named.
 
 {\bf Cross-compilation.}
-To support cross-compilation, Spack includes the platform in the package spec (line 5).
+To support cross-compilation, specs includes the system architecture (line 5).
 Platforms begin with {\tt =} and take names like {\tt linux-ppc64} or {\tt bgq}.  They are
@@
 
-So far, we have shown examples of specs being used to request constraints from the
+So far, we have shown examples of specs used to request constraints from the
 command line, when {\tt spack install} is invoked.  However, the user is not the only
@@
 example, the ROSE compiler only builds with a certain version of the {\tt boost} library.
-At first glance, the {\tt depends\_on()} directives in Figure~\ref{fig:mpileaks} look
-like they take simple package names.  However, a package name is also a spec, and
+The {\tt depends\_on()} directives in Figure~\ref{fig:mpileaks} contain
+simple package names.  However, a package name is also a spec, and
 the same constraint syntax usable from the command line can be applied inside directives.
@@
 %
-\begin{figure}
-\begin{minted}[linenos,
-               numbersep=5pt,
-			   fontsize=\scriptsize,
-               frame=lines,
-               framesep=2mm]{python}
-    def install(self, spec, prefix):  # default build uses cmake
-        with working_dir('spack-build', create=True):
-            cmake('..', *std_cmake_args)
-            make()
-            make("install")
-
-    @when('@:8.1')                    # <= 8.1 uses autotools
-    def install(self, spec, prefix):
-        configure("--prefix=" + prefix)
-        make()
-        make("install")
-\end{minted}
-\caption{
-    Specialized {\tt install} method in Dyninst.
-	\label{fig:specialization}
-}
-\end{figure}
 

### 2015-07-23 usecase-ares.tex
Update ares table.

@@
 including MPI, BLAS, Python, and many other libraries.  Together, these 
-packages ares written in a diverse set of languages including C, C++, 
+packages are written in a diverse set of languages including C, C++, 
 Fortran, Python and tcl and uses MPI and OpenMP for parallelism.
@@
 
-\begin{table}\centering
-\begin{tabular}{|r|c|c|c|c|}
+\newcommand{\hfmt}[1]{\textit{\scriptsize #1}}
+\newcommand{\cfmt}[1]{\texttt{\scriptsize #1}}
+
+\begin{table}\centering % mvapich2 for intel 15
+\footnotesize
+\begin{tabular}{|r|c|c|c|c|c|}
 \hline
-\multirow{2}{*}{} & \multicolumn{2}{|c|}{\bf Linux} & {\bf BG/Q}     & {\bf Cray XE6} \\\cline{2-5}
-                  & {\it MVAPICH} & {\it OpenMPI}   & {\it BG/Q MPI} & {\it Cray MPI} \\\hline
-{\bf GCC}         & {\tt C P L D} &                 &                &                \\\hline
-{\bf Intel}       & {\tt C P L D} & {\tt C P L D}   &                &                \\\hline
-{\bf PGI}         & {\tt C P L D} &                 &                & {\tt C P L D}  \\\hline
-{\bf Clang}       & {\tt C P L D} &                 & {\tt C P L D}  &                \\\hline
-{\bf XL}          &               &                 & {\tt C P L D}  &                \\\hline
+\multirow{2}{*}{} & \multicolumn{3}{|c|}{\hfmt{Linux}}                      & {\hfmt{BG/Q}}     & {\hfmt{Cray XE6}} \\\cline{2-6}
+                  & {\hfmt{MVAPICH}} & {\hfmt{MVAPICH2}} & {\hfmt{OpenMPI}} & {\hfmt{BG/Q MPI}} & {\hfmt{Cray MPI}} \\\hline
+{\hfmt{GCC}}      & {\cfmt{C P L D}} &                   &                  &                   &                   \\\hline
+{\hfmt{Intel 14}} & {\cfmt{C P L D}} &                   &                  &                   &                   \\\hline
+{\hfmt{Intel 15}} & {\cfmt{C P L D}} & {\cfmt{D}}        &                  &                   &                   \\\hline
+{\hfmt{PGI}}      &                  & {\cfmt{D}}        & {\cfmt{C P L D}} &                   & {\cfmt{C L D}}    \\\hline
+{\hfmt{Clang}}    & {\cfmt{C P L D}} &                   &                  & {\cfmt{C L D}}    &                   \\\hline
+{\hfmt{XL}}       &                  &                   &                  & {\cfmt{C P L D}}  &                   \\\hline
 \end{tabular}

### 2015-07-23 usecase-ares.tex
update ARES text to go with table.

@@
 
-Table~\ref{tab:ares-configs} shows the configurations of ARES that we have packaged.
-The rows and columns show architectures, compilers, and MPI versions.
+Table~\ref{tab:ares-configs} shows configurations of ARES that the ARES
+team tests nightly.  The rows and columns show architectures, compilers, and MPI versions.
 The ARES Spack package supports four different code configurations:
-the current and previous production versions, a ``lite'' version that includes
-a smaller set of features and dependencies, and a development version.  We 
-denote these as C, P, L, and D. Each cell in the table indicates the
-ARES configurations built for an architecture, compiler, and MPI combination.
-Each configuration requires a slightly different set of dependencies and 
-dependency versions, but one common ARES package supports all of them with 
-conditional logic on versions and variants.
+the current (C) and previous (P) production versions, a ``lite'' version (L) that includes
+a smaller set of features and dependencies, and a development version (D).
+Each cell in the table indicates the ARES configurations built for an architecture,
+compiler, and MPI combination. Each configuration requires a slightly different
+set of dependencies and dependency versions, but one common ARES package supports
+all of them with conditional logic on versions and variants.
 
 Altogether, the initial packaging effort for ARES took roughly two months,
-half time, for an experienced build engineer. It supports 64 different builds
-(4 configurations and 8 architecture-compiler-MPI combinations). Prior to 
-using Spack, only the Linux/Intel configurations were automated. The ARES 
+half time, for an experienced build engineer. As shown in the table, 32
+different configurations have been run using Spack
+(some of 4 versions on each of 10 architecture-compiler-MPI combinations).
+Prior to using Spack, only Linux/Intel configurations were automated. The ARES 
 team listed a number of key features that enabled the increased automation:

### 2015-07-28 motivation.tex
Address shepherd comments through motivatoin.

@@
 are installed before packages that require them.
-Others, such as Anaconda~\cite{anaconda,conda}, use the same approach but
-are designed to run on top of a host OS.
+Anaconda~\cite{anaconda,conda}, uses the same approach but
+is designed to run on top of a host OS.
 These tools largely solve the problem of managing a {\it single} software
@@
 common to all applications on a system.
-These tools assume that each package only has a single version
-and most of the tools install packages in a single, inflexible location.
-To install multiple configurations, the user must create a custom, combinatorial
-naming scheme to avoid conflicts. They typically require root
+These tools assume that each package has only a single version,
+and most install packages in a single, inflexible location.
+To install multiple configurations, users must create custom, combinatorial
+naming schemes to avoid conflicts. They typically require root
 privileges and do not optimize for specific hardware.
@@
 for manual installs.
-ORNL uses hand installs but adheres to strict scripting conventions
+Oak Ridge National Laboratory (ORNL) uses hand installs but adheres
+to strict scripting conventions
 to reproduce each build~\cite{jones+:cug08}.
-TACC relies heavily on locally maintained RPMs.
+The Texas Advanced Computing Center (TACC) relies heavily on locally maintained RPMs.
 
@@
 {\tt LD\_LIBRARY\_PATH} (or a similar environment variable) so that the loader
-can find them.  The user is often not the same person who installed the binary,
-so determining what paths to add can be difficult, especially for inexperienced users.
+can find them.  Often, the user is not the same person who installed the library,
+and even advanced users may have difficulty determining which paths to add.
 
@@
 commands. Environment modules emerged in 1991, and there are many implementations~\cite{dotkit,furlani+:lisa91,furlani+:lisa96,mclay:lmod,mclay:lmod-tutorial}.
-The most advanced of these, such as Lmod~\cite{mclay:lmod,mclay:lmod-tutorial},
-provide software hierarchies similar to the naming conventions in
-Table~\ref{tab:naming-conventions}. They allow users to load a software stack
+The most advanced of these, Lmod~\cite{mclay:lmod,mclay:lmod-tutorial},
+provides software hierarchies similar to the naming conventions in
+Table~\ref{tab:naming-conventions}. This allows users to load a software stack
 quickly if they know which one is required.
@@
 path is called an {\tt RPATH}. {\tt RPATHs} and environment modules are not
-mutually exclusive. Modules can still be used to set {\tt MANPATH}, {\tt PATH},
-etc., while adding {\tt RPATHs} ensures that binaries run correctly
+mutually exclusive. Modules can still be used to set variables that are
+unrelated to linking, such as {\tt MANPATH} and {\tt PATH}.  Adding
+{\tt RPATHs} still ensures that binaries run correctly,
 regardless of whether the right module is loaded. LC installs software with
@@
 The EasyBuild~\cite{hoste+:pyhpc12} tool is in production use at
-the University of Ghent and several other sites.  It allows multiple versions
-to be installed at once.  Rather than setting {\tt RPATHs}, it
-auto-generates module files
-to manage each package's environment, and it is closely coupled with
+the University of Ghent and the J\"ulich Supercomputing Center.
+It allows multiple versions to be installed at once.  Rather
+than setting {\tt RPATHs}, it auto-generates module files
+to manage the environment, and it is closely coupled with
 Lmod~\cite{geimer+:hust14}.  EasyBuild groups the compiler, MPI, FFT, and
-BLAS libraries together in a {\it toolchain} that can be combined with
-any package file. This provides some degree of composability and
+BLAS libraries together in a {\it toolchain} that can be used by
+package files. This provides some degree of composability and
 separates compiler flags and MPI concerns from client packages.
@@
 HashDist~\cite{hashdist} is a meta-build system and package manager
-for HPC.  Of the existing solutions, it is the most similar to spack.
+for HPC.  Of the existing solutions, it is the most similar to Spack.
 Like Nix, it uses cryptographic versioning and stores installations in
@@
 The full set of package versions is combinatorial, and arbitrary combinations 
-of compiler, MPI version, build options, and dependency versions requires 
-tedious modifications of many package files. Indeed, the number of package 
+of compiler, MPI version, build options, and dependency versions require 
+non-trivial modifications to many package files. Indeed, the number of package 
 files required for most existing systems scales with the number of version 
@@
 %
-HPC sites need better mechanisms to {\it parameterize} packages so that new
+HPC sites need better ways to {\it parameterize} packages so that new
 builds can be {\it composed} in response to user needs.

### 2015-07-31 figs/concretization-overhead/concretization-overhead.py
New overhead figure.

@@ -23,7 +23,7 @@ for name in spack.db.all_package_names():
         end = time.clock()
         times.append(end - start)
 
-    record = (name, s.size, times)
+    record = (name, len([s for s in s.traverse()]), times)
     print record
     timings.append(record)
 

### 2015-07-31 figs/concretization-overhead/concretization-times.py

@@ -1,0 +1,43 @@
+#!/usr/bin/env python
+
+import yaml
+import matplotlib.pyplot as plt
+
+files = (
+#    ('macbook.yaml' , 'MacBook Pro, Haswell, 2.6Ghz'),
+    ('rzuseqlac2-timings.yaml'    , 'Linux, IBM Power7, 3.6Ghz'),
+    ('rzmerl156-timings.yaml'     , 'Linux, Intel Sandy Bridge, 2.6Ghz'),
+    ('rzoz14-timings.yaml'        , 'Linux, Intel Haswell, 2.3Ghz'),
+#    ('cab688-timings.yaml'       , 'Linux, Intel Sandy Bridge, 2.6Ghz'),
+#    ('catalyst160-timings.yaml'  , 'Linux, Intel Ivy Bridge, 2.4Ghz'),
+#    ('vulcanlac3-timings.yaml'   , 'Linux, IBM Power7, 3.6Ghz'),
+)
+
+
+styles = ['bd', 'r^', 'gv', 'bs', 'b*']
+
+#plt.axes().set_aspect(3)
+
+
+for i, (f, label) in enumerate(files):
+    with open(f) as stream:
+        timings = yaml.load(stream)
+
+    timings.sort(key=lambda x:int(x[1]))
+
+    timings = [(name, size, sum(t)/len(t)) for name, size, t in timings]
+    sizes = [t[1] for t in timings]
+    times = [t[2] for t in timings]
+
+    plt.plot(sizes, times, styles[i], label=label)
+
+plt.ylabel('Time (seconds)', fontsize=14)
+plt.xlabel('Package DAG size (nodes)', fontsize=14)
+l = plt.legend(loc=2, framealpha=0)
+l.get_frame().set_linewidth(0)
+
+plt.gca().set_xlim(right=55)
+
+plt.axes().set_aspect(.3 * 1/plt.axes().get_data_ratio())
+
+plt.savefig('concretization-times.pdf')

### 2015-07-31 figs/concretization-overhead/package-sizes.py

@@ -1,4 +1,4 @@
-#!/usr/bin/env python
+#!/usr/bin/env spack-python
 
 from external import yaml
 import matplotlib.pyplot as plt

### 2015-07-31 figs/concretization-overhead/timings.py

@@ -1,30 +1,0 @@
-#!/usr/bin/env python
-
-import yaml
-import matplotlib.pyplot as plt
-
-files = { 'macbook.yaml' : 'MacBook Pro, Haswell, 2.6Ghz',
-          'merl.yaml'    : 'Linux, Sandy Bridge, 2.6Ghz'}
-
-
-styles = ['bd', 'r^']
-
-plt.axes().set_aspect(3)
-
-for i, (f, label) in enumerate(files.items()):
-    with open(f) as stream:
-        timings = yaml.load(stream)
-
-    timings.sort(key=lambda x:int(x[1]))
-
-    timings = [(name, size, sum(t)/len(t)) for name, size, t in timings]
-    sizes = [t[1] for t in timings]
-    times = [t[2] for t in timings]
-
-    plt.plot(sizes, times, styles[i], label=label)
-
-plt.ylabel('Time (seconds)')
-plt.xlabel('Package DAG size (nodes)')
-plt.legend(loc=2)
-
-plt.savefig('concretization-times.pdf')

### 2015-07-31 figs/overhead/overhead.py

@@ -1,0 +1,139 @@
+#!/usr/bin/env python
+import sys
+import yaml
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.ticker import ScalarFormatter
+
+with open('overhead.yaml') as f:
+    data = yaml.load(f)
+
+N = len(data)
+keys = ['libelf', 'libpng', 'mpileaks',
+        'libdwarf', 'python', 'dyninst', 'lapack']
+
+#nfs_color='RoyalBlue'
+#wrap_color='Orange'
+#orig_color='SeaGreen'
+
+nfs_color='Black'
+wrap_color='Gray'
+orig_color='White'
+
+nfs_label='Wrappers, NFS'
+wrap_label='Wrappers, Temp FS'
+orig_label='No Wrappers, Temp FS'
+
+fsize = 14
+
+wrapper_means = []
+tmp_means = []       # same configuration as 'wrapper'
+no_wrap_means = []
+nfs_means = []
+for pkg in keys:
+    wrapper = [float(f) for f in data[pkg]['wrapper']]
+    tmp     = [] #[float(f) for f in data[pkg]['tmp']]
+    no_wrap = [float(f) for f in data[pkg]['no_wrapper']]
+    nfs     = [float(f) for f in data[pkg]['home']]
+
+    no_wrap_means.append(sum(no_wrap) / len(no_wrap))
+    nfs_means.append(sum(nfs) / len(nfs))
+    wrapper_means.append(
+        (sum(wrapper) + sum(tmp))
+        / (len(wrapper) + len(tmp)))
+
+wrap = np.array(wrapper_means)
+orig = np.array(no_wrap_means)
+nfs = np.array(nfs_means)
+
+nfs_overhead = (nfs - orig) / orig
+wrap_overhead = (wrap - orig) / orig
+
+ind = np.arange(N,dtype=float)  # the x locations for the groups
+ind += .1
+
+
+# ==============================================================
+# first figure: build times.
+# ==============================================================
+plt.figure()
+fig, ax = plt.subplots()
+width = 0.25       # the width of the bars
+
+nfs_rects     = ax.bar(ind,         nfs_means, width,     color=nfs_color)
+wrapper_rects = ax.bar(ind+width,   wrapper_means, width, color=wrap_color)
+no_wrap_rects = ax.bar(ind+2*width, no_wrap_means, width, color=orig_color)
+
+# add some text for labels, title and axes ticks
[…]

### 2015-07-31 figs/wrappers/wrapper-overhead.py

@@ -1,63 +1,0 @@
-#!/usr/bin/env python
-import sys
-import yaml
-import numpy as np
-import matplotlib.pyplot as plt
-
-with open('overhead.yaml') as f:
-    data = yaml.load(f)
-
-N = len(data)
-
-keys = sorted(data.keys())
-
-wrapper_means = []
-no_wrap_means = []
-for pkg in keys:
-    wrapper = [float(f) for f in data[pkg]['wrapper']]
-    no_wrap = [float(f) for f in data[pkg]['no_wrapper']]
-
-    wrapper_means.append(sum(wrapper) / len(wrapper))
-    no_wrap_means.append(sum(no_wrap) / len(no_wrap))
-
-wrap = np.array(wrapper_means)
-orig = np.array(no_wrap_means)
-overhead = (wrap - orig) / orig
-for k, z in zip(keys, overhead):
-    print k, overhead
-
-ind = np.arange(N,dtype=float)  # the x locations for the groups
-ind += .1
-width = 0.35       # the width of the bars
-
-fig, ax = plt.subplots()
-
-
-wrapper_rects = ax.bar(ind,       wrapper_means, width, color='r')
-no_wrap_rects = ax.bar(ind+width, no_wrap_means, width, color='b')
-
-# add some text for labels, title and axes ticks
-ax.set_xlabel('Libraries')
-ax.set_ylabel('Build time (s)')
-ax.set_xticks(ind+width)
-ax.set_xticklabels(keys)
-
-ax.legend( (wrapper_rects[0], no_wrap_rects[0]), ('Wrapper', 'No Wrapper') )
-
-def autolabel(rects):
-    # attach some text labels
-    for i, rect in enumerate(rects):
-        height = rect.get_height()
-        ax.text(rect.get_x()+rect.get_width()/2. + .1, 1.05*height,
-                #'%d'%int(height),
-                '%.1f%%' % (overhead[i] * 100),
-                ha='center', va='bottom')
-
-autolabel(wrapper_rects)
-#autolabel(no_wrap_rects)
-
-plt.gca().set_ylim(top=400)
-
-ax.set_aspect(.25 * 1/ax.get_data_ratio())
-
-plt.savefig('wrapper-overhead.pdf')

### 2015-07-31 usecase-ares.tex
Make figures color-agnostic

@@
 Figure~\ref{fig:ares} shows the DAG for the current production configuration 
-of ARES. At the top is ARES itself.  ARES depends on 11 LLNL physics packages 
-(red), 4 LLNL math/meshing libraries (gold), and 8 LLNL utility libraries 
-(green). The utility libraries handle tasks including logging, I/O, and 
+of ARES. At the top is ARES itself.  ARES depends on 11 LLNL physics packages,
+4 LLNL math/meshing libraries, and 8 LLNL utility libraries.
+The utility libraries handle tasks including logging, I/O, and 
 performance measurement. ARES also uses 23 external software packages, 
@@
 
-Altogether, the initial packaging effort for ARES took roughly two months,
-half time, for an experienced build engineer. As shown in the table, 32
-different configurations have been run using Spack
+Altogether, the initial packaging effort required roughly two months 
+for an experienced build engineer working 20 hours per week.
+As shown in the table, 32 different configurations have been run using Spack
 (some of 4 versions on each of 10 architecture-compiler-MPI combinations).
@@
 \item The spec syntax allowed build scripts to concisely test compiler,
-      compiler version, and dependency versions: a necessity
+      compiler version, and dependency versions---a necessity
       for handling the different architectures;

### 2015-07-31 figs/overhead/overhead.py

@@ -12,13 +12,9 @@ N = len(data)
 keys = ['libelf', 'libpng', 'mpileaks',
         'libdwarf', 'python', 'dyninst', 'lapack']
 
-#nfs_color='RoyalBlue'
-#wrap_color='Orange'
-#orig_color='SeaGreen'
-
-nfs_color='Black'
-wrap_color='Gray'
-orig_color='White'
+nfs_color='RoyalBlue'
+wrap_color='Orange'
+orig_color='SeaGreen'
 
 nfs_label='Wrappers, NFS'
 wrap_label='Wrappers, Temp FS'

### 2015-07-31 implementation-concretization.tex
Revisions and better overhead plot.

@@
 The next part of the process is iterative.
-If any node is a virtual dependency, spack replaces it with with a
+If any node is a virtual dependency, Spack replaces it with with a
 suitable interface provider.  It does this by building a reverse
@@
 and less than 1 second on a Haswell MacBook Pro.  For larger DAGs, we begin
-to see superlinear scaling, but even for 50 nodes or more, concretization
+to see a quadratic trend, but even for 50 nodes or more, concretization
 takes less than 6 seconds.
@@
 satisfaction problem, which is NP-complete. Our greedy solver is sufficient for the
-packages we have built so far, and its running time is currently negligible. We do
-not expect to see software packages with thousands of dependencies in the near
+packages we have built so far.  Its running time is insignificant
+compared to the build time of most packages, and it only executes
+once per build.
+We do not expect to see software packages with thousands of dependencies in the near
 future, and we do not expect concretization to become a bottleneck, even if we

### 2015-07-31 implementation-environment.tex
Tighten text.
Revisions and better overhead plot.

--- 2015-07-19 implementation-environment.tex
@@
 library. One is distributed with RedHat Linux, while the
-other is publicly available, and they have the same API but an incompatible ABI.
+other is publicly available. They have the same API but an incompatible ABI.
 Specifying the wrong version at build time has caused many
@@
 Spack manages the build environment by running each call to {\tt install}
-in its own process.  In this environment, Spack sets
+in its own process.  It also sets
 {\tt PATH}, {\tt PKG\_CONFIG\_PATH}, {\tt CMAKE\_PREFIX\_PATH}, and
 {\tt LD\_LIBRARY\_PATH} to include the dependencies of the current build.
-These variables are commonly used by build systems to locate dependencies,
-and setting them helps to ensure that the correct libraries are detected by
-build systems.
-The isolated build environment also gives package authors 
+Build systems commonly use these variables to locate dependencies,
+and setting them helps to ensure that the correct libraries are detected.
+Using a separate process also gives package authors 
 free reign to set build-specific environment variables without interfering
@@
 for binaries to be able to find dependency libraries at {\it runtime}.
-One of the most common user errors at LC is improper library configuration.
+One of the most common user errors at LLNL is improper library configuration.
 Users frequently do not know what libraries a package was built with, and
@@
 
+\begin{figure}%{\linewidth}
+	\centering
+	\includegraphics[width=\columnwidth]{figs/overhead/build-time.pdf}
+	\caption{
+		Build time on NFS and temp, with and without wrappers.
+		\label{fig:build-times}
+	}
+\end{figure}
+\begin{figure}%{\linewidth}
+	\centering
+	\includegraphics[width=\columnwidth]{figs/overhead/overhead.pdf}
+	\caption{
+		Build overhead of using NFS and compiler wrappers.
+		\label{fig:overhead}`
+	}
+\end{figure}
+
+\subsubsection{Build Overhead}
+
+Argument parsing and indirection cause Spack's compiler wrappers to incur a small
+but noticeable performance overhead.  Figure~\ref{fig:build-time} shows
+build times with and without compiler wrappers for seven different packages,
+and Figure~\ref{fig:overhead} shows the corresponding overhead as a percentage
+of the wrapper-less runtime.  {\tt Dyninst} and {\tt lapack}
+use CMake to build, and the rest use autotools.  We ran
+each build three times on an Intel Sandy Bridge cluster node, and we
+report the average here.
+In the case of {\tt dyninst}, the overhead is negligible (-0.4\%),
+but it is as high as 12.3\% for shorter builds, like
+{\tt mpileaks}
+
+To increase build speed, by default Spack runs each {\tt install}
+method in the default temporary directory. On most cluster nodes,
+this is a fast, locally-mounted filesystem. In our experience,
+many users build manually in their home directories out of habit.  At most sites, home
+directories are remotely mounted volumes (e.g., NFS).  To compare Spack builds
+with {\it typical} manual builds, we timed the same seven builds using
+NFS.  Figure~\ref{fig:overhead} shows that building this way can be as much
+as 62.7\% slower than using a temporary filesystem and
+33\% slower on average. We therefore expect many users to notice
+a speedup when Spack uses temporary space to build.
+
 \subsubsection{Environment Module Integration}

### 2015-07-31 implementation-packages.tex
Name change spack-paper-sc15 -> spack-sc15
Revisions and better overhead plot.

@@
 at LLNL to find un-released MPI handle objects in parallel programs.
-The {\tt MpiLeaks} class first provides simple metadata: a text description,
-a homepage, and a download URL.
-Next, two {\tt version} directives identify known versions and provide
-MD5 checksums used to verify downloads.
-Below this, two {\tt depends\_on}
-directives indicate prerequisite packages that must be installed before \mpileaks.
-Last, the {\tt install} method contains the commands used to build
-the tool.  Spack's DSL allows shell commands to be invoked as Python functions,
+The {\tt MpiLeaks} class provides simple metadata on lines 4-8: a text
+description, a homepage, and a download URL.
+Two {\tt version} directives on lines 10-11 identify known versions and provide
+MD5 checksums to verify downloads.
+The two {\tt depends\_on} directives on lines 13-14 indicate prerequisite
+packages that must be installed before \mpileaks.
+Last, the {\tt install} method on line 16 contains the commands for building.
+Spack's DSL allows shell commands to be invoked as Python functions,
 and the {\tt install} method invokes {\tt configure},
-{\tt make}, and {\tt make install} commands as a shell script would.
+{\tt make}, and {\tt make install} as a shell script would.

### 2015-07-31 implementation-specs.tex
Revisions and better overhead plot.

@@
 different ways, according to a set of {\it parameters}.
-Spack calls a single build configuration a {\it spec}. The {\tt spec} object
-passed to the {\tt install} method is how Spack communicates dependencies and parameters to package authors.
+Spack calls a single build configuration a {\it spec}. 
+Spack communicates dependencies and parameters to package authors using
+the {\tt spec} argument of the {\tt install} method.
 
@@
 %
-To guarantee a consistent build and to avoid ABI compatibility issues, we construct
-the DAG so that there is only {\it one} version of any package.  Note
-the distinction: while Spack can install arbitrarily many configurations of any package,
+To guarantee a consistent build and to avoid 
+Application Binary Interface (ABI) incompatibility, we construct
+the DAG with only {\it one} version of each package.  Thus, while
+Spack can install arbitrarily many configurations of any package,
 no two configurations of the same package will ever appear in the same build DAG.
@@
 Recall that Spack guarantees that there is only a single version of any package in
-the spec DAG.  Within the same DAG, each dependency can be uniquely
-identified by its package name alone.  Therefore, the user need not think about DAG
-connectivity to add constraints.  She need only know that {\tt mpileaks}
+the spec DAG.  Within a spec, each dependency can be uniquely
+identified by its package name alone.  Therefore, the user does not need to consider 
+DAG connectivity to add constraints.  She need only know that {\tt mpileaks}
 somehow depends on {\tt callpath}.
@@
 and it is often desirable to write constraints like this into a package file.  For
-example, the ROSE compiler only builds with a certain version of the {\tt boost} library.
+example, the ROSE compiler~\cite{rose} only builds with a certain version of the {\tt boost} library.
 The {\tt depends\_on()} directives in Figure~\ref{fig:mpileaks} contain

### 2015-07-31 implementation.tex
Revisions and better overhead plot.

@@
 Based on our experiences at LLNL, we have developed
-{\it Spack}, the Supercomputing PACKage manager.
-Spack is written in Python.  We chose Python for its flexibility
-and its wide use in the HPC community.
+{\it Spack}, the \underline{S}upercomputing \underline{Pack}age manager.
+Spack is written in Python, which we chose for its flexibility
+and its increasing use in HPC.
 %
@@
 \begin{enumerate}
-\item Packages explicitly {\bf parameterized} by version, platform,
-      compiler, options, and dependencies, to allow easy composition.
+\item Composable packages, explicitly {\bf parameterized} by version, platform,
+      compiler, options, and dependencies.
 \item A novel, recursive {\bf spec syntax} for dependency graphs and constraints,
@@
       specification into a full, concrete build specification.
-\item A build environment with {\bf compiler wrappers} that enforce build
+\item A build environment that uses {\bf compiler wrappers} to enforce build
       consistency and simplify package writing.

### 2015-07-31 introduction.tex
Revisions and better overhead plot.

@@
 \item Spack: An implementation of these concepts; and
-\item Four use cases that detail LLNL's use of Spack to comcpose
-      software configurations rapidly, to manage Python installations, and
-      to implement site build policies, and to build an LLNL production code.
+\item Four use cases detailing LLNL's use of Spack to compose
+      software configurations rapidly, to manage Python installations,
+      to implement site build policies, and to build an LLNL production
+      code with 47 dependencies in 32 configurations.
 \end{enumerate}
 
-\noindent Spack is in active development at LLNL, and our use cases outline some of
-its first applications in our environment.  Spack solves software problems
-that are pervasive at large, multi-user HPC centers, and our experiences
-are relevant to the full range of HPC facilities.  Spack improves
+\noindent Spack is in active development at LLNL, and our use cases
+outline some of its first applications in our environment.  
+Spack solves software problems that are pervasive at large,
+multi-user HPC centers, and our experiences are relevant to the 
+full range of HPC facilities.  Spack improves
 operational efficiency by simplifying the build and deployment of

### 2015-07-31 limitations.tex
Revisions and better overhead plot.

@@
 
-Spack can require more disk space than a module-based system, as otherwise 
-identical packages with different dependencies must be built separately. In 
-our view, the significant reduction in complexity for the end user justifies 
-this cost.
+Spack requires more disk space than a module-based system, as otherwise 
+identical packages with different dependencies must be built separately. 
+The exact space overhead depends on the structure of the installed
+software; some builds can share more dependency libraries than others
+(see Figure~\ref{fig:reuse}).
+In our view, the significant reduction in complexity for the end user
+justifies this cost.
 
@@
 Spack packages to suit their needs. Such customizations frequently
-guide our priorities for new core features. The
-benefits of using a popular, rapidly emerging HPC language for our
-embedded DSL outweigh any costs of the learning curve.
+guide our priorities for new core features. Some users have found the 
+learning curve to be too steep, but this only prevents them from 
+packaging software. They still use Spack as a command-line tool.
 
 In the near term, we are actively working to grow a community around Spack 
-and to build a larger base of contributors. On the technical side, we will
+and to build a larger base of contributors. We plan to
 increase build robustness by deploying a continuous testing system
@@
 capabilities to Spack that allow packages to depend on particular
-compiler features.  More and more, our codes are beginning to rely
+compiler features.  More and more, our codes are relying
 on advanced compiler capabilities, like {\tt C++11} language features,

### 2015-07-31 usecase-python.tex
Name change spack-paper-sc15 -> spack-sc15
Revisions and better overhead plot.

@@
 %
-LC supports Python installations for several application teams,
-and maintaining these repositories has grown increasingly complex over
+LC supports Python installations for several application teams.
+Maintaining these repositories has grown increasingly complex over
 time. The problems are similar to those that drove us to create
@@
 
-Existing Python package managers are either language-specific~\cite{eby:setuptools}, 
-or they do not support building from source~\cite{anaconda,conda}. None
-handles multi-configuration builds.  More glaringly, Python extensions
+Existing Python package managers either do not support building from
+source~\cite{anaconda,conda}, or they are language-specific~\cite{eby:setuptools}.
+None handles multi-configuration builds.  More glaringly, Python extensions
 are usually installed into the Python interpreter's prefix.
@@
 and each module depends on a particular Python installation.
-But extensions can be {\tt activated} or {\tt deactivated} an
+But extensions can be {\tt activated} or {\tt deactivated}
 within the dependent Python installation.  The {\tt activate} operation

### 2015-07-31 usecases.tex
edits to usecases
Change name of tex file to 'spack-paper-sc15'
combinatoric -> combinatorial
Add ARES dependency figure.
Name change spack-paper-sc15 -> spack-sc15
Revisions and better overhead plot.

@@
 
-We have recently using Spack in production at 
-Livermore Computing.  We have already outlined a number of the
+We have recently begun using Spack in production at
+Livermore Computing (LC).  We have already outlined a number of the
 experiences that drove us to develop Spack.
 In this section, we describe real-world use cases that Spack
-has been used to address.  In some cases, Spack had to be 
+has addressed.  In some cases, Spack had to be
 adapted to meet production needs, and we describe how its
@@
 
-\input{usecase-combinatoric}
+\input{usecase-combinatorial}
 \input{usecase-python}
 \input{usecase-policy}
+\input{usecase-ares}

### 2015-07-31 implementation-concretization.tex
New concretization textx

@@
 package DAG size in nodes.  We generated this plot by running {\tt concretize}
-on all of Spack's 245 packages.  We see that for all but the 9 largest
-packages, concretization takes less than 2 seconds on a Sandy Bridge machine, 
-and less than 1 second on a Haswell MacBook Pro.  For larger DAGs, we begin
-to see a quadratic trend, but even for 50 nodes or more, concretization
-takes less than 6 seconds.
+on all of Spack's 245 packages. We tested on three LLNL cluster front-end nodes:
+a 2.3Ghz Intel Haswell node and a 2.6 GHz Intel Sandy Bridge node on our Linux
+clusters, and the 3.6 GHz IBM Power7 front-end node of a Blue Gene/Q system.
+Each point is the average of 10 trials.  We see that for all but the 10
+largest packages, concretization takes less than 2 seconds on all machines tested.
+For larger DAGs, we begin to see a quadratic trend, but even for 50 nodes or more, 
+concretization takes less than 4 seconds on the Haswell machine and 9 seconds on the 
+Power7.
 %
-This performance is expected from a greedy, fixed-point method.
-The algorithm repeats, evaluating conditions and adding dependencies and constraints, until
-there are no more changes to be made.
-%
-%  Additionally, our current compiler concretization
-% policy requires a search in the DAG; it prefers the compiler of the nearest ancestor
-% if the current node's compiler is not set. 
-%
+This performance is expected from our greedy, fixed-point method, and
+it is sufficient for the packages we have built so far.
+Its running time is insignificant compared to the build time of
+most packages, and it only executes
+once per build. 
 More generally, concretization is an instance of the constraint
-satisfaction problem, which is NP-complete. Our greedy solver is sufficient for the
-packages we have built so far.  Its running time is insignificant
-compared to the build time of most packages, and it only executes
-once per build.
-We do not expect to see software packages with thousands of dependencies in the near
-future, and we do not expect concretization to become a bottleneck, even if we
-use a full constraint solver.
+satisfaction problem, which is NP-complete, so concretization could become
+more costly.  However, we do not expect to see software packages with
+thousands of dependencies in the near future, so we do not expect
+concretization to become a bottleneck, even if we use a full constraint solver.
 

### 2015-07-31 introduction.tex
Content complete based on revisions.

--- 2015-07-20 introduction.tex
@@
 The simulation software that runs on these machines is very complex; some
-codes depend on specific versions of over 50 dependency libraries.
+codes depend on specific versions of over 70 dependency libraries.
 They require specific compilers, build options and MPI implementations to
@@
 typically require that package files be created for each combination of
-options~ \cite{digirolamo:smithy,dolstra+:icfp08,dolstra+:lisa04,hashdist,hoste+:pyhpc12},
+options~ \cite{digirolamo:smithy,dolstra+:icfp08,dolstra+:lisa04,hoste+:pyhpc12},
 leading to a profusion of files and maintenance issues.
@@
 \begin{enumerate}
-\item A novel, recursive syntax for concisely specifying and querying
+\item A novel, recursive syntax for concisely specifying constraints within
       the large parameter space of HPC packages;
@@
 \item Spack: An implementation of these concepts; and
-\item Four use cases that detail LLNL's use of Spack to comcpose
-      software configurations rapidly, to manage Python installations, and
-      to implement site build policies, and to build an LLNL production code.
+\item Four use cases detailing LLNL's use of Spack in production.
 \end{enumerate}
+\noindent
+Spack is in active development at LLNL, and our four use cases
+highlight its ability to manage complex software.
+%
+Spack can rapidly compose new package configurations,
+manage Python installations, and implement site-specific build policies. 
+%
+It also automates 36 different build configurations of an
+LLNL production code with 46 dependencies, across three platforms.  Despite
+this complexity, we show that Spack's {\it concretization} algorithm
+for managing
+dependency constraints runs in seconds, even for large packages.
+We further show that Spack's install environment incurs around
+10\% build-time overhead compared to a native install.
 
-\noindent Spack is in active development at LLNL, and our use cases outline some of
-its first applications in our environment.  Spack solves software problems
-that are pervasive at large, multi-user HPC centers, and our experiences
-are relevant to the full range of HPC facilities.  Spack improves
+Spack solves software problems that are pervasive at large,
+multi-user HPC centers, and our experiences are relevant to the 
+full range of HPC facilities.  Spack improves
 operational efficiency by simplifying the build and deployment of

### 2015-07-31 usecase-ares.tex
Content complete based on revisions.

@@
                   & {\hfmt{MVAPICH}} & {\hfmt{MVAPICH2}} & {\hfmt{OpenMPI}} & {\hfmt{BG/Q MPI}} & {\hfmt{Cray MPI}} \\\hline
-{\hfmt{GCC}}      & {\cfmt{C P L D}} &                   &                  &                   &                   \\\hline
+{\hfmt{GCC}}      & {\cfmt{C P L D}} &                   &                  & {\cfmt{C P L D}}  &                   \\\hline
 {\hfmt{Intel 14}} & {\cfmt{C P L D}} &                   &                  &                   &                   \\\hline
@@
 for an experienced build engineer working 20 hours per week.
-As shown in the table, 32 different configurations have been run using Spack
+As shown in the table, 36 different configurations have been run using Spack
 (some of 4 versions on each of 10 architecture-compiler-MPI combinations).

### 2015-07-31 introduction.tex
tighten new intro.

@@
 codes depend on specific versions of over 70 dependency libraries.
-They require specific compilers, build options and MPI implementations to
+They require specific compilers, build options and Message Passing Interface (MPI)
+implementations to
 achieve the best performance, and users may run several
@@
 To support the diverse needs of applications, system administrators
-and application developers frequently build, install, and support many
-different configurations of math and physics libraries, as well as
+and developers frequently build, install, and support many
+different configurations of math and physics libraries, and
 other software.  Frequently, applications must be rebuilt to fix bugs
-and to support new versions of the operating system (OS), Message
-Passing Interface (MPI) implementation, compiler, and other
+and to support new versions of the operating system (OS), 
+MPI implementation, compiler, and other
 dependencies.  Unfortunately, building scientific software is
@@
 process~\cite{bsdports,digirolamo:smithy,dolstra+:icfp08,dolstra+:lisa04,hashdist,homebrew,hoste+:pyhpc12,macports,thiruvathukal:gentoo04}.
-For the most part, these systems focus on keeping a single, stable set of
-packages up to date, and they do not handle installation of multiple
-versions or configurations.  Those that {\it do} handle multiple configurations
+For the most part, they focus on keeping a single, stable set of
+packages up to date, and they do not handle installation of multiple versions or
+configurations.  Those that {\it do} handle multiple configurations
 typically require that package files be created for each combination of
@@
 leading to a profusion of files and maintenance issues.
-To our knowledge, no existing tool allows new configurations to be composed
-and assembled on demand.  Some allow limited forms of
+Some allow limited forms of 
 composition~\cite{hoste+:pyhpc12,dolstra+:icfp08,dolstra+:lisa04}, but their
@@
 \noindent
-Spack is in active development at LLNL, and our four use cases
-highlight its ability to manage complex software.
+Our use cases highlight Spack's ability to manage complex software.
 %
-Spack can rapidly compose new package configurations,
+Spack allows users to rapidly compose new package configurations,
 manage Python installations, and implement site-specific build policies. 
 %
-It also automates 36 different build configurations of an
-LLNL production code with 46 dependencies, across three platforms.  Despite
-this complexity, we show that Spack's {\it concretization} algorithm
-for managing
-dependency constraints runs in seconds, even for large packages.
-We further show that Spack's install environment incurs around
+It automates 36 different build configurations of an LLNL
+production code with 46 dependencies.
+Despite this complexity, Spack's {\it concretization} algorithm
+for managing constraints runs in seconds, even for
+large packages. Spack's install environment incurs only around
 10\% build-time overhead compared to a native install.
@@
 multi-user HPC centers, and our experiences are relevant to the
-full range of HPC facilities.  Spack improves
-operational efficiency by simplifying the build and deployment of
-bleeding-edge scientific software.
+full range of HPC facilities.
+Spack improves operational efficiency by simplifying the
+build and deployment of bleeding-edge scientific software.

### 2015-07-31 implementation-concretization.tex
Bibliography formatting, tighten to 12 pages.

@@
 description to a concrete build.
-
-\begin{figure*}
-	\centering
-	\includegraphics[width=\textwidth]{figs/concretization.pdf}
-	\caption{
-		Spack's concretization process.
-		\label{fig:concretization}
-	}
-\end{figure*}
 
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figs/concretization-overhead/concretization-times.pdf}
+	\includegraphics[width=.95\columnwidth]{figs/concretization-overhead/concretization-times.pdf}
 	\caption{
@@
 clusters, and the 3.6 GHz IBM Power7 front-end node of a Blue Gene/Q system.
-Each point is the average of 10 trials.  We see that for all but the 10
-largest packages, concretization takes less than 2 seconds on all machines tested.
+Each point is the average of 10 trials.  For all but the 10
+largest packages, concretization takes less than 2 seconds on all machines.
 For larger DAGs, we begin to see a quadratic trend, but even for 50 nodes or more, 
@@
 %
-This performance is expected from our greedy, fixed-point method, and
+We expect this performance from our greedy, fixed-point method, and
 it is sufficient for the packages we have built so far.
@@
 More generally, concretization is an instance of the constraint
-satisfaction problem, which is NP-complete, so concretization could become
-more costly.  However, we do not expect to see software packages with
-thousands of dependencies in the near future, so we do not expect
-concretization to become a bottleneck, even if we use a full constraint solver.
+satisfaction problem, which is NP-complete. While concretization could become
+more costly, we do not expect to see packages with
+thousands of dependencies in the near future. We do not expect
+it to become a bottleneck, even if we use a full constraint solver.
 
@@
 \label{sec:directory-layout}
-
-\begin{figure}\centering
-   \includegraphics[width=\linewidth]{specs/rpaths.pdf}
-   \caption{
-       {\tt mpileaks} built with {\tt mpich}, then {\tt openmpi}.
-       \label{fig:reuse}
-   }
-\end{figure}
 
@@
 separates these concerns.
+
+\begin{figure}\centering
+   \includegraphics[width=.9\linewidth]{specs/rpaths.pdf}
+   \caption{
+       {\tt mpileaks} built with {\tt mpich}, then {\tt openmpi}.
+       \label{fig:reuse}
+   }
+\end{figure}

### 2015-07-31 implementation-environment.tex
Bibliography formatting, tighten to 12 pages.

@@
 
-Spack is designed to build a consistent HPC stack for our multi-user
+Spack is designed to build a consistent HPC stack for our
 environment, and reproducible builds are one of our design goals.
@@
 %
-Many packages used at LLNL have a profusion of build options, and specifying
+Many packages have a profusion of build options, and specifying
 them correctly often requires tedious experimentation.  This is due to lack of
@@
 separated by a comma.
-The {\tt install} method in Spack's package files encodes precise build
+The {\tt install} method in Spack's packages encodes precise build
 incantations for later reuse.
@@
 it is difficult for them to construct a suitable {\tt LD\_LIBRARY\_PATH} for
-a package that was built by someone else.  To reduce the number of support calls,
+a package that was built by someone else.  To reduce the number of support calls
+we receive,
 we typically add {\tt RPATHs} to public software installations, so that paths
@@
 
-Spack manages {\tt RPATH} settings and other build policies with
+Spack manages {\tt RPATHs} and other build policies with
 {\it compiler wrappers}.
-In each isolated {\tt install} environment, Spack sets the standard
-environment variables
-{\tt CC}, {\tt CXX}, {\tt F77}, and {\tt FC} to point to its own compiler
+In each isolated {\tt install} environment, Spack sets the environment variables
+{\tt CC}, {\tt CXX}, {\tt F77}, and {\tt FC} to point to its compiler
 wrapper scripts.  These variables are used by most build systems to select 
-C, C++, and Fortran compilers, so they are generally picked up
-automatically.\footnote{If builds do not respect {\tt CC}, {\tt CXX}, etc.,
+C, C++, and Fortran compilers, so they are typically picked up
+automatically.\footnote{If builds do not respect {\tt CC}, {\tt CXX}, {\tt FC},
+or {\tt F77},
 wrappers can be added as arguments or inserted into Makefiles
@@
 These point to the {\tt include} and {\tt lib} directories of dependency
-library installations, where needed headers and libraries are located.
-The wrappers then invoke the real compiler with the modified arguments.
+installations, where needed headers and libraries are located.
+The wrappers invoke the real compiler with the modified arguments.
 
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figs/overhead/build-time.pdf}
+	\includegraphics[width=.95\columnwidth]{figs/overhead/build-time.pdf}
 	\caption{
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figs/overhead/overhead.pdf}
+	\includegraphics[width=.95\columnwidth]{figs/overhead/overhead.pdf}
 	\caption{
@@
 but it is as high as 12.3\% for shorter builds, like
-{\tt mpileaks}
+{\tt mpileaks}.
 

### 2015-07-31 implementation-packages.tex
Bibliography formatting, tighten to 12 pages.

@@
                framesep=2mm]{python}
-from spack import *
-
 class Mpileaks(Package):
-    """Tool to detect and report leaked MPI objects like
-       MPI_Requests and MPI_Datatypes."""
+    """Tool to detect and report leaked MPI objects."""
 
@@
 at LLNL to find un-released MPI handle objects in parallel programs.
-The {\tt MpiLeaks} class provides simple metadata on lines 4-8: a text
+The {\tt MpiLeaks} class provides simple metadata on lines 2-5: a text
 description, a homepage, and a download URL.
-Two {\tt version} directives on lines 10-11 identify known versions and provide
+Two {\tt version} directives on lines 7-8 identify known versions and provide
 MD5 checksums to verify downloads.
-The two {\tt depends\_on} directives on lines 13-14 indicate prerequisite
+The two {\tt depends\_on} directives on lines 10-11 indicate prerequisite
 packages that must be installed before \mpileaks.
-Last, the {\tt install} method on line 16 contains the commands for building.
+Last, the {\tt install} method on line 13 contains the commands for building.
 Spack's DSL allows shell commands to be invoked as Python functions,

### 2015-07-31 implementation-specs.tex
Bibliography formatting, tighten to 12 pages.

@@
 		\centering
-		\includegraphics[width=\columnwidth]{specs/mpileaks.pdf}
+		\includegraphics[width=.9\columnwidth]{specs/mpileaks.pdf}
 		\caption{
@@
 		\centering
-		\includegraphics[width=\columnwidth]{specs/mpileaks-version}
+		\includegraphics[width=.9\columnwidth]{specs/mpileaks-version}
 		\caption{
@@
 		\centering
-		\includegraphics[width=\columnwidth]{specs/mpileaks-abstract.pdf}
+		\includegraphics[width=.9\columnwidth]{specs/mpileaks-abstract.pdf}
 		\caption{
@@
 placed on the {\tt mpileaks} node in the DAG, but the rest of the DAG remains unconstrained.
-If the user does not need a specific version of {\tt mpileaks}, but does require
-particular minimum version, then the user could use {\it version range} syntax
-and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would use
-{\tt @2.3:2.5.6} to designate the range. In these cases, the user can save time if
+If the user only requires a particular minimum version, then she could use
+{\it version range} syntax
+and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would write
+{\tt @2.3:2.5.6}. In these cases, the user can save time if
 Spack already has a version installed that satisfies the spec -- Spack will use
@@
 Recall that Spack guarantees that there is only a single version of any package in
-the spec DAG.  Within a spec, each dependency can be uniquely
+a spec.  Within a spec, each dependency can be uniquely
 identified by its package name alone.  Therefore, the user does not need to consider 
-DAG connectivity to add constraints.  She need only know that {\tt mpileaks}
+DAG connectivity to add constraints.  She must only know that {\tt mpileaks}
 somehow depends on {\tt callpath}.
-For the same reason, the constraint order does not matter; dependency constraints
-can appear in arbitrary order.
+Thus, dependency constraints can appear in arbitrary order.
 
@@
 the same constraint syntax usable from the command line can be applied inside directives.
-So, the ROSE compiler can simply write:
+So, we can simply write:
 %
@@
 %
+\vfill\eject
+\noindent
 The same notation is used to optionally build with libraries like MPI:
@@
           # Handle XL compilers
-      ...
 \end{minted}

### 2015-07-31 conclusion.tex
Revisions and better overhead plot.
Typos.

@@
 process, which converts an abstract build DAG into a concrete,
-buildable spec.
+build-able spec.
 
@@
 to a broad range of HPC facilities.
-Spack is available online at http://github.com/scalability-llnl/spack.
+Spack is available online at \url{http://github.com/scalability-llnl/spack}.

### 2015-07-31 implementation-concretization.tex
Typos.

@@
 on all of Spack's 245 packages. We tested on three LLNL cluster front-end nodes:
-a 2.3Ghz Intel Haswell node and a 2.6 GHz Intel Sandy Bridge node on our Linux
+a 2.3GHz Intel Haswell node and a 2.6 GHz Intel Sandy Bridge node on our Linux
 clusters, and the 3.6 GHz IBM Power7 front-end node of a Blue Gene/Q system.

### 2015-07-31 implementation-environment.tex
Changes from Greg.

@@
 Argument parsing and indirection cause Spack's compiler wrappers to incur a small
-but noticeable performance overhead.  Figure~\ref{fig:build-time} shows
+but noticeable performance overhead.  Figure~\ref{fig:build-times} shows
 build times with and without compiler wrappers for seven different packages,
 and Figure~\ref{fig:overhead} shows the corresponding overhead as a percentage
-of the wrapper-less runtime.  {\tt Dyninst} and {\tt lapack}
+of the wrapper-less runtime.  {\tt Dyninst} and {\tt LAPACK}
 use CMake to build, and the rest use autotools.  We ran
@@
 method in the default temporary directory. On most cluster nodes,
-this is a fast, locally-mounted filesystem. In our experience,
+this is a fast, locally-mounted file system. In our experience,
 many users build manually in their home directories out of habit.  At most sites, home
@@
 NFS.  Figure~\ref{fig:overhead} shows that building this way can be as much
-as 62.7\% slower than using a temporary filesystem and
+as 62.7\% slower than using a temporary file system and
 33\% slower on average. We therefore expect many users to notice
@@
 users to setup their runtime environment using familiar systems.  While Spack
-packages do not {\tt LD\_LIBRARY\_PATH} to run, we set it in our generated module
+packages do not need {\tt LD\_LIBRARY\_PATH} to run, we set it in our generated module
 files as well, because it can be used by build systems to find libraries, as well

### 2015-07-31 implementation-packages.tex
Changes from Greg.

@@
 handle managing the install location.  Rather, Spack passes the {\tt install}
-method a {\tt prefix} parameter.  The package implementor must ensure that
+method a {\tt prefix} parameter.  The package implementer must ensure that
 her {\tt install} function installs the package {\it into} the {\tt prefix},
-but Spack ensures that {\tt prefix} is computed in such a way that it is
+but Spack ensures that {\tt prefix} is computed so that it is
 unique for every configuration of a package.  To further simplify
@@
 language (DSL).
-The DSL provides special directives such as {\tt depends\_on},
-{\tt version}, {\tt patch}, and {\tt provides} that add metadata
+The DSL provides directives, such as {\tt depends\_on},
+{\tt version}, {\tt patch}, and {\tt provides}, that add metadata
 to the package class.

### 2015-07-31 implementation-specs.tex
Changes from Greg.

@@
 have named flags, or {\it variants}.  Variants are associated with the package,
-so the {\tt mpileaks} package implementor must explicitly handle cases
+so the {\tt mpileaks} package implementer must explicitly handle cases
 where debug is enabled ({\tt +debug}) or disabled ({\tt -debug}

### 2015-07-31 implementation-virtual-deps.tex
Name change spack-paper-sc15 -> spack-sc15
Bibliography formatting, tighten to 12 pages.
Changes from Greg.

@@
 \end{figure}
+
+\begin{figure*}[ht!]
+	\centering
+	\includegraphics[width=.9\textwidth]{figs/concretization.pdf}
+	\caption{
+		Spack's concretization process.
+		\label{fig:concretization}
+	}
+\end{figure*}
 
@@
 a specific implementation; they can depend on the virtual name, for which
-the user or Spack can select an an implementation at build time.
+the user or Spack can select an implementation at build time.
 Other package managers support the notion of virtual dependencies, but Spack

### 2015-07-31 introduction.tex
Changes from Greg.

@@
 in size from a 1.6 teraflop, 256-core cluster to the
-20 petaflop, 1.6 million-core Sequoia machine, currently ranked first and
+20 petaflop, 1.6 million-core Sequoia machine, currently ranked second and
 third on the Graph500~\cite{graph500} and Top500~\cite{top500}

### 2015-07-31 usecase-policy.tex
Name change spack-paper-sc15 -> spack-sc15
Revisions and better overhead plot.
Changes from Greg.

@@
 While Spack can easily create many installations of a package like {\tt mpileaks}
-for different compilers and MPI implementations, end users may find Spack's directory layout confusing, and they may not be able to find libraries intuitively.
+for different compilers and MPI implementations, end-users may find Spack's directory layout confusing, and they may not be able to find libraries intuitively.
 Spack installs packages into paths based on concretized specs.
@@
 Even when modules are available, many users still
-navigate the filesystem to access packages.
+navigate the file system to access packages.
 
@@
 user level.
-For each package or set of packages, configuration file contains rules
+For each package or set of packages, the configuration file contains rules
 describing the links that should point into that package.
@@
 %
-to the site's {\tt spackconfig} file, which would cause the ambiguous
+to the site's configuration file, which would cause the ambiguous
 {\tt mpileaks} link to point to an installation compiled with {\tt icc}.
@@
 Custom packages can inherit from and replace Spack's default
-packages, so other sites to either tweak or completely replace Spack's
+packages, so other sites can either tweak or completely replace Spack's
 build recipes.  To continue the previous example, a site can write

### 2015-07-31 figs/overhead/overhead.py

@@ -10,7 +10,7 @@ with open('overhead.yaml') as f:
 
 N = len(data)
 keys = ['libelf', 'libpng', 'mpileaks',
-        'libdwarf', 'python', 'dyninst', 'lapack']
+        'libdwarf', 'python', 'dyninst', 'LAPACK']
 
 nfs_color='RoyalBlue'
 wrap_color='Orange'

### 2015-08-03 introduction.tex
small tweaks to intro

--- 2015-07-31 introduction.tex
@@
 in size from a 1.6 teraflop, 256-core cluster to the
-20 petaflop, 1.6 million-core Sequoia machine, currently ranked first and
-third on the Graph500~\cite{graph500} and Top500~\cite{top500}
-lists, respectively.
+20 petaflop, 1.6 million-core Sequoia machine, currently ranked second on the 
+Graph500~\cite{graph500} and third on the Top500~\cite{top500}.
 %
@@
 %
-Spack allows users to rapidly compose new package configurations,
-manage Python installations, and implement site-specific build policies. 
+Spack supports rapid composition of package configurations, management 
+of Python installations, and site-specific build policies. 
 %

### 2015-08-03 motivation.tex
Changes from Greg.
tweaks to existing practice discussion

@@
 Packaging problems arise in HPC because a supercomputer's hardware, OS, and
-filesystem are shared by many users with different requirements.  The classic
+file system are shared by many users with different requirements.  The classic
 solution to this problem is to use virtual
@@
 The most advanced of these, Lmod~\cite{mclay:lmod,mclay:lmod-tutorial},
-provides software hierarchies similar to the naming conventions in
-Table~\ref{tab:naming-conventions}. This allows users to load a software stack
+provides software hierarchies that are similar to the naming conventions in
+Table~\ref{tab:naming-conventions} and allow users to load a software stack
 quickly if they know which one is required.
@@
 BLAS libraries together in a {\it toolchain} that can be used by
-package files. This provides some degree of composability and
+package files. The grouping provides some composability and
 separates compiler flags and MPI concerns from client packages.

### 2015-08-03 implementation-packages.tex
tweaks to package discussion

@@
 method a {\tt prefix} parameter.  The package implementer must ensure that
-her {\tt install} function installs the package {\it into} the {\tt prefix},
+the {\tt install} function installs the package {\it into} the {\tt prefix},
 but Spack ensures that {\tt prefix} is computed so that it is
-unique for every configuration of a package.  To further simplify
-package implementation, Spack implements an embedded domain-specific
+unique for every configuration of a package.  To simplify package 
+implementation further, Spack implements an embedded domain-specific
 language (DSL).

### 2015-08-03 implementation-specs.tex
many changes to discussion of specs

@@
 To understand specs, consider the {\tt mpileaks} package structure.
-Metadata in the {\tt Mpileaks} class ({\tt version}, {\tt depends\_on}, etc.) describe
-its relationships with other packages.  There are two direct dependencies:
+Metadata in the {\tt Mpileaks} class (e.g., {\tt version} or 
+{\tt depends\_on}) describe its relationships with other packages.  
+The tool has two direct dependencies:
 the {\tt callpath} library and {\tt mpi}.  Spack recursively inspects the class definitions
@@
 Each node represents a package, and each package has five configuration parameters that control
-how it will be built: 1) the package version, 2) the compiler to
-build with, 3) the compiler version, 4) named compile-time build options, or {\it variants},
-and 5) the target architecture.
+how it will be built: 1) the package version, 2) the compiler with which to
+build, 3) the compiler version, 4) named compile-time build options, or 
+{\it variants}, and 5) the target architecture.
 
 \subsubsection{Configuration Complexity}
-A spec DAG has many degrees of freedom, and it is not reasonable to expect users to
-understand or specify all of them.  In our experience at LLNL, the typical user
-only cares about a small number of build constraints (if any), and does not know enough to
+A spec DAG has many degrees of freedom, and iusers cannot reasonably be 
+expected to understand or to specify all of them.  In our experience at LLNL, 
+the typical user only cares about a small number of build constraints (if any),
+and does not know enough to
 specify the rest. For example, a user may know that a certain version of a library like
@@
 %
-Configuration complexity makes the HPC software ecosystem difficult to manage: there are
-simply too many parameters. However, the small set of important build constraints can be
-very specific, so we have two competing concerns.  We need the ability to specify details
-of the configuration, without the complexity of remembering all of it.
+Configuration complexity makes the HPC software ecosystem difficult to manage: 
+too many parameters exist to know all of them. However, the known and 
+important ones often provide detailed build constraints. Thus, we have two 
+competing concerns.  We need the ability to specify details
+without having to remember all of them.
 
@@
 We have developed a syntax for specs that allows users to specify
-only those constraints they care about, but to be specific when necessary.
-Our syntax is expressive enough to represent software DAGs  but concise enough to use
-on the command line. The spec syntax is recursively defined to allow users to specify
-parameters on dependencies as well as on the root of the DAG.  The EBNF grammar we use
-to implement this in Spack is shown in Figure~\ref{fig:grammar}.
+only the constraints that matter. Our syntax can represent software 
+DAGs concisely enough to use it on the command 
+line. The spec syntax is recursively defined to allow users to specify
+parameters on dependencies as well as on the root of the DAG. 
+Figure~\ref{fig:grammar} showsh the EBNF grammar through which Spack 
+flexibly supports constraints.
 
@@
 Note that even though the spec contains no dependency information, it is still
-converted to a full DAG, based on directives supplied in package files. Since there
-are no constraints on the nodes, Spack has leeway when building the
+converted to a full DAG, based on directives supplied in package files. Since 
+the spec does not constrain the nodes, Spack has leeway when building the
 package, and we say that it is {\it unconstrained}.
 %
-Suppose instead that the user wants a specific {\tt mpileaks} version.
-She can request it with a version constraint after the package name:
+However, a user who wants a specific {\tt mpileaks} version can 
+request it with a version constraint after the package name:
 %
@@
 %
-We see from Figure~\ref{fig:specs-mpileaks-version} that the specific version constraint is
-placed on the {\tt mpileaks} node in the DAG, but the rest of the DAG remains unconstrained.
-If the user only requires a particular minimum version, then she could use
+Figure~\ref{fig:specs-mpileaks-version} shows that the specific version 
+constraint is placed on the {\tt mpileaks} node in the DAG, which otherwise 
+remains unconstrained.
+A user who only requires a particular minimum version could use
 {\it version range} syntax
-and write {\tt @2.3:}.  Likewise, for a version between 2.3 and 2.5.6, she would write
-{\tt @2.3:2.5.6}. In these cases, the user can save time if
+and write {\tt @2.3:}.  Likewise, {\tt @2.3:2.5.6} would specify a 
+version between 2.3 and 2.5.6. In these cases, the user can save time if
 Spack already has a version installed that satisfies the spec -- Spack will use
@@
 %
-The caret (\verb|^|) denotes constraints for a particular dependency.  In the DAG,
-there are now version constraints on {\tt callpath} {\it and} {\tt libelf},
+The caret (\verb|^|) denotes constraints for a particular dependency. The DAG,
+now has version constraints on {\tt callpath} {\it and} {\tt libelf},
 and the user has requested the {\tt debug} variant of the {\tt callpath} library.
 
-Recall that Spack guarantees that there is only a single version of any package in
-a spec.  Within a spec, each dependency can be uniquely
+Recall that Spack guarantees that only a single version of any package occurs 
+in a spec.  Within a spec, each dependency can be uniquely
 identified by its package name alone.  Therefore, the user does not need to consider
-DAG connectivity to add constraints.  She must only know that {\tt mpileaks}
-somehow depends on {\tt callpath}.
-Thus, dependency constraints can appear in arbitrary order.
+DAG connectivity to add constraints but instead only must only know that 
+{\tt mpileaks} depends on {\tt callpath}.
+Thus, dependency constraints can appear in an arbitrary order.
 
-Table~\ref{tab:specs} shows further examples of specs, ranging from very
-simple to more complex. These examples show how Spack's constraint
+Table~\ref{tab:specs} shows further examples of specs, ranging from
+simple to complex. These examples show how Spack's constraint
 notation covers the rest of the HPC package parameter space.
@@
 version on the command line that is unknown to Spack,
-Spack will attempt to fetch and install it.  Spack also uses the same
-model to scrape webpages and find new versions as they become available.
+Spack will attempt to fetch and to install it.  Spack uses the same
+model to scrape webpages and to find new versions as they become available.
 
@@
 specifier.  Spack compiler names like {\tt gcc} refer to the full compiler {\it toolchain},
-i.e. the C, C++, and Fortran 77 and 90 compilers.  Spack can auto-detect
+i.e., the C, C++, and Fortran 77 and 90 compilers.  Spack can auto-detect
 compiler toolchains in the user's {\tt PATH}, or they can be registered manually
@@
 {\bf Cross-compilation.}
-To support cross-compilation, specs includes the system architecture (line 5).
-Platforms begin with {\tt =} and take names like {\tt linux-ppc64} or {\tt bgq}.  They are
-specified per-package; this allows front-end tools to depend on their back-end measurement
+To support cross-compilation, spec can include the system architecture 
+(line 5). Platforms begin with {\tt =} and take names like {\tt linux-ppc64} 
+or {\tt bgq}.  They are specified per-package. This mechanism allows front-end 
+tools to depend on their back-end measurement
 libraries with a {\it different} architecture on cross-compiled machines.
@@
 So far, we have shown examples of specs used to request constraints from the
-command line, when {\tt spack install} is invoked.  However, the user is not the only
-source of constraints.  Applications may require specific versions of dependencies,
-and it is often desirable to write constraints like this into a package file.  For
+command line, when {\tt spack install} is invoked.  However, the user is not 
+the only source of constraints.  Applications may require specific versions 
+of dependencies. Often, these constraints should be specified in a package 
+file.  For
 example, the ROSE compiler~\cite{rose} only builds with a certain version of the {\tt boost} library.
@@
 the {\tt ROSE} package builds with different versions of boost, depending on the
-compiler version. To handle this, directives also accept spec syntax as a
+compiler version. So, directives also accept spec syntax as a
 predicate in an optional {\tt when} parameter:
@@
 \noindent
-The same notation is used to optionally build with libraries like MPI:
+The same notation is used to build optionally with libraries like MPI:
 %
@@
 
-In our experience maintaining packages at LLNL, we often have to change
+In our experience maintaining packages at LLNL, we often must change
 entire package build scripts due to large changes in the way certain packages build.
-This can be very cumbersome, and it is difficult to maintain both the old and new
-version of a build script, which we must do for users who rely on the old version.
+These changes are cumbersome, particularly since we often must maintain both 
+the old and new version of a build script, as some users rely on the old version.
 
@@
 
-For cases like this, Spack provides functionality that allows Python functions to have
+Spack provides functionality that allows Python functions to have
 multiple definitions, each specialized for particular configurations of the package.
-This allows us to have two separate implementations of {\tt install} or {\it any} method
-in a package class. Figure~\ref{fig:specialization} shows how this is used in the Dyninst
+This capability allows us to have two separate implementations of {\tt install} or {\it any} method
+in a package class, as Figure~\ref{fig:specialization} shows for the Dyninst
 package.  The {\tt @when} directive is a Python decorator: a higher order function that
@@
 We replace the function with a callable multi-function dispatch object, and we
-integrate the predicate check into the function dispatch mechanism.  Here, the
-{\tt @when} condition is true when Dyninst is at version 8.1 or lower, and in those cases
-the package will use the {\tt configure}-based build.  By default if no predicate matches,
+integrate the predicate check into the function dispatch mechanism.  The
+{\tt @when} condition is true when Dyninst is at version 8.1 or lower, in which
+case the package will use the {\tt configure}-based build.  By default if no predicate matches,
 {\tt install} will use the default CMake-based implementation. The simple {\tt @when}

### 2015-08-03 implementation-virtual-deps.tex
a few tweaks to virtul dependency wording

@@
 
-Many libraries share a common source interface and can be interchanged
-within a build.
-The archetypal example in HPC is the MPI interface, which has a number of
-open source implementations (MPICH, OpenMPI, MVAPICH, etc.) and vendor-specific
-implementations. An application that can be built with one MPI can generally be
+Many libraries share a common interface and can be interchanged within a build.
+The archetypal examples in HPC are MPI libraries, which has several
+open source (e.g., MPICH, OpenMPI, and MVAPICH) and 
+vendor-specific implementations. An application that can be built with 
+one MPI implementation can generally be
 built with another.  Another example is the Basic Linear Algebra
 Subroutines (BLAS), which has many fungible implementations
-(ATLAS, LAPACK-BLAS, MKL, etc.).
+(e.g., ATLAS, LAPACK-BLAS, and MKL).
 
@@
 re-linked with a new version. They must be recompiled and reinstalled.
-At LLNL, we frequently need to build tools with many versions of MPI to support
+At LLNL, we must frequently build tools with many versions of MPI to support
 the many different applications that run at our center.
@@
 To allow rapid composition of libraries using an interface, Spack supports
-{\it virtual dependencies}.  A virtual dependency is an abstract name
-representing a library interface or capability instead of a library
-implementation.  Packages that need this capability need not depend on
+{\it virtual dependencies}.  A virtual dependency is an abstract name that
+represents a library interface (or capability) instead of a library
+implementation.  Packages that need this interface do need not to depend on
 a specific implementation; they can depend on the virtual name, for which
@@
 Other package managers support the notion of virtual dependencies, but Spack
-adds {\it versioning} to its interfaces.  This allows concepts like MPI
-interface versions and BLAS levels to be represented directly.  Spack
+adds {\it versioning} to its interfaces, which directly supports concepts 
+like MPI versions and BLAS levels.  Spack
 handles the details of managing and checking complex versioning constraints.
@@
 Figure~\ref{fig:virtual-deps} shows how packages provide
-virtual interfaces in Spack.  The spec syntax is used to concisely
-associate ranges of {\tt mpi} versions for the {\tt mvapich2} and {\tt mpich} packages.
+virtual interfaces in Spack.  The spec syntax concisely associates ranges 
+of {\tt mpi} versions for the {\tt mvapich2} and {\tt mpich} packages.
 The {\tt mpileaks} package requires {\tt mpi}, but it does not constrain the version.

### 2015-08-03 implementation-concretization.tex
Changes from Greg.
various changes to concretization discussion including continued removal of non-neutral gender users

@@
 We have discussed Spack's internal software DAG model, and we have shown
-how the spec syntax can be used to quickly specify a partially constrained
-software DAG.  We say such a DAG is {\it abstract}, as it could potentially
+how the spec syntax can be used to specify a partially constrained
+software DAG.  This DAG is {\it abstract}, as it could potentially
 describe more than one software configuration. Before Spack builds a spec, it
@@
 \begin{enumerate}
-\item No package in the spec DAG is missing dependencies.
-\item No package in the spec DAG is virtual.
+\item No package in the spec DAG is missing dependencies;
+\item No package in the spec DAG is virtual;
 \item All parameters are set for all packages in the DAG.
 \end{enumerate}
-If a spec meets all of these criteria, we say it is {\it concrete}.
+If a spec meets these criteria then it is {\it concrete}.
 {\it Concretization} is the central component of Spack's build
@@
 
-Spack's concretization algorithm is shown in Figure~\ref{fig:concretization}.
+Figure~\ref{fig:concretization} shows Spack's concretization algorithm.
 The process starts when a user invokes {\tt spack install} and requests that a
@@
 Spack intersects the constraints of the two DAGs package by package, and it
-checks each parameter for inconsistencies.  Inconsistencies can arise if, for example,
+checks each parameter for inconsistencies.  Inconsistencies can arise if, 
+for example,
 the user inadvertently requests two versions of the same package, or if a
@@
 the user of the conflict. If the user or package specifies version ranges,
-they are intersected, and if the ranges do not overlap, an error is raised.
+they are intersected, and if the ranges do not overlap, Spack raises an error.
 When the intersection succeeds, Spack has a single DAG with the merged
@@
 The next part of the process is iterative.
-If any node is a virtual dependency, Spack replaces it with with a
-suitable interface provider.  It does this by building a reverse
+If any node is a virtual dependency, Spack replaces it with a
+suitable interface provider by building a reverse
 index from virtual packages to providers using the {\tt provides when}
@@
 provider.  The selected provider may {\it itself} have virtual dependencies,
-this process is repeated until there are no more virtual packages in the DAG.
+so this process is repeated until the DAG has no more virtual packages.
 
@@
 libraries (as in Section~\ref{sec:constraints}). The concretization algorithm
-evaluates conditions from {\tt when} clauses on the DAG, and if there are
+evaluates conditions from {\tt when} clauses on the DAG; if they result in
 new libraries or other changes, the cycle begins again.
@@
 to an inconsistency.  Rather, it will raise an error and the user must resolve
-the issue by being more explicit.  The user might toggle a variant, or she might
-force the build to use a {\it particular} MPI implementation
+the issue by being more explicit.  The user might toggle a variant or force 
+the build to use a {\it particular} MPI implementation
 by supplying \verb|^openmpi| or \verb|^mpich|.  We leave automatic constraint
@@
 To prevent the directory name from growing too long, Spack uses a SHA hash of
-dependencies' specs as the last directory component.  This
-does {\it not} mean that Spack rebuilds every library for each new configuration.
-If two configurations share a sub-DAG, then the sub-DAG's configuration will
-be reused.  Figure~\ref{fig:reuse} shows how the {\tt dyninst} sub-DAG is used for
-both the {\tt mpich} and {\tt openmpi} builds of {\tt mpileaks}.
+dependencies' specs as the last directory component.  However, Spack
+does {\it not} rebuild every library for each new configuration.
+If two configurations share a sub-DAG, then Spack reuses the sub-DAG's 
+configuration.  Figure~\ref{fig:reuse} shows how the {\tt dyninst} sub-DAG 
+is used for both the {\tt mpich} and {\tt openmpi} builds of {\tt mpileaks}.
 
@@
 built.  These include the {\tt package.py} file used to build, a build log
-containing output and error messages, and a file containing the complete
+that contains output and error messages, and a file that contains the complete
 concrete spec for the package {\it and} its dependencies. The spec file can be
-used later to reproduce the build, even if concretization preferences have changed.
+used later to reproduce the build, even if concretization preferences have 
+changed.
 
@@
 of building HPC software comes from the size of the build parameter space.
-As mentioned, there are only a few parameters that typical users care about.
+As previously mentioned, typical users only care about a few parameters.
 The rest add unneeded complexity.  When building manually, LLNL staff
@@
 request builds with a minimal spec expression, while still providing a
-mechanism for the site and the user to make consistent, repeatable choices about
-other build parameters.  For example, the site or the user can set default versions
-to use for any library that is not specified explicitly.
+mechanism for the site and the user to make consistent, repeatable choices 
+for other build parameters.  For example, the site or the user can set 
+default versions to use for any library that is not specified explicitly.
 %
 Second, concretization reduces the burden of packaging software, because
-package authors do not have to make these decisions.  There is no
-need for packages to contain complicated checks, or for packages to be overly
-specific about versions.  Other multi-configuration systems like Nix, EasyBuild,
-and HashDist require the package author to write a mostly concrete build spec in advance.
-We believe that this puts undue burden on the package author, and it makes the task
+package authors do not have to make these decisions. Packages do not need 
+to contain complicated checks, or to be overly specific about versions.  
+Other multi-configuration systems like Nix, EasyBuild, and HashDist require 
+the package author to write a mostly concrete build spec in advance.
+This requirement puts undue burden on the package author, and it makes the task
 of changing site policies within a software stack difficult.  Spack

### 2015-08-03 implementation-environment.tex
many changes to improve wording

@@
 environment, and reproducible builds are one of our design goals.
-Experience at LLNL has shown that it is vexingly difficult to reproduce
-a build manually.
+Experience at LLNL has shown that reproducing a build manually
+is vexingly difficult.
 %
-Many packages have a profusion of build options, and specifying
-them correctly often requires tedious experimentation.  This is due to lack of
+Many packages have a profusion of build options. Specifying
+them correctly often requires tedious experimentation due to lack of
 build standards, as well as the diversity of HPC environments.
 For example, in some packages that depend on the {\tt Silo} library,
-the \verb|--with-silo| parameter takes a path to {\tt Silo}'s installation prefix.
-In others, it takes the {\tt include} and {\tt lib} subdirectories,
+the \verb|--with-silo| parameter takes a path to {\tt Silo}'s installation 
+prefix. In others, it takes the {\tt include} and {\tt lib} subdirectories,
 separated by a comma.
@@
 \subsubsection{Environment isolation}
-In addition to inconsistencies that arise from building manually, we
-frequently encounter errors due to inconsistencies between the package
-installation environment and the environment of the package user.
+Inconsistencies arise not only from building manually but also due to 
+package installation and user environment differences.
 %
@@
 Finding dependencies at build time is not the only obstacle to reproducible
-behavior.  As mentioned in Section~\ref{sec:motivation}, it is also important
-for binaries to be able to find dependency libraries at {\it runtime}.
+behavior.  As mentioned in Section~\ref{sec:motivation}, binaries must be 
+able to find dependency libraries at {\it runtime}.
 One of the most common user errors at LLNL is improper library configuration.
-Users frequently do not know what libraries a package was built with, and
-it is difficult for them to construct a suitable {\tt LD\_LIBRARY\_PATH} for
-a package that was built by someone else.  To reduce the number of support calls
-we receive,
-we typically add {\tt RPATHs} to public software installations, so that paths
-to dependencies are embedded in binaries. This way, users do not need to know
-details of dependencies to run installed software correctly.
+Users frequently do not know the libraries with which a package was built, and
+constructing a suitable {\tt LD\_LIBRARY\_PATH} for a package that was built 
+by someone else is difficult. To reduce the number of support calls that we 
+receive, we typically add {\tt RPATHs} to public software installations, so 
+that paths to dependencies are embedded in binaries. Thus, users do not need 
+to know details of dependencies to run installed software correctly.
 
@@
 
-Spack's compiler wrappers have a number of benefits.  First, they allow
-Spack to transparently switch compilers in most builds.  This is how
-compiler options like {\tt \%gcc} are implemented.  Second, they enforce the
-use of {\tt RPATHs} in
-installed binaries.  This causes applications built by Spack to run correctly
-{\it regardless of the environment}.  Third, because compiler wrappers add
-header and library search paths for dependencies, header and library detection
-tests used by most build systems succeed automatically, without
-the need to use special arguments for nonstandard locations.  {\tt configure}
+Spack's compiler wrappers have several benefits.  First, they allow
+Spack to switch compilers transparently in most builds, which is how we
+implement compiler options like {\tt \%gcc}.  Second, they enforce the
+use of {\tt RPATHs} in installed binaries, which allows applications that 
+Spack builds to run correctly {\it regardless of the environment}.  Third, 
+because compiler wrappers add header and library search paths for dependencies,
+header and library detection tests that most build systems use succeed without
+using special arguments for nonstandard locations.  {\tt configure}
 commands in Spack's {\tt install} function can have fewer arguments, and can
-be written as they would be for system installs.  This reduces complexity
+be written as for system installs, which reduces complexity
 for package maintainers and helps to ensure consistent, reproducible
-build policies across packages.  Finally, because Spack has control over the
+build policies across packages.  Finally, because Spack controls the
 wrappers, package authors can programmatically filter the compiler flags
-used by software build systems, a useful last resort when porting to
-bleeding-edge platforms or new, esoteric compilers.
+used by build systems, which facilitates porting to bleeding-edge platforms 
+or new compilers.
 
@@
 and Figure~\ref{fig:overhead} shows the corresponding overhead as a percentage
-of the wrapper-less runtime.  {\tt Dyninst} and {\tt LAPACK}
-use CMake to build, and the rest use autotools.  We ran
-each build three times on an Intel Sandy Bridge cluster node, and we
-report the average here.
-In the case of {\tt dyninst}, the overhead is negligible (-0.4\%),
-but it is as high as 12.3\% for shorter builds, like
-{\tt mpileaks}.
+of the wrapper-less runtime.  {\tt Dyninst} and {\tt LAPACK} use CMake to 
+build, and the rest use autotools.  We ran each build three times on an Intel 
+Sandy Bridge cluster node, and we report the average. In the case of 
+{\tt dyninst}, the overhead is negligible (-0.4\%), but it is as high 
+as 12.3\% for shorter builds, like {\tt mpileaks}.
 
-To increase build speed, by default Spack runs each {\tt install}
-method in the default temporary directory. On most cluster nodes,
-this is a fast, locally-mounted file system. In our experience,
-many users build manually in their home directories out of habit.  At most sites, home
+To increase build speed, by default Spack runs each {\tt install} method 
+in the default temporary directory, which is a fast, locally-mounted file 
+system on most cluster nodes. In our experience, many users build manually 
+in their home directories out of habit.  At most sites, home
 directories are remotely mounted volumes (e.g., NFS).  To compare Spack builds
@@
 Future versions of Spack may also allow the creation of Lmod~\cite{mclay:lmod}
-hierarchies, as discussed in Section~\ref{sec:env-rpath}. Spack's rich
-dependency information would allow such hierarchies to be generated automatically.
+hierarchies.
+%% We discussed Lmod hierarchies generall; we did not discuss using
+%% them with Spack so the statment is misleading and unnecessary
+%%, as discussed in Section~\ref{sec:env-rpath}.
+Spack's rich dependency information would allow automatic generation of 
+such hierarchies.

### 2015-08-03 implementation-specs.tex

@@
 \subsubsection{Configuration Complexity}
-A spec DAG has many degrees of freedom, and iusers cannot reasonably be 
+A spec DAG has many degrees of freedom, and users cannot reasonably be
 expected to understand or to specify all of them.  In our experience at LLNL,
@@
 parameters on dependencies as well as on the root of the DAG.
-Figure~\ref{fig:grammar} showsh the EBNF grammar through which Spack 
+Figure~\ref{fig:grammar} shows the EBNF grammar through which Spack
 flexibly supports constraints.
@@
 %
-The caret (\verb|^|) denotes constraints for a particular dependency. The DAG,
+The caret (\verb|^|) denotes constraints for a particular dependency. The DAG
 now has version constraints on {\tt callpath} {\it and} {\tt libelf},
@@
 identified by its package name alone.  Therefore, the user does not need to consider
-DAG connectivity to add constraints but instead only must only know that 
+DAG connectivity to add constraints but instead must only know that
 {\tt mpileaks} depends on {\tt callpath}.
@@
 {\bf Cross-compilation.}
-To support cross-compilation, spec can include the system architecture 
+To support cross-compilation, specs can include the system architecture
 (line 5). Platforms begin with {\tt =} and take names like {\tt linux-ppc64}

### 2015-08-03 usecase-combinatorial.tex
Name change spack-paper-sc15 -> spack-sc15
a few small tweaks

@@
 Spack presented a solution to both problems.  Package administrators can use Spack to
-easily maintain a central install of {\tt gperftools} across combinations of
-compilers and compiler versions.  Spack's {\tt gperftools} package also serves as a central
+maintain a central install of {\tt gperftools} across combinations of
+compilers and compiler versions easily.  Spack's {\tt gperftools} package 
+also serves as a central
 institutional knowledge repository, as its package files encode
@@
 configurations of {\tt mpileaks} using Spack.  Moreover, Spack's virtual dependency
-system allows us to quickly compose a new {\tt mpileaks} build when a new MPI library is 
+system allows us to compose a new {\tt mpileaks} build quickly when a new MPI library is
 deployed at LC, {\it without} modifying the {\tt mpileaks} package itself.

### 2015-08-04 usecase-python.tex
a few small tweaks

@@
 None handles multi-configuration builds.  More glaringly, Python extensions
-are usually installed into the Python interpreter's prefix.
-This makes it impossible to install multiple versions.\footnote{{\tt setuptools}
+are usually installed into the Python interpreter's prefix, which makes it 
+impossible to install multiple versions.\footnote{{\tt setuptools}
 has support for multiple versions via {\tt pkg_resources},
@@
 
-Per Spack's design philosophy, we wanted a way to easily manage many different
-versions, but {\it also} to provide a baseline set of extensions {\it without}
-requiring environment settings.
+Per Spack's design philosophy, we wanted a way to manage many different
+versions easily, but {\it also} to provide a baseline set of extensions 
+{\it without} requiring environment settings.
 %
@@
 and each module depends on a particular Python installation.
-But extensions can be {\tt activated} or {\tt deactivated}
+However extensions can be {\tt activated} or {\tt deactivated}
 within the dependent Python installation.  The {\tt activate} operation
 symbolically links each file in the extension prefix into the Python
-installation prefix, as though it were installed directly. If any file
+installation prefix, as if it were installed directly. If any file
 conflict would arise from this operation, {\tt activate} fails.
@@
 
-There were additional complications because many Python packages {\it install their own
-package manager} if they do not find one in the Python installation.
-There are also many ways that Python packages add themselves to the
-interpreter's default path, and some of them conflict. We modified
-Spack so that extendable packages, like Python, can supply custom code
-in the package file that that specializes {\tt activate} and {\tt deactivate}
-for the particular package. Python uses this feature to merge conflicting
-files during activation.  The end result is that Python extensions can
+Additional complications arose because many Python packages 
+{\it install their own package manager} if they do not find one in the 
+Python installation. Also, Python packages use a wide range of mechanisms
+to add themselves to the interpreter's default path, some of which conflict. 
+We modified Spack so that extendable packages, like Python, can supply 
+custom code in the package file that specializes {\tt activate} and 
+{\tt deactivate} for the particular package. For Python, this feature 
+merges conflicting files during activation. Overall, Python extensions can
 be installed automatically in their own prefixes, and they can be composed
@@
 instance, which can coexist with Spack's normal installation model.
-This has allowed us to efficiently support our application teams, for whom we can
+This mechanism efficiently supports our application teams, for whom we can
 now rapidly construct custom Python installations.  We have also reduced
-the amount of time that LC staff spend installing Python modules.
+the time that LC staff spend installing Python modules.
 Because Spack packages can extend the activation and deactivation mechanisms,
-we believe the same mechanism could be used with other
-interpreted languages with similar extension models, such as R, Ruby, or Lua.
+the same mechanism could be used with other interpreted languages with 
+similar extension models, such as R, Ruby, or Lua.

### 2015-08-04 implementation-virtual-deps.tex
Makefile update for new diff; minor wording quibble.

@@
 represents a library interface (or capability) instead of a library
-implementation.  Packages that need this interface do need not to depend on
+implementation.  Packages that need this interface do need not depend on
 a specific implementation; they can depend on the virtual name, for which

### 2015-08-04 usecase-policy.tex
many changes to make clearer; also removed reference to non-existent previous example about Spindle since Spindle was not mentioned anywhere else

@@
 
-Spack makes it easy to create and organize package installations,
-but it must also be easy for end-users to find and use packages on an HPC
+Spack makes it easy to create and to organize package installations,
+but it must also be easy for end-users to find and to use packages on an HPC
 system. Different end-users have different expectations about how packages
@@
 
-While Spack can easily create many installations of a package like {\tt mpileaks}
-for different compilers and MPI implementations, end-users may find Spack's directory layout confusing, and they may not be able to find libraries intuitively.
-Spack installs packages into paths based on concretized specs.
-This is ideal for maintaining multiple package installations,
-but it may be difficult for an end-user to navigate.
+While Spack can easily create many installations of a package like 
+{\tt mpileaks}, end-users may find Spack's 
+directory layout confusing, and they may not be able to find libraries 
+intuitively. Spack installs packages into paths based on concretized specs,
+which is ideal for maintaining multiple package installations. However, 
+end-users would find these paths difficult to navigate. 
 For example, an {\tt mpileaks} installation prefix might be:
@@
 environment modules are commonly used on HPC systems to solve this problem,
-and Spack allows the package author to automatically create environment modules
-for packages that Spack installs.
+and Spack allows the package author to create environment modules 
+automatically for packages that Spack installs.
 Even when modules are available, many users still
@@
 
-Spack thus allows the creation of views, which are a symbolic-link based directory layout of packages.
-Views provide a human-readable directory layout,
-which can be adapted to match legacy directory layouts that users rely on.
-For example, the above {\tt mpileaks} package may have a view that creates a link in
-{\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpileaks}.
+Spack allows the creation of views, which are a symbolic-link based directory 
+layout of packages. Views provide a human-readable directory layout that
+can be adapted to match legacy directory layouts on which users rely. For 
+example, the above {\tt mpileaks} package may have a view that creates a link 
+in {\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpileaks}.
 The same package install may be referenced by multiple links and views,
 so the above package could also be linked from a more generic {\tt /opt/mpileaks-openmpi}
-link. This is useful for users who always want to use the latest version.
+link. This reuse supports users who always want to use the latest version.
 Views can also be used to create symbolic links to specific executables or libraries in an install,
@@
 For each package or set of packages, the configuration file contains rules
-describing the links that should point into that package.
+that describe the links that should point into that package.
 The link names can be parameterized.
@@
 By default, Spack prefers newer versions of packages compiled with newer compilers
-to older packages built with older compilers. It has well-defined, but not
+to older packages built with older compilers. It has a well-defined, but not
 necessarily meaningful, order of preference for deciding between MPI
@@
 {\tt mpileaks} link to point to an installation compiled with {\tt icc}.
-Any compiler not in the compiler\_order setting is treated as less preferred
-than those explicitly provided.
+Any compiler not in the {\tt compiler\_order} setting is less 
+preferred than those explicitly provided.
 %
 In a similar manner, Spack can be configured to give specific package
-configurations priority over others.
-This can be useful if a new version is unstable and untested.
+configurations priority over others, which can be useful with a new, unstable 
+and untested version.
 
@@
 organization.  Putting this type of code back into a public repository
-does not often make sense, and if it makes the
+does not often make sense and, if it makes the
 mainline less stable, it can actually make sharing code between sites more difficult.
@@
 Spack allows the creation of site-specific variants of packages.
-Via configuration files, users can specify additional search
-directories for finding additional {\tt Package} classes.
+Users can specify additional search directories for finding additional 
+{\tt Package} classes via configuration files.
 %
-The additional packages are like the {\tt mpileaks} package shown in Figure~\ref{fig:mpileaks}.  However, the extension packages can extend from not
+The additional packages are like the {\tt mpileaks} package shown in 
+Figure~\ref{fig:mpileaks}.  However, the extension packages can extend from not
 only {\tt Package}, but also any of Spack's built-in packages.
@@
 packages, so other sites can either tweak or completely replace Spack's
-build recipes.  To continue the previous example, a site can write
-a {\tt LocalSpindle} Python class, which inherits from Spack's
-{\tt Spindle} class.  {\tt LocalSpindle} may simply add additional
-configure flags to the {\tt Spindle} class, while leaving the
-dependencies and most of the build instructions from its parent
-class.  For reproducibility, Spack also tracks the {\tt Package}
-class that drove a specific build.
+build recipes. For example, a site can write a local Python class that 
+inherits from Spack's base class. The local class may simply add configure 
+flags to the Spackclass, while leaving the dependencies and most of the 
+build instructions from its parent class. For reproducibility, Spack also 
+tracks the {\tt Package} class that drove a specific build.

### 2015-08-04 usecase-ares.tex
a few small tweaks

@@
 packages are written in a diverse set of languages including C, C++, 
-Fortran, Python and tcl and uses MPI and OpenMP for parallelism.
+Fortran, Python and tcl.
+%% Sense of the following has been lost and seems unnecessary
+%% and uses MPI and OpenMP for parallelism.
 
@@
 As shown in the table, 36 different configurations have been run using Spack
-(some of 4 versions on each of 10 architecture-compiler-MPI combinations).
+(up to 4 versions on each of 10 architecture-compiler-MPI combinations).
 Prior to using Spack, only Linux/Intel configurations were automated. The ARES 
@@
       build the four configurations with correct libraries;
-\item The spec syntax allowed build scripts to concisely test compiler,
-      compiler version, and dependency versions---a necessity
+\item The spec syntax allowed build scripts to test compiler,
+      compiler version, and dependency versions concisely---a necessity
       for handling the different architectures;

### 2015-08-04 conclusion.tex
two small fixes

@@
 of the software configuration space directly. None of them allows
-a user to rapidly {\it compose} new builds with version, compiler,
-and dependency parameters.
+a user to {\it compose} new builds with version, compiler,
+and dependency parameters rapidly.
 
@@
 We showed through four use cases that Spack is already increasing
-operational efficiency in production at LLNL.  We believe that the
+operational efficiency in production at LLNL. The
 software management techniques implemented in Spack are applicable

### 2015-08-10 conclusion.tex
Minor tweaks, Final version.

@@
 across a number of HPC sites, with particular focus on Livermore
-Computing (LC). While tools exist that can handle multi-configuration
-installs, none of them addresses the combinatorial nature
-of the software configuration space directly. None of them allows
+Computing (LC). While existing tools can handle multi-configuration
+installs, none of them sufficiently addresses the combinatorial nature
+of the software configuration space. None of them allows
 a user to {\it compose} new builds with version, compiler,

### 2015-08-10 implementation-environment.tex
Minor tweaks, Final version.

@@
 
-\subsubsection{Environment isolation}
+\subsubsection{Environment Isolation}
 Inconsistencies arise not only from building manually but also due to 
-package installation and user environment differences.
+differences between the package installation environment and the user
+environment.
 %
@@
 Specifying the wrong version at build time has caused many
-unexpected crashes.
+unexpected crashes at runtime.
 
@@
 
-\subsubsection{Compiler wrappers and RPATHs}
+\subsubsection{Compiler Wrappers and RPATHs}
 Finding dependencies at build time is not the only obstacle to reproducible

### 2015-08-10 implementation-specs.tex
Minor tweaks, Final version.

@@
 Configuration complexity makes the HPC software ecosystem difficult to manage:
-too many parameters exist to know all of them. However, the known and
+too many parameters exist to specify them all. However, the known and
 important ones often provide detailed build constraints. Thus, we have two
@@
 We have developed a syntax for specs that allows users to specify
-only the constraints that matter. Our syntax can represent software
-DAGs concisely enough to use it on the command
-line. The spec syntax is recursively defined to allow users to specify
+only constraints that matter to them. Our syntax can represent
+DAGs concisely enough for command line use.
+The spec syntax is recursively defined to allow users to specify
 parameters on dependencies as well as on the root of the DAG.
@@
 
-\subsubsection{Constraints in packages}\label{sec:constraints}
+\subsubsection{Constraints in Packages}\label{sec:constraints}
 
@@
 
-\subsubsection{Build specialization}
+\subsubsection{Build Specialization}
 

### 2015-08-10 introduction.tex
Minor tweaks, Final version.

@@
 and developers frequently build, install, and support many
-different configurations of math and physics libraries, and
+different configurations of math and physics libraries, as well as
 other software.  Frequently, applications must be rebuilt to fix bugs

### 2015-08-10 limitations.tex
Minor tweaks, Final version.

@@
 cases have been rare so far. However, we plan to add better constraint
-satisfaction solving to Spack in future work.
+solving to Spack in future work.
 
@@
 compiler features.  More and more, our codes are relying
-on advanced compiler capabilities, like {\tt C++11} language features,
+on advanced compiler capabilities, like C++11 language features,
 OpenMP versions, and GPU compute capabilities. Ideally, Spack

### 2015-08-10 motivation.tex
Minor tweaks, Final version.

@@
 
-\paragraph{Traditional package managers}
+\paragraph{Traditional Package Managers}
 Package managers automate the installation of complex sets of software packages.
@@
  user to have a personalized environment with its own package manager, and it
-has been extremely successful for servers at cloud data centers. VMs typically
+has been very successful for servers at cloud data centers. VMs typically
 have near-native compute performance but low-level HPC network drivers still
@@
 so running many configurations still requires a large number of VMs.
-For facilities, this profusion of VMs is a security concern since it
+For facilities, this profusion of VMs is a security concern because it
 complicates mandatory patching. Also, managing  a large number of VM
@@
 \hline
-Oak Ridge~\cite{jones+:cug08}  & {\tt / \$arch / \$package / \$version / \$build} \\
+ORNL~\cite{jones+:cug08}  & {\tt / \$arch / \$package / \$version / \$build} \\
 \hline
-TACC/Lmod~\cite{mclay:lmod-tutorial}& {\tt / \$compiler-\$comp\_version / \$mpi / \$mpi\_version / \$package / \$version} \\
+TACC / Lmod~\cite{mclay:lmod-tutorial}& {\tt / \$compiler-\$comp\_version / \$mpi / \$mpi\_version / \$package / \$version} \\
 \hline
@@
 It allows multiple versions to be installed at once.  Rather
-than setting {\tt RPATHs}, it auto-generates module files
+than setting {\tt RPATHs}, it generates module files
 to manage the environment, and it is closely coupled with
@@
 Both Nix and HashDist use {\tt RPATHs} in their packages to ensure that
-libraries are found correctly.
+libraries are found correctly at runtime.
 
-\paragraph{Gaps in current practice}
+\paragraph{Gaps in Current Practice}
 The flexible cryptographic versioning of Nix and HashDist manages
@@
 unmanageable.  As an example, the EasyBuild system has over 3,300 files for
-several permutations of only 600 packages. A slightly different dependency
+several permutations of around 600 packages. A slightly different dependency
 graph requires an entire new package file hierarchy.  HashDist supports

### 2015-08-10 usecase-ares.tex
Minor tweaks, Final version.

@@
 We have configured Spack to build ARES with external MPI implementations, 
-depending on the host system. This configuration exploits the vendor- 
-or site-supplied MPI installation that often uses host-specific optimized 
-network drivers. MPI is shown as a virtual dependency in the figure, as 
-the implementation differs according to the host machine.  ARES builds its 
-own Python version in order to run on machines where Python is not well 
-supported, like Blue Gene/Q.  In particular, ARES builds a version of 
-Python 2.7 for Blue Gene/Q, which the native software stack does not support.
+depending on the host system. This configuration exploits vendor- 
+or site-supplied MPI installations that often use host-specific, optimized 
+network drivers. MPI is thus shown as a virtual dependency in the figure.
+ARES builds its own Python version, so that it can run on machines
+where Python is not well supported, like Blue~Gene/Q.  Specifically, for
+Blue~Gene/Q, it builds Python 2.7.9, which the native software stack
+does not support.
 
@@
 Thus, the ARES team already had some experience supporting 
-automated builds of dependencies. We developed Spack packages for the LLNL 
+automated builds of dependencies. We developed Spack packages for the 
 packages in Figure~\ref{fig:ares}. Many of the external packages were already 
@@
 Altogether, the initial packaging effort required roughly two months 
-for an experienced build engineer working 20 hours per week.
+for an experienced build engineer working 20 hours per week. This includes
+time spent learning to use Spack.
 As shown in the table, 36 different configurations have been run using Spack

### 2015-08-10 usecase-policy.tex
Minor tweaks, Final version.

@@
 Spack makes it easy to create and to organize package installations,
-but it must also be easy for end-users to find and to use packages on an HPC
+but it must also be easy for end-users to find and use packages on an HPC
 system. Different end-users have different expectations about how packages
@@
 
-Spack allows the creation of views, which are a symbolic-link based directory 
-layout of packages. Views provide a human-readable directory layout that
+Spack allows the creation of {\it views}, which are symbolic-link based directory 
+layouts of packages. Views provide a human-readable directory layout that
 can be adapted to match legacy directory layouts on which users rely. For 
@@
 
-Spack's views are a projection from a point in a high-dimensional space
-(the concretized spec, which fully specifies all parameters) to point
+Spack's views are a projection from points in a high-dimensional space
+(concretized specs, which fully specify all parameters) to points
 in a lower-dimensional space
-(the link name, which may only contain a few parameters).
-Thus, several installations may map to the same link.
+(link names, which may only contain a few parameters).
+Several installations may map to the same link.
 For example, the above {\tt mpileaks} link could point to an {\tt mpileaks} compiled with
-{\tt gcc} or {\tt icc} -- the compiler parameter is not part of the link.
+{\tt gcc} or {\tt icc}---the compiler parameter is not part of the link.
 To keep  package installations consistent and reproducible,
@@
 inherits from Spack's base class. The local class may simply add configure 
-flags to the Spackclass, while leaving the dependencies and most of the 
+flags to Spack's version, while leaving the dependencies and most of the 
 build instructions from its parent class. For reproducibility, Spack also 

### 2015-08-10 usecase-python.tex
Minor tweaks, Final version.

@@
 
-\subsection{Support for interpreted languages}
+\subsection{Support for Interpreted Languages}
 \label{sec:usecase-python}
@@
 To support this mode of operation, we added the concept of {\tt extension} packages
-to Spack. Python modules use the {\tt extends('python')} directive instead of
-{\tt depends\_on('python')}.
+to Spack. Python modules use the \verb|extends('python')| directive instead of
+\verb|depends_on('python')|.
 Each module installs into its own prefix like any other package,
@@
 Spack essentially implements a ``meta package-manager'' for each Python
-instance, which can coexist with Spack's normal installation model.
-This mechanism efficiently supports our application teams, for whom we can
+instance, which coexists with Spack's normal installation model.
+This mechanism efficiently supports application teams, for whom we can
 now rapidly construct custom Python installations.  We have also reduced
 the time that LC staff spend installing Python modules.
-Because Spack packages can extend the activation and deactivation mechanisms,
-the same mechanism could be used with other interpreted languages with 
-similar extension models, such as R, Ruby, or Lua.
+Because Spack packages can extend activation and deactivation mechanisms,
+this design could also be used with other languages
+with similar extension models, such as R, Ruby, or Lua.

### 2015-08-28 usecase-combinatorial.tex
Slightly change quoting in code and '[', ']', '*' in grammar.

@@
                framesep=2mm]{python}
-patch("patch.gpeftools2.4_xlc", when="@2.4 %xlc")
+patch('patch.gpeftools2.4_xlc', when='@2.4 %xlc')
 
 def install(self, spec, prefix):
-    if spec.architecture == "bgq" and self.compiler.satisfies("xlc"):
+    if spec.architecture == 'bgq' and self.compiler.satisfies('xlc'):
         configure("--prefix=" + prefix,  "LDFLAGS=-qnostaticlink")
-    elif spec.architecture == "bgq":
+    elif spec.architecture == 'bgq':
         configure("--prefix=" + prefix,  "LDFLAGS=-dynamic")

### 2015-08-28 figs/concretization-overhead/concretization-times.py
Get rid of Type 3 fonts.

@@ -1,6 +1,11 @@
 #!/usr/bin/env python
 
 import yaml
+
+from matplotlib import rcParams as rc
+rc['ps.useafm'] = True
+rc['pdf.use14corefonts'] = True
+rc['text.usetex'] = True
 import matplotlib.pyplot as plt
 
 files = (

### 2015-08-28 figs/concretization-overhead/package-sizes.py

@@ -1,8 +1,13 @@
 #!/usr/bin/env spack-python
 
 from external import yaml
-import matplotlib.pyplot as plt
 import numpy as np
+from matplotlib import rcParams as rc
+rc['ps.useafm'] = True
+rc['pdf.use14corefonts'] = True
+rc['text.usetex'] = True
+import matplotlib.pyplot as plt
+
 
 with open('merl.yaml') as stream:
     timings = yaml.load(stream)

### 2015-08-28 figs/overhead/overhead.py

@@ -2,6 +2,11 @@
 import sys
 import yaml
 import numpy as np
+
+from matplotlib import rcParams as rc
+rc['ps.useafm'] = True
+rc['pdf.use14corefonts'] = True
+rc['text.usetex'] = True
 import matplotlib.pyplot as plt
 from matplotlib.ticker import ScalarFormatter
 
@@ -62,7 +67,7 @@ no_wrap_rects = ax.bar(ind+2*width, no_wrap_means, width, color=orig_color)
 
 # add some text for labels, title and axes ticks
 ax.set_xlabel('Libraries', fontsize=fsize)
-ax.set_ylabel('Build time (s)', fontsize=fsize)
+ax.set_ylabel('Build time (seconds)', fontsize=fsize)
 ax.set_xticks(ind+width)
 ax.set_xticklabels(keys, fontsize=(fsize-1))
 
@@ -104,7 +109,7 @@ wrapper_rects = ax.bar(ind+width, wrap_overhead * 100, width, color=wrap_color)
 plt.axhline(y=0, color='k')
 
 ax.set_xlabel('Libraries', fontsize=fsize)
-ax.set_ylabel('Overhead (% time)', fontsize=fsize)
+ax.set_ylabel('Overhead (\% time)', fontsize=fsize)
 ax.set_xticks(ind+width)
 ax.set_xticklabels(keys, fontsize=(fsize-1))
 
